### 2017-05-05 R/olfactory.Rmd
zinbwave and clusterExp first round

---
title: "Analysis of the olfactory data"
author: "Davide Risso, Fanny Perraudeau"
date: "05/04/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",mesasge=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(SummarizedExperiment)
library(scRNAseq)
library(zinbwave)
library(ggplot2)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(scone)
library(RColorBrewer)
library(digest)
library(scone)
library(rARPACK)
set.seed(20)
```

## TL;DR

Applying our zero-inflated model to the olfactory data.

## Unnormalized data

We select only the cells that pass QC, color coded by the experimental time point. Note that the data are not public, hence it should not work other than on Davide's computer.

**I think Kelly said the data are now publicly available. How do we get the publicly available dataset?**

```{r datain}
load("../data/Expt4c2b_filtdata.Rda")
load("../data/E4c2b_slingshot_wsforkelly.RData")
```

```{r filtering}
names(batch) <- colnames(counts)

counts <- counts[,names(clus.labels)]
batch <- droplevels(batch[names(clus.labels)])
qc <- qc[names(clus.labels),]

vars <- rowVars(log1p(counts))
names(vars) <- rownames(counts)
vars <- sort(vars, decreasing = TRUE)
core <- counts[names(vars)[1:1000],]
```

We have `r ncol(counts)` cells. To speed up the computations, we will focus on the top 1,000 most variable genes.
```{r lookatdata}
dim(core)
core[1:3, 1:3]
```

Cells have been processed in `r length(unique(batch))` different batches
```{r tabbatch}
table(batch)
```

Cells have been clustered into `r length(unique(clus.labels))` different clusters
```{r tabclus}
table(clus.labels)
```

Batches are kind of confounded with the biology
```{r}
table(data.frame(batch = as.vector(batch),
                 cluster = clus.labels))

```

**Could you remind me where the clusters come from?**

For each cell, we dispose of qc measures
```{r}
head(qc, 2)
```

## PCA

First, let's look at PCA (of the log counts) for reference.

```{r pca}
par(mfrow=c(1, 2))
detection_rate <- colSums(core>0)
coverage <- colSums(core)
colMerged <- cc_rev[clus.labels]

pca <- prcomp(t(log1p(core)))
plot(pca$x, col=colMerged, pch=19, main="PCA of log-counts, centered not scaled")

fq <- EDASeq::betweenLaneNormalization(counts, which="full")
pcafq <- prcomp(t(log1p(fq)))
plot(pcafq$x, col=colMerged, pch=19, main="PCA of FQ log-counts, centered not scaled")
```

The first plot is the unnormalized data and the second plot is after full-quantile normalization, which is what Russell and Diya used for the paper.

**Do we have a reference for this paper?**

**Is it a fair comparison?? for pca on log counts, we use the 1000 most variable genes, but for full quantile normalization, we use all the genes. I understand that to perform full quantile normalization we need all the genes, but then should we look at only the 1000 most variable genes defined above, no?**

They found that to fully explain the differences between clusters, we need three dimensions.

```{r pca3}
pairs(pcafq$x[,1:3], col=colMerged, pch=19,
      main="PCA of FQ log-counts, centered not scaled")
```

```{r pca_cor}
df <- data.frame(PC1 = pcafq$x[, 1],
                 PC2 = pcafq$x[, 2],
                 detection_rate = detection_rate,
                 coverage = coverage)
pairs(df, pch = 19, col = colMerged)

print(cor(df, method = "spearman"))
```

Even after full-quantile normalization, there is some residual correlation between PC2 and detection rate. Note that, as expected, detection rate and coverage are correlated.

```{r pca_tosave}
totalcount = function (ei) {
  sums = colSums(ei)
  eo = t(t(ei)*mean(sums)/sums)
  return(eo)
}
raw <- as.matrix(counts)
tc <- totalcount(raw)
fq <- FQT_FN(raw)
tmm <- TMM_FN(raw)

fastpca <- function(expr, scale=FALSE) {
  svd_raw <- svds(scale(t(expr), center=TRUE, scale=scale), k=3, nu=3, nv=0)
  pc_raw <- svd_raw$u %*% diag(svd_raw$d[1:3])
  return(pc_raw)
}
vargenes <- rownames(core)
pc_raw <- fastpca(log1p(raw[vargenes,]))
pc_tc <- fastpca(log1p(tc[vargenes,]))
pc_fq <- fastpca(log1p(fq[vargenes,]))
pc_tmm <- fastpca(log1p(tmm[vargenes,]))
```

## ZIFA

Let's now look at zifa.

```{r zifa}
wrapRzifa <- function(Y, block = TRUE, k=2){
  # wrapper R function for ZIFA.
  # md5 hashing and temporary files are used not to re-run zifa 
  # if it has already be run on this computer.
  d = digest(Y, "md5")
  tmp = paste0(tempdir(), '/', d)
  write.csv(Y, paste0(tmp, '.csv'))

  if (!file.exists(paste0(tmp, "_", k, '_zifa.csv'))){
    print('run ZIFA')
    bb = ifelse(block, '-b ', '')
    cmd = sprintf('python run_zifa.py -d %d %s%s.csv %s_%d_zifa.csv', k, bb, tmp, tmp, k)
    system(cmd)
  }
  read.csv(sprintf("%s_%d_zifa.csv", tmp, k), header=FALSE)
}

zifa_raw <- wrapRzifa(log1p(raw[vargenes,]), k=3)

zifa_tc <- wrapRzifa(log1p(tc[vargenes,]), k=3)

zifa_tmm <- wrapRzifa(log1p(tmm[vargenes,]), k=3)

zifa_fq <- wrapRzifa(log1p(fq[vargenes,]), k=3)
```

```{r zifaplots}
pairs(zifa_raw[,1:3], col=colMerged, pch=19,
      main="ZIFA of raw log-counts")

pairs(zifa_fq[,1:3], col=colMerged, pch=19,
      main="ZIFA of FQ log-counts")

```

## ZINB

### Three dimensions
```{r zinb}
print(system.time(zinb_3 <- zinbFit(core, ncores = 3, K = 3)))
```

```{r zinb_plot}
pairs(zinb_3@W, col = colMerged, pch = 19, main = "ZINB")
```

```{r gamma}
qcpca <- prcomp(qc, center = TRUE, scale. = TRUE)
df <- data.frame(W1 = zinb_3@W[,1],
                 W2 = zinb_3@W[,2], 
                 W3 = zinb_3@W[,3],
                 gamma_mu = zinb_3@gamma_mu[1,],
                 gamma_pi = zinb_3@gamma_pi[1,],
                 detection_rate=detection_rate,
                 coverage = coverage, 
                 quality = qcpca$x[,1])
pairs(df, pch = 19, col = colMerged)

print(cor(df, method = "spearman"))
```

### Five dimensions

```{r}
print(system.time(zinb_5 <- zinbFit(core, ncores = 3, K = 5)))
```

```{r}
pairs(zinb_5@W, col = colMerged, pch = 19, main = "ZINB")
```

### Ten dimensions

```{r}
print(system.time(zinb_10 <- zinbFit(core, ncores = 3, K = 10)))
```

```{r}
pairs(zinb_10@W, col = colMerged, pch = 19, main = "ZINB")
```

### Accounting for quality

```{r qpca}
mod <- model.matrix(~ qcpca$x[, 1:5])
```

```{r}
print(system.time(zinb_3_qc <- zinbFit(core, ncores = 3, K = 3, X = mod)))
```

```{r}
pairs(zinb_3_qc@W, col=colMerged, pch=19, main="ZINB")
```

```{r}
#total number of detected genes in the cell
df <- data.frame(W1 = zinb_3_qc@W[,1],
                 W2 = zinb_3_qc@W[,2],
                 W3 = zinb_3_qc@W[,3],
                 gamma_mu = zinb_3_qc@gamma_mu[1,],
                 gamma_pi = zinb_3_qc@gamma_pi[1,],
                 detection_rate = detection_rate,
                 coverage = coverage)
pairs(df, pch = 19, col = colMerged)

print(cor(df, method = "spearman"))
```

K = 5
```{r}
print(system.time(zinb_5_qc <- zinbFit(core, ncores = 3, K = 5, X = mod)))
```

```{r}
pairs(zinb_5_qc@W, col=colMerged, pch=19, main="ZINB")
```

K = 10
```{r}
print(system.time(zinb_10_qc <- zinbFit(core, ncores = 3, K = 10, X = mod)))
```

```{r}
pairs(zinb_10_qc@W, col=colMerged, pch=19, main="ZINB")
```

### Accounting for batch

```{r}
class(batch)
mod <- model.matrix(~ batch)
dim(mod)
```

```{r}
print(system.time(zinb_3_batch <- zinbFit(core, ncores = 3, K = 3, X = mod)))
```

```{r}
pairs(zinb_3_batch@W, col=colMerged, pch=19, main="ZINB")
```

```{r}
#total number of detected genes in the cell
df <- data.frame(W1 = zinb_3@W[,1],
                 W2 = zinb_3@W[,2],
                 W3 = zinb_3@W[,3],
                 gamma_mu = zinb_3@gamma_mu[1,],
                 gamma_pi = zinb_3@gamma_pi[1,],
                 detection_rate = detection_rate,
                 coverage = coverage)
pairs(df, pch = 19, col = colMerged)

print(cor(df, method = "spearman"))
```

K = 5
```{r}
print(system.time(zinb_5_batch <- zinbFit(core, ncores = 3, K = 5, X = mod)))
```

```{r}
pairs(zinb_5_batch@W, col=colMerged, pch=19, main="ZINB")
```

K = 10
```{r}
print(system.time(zinb_10_batch <- zinbFit(core, ncores = 3, K = 10, X = mod)))
```

```{r}
pairs(zinb_10_batch@W, col=colMerged, pch=19, main="ZINB")
```

### Accounting for batch and quality

```{r}
mod <- model.matrix(~ qcpca$x[, 1:5] + batch)
```

```{r}
print(system.time(zinb_3_qc_batch <- zinbFit(core, ncores = 3, K = 3, X = mod)))
```

```{r}
pairs(zinb_3_qc_batch@W, col=colMerged, pch=19, main="ZINB")
```

K = 5
```{r}
print(system.time(zinb_5_qc_batch <- zinbFit(core, ncores = 3, K = 5, X = mod)))
```

```{r}
pairs(zinb_5_qc_batch@W, col=colMerged, pch=19, main="ZINB")
```

K = 10
```{r}
print(system.time(zinb_10_qc_batch <- zinbFit(core, ncores = 3, K = 10, X = mod)))
```

```{r}
pairs(zinb_10_qc_batch@W, col=colMerged, pch=19, main="ZINB")
```

```{r save, dependson="zifa"}
save(zinb_3, zinb_5, zinb_10,
     zinb_3_qc, zinb_5_qc, zinb_10_qc, 
     zinb_3_batch, zinb_5_batch, zinb_10_batch,
     zinb_3_qc_batch, zinb_5_qc_batch, zinb_10_qc_batch,
     pc_tmm, pc_fq, pc_tc, pc_raw,
     zifa_fq, zifa_tmm, zifa_tc, zifa_raw,
     file="olfactory.rda")
```

### 2017-05-05 R/clusterExperiment_olfactory.R

@@ -1,0 +1,102 @@
+require(clusterExperiment)
+require(SummarizedExperiment)
+require(limma)
+
+load('data/Expt4c2b_filtdata.Rda')
+load('data/E4c2b_slingshot_wsforkelly.RData')
+load('R/olfactory.rda')
+
+names(batch) <- colnames(counts)
+counts <- counts[,names(clus.labels)]
+batch <- droplevels(batch[names(clus.labels)])
+qc <- qc[names(clus.labels),]
+
+se = SummarizedExperiment(assays = list(counts = counts),
+                           colData = qc)
+
+assay(se)[1:2,1:5]
+colData(se)[,1:5]
+pass_filter = apply(assay(se), 1, function(x) length(x[x >= 10]) >= 10)
+dim(se)
+sum(!pass_filter)
+se <- se[pass_filter,]
+dim(se)
+fq <- round(limma::normalizeQuantiles(assay(se)))
+assays(se) <- list(normalized_counts=fq)
+
+
+# clusterMany
+# Parameters that seem relevant when input is zinbwave W matrix
+# isCount = FALSE
+# no dimReduce
+ce <- clusterMany(se, clusterFunction = "pam", ks = 5:10,
+                  isCount = TRUE, dimReduce = c("PCA", "var"),
+                  nVarDims = c(100, 500, 1000), nPCADims = c(5, 15, 50),
+                  run = TRUE)
+
+defaultMar <- par("mar")
+plotCMar <- c(1.1, 16.1, 4.1, 1.1)
+par(mar = plotCMar)
+plotClusters(ce, main = "Clusters from clusterMany",  axisLine = -1)
+
+# remove 'features' from the names of the labels
+cl<-clusterLabels(ce)
+cl<-gsub("Features","",cl)
+clusterLabels(ce)<-cl
+
+# different ordering of the clusterings
+cl <- clusterLabels(ce)
+ndims = sapply(cl, function(x) strsplit(x, '=|,')[[1]][2])
+ord <- order(as.numeric(ndims))
+par(mar = plotCMar)
+plotClusters(ce, main = "Clusters from clusterMany", whichClusters = ord, axisLine=-1)
+
+# combineMany
+clusterMatrix(ce)[1:5,1:10]
+ce<-combineMany(ce)
+par(mar=plotCMar)
+plotClusters(ce,whichClusters="workflow")
+wh<-which(clusterLabels(ce)=="combineMany")
[…]

### 2017-05-05 R/run_zifa.py

@@ -1,0 +1,35 @@
+#!/usr/bin/env python
+
+import sys
+sys.path.insert(1, '/usr/local/lib/python2.7/site-packages/') ##change the path if needed
+
+from argparse import ArgumentParser
+from pandas import read_csv
+from ZIFA import ZIFA,block_ZIFA
+import numpy as np
+
+def main():
+    parser = ArgumentParser(description="Fit a ZIFA model on the data.")
+    parser.add_argument('-b', '--block', action='store_true', default=False, help="Whether the block algorithm should be used.")
+    parser.add_argument('-d', '--dimensions', type=int, default=2, help="The number of dimensions [2].")
+    parser.add_argument('input_file', type=str, help="The input CSV file.")
+    parser.add_argument('output_file', type=str, help="The output CSV file.")
+
+    args = parser.parse_args()
+
+    df = read_csv(args.input_file)
+    del df['Unnamed: 0']
+
+    lc = np.array(df)
+    Y = np.transpose(lc)
+
+    if(args.block):
+        Z, model_params  = block_ZIFA.fitModel(Y, args.dimensions)
+    else:
+        Z, model_params  = ZIFA.fitModel(Y, args.dimensions)
+
+    np.savetxt(args.output_file, Z, delimiter=',')
+
+
+if __name__ == '__main__':
+        main()

### 2017-05-11 R/clusterExperiment_olfactory.Rmd
clusterExp html

---
title: "clusterExperiment on olfactory data"
author: "Fanny Perraudeau"
date: "05/10/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: show
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",mesasge=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(SummarizedExperiment)
library(clusterExperiment)
library(limma)
set.seed(20)
```

## TL;DR

The goal of this document is to create a worflow to cluster scRNASeq data with an example on olfactory data. It is not meant to be the final document where we would expect more written explanations. It is meant to choose the functions, parameters, and plots we want to show. Another goal of this document is also to see how we could use clusterExperiment in combinaison with zinbwave.

```{r dataIn}
load('../data/Expt4c2b_filtdata.Rda')
load('../data/E4c2b_slingshot_wsforkelly.RData')
load('olfactory.rda')
```

Let's create a SummarizedExperiment object. In colData, we add a column for the batches and another column for the clus.labels.

**Could you remind me where are the clus.labels from? I think we should have a more explicit name, like bio_condition or paper_clusters.**

```{r}
names(batch) <- colnames(counts)
counts <- counts[,names(clus.labels)]
batch <- droplevels(batch[names(clus.labels)])
qc <- qc[names(clus.labels),]
qc$batch = batch
qc$clusters = clus.labels

se = SummarizedExperiment(assays = list(counts = counts),
                          colData = qc)

assay(se)[1:2,1:5]
colData(se)[,1:5]
```

Let's filter out genes with too many zeros and normalize counts using FQ limma normalization.
```{r}
pass_filter = apply(assay(se), 1, function(x) length(x[x >= 10]) >= 10)
se <- se[pass_filter,]
dim(se)
fq <- round(limma::normalizeQuantiles(assay(se)))
assays(se) <- list(normalized_counts=fq)
```

## clusterMany

We included the clus.labels and batch. It is wanted that the cells don't cluster by batch.

**I think this is where we could include zinbwave. We could add zinbwave as dimReduce in clusterExperiment with an argument similar to nPCADims (something like nZINBDims) which would correspond to the K in the W matrix. We could then skip the normalization of the counts using limma (see above). We would need to set isCount = FALSE. Thoughts?**

```{r}
ce <- clusterMany(se,
                  clusterFunction = "pam",
                  ks = seq(9, 15, 3),
                  isCount = TRUE,
                  dimReduce = c("PCA", "var"),
                  nVarDims = c(100, 500, 1000),
                  nPCADims = c(5, 15, 50),
                  run = TRUE)

defaultMar <- par("mar")
plotCMar <- c(1.1, 16.1, 4.1, 1.1)
par(mar = plotCMar)
plotClusters(ce,
             main = "Clusters from clusterMany",
             axisLine = -1,
             sampleData=c("batch","clusters"))
```

Remove 'features' from the clusters labels to have more succint names.
```{r}
cl = clusterLabels(ce)
cl = gsub("Features", "", cl)
clusterLabels(ce) = cl
```

Re-order the clusters by dimension and not by k.
```{r}
cl = clusterLabels(ce)
ndims = sapply(cl, function(x) strsplit(x, '=|,')[[1]][2])
ord <- order(as.numeric(ndims))
par(mar = plotCMar)
plotClusters(ce,
             main = "Clusters from clusterMany",
             whichClusters = ord, axisLine = -1,
             sampleData=c("batch","clusters"))
```

The output matrix of clusterMany is
```{r}
clusterMatrix(ce)[1:3,1:3]
```

## CombineMany
```{r}
ce = combineMany(ce)
head(clusterMatrix(ce)[,1:3])
```

Look at the new row added for combineMany
```{r}
par(mar = plotCMar)
plotClusters(ce, whichClusters = "workflow",
             axisLine = -1, main = "Clusters from clusterMany",
             sampleData=c("batch","clusters"))
```

clusterMany requires samples to be in the same cluster in every clustering to be assigned a cluster. We need to change the minimum proportion of times they should be together with other samples in the cluster they are assigned to.
```{r}
wh = which(clusterLabels(ce) == "combineMany")
if (length(wh) != 1){
  stop()
}else{
  clusterLabels(ce)[wh] = "combineMany,default"
} 
# change proportion
ce = combineMany(ce, proportion = 0.7,
                 clusterLabel = "combineMany,0.7")
# plot
par(mar=plotCMar)
plotClusters(ce, whichClusters = "workflow",
             axisLine = -1, main = "Clusters from clusterMany",
             sampleData=c("batch","clusters"))
```
We can also play with the minSize parameter.
```{r}
# change minsize
ce = combineMany(ce, proportion = 0.7, minSize = 5, 
                 clusterLabel = "combineMany,final")
# plot
par(mar=plotCMar)
plotClusters(ce, whichClusters = "workflow",
             axisLine = -1, main = "Clusters from clusterMany",
             sampleData=c("batch","clusters"))
```

The proportion of times these clusters were together across these clusterings
```{r}
plotCoClustering(ce)
```

## mergeClusters
Cluster hierarchy for combineMany,final where before clustering PCA is performed and 15 most variable genes are kept 
```{r}
# makeDendrogram
ce = makeDendrogram(ce, whichCluster = "primaryCluster",
                    dimReduce = "PCA", ndims = 15)
plotDendrogram(ce)
```

We have not merged the clusters yet, but we have made the dendrogram
```{r}
ce
```

Before to actually merge the clusters, we are going to visualize the merge at the default cutoff (0.1)
```{r}
mergeClusters(ce, mergeMethod = "adjP", plot = "mergeMethod")
```

```{r}
ce = mergeClusters(ce, mergeMethod = "adjP", cutoff = 0.1)
```

```{r}
par(mar = plotCMar)
plotClusters(ce, whichClusters = "workflow", 
             axisLine = -1, main = "Clusters from clusterMany",
             sampleData=c("batch","clusters"))
```

```{r}
plotCoClustering(ce, sampleData = c("clusters"),
                 whichClusters = c("mergeClusters", 
                                   "combineMany"),
                 annLegend = FALSE)
```

```{r}
plotHeatmap(ce, clusterSamplesData = "dendrogramValue",
            breaks = .99,
            sampleData = c("clusters"))
```

## getBestFeatures
DE genes between clusters using limma voom
```{r}
pairsAll = getBestFeatures(ce, contrastType = "Pairs", 
                           p.value = 0.05,
                           number = 10,
                           isCount = TRUE)
head(pairsAll)
length(pairsAll$Feature) == length(unique(pairsAll$Feature))
```

```{r}
main = "Heatmap of features w/ significant pairwise differences"
clusterFeaturesData = unique(pairsAll[,"IndexInOriginal"])
plotHeatmap(ce, clusterSamplesData = "dendrogramValue",
            clusterFeaturesData = clusterFeaturesData,
            main = main, breaks = .99)
```

### 2017-05-11 R/zinbwave_olfactory.Rmd
clusterExp html

--- 2017-05-05 R/olfactory.Rmd
@@
 
-### Three dimensions
+We are going to look at different parameters for ZINB
+
+- K (3, 5, and 10),
+- X (intercept only, accounting for QC, accounting for batch, accounting for QC and batch).
+
+### No cell covariate
+
+Using only intercepts in X and V.
+
+#### Three dimensions
 ```{r zinb}
@@
 
-### Five dimensions
+#### Five dimensions
 
@@
 
-### Ten dimensions
+#### Ten dimensions
 
@@
 
+X has an intercept and 5 additional columns with the first five PC on qc measures.
+
 ```{r qpca}
@@
 
+#### Three dimensions
 ```{r}
@@
 
-K = 5
+#### Five dimensions
 ```{r}
@@
 
-K = 10
+#### Ten dimensions
 ```{r}
@@
 
+X has an intercept and columns for the batchs.
+
 ```{r}
@@
 ```
+
+#### Three dimensions
 
@@
 
+X has an intercept, columns for the batches, and columns with the first five PC on QC measures.
+
 ```{r}

### 2017-05-11 R/clusterExperiment_olfactory.R

@@ -1,102 +1,0 @@
-require(clusterExperiment)
-require(SummarizedExperiment)
-require(limma)
-
-load('data/Expt4c2b_filtdata.Rda')
-load('data/E4c2b_slingshot_wsforkelly.RData')
-load('R/olfactory.rda')
-
-names(batch) <- colnames(counts)
-counts <- counts[,names(clus.labels)]
-batch <- droplevels(batch[names(clus.labels)])
-qc <- qc[names(clus.labels),]
-
-se = SummarizedExperiment(assays = list(counts = counts),
-                           colData = qc)
-
-assay(se)[1:2,1:5]
-colData(se)[,1:5]
-pass_filter = apply(assay(se), 1, function(x) length(x[x >= 10]) >= 10)
-dim(se)
-sum(!pass_filter)
-se <- se[pass_filter,]
-dim(se)
-fq <- round(limma::normalizeQuantiles(assay(se)))
-assays(se) <- list(normalized_counts=fq)
-
-
-# clusterMany
-# Parameters that seem relevant when input is zinbwave W matrix
-# isCount = FALSE
-# no dimReduce
-ce <- clusterMany(se, clusterFunction = "pam", ks = 5:10,
-                  isCount = TRUE, dimReduce = c("PCA", "var"),
-                  nVarDims = c(100, 500, 1000), nPCADims = c(5, 15, 50),
-                  run = TRUE)
-
-defaultMar <- par("mar")
-plotCMar <- c(1.1, 16.1, 4.1, 1.1)
-par(mar = plotCMar)
-plotClusters(ce, main = "Clusters from clusterMany",  axisLine = -1)
-
-# remove 'features' from the names of the labels
-cl<-clusterLabels(ce)
-cl<-gsub("Features","",cl)
-clusterLabels(ce)<-cl
-
-# different ordering of the clusterings
-cl <- clusterLabels(ce)
-ndims = sapply(cl, function(x) strsplit(x, '=|,')[[1]][2])
-ord <- order(as.numeric(ndims))
-par(mar = plotCMar)
-plotClusters(ce, main = "Clusters from clusterMany", whichClusters = ord, axisLine=-1)
-
-# combineMany
-clusterMatrix(ce)[1:5,1:10]
-ce<-combineMany(ce)
-par(mar=plotCMar)
-plotClusters(ce,whichClusters="workflow")
-wh<-which(clusterLabels(ce)=="combineMany")
[…]

### 2017-05-16 R/zinb_clustering_over_k.Rmd
zinbwave + clustering + slingshot

---
title: "Consensus clustering after ZINB-WAVE across different k"
author: "Davide Risso"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)

library(zinbwave)
library(clusterExperiment)

run_zinb <- FALSE
run_cluster <- FALSE
```

This workflow should help us decide if it's meaningful to run ZINB-WAVE with 
different values of k to get a consensus clustering or if it's better to pick "the right k."

## INPUT DATA

For now I'm using the data that I have stored in my computer, but eventually we
should start with the data from GEO. Question: which one? RSEM? Cufflinks? Kallisto?

```{r datain}
load("~/git/scone_analysis/data/olfactory_for_scone.rda")
se <- as(scone_obj, "SummarizedExperiment")
colData(se)

labels <- read.table("https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt", row.names = 1)
cl <- labels[,1]
names(cl) <- rownames(labels)
table(cl)

original <- rep(-1, NCOL(se))
names(original) <- colnames(se)
original[intersect(names(cl), colnames(se))] <- cl[intersect(names(cl), colnames(se))]

palette <- read.table("https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_colorPalette.txt", comment.char = "%", stringsAsFactors = FALSE)
pal <- c("transparent", palette[,2])
names(pal) <- c("-1", palette[,1])

colData(se)$original_cluster <- as.factor(original)
```

# ZINB-WaVE

First of all, we check that with K=2 we get something reasonable.

For now, let's focus on the 1000 most variable genes, but eventually one question
is which genes to use and how to decide.

```{r vargenes}
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
core
```

```{r zinb}
zinb2 <- zinbFit(core, X = "~batch", K = 2, ncores = 7)
plot(getW(zinb2), pch=19, col=pal[colData(core)$original_cluster])
```

This seems fine.

Since we are running clusterExperiment on the first 50 principal components,
what happens if we run zinb with k=50?

```{r zinb_50}
zinb50 <- zinbFit(core, X = "~batch", K = 50, ncores = 7)
plot(getW(zinb50), pch=19, col=pal[colData(core)$original_cluster])
```

## Run for multiple k

```{r zinb_k}
ks <- 3:10

if(run_zinb) {
  zinb_res <- lapply(ks, function(k) zinbFit(core, X = "~batch", K = k, ncores=7))
  save(zinb_res, file="../data/zinb_k3_10.rda")
} else {
  load("../data/zinb_k3_10.rda")
}
```

```{r zinb_plot, eval=FALSE}
for(i in seq_along(ks)) {
  pairs(getW(zinb_res[[i]]), pch=19, col=pal[colData(core)$original_cluster])
}
```

# ClusterExperiment

```{r clustering}
ws <- lapply(zinb_res, function(x) t(getW(x)))
if(run_cluster) {
  cl_res <- clusterMany(ws, ks = 5:15, alphas = c(0.1, 0.3), betas = 0.9,
                        clusterFunction = "hierarchical01", minSizes=5,
                        sequential = TRUE, subsample = TRUE, ncores = 7)
  save(cl_res, file="../data/cl_k3_10.rda")
} else {
  load("../data/cl_k3_10.rda")
}
```

## Consensus across all k

```{r consensus}
combined <- combineMany(cl_res$clMat, proportion = 0.5)
dendro <- makeDendrogram(ws[[8]], combined$clustering)
merged <- mergeClusters(assay(core), combined$clustering, dendro$clusters, mergeMethod = "adjP")

clmat <- cbind(merged$clustering, combined$clustering, cl_res$clMat)

plotClusters(clmat)
```

## Consensus using only one k

### k=15

```{r consensus_k}
combined <- combineMany(cl_res$clMat[,grep("dataset8", colnames(cl_res$clMat))],
                        proportion = 0.5)
dendro <- makeDendrogram(ws[[8]], combined$clustering)
merged <- mergeClusters(assay(core), combined$clustering, dendro$clusters, mergeMethod = "adjP")

clmat <- cbind(merged$clustering, combined$clustering, cl_res$clMat)

plotClusters(clmat)
```

### k=50

```{r consensus_50}
## matching the parameters used by Russell
W <- t(getW(zinb50))

if(run_cluster) {
  cl_res <- clusterMany(W, ks = 4:15, alphas = c(0.1), betas = 0.8,
                        clusterFunction = "hierarchical01", minSizes=1,
                        sequential = TRUE, subsample = TRUE, ncores = 7,
                        subsampleArgs = list(resamp.num=100,
                                             clusterFunction="kmeans",
                                             clusterArgs=list(nstart=10)),
                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
  save(cl_res, file="../data/cl_k50.rda")
} else {
  load("../data/cl_k50.rda")
}

```

```{r combine_50}
combined <- combineMany(cl_res, proportion = 0.7)

plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))

table(primaryClusterNamed(combined))
sum(primaryCluster(combined) == -1)
sum(colData(core)$original_cluster == -1)

table(primaryClusterNamed(combined), colData(core)$original_cluster)

plot(getW(zinb50), pch=19, col=pal[colData(core)$original_cluster])
plot(getW(zinb50), pch=19, col=pal[as.factor(primaryCluster(combined))])

combined <- makeDendrogram(combined)
plotDendrogram(combined)
merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)

# 
# plot(getW(zinb50), pch=19, col=pal[as.factor(primaryCluster(merged))])
```

# Slingshot

Let's try to run slingshot on W, with different values of k to see how stable the
lineages are. We run slingshot in two modes, one in which we specify only the starting
cluster (unsupervised) and one in which we specify the end clusters too (supervised).

A note on the interpretation of the clusters.

Cluster name | Description | Correspondence
-------------|-------------|-----------------
c1 | resting HBC | original 1, 5
c2 | activated HBC | original 5
c3 | GBC / immature neurons / MV 1 | original 2, 3, 11
c4 | Sus | original 4, 7
c5 | new | new
c6 | new | new
c7 | Neuron | original 9, 12
c8 | Immature Neuron | original 10, 14
c9 | Microvillus | original 15
c10 | Immature Neuron | original 14

Cluster name | Correspondence | Description
-------------|-----------------|-------------
m1 | c1 + c2 | HBC
m2 | c3 + c10 | GBC + immature nerons
m3 | c4 | Sus
m4 | c5 + c6 | new
m5 | c7 | Neuron
m6 | c8 | Immature Neuron
m7 | c9 | Microvillus

## Unsupervised

We try with k=3, k=4, k=5.

```{r slingshot}
library(slingshot)
pal2 <- pal[-1][c(1, 5, 3, 4, 2, 10, 11, 9, 13)]
pal3 <- pal[-1][c(1, 3, 4, 2, 11, 9, 13)]
cl_labs <- as.factor(primaryClusterNamed(merged))
cl <- droplevels(cl_labs[cl_labs != "-1"])

for(k in c(3, 4, 5)) {
  zinb_k <- zinbFit(core[,cl_labs != "-1"], X = "~batch", K = k, ncores = 7)
  X <- getW(zinb_k)

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal3)
  plot_tree(X, cl, lineages, col.clus = pal3)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

## Supervised

We try with k=3, k=4, k=5.

```{r slingshot_sup}
for(k in c(3, 4, 5)) {
  zinb_k <- zinbFit(core[,cl_labs != "-1"], X = "~batch", K = k, ncores = 7)
  X <- getW(zinb_k)

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1", end.clus = c("m3", "m5", "m7"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal3)
  plot_tree(X, cl, lineages, col.clus = pal3)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

# Session Info

```{r}
sessionInfo()
```

### 2017-05-25 R/zinbwave_residuals.Rmd
residuals

---
title: "Residuals in zinbwave"
author: "Fanny Perraudeau, Davide Risso"
date: "05/25/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(SummarizedExperiment)
library(scRNAseq)
library(zinbwave)
library(ggplot2)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(scone)
library(RColorBrewer)
library(digest)
library(scone)
library(rARPACK)
set.seed(20)
```

# TL;DR

The goal of this document is to look at residuals and see if we could use the residuals to look at some sort of adjusted "counts" (residuals can be negative, so we should probably find another name than "counts"). First, I fitted zinbwave without the batches in the X, then I included the batches in the X. For each scenario, I looked at naive, naive on the log scale, Pearson, and deviance residuals. I plotted the PCA of the residuals with two coloring (batch and clusters) and boxplots of the residuals for each cell.

# Unnormalized data
I am still using the data Davide gave me a while ago. And I think now it is not the same data Davide is using in its most recent vignette (that is zinb_clustering_over_k.Rmd). It should not be very important but it would be nice if we have a common dataset we can work on.

```{r datain}
load("../data/Expt4c2b_filtdata.Rda")
load("../data/E4c2b_slingshot_wsforkelly.RData")
```

Here we only look at the 1000 most variable genes.
```{r filtering}
names(batch) <- colnames(counts)

counts <- counts[,names(clus.labels)]
batch <- droplevels(batch[names(clus.labels)])
qc <- qc[names(clus.labels),]

vars <- rowVars(log1p(counts))
names(vars) <- rownames(counts)
vars <- sort(vars, decreasing = TRUE)
core <- counts[names(vars)[1:1000],]
```

We have `r ncol(counts)` cells. To speed up the computations, we will focus on the top 1,000 most variable genes.
```{r lookatdata}
dim(core)
core[1:3, 1:3]
```

Cells have been processed in `r length(unique(batch))` different batches
```{r tabbatch}
table(batch)
```

Cells have been clustered into `r length(unique(clus.labels))` different clusters
```{r tabclus}
table(clus.labels)
```

Batches are kind of confounded with the biology
```{r}
table(data.frame(batch = as.vector(batch),
                 cluster = clus.labels))

```

```{r}
plotBoxplot <- function(y, main = '', col, log = F){
  if (log) y = log2(y + 1)
  #median = apply(y, 1, median)
  #rle = apply(y, 2, function(x) x - median)
  boxplot(y, main=main, col=col, staplewex=0, outline=0, 
          border=col, xaxt='n')
  abline(h=0)
}

compute.naive.residuals <- function(Y, zinb){
  mu_hat = t(getMu(zinb)) 
  pi_hat = t(getPi(zinb)) 
  Y_hat = (1 - pi_hat) * mu_hat
  Y - Y_hat 
}

compute.log.naive.residuals <- function(Y, zinb){
  mu = t(getMu(zinb)) 
  pi = t(getPi(zinb)) 
  phi = matrix(rep(getPhi(zinb), ncol(Y)), ncol = ncol(Y))
  var_hat = (1 - pi) * mu * (1 + mu * (phi + pi))
  Y_hat = (1 - pi) * mu
  log_Y_hat_plus_1 = log(1 + Y_hat) - var_hat / (2*(1 + Y_hat)^2)
  log(Y + 1) - log_Y_hat_plus_1
}

compute.pearson.residuals <- function(Y, zinb){
  num = compute.naive.residuals(Y, zinb)
  mu = t(getMu(zinb)) 
  pi = t(getPi(zinb)) 
  phi = matrix(rep(getPhi(zinb), ncol(Y)), ncol = ncol(Y))
  var_hat = (1 - pi) * mu * (1 + mu * (phi + pi))
  num / sqrt(var_hat)  
}

compute.zinb.loglik <- function(Y, zinb){
  mu = t(getMu(zinb)) 
  theta = getTheta(zinb)
  theta_mat = matrix(rep(theta, ncol(Y), ncol = ncol(Y)))
  pi = t(getPi(zinb))
  log( pi * (Y == 0) + (1 - pi) * dnbinom(Y, size = theta, mu = mu) )
}

compute.deviance.residuals <- function(Y, zinb){
  mu_hat = t(getMu(zinb)) 
  pi_hat = t(getPi(zinb)) 
  Y_hat = (1 - pi_hat) * mu_hat
  ll = compute.zinb.loglik(Y, zinb)
  sign = as.numeric(Y - Y_hat > 0)
  sign[sign == 0] = -1
  sign * sqrt(-2 * ll)
}

```

# X intercept, no batch
Let's run zinbwave with K = 0 and X with an intercept only (no batch).
```{r}
print(system.time(zinb <- zinbFit(core, ncores = 3, K = 0)))
```

## Naive residuals
```{r}
Rn = compute.naive.residuals(core, zinb)
Rn[1:3,1:3]
```

PCA (centered not scale) of residuals, first colored by batch (left), then by clusters (right). On the left side, we should not see patterns whereas on the right hand side we should.
```{r}
pca = prcomp(t(Rn))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

Each boxplot is for a cell. Colors correspond to the batches. When colors are the same but boxplots are not next to each others it corresponds to different batches. We would expect that the residuals look similar for the different batches.
```{r}
Rn_order = Rn[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rn_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

## Naive residuals of log(Y+1)
```{r}
Rn = compute.log.naive.residuals(core, zinb)
Rn[1:3,1:3]
```

```{r}
pca = prcomp(t(Rn))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```
```{r}
Rn_order = Rn[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rn_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

## Pearson residuals
```{r}
Rp = compute.pearson.residuals(core, zinb)
Rp[1:3,1:3]
```

```{r}
pca = prcomp(t(Rp))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

```{r}
Rp_order = Rp[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rp_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

## Deviance residuals
```{r}
Rd = compute.deviance.residuals(core, zinb)
Rd[1:3,1:3]
```

```{r}
pca = prcomp(t(Rd))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

```{r}
Rd_order = Rd[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rd_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

# X batches
Let's run zinbwave with K = 0 and X with an intercept only (no batch).
```{r}
mod = model.matrix( ~ batch)
print(system.time(zinb_batch <- zinbFit(core, ncores = 3, K = 0,
                                  X = mod)))
```

## Naive residuals
```{r}
Rn = compute.naive.residuals(core, zinb_batch)
Rn[1:3,1:3]
```
```{r}
pca = prcomp(t(Rn))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

```{r}
Rn_order = Rn[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rn_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

## Naive residuals of log(Y+1)
```{r}
Rn = compute.log.naive.residuals(core, zinb)
Rn[1:3,1:3]
```

```{r}
pca = prcomp(t(Rn))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

```{r}
Rn_order = Rn[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rn_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

## Pearson residuals
```{r}
Rp = compute.pearson.residuals(core, zinb_batch)
Rp[1:3,1:3]
```

```{r}
pca = prcomp(t(Rp))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

```{r}
Rp_order = Rp[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rp_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

## Deviance residuals
```{r}
Rd = compute.deviance.residuals(core, zinb_batch)
Rd[1:3,1:3]
```

```{r}
pca = prcomp(t(Rd))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

```{r}
Rd_order = Rd[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
plotBoxplot(Rd_order, main = 'Boxplot of residuals\ncolor=batch',
            col = col_order)
```

# Conclusions

As batches are somehow confounded with the biology, I don't know if we should adjust or not for the batches. For the residuals, it seems that the Pearson residuals when not adjusting for batches is the most informative (at least, when looking at PC 1 and 2).

# Session Info

```{r}
sessionInfo()
```

### 2017-05-26 R/workflow.Rmd
worflow

---
title: "Worflow"
author: "Fanny Perraudeau"
date: "05/25/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
library(SummarizedExperiment)
library(scRNAseq)
library(zinbwave)
library(ggplot2)
library(gplots)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(scone)
library(RColorBrewer)
library(digest)
library(scone)
library(slingshot)
library(rARPACK)
library(clusterExperiment)
set.seed(20)
```

```{r}
runZinb = F
runClus = F
NCORES = 2
```

# TL;DR

This document is a temptative of workflow with the following steps

- Fit zinbwave with K = 50, X including the batches,   
- Get W which should capture the biology,  
- Use clusterExperiment on W,   
- Get the cluster labels,   
- Use slingshot on W from zinbwave and with cluster labels from clusterExperiment,  
- Get lineages,  
- Fit again zinbwave but with X including batches and lineages from slingshot,  
- Get DE genes between lineages.

Along the worflow, use deviance residuals as adjusted "counts".

# Unnormalized data
I am still using the data Davide gave me a while ago. And I think now it is not the same data Davide is using in its most recent vignette (that is zinb_clustering_over_k.Rmd). It should not be very important but it would be nice if we have a common dataset we can work on.

```{r datain}
load("../data/Expt4c2b_filtdata.Rda")
load("../data/E4c2b_slingshot_wsforkelly.RData")
```

Here we only look at the 1000 most variable genes.
```{r filtering}
names(batch) <- colnames(counts)

counts <- counts[,names(clus.labels)]
batch <- droplevels(batch[names(clus.labels)])
qc <- qc[names(clus.labels),]

vars <- rowVars(log1p(counts))
names(vars) <- rownames(counts)
vars <- sort(vars, decreasing = TRUE)
core <- counts[names(vars)[1:1000],]
```

We have `r ncol(counts)` cells.
```{r lookatdata}
dim(core)
core[1:3, 1:3]
```

Cells have been processed in `r length(unique(batch))` different batches
```{r tabbatch}
table(batch)
```

Cells have been clustered into `r length(unique(clus.labels))` different clusters
```{r tabclus}
table(clus.labels)
```

Batches are kind of confounded with the biology
```{r}
table(data.frame(batch = as.vector(batch),
                 cluster = clus.labels))

```

```{r}
compute.naive.residuals <- function(Y, zinb){
  mu_hat = t(getMu(zinb)) 
  pi_hat = t(getPi(zinb)) 
  Y_hat = (1 - pi_hat) * mu_hat
  Y - Y_hat 
}

compute.pearson.residuals <- function(Y, zinb){
  num = compute.naive.residuals(Y, zinb)
  mu = t(getMu(zinb)) 
  pi = t(getPi(zinb)) 
  phi = matrix(rep(getPhi(zinb), ncol(Y)), ncol = ncol(Y))
  var_hat = (1 - pi) * mu * (1 + mu * (phi + pi))
  num / sqrt(var_hat)  
}

compute.zinb.loglik <- function(Y, zinb){
  mu = t(getMu(zinb)) 
  theta = getTheta(zinb)
  theta_mat = matrix(rep(theta, ncol(Y), ncol = ncol(Y)))
  pi = t(getPi(zinb))
  log( pi * (Y == 0) + (1 - pi) * dnbinom(Y, size = theta, mu = mu) )
}

compute.deviance.residuals <- function(Y, zinb){
  mu_hat = t(getMu(zinb)) 
  pi_hat = t(getPi(zinb)) 
  Y_hat = (1 - pi_hat) * mu_hat
  ll = compute.zinb.loglik(Y, zinb)
  sign = 1*(Y - Y_hat > 0)
  sign[sign == 0] = -1
  sign * sqrt(-2 * ll)
}
```

# 0. Residuals for visualization
## Fit zinbwave X intercept, no batch
Let's run zinbwave with K = 0 and X batches.
```{r}
fn = '../data/zinb_k0_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb0 <- zinbFit(core, K = 0, X = mod, ncores = NCORES)))
  save(zinb0, file = fn)
}else{
  load(fn)
}
```

We use deviance residuals for visualization. Let's check that deviance residuals look ok.
```{r}
res = compute.deviance.residuals(core, zinb0)
res[1:3,1:3]
```

PCA on the residuals where color are for batches on the left and previously found clusters on the right.
```{r}
pca = prcomp(t(res))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

Boxplot of the residuals for each cell.
```{r}
res_order = res[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
boxplot(res_order, main='Boxplot of residuals\ncolor=batch', col = col_order,         staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
```

Deviance residuals seem ok.

# 1. zinbwave with raw counts

Let's run zinbwave with K = 50 and X including batches.
```{r}
fn = '../data/zinb_k50_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50,
                                      X = mod)))
  save(zinb50, file = fn)
}else{
  load(fn)
}
```

# 2. clusterExperiment with W

I use the code of Davide from zinb_clutering_over_k.Rmd
```{r consensus_50}
## matching the parameters used by Russell
W = zinb50@W
fn = '../data/clustExp_W_k50.rda'
if(runClus) {
  cl_res <- clusterMany(t(W), ks = 4:15, alphas = c(0.1), betas = 0.8,
                        clusterFunction = "hierarchical01", minSizes=1,
                        sequential = TRUE, subsample = TRUE, ncores = 7,
                        subsampleArgs = list(resamp.num=100,
                                             clusterFunction="kmeans",
                                             clusterArgs=list(nstart=10)),
                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
  save(cl_res, file = fn)
} else {
  load(fn)
}
```

```{r combine_50}
combined <- combineMany(cl_res, proportion = 0.7)
plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(combined)
table(primaryClusterNamed(combined))
sum(primaryCluster(combined) == -1)
```

```{r new_clusters}
plot(getW(zinb50), col=clus.labels, main = 'color = previous')
plot(getW(zinb50), col=as.factor(primaryCluster(combined)), main = 'color = new')
table(data.frame(previous = clus.labels, new = primaryCluster(combined)))
```

```{r merged}
combined <- makeDendrogram(combined)
plotDendrogram(combined)
merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)
```

Just to see what it looks like, let's look at the heatmap on the W (number of cells x K=50). Now, we would like to plot a heatmap using the residuals (number of cells x number of genes). When I call plotHeatmap using argument visualizeData = residuals_matrix, it does not work and says that if give separate visualizeData, must be of same dimensions as assay(data).
```{r heatmap}
plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
```

So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
```{r}
palDF = merged@clusterLegend[[1]]
pal = palDF[, 'color']
names(pal) = palDF[, 'name']

heatmap.2(res, Colv = merged@dendro_samples,
          scale = 'none', trace = 'none', 
          col = colorRampPalette(c('blue', 'yellow'))(51),
          ColSideColors = pal[primaryClusterNamed(merged)],
          labCol = '', main = 'Deviance Residuals, 1000 most variable genes')
legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
       horiz = T, cex = 0.5)
```

# 3. slingshot with W and cluster labels
I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.

```{r slingshot}
fn = '../data/zinb_k4_batch.rda'
if (runZinb){
  cl_labs = primaryCluster(merged)
  batch_red = as.vector(batch)
  batch_red = batch_red[cl_labs != -1]
  cl_labs = cl_labs[cl_labs != -1]
  mod = model.matrix( ~ batch_red)
  zinb4 = zinbFit(core[,primaryCluster(merged) != -1],
                  X = mod, K = 4, ncores = NCORES)
  save(zinb4, file = fn)
}else{
  cl_labs = primaryCluster(merged)
  batch_red = as.vector(batch)
  batch_red = batch_red[cl_labs != -1]
  cl_labs = cl_labs[cl_labs != -1]
  mod = model.matrix( ~ batch_red)
  load(fn)
}
```

```{r}
X <- zinb4@W

lineages <- get_lineages(X, clus.labels = cl_labs, start.clus = 1)
curves <- get_curves(X, clus.labels = cl_labs, lineages = lineages)
plot_curves(X, cl_labs, curves, col.clus = unique(cl_labs))

print(lineages$lineage1)
print(lineages$lineage2)
print(lineages$lineage3)
print(lineages$lineage4)
```

# 4. zinbwave with lineages

Which clusters would it make sense to test one versus the other? Would it be interesting to do two by two comparison between clusters next to each other in the different lineages?

# 5. DE genes with betas

DE analysis using ZINB-Wave has not been implemented yet and we are still thinking about it. Would it make sense just to test the betas between clusters making the assumption that the genes are independent? Below is just a visualization of what we could have at the end of the worflow: a heatmap of the deviance residuals for DE genes between clusters.

```{r}
de = read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
de = de$V1

breaks = setBreaks(res[rownames(res) %in% de, ], 0.99)
heatmap.2(res[rownames(res) %in% de, ], breaks = breaks,
          Colv = merged@dendro_samples,
          scale = 'none', trace = 'none', 
          col = colorRampPalette(c('blue', 'yellow'))(51),
          ColSideColors = pal[primaryClusterNamed(merged)],
          labCol = '', main = 'Deviance Residuals, DE genes (Russell)')
legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
       horiz = T, cex = 0.5)
```

# Session Info

```{r}
sessionInfo()
```

### 2017-05-30 R/workflow.Rmd
update to RSEC code

@@
 ```{r}
-runZinb = F
-runClus = F
+runZinb = FALSE
+runClus = FALSE
 NCORES = 2
@@
 We have `r ncol(counts)` cells.
+
 ```{r lookatdata}
@@
 Cells have been processed in `r length(unique(batch))` different batches
+
 ```{r tabbatch}
@@
 Cells have been clustered into `r length(unique(clus.labels))` different clusters
+
 ```{r tabclus}
@@
 Batches are kind of confounded with the biology
+
 ```{r}
@@
                  cluster = clus.labels))
-
 ```
@@
 
+EAP: Fanny, we should really be using the RSEC function for all the steps of the clustering.
+I've put in the code, but I set `eval=FALSE` because I don't have the data to test it.
+I also am not sure that we want to set the options for `subsampleArgs` and `seqArgs` since those are suppose to be sort of esoteric concerns...
+
+```{r rsec_50, eval=FALSE}
+fn = '../data/clustExp_W_k50.rda'
+if(runClus) {
+ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
+                      clusterFunction = "hierarchical01", minSizes=1,
+                      sequential = TRUE, subsample = TRUE, ncores = NCORES,
+					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01
+                      subsampleArgs = list(resamp.num=100,
+                                           clusterFunction="kmeans",
+                                           clusterArgs=list(nstart=10)),
+                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
+save(ceObj, file = fn)
+}
+else{
+	load(fn)
+}
+```
+
+```{r examineCombineMany, eval=FALSE}
+plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
+plotCoClustering(ceObj)
+table(primaryClusterNamed(ceObj))
+sum(primaryCluster(ceObj) == -1)
+```
+
+```{r examineMergeClusters, eval=FALSE}
+#re-does merging simpling to make plot 
+mergeClusters(ceObj, mergeMethod = "locfdr", plot="mergeMethod",cutoff = 0.01)
+```
+
+```{r heatmap,eval=FALSE}
+plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
+```
+
+## Original ClusterExperiment code (in Steps)
 I use the code of Davide from zinb_clutering_over_k.Rmd
+
 ```{r consensus_50}
@@
 When I call plotHeatmap using argument visualizeData = residuals_matrix, it does not work and says that if give separate visualizeData, must be of same dimensions as assay(data).
+
 ```{r heatmap}

### 2017-05-31 R/workflow.Rmd
add data

@@
 #re-does merging simpling to make plot 
-mergeClusters(ceObj, mergeMethod = "locfdr", plot="mergeMethod",cutoff = 0.01)
+if(packageVersion("clusterExperiment")>='1.3.0.9001') mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
 ```
@@
 ```{r heatmap,eval=FALSE}
-plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
+if(packageVersion("clusterExperiment")>='1.3.0.9002') plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
 ```

### 2017-05-31 R/createData.R

@@ -1,0 +1,35 @@
+load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet")
+## count matrix
+E <- assayData(Cufflinks_eSet)$counts_table
+
+# Remove undetected genes
+E <- na.omit(E)
+E <- E[rowSums(E)>0,]
+
+# Remove ERCC and CreER
+cre <- E["CreER",]
+ercc <- E[grep("^ERCC-", rownames(E)),]
+E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
+E <- E[-which(rownames(E)=="CreER"), ]
+
+# Extract QC metrics
+qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
+qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))
+
+# Extract metadata
+batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
+bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
+
+
+metaData<-data.frame("Experiment"=bio,"Batch"=batch,qc)
+
+# filtering to top 1000 most variable:
+library(matrixStats)
+vars <- rowVars(log1p(E))
+names(vars) <- rownames(E)
+vars <- sort(vars, decreasing = TRUE)
+core <- E[names(vars)[1:1000],]
+
+
+write.table(core,file="../data/oeCufflinkCountData_1000Var.txt",sep="\t",row.names=TRUE,col.names=TRUE)
+write.table(metaData,file="../data/oeMetadata.txt",sep="\t",row.names=TRUE,col.names=TRUE)

### 2017-05-31 R/workflow.Rmd
update code for new data and add makefile

--- 2017-05-30 R/workflow.Rmd
@@
 ```{r}
-runZinb = FALSE
-runClus = FALSE
+runZinb = TRUE
+runClus = TRUE
 NCORES = 2
@@
 
+EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
+
 # Unnormalized data
@@
 
+EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file.
+For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article.
+This will be slightly different data from Russell's, which didn't use all of the samples.
+We can adjust that decision later, or just compare the samples that are the same.
+(Russell's clusterLabels are in the meta data)
+
 ```{r datain}
-load("../data/Expt4c2b_filtdata.Rda")
-load("../data/E4c2b_slingshot_wsforkelly.RData")
+#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
+core<-read.table("../data/oeCufflinkCountData_1000Var.txt",sep="\t",header=TRUE)
+metadata<-read.table("../data/oeMetadata.txt",sep="\t",header=TRUE)
+# load("../data/Expt4c2b_filtdata.Rda")
+# load("../data/E4c2b_slingshot_wsforkelly.RData")
 ```
 
-Here we only look at the 1000 most variable genes.
+Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.
+
 ```{r filtering}
-names(batch) <- colnames(counts)
-
-counts <- counts[,names(clus.labels)]
-batch <- droplevels(batch[names(clus.labels)])
-qc <- qc[names(clus.labels),]
-
-vars <- rowVars(log1p(counts))
-names(vars) <- rownames(counts)
-vars <- sort(vars, decreasing = TRUE)
-core <- counts[names(vars)[1:1000],]
+batch <- metadata$Batch
+qc <- metadata[,!names(metadata)%in% c("Batch","Experiment","clusterLabels")]
+clus.labels<-metadata[,"clusterLabels"]
+# vars <- rowVars(log1p(counts))
+# names(vars) <- rownames(counts)
+# vars <- sort(vars, decreasing = TRUE)
+# core <- counts[names(vars)[1:1000],]
 ```
 
-We have `r ncol(counts)` cells.
+We have `r ncol(core)` cells.
 
@@
 ```
+
+EAP: if we are going to ultimately use these functions, it would be good to add them to the package for zinbwave.
 
@@
 Let's run zinbwave with K = 0 and X batches.
+
 ```{r}
@@
 We use deviance residuals for visualization. Let's check that deviance residuals look ok.
+
 ```{r}
@@
 PCA on the residuals where color are for batches on the left and previously found clusters on the right.
+
 ```{r}
@@
 Boxplot of the residuals for each cell.
+
 ```{r}
@@
 Let's run zinbwave with K = 50 and X including batches.
+
 ```{r}
@@
 
-```{r rsec_50, eval=FALSE}
+```{r rsec_50}
 fn = '../data/clustExp_W_k50.rda'
@@
 
-```{r examineCombineMany, eval=FALSE}
+```{r examineCombineMany}
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
@@
 
-```{r examineMergeClusters, eval=FALSE}
+```{r examineMergeClusters}
 #re-does merging simpling to make plot 
-mergeClusters(ceObj, mergeMethod = "locfdr", plot="mergeMethod",cutoff = 0.01)
+if(packageVersion("clusterExperiment")>='1.3.0.9001') mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
 ```
 
-```{r heatmap,eval=FALSE}
-plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
+```{r heatmap }
+if(packageVersion("clusterExperiment")>='1.3.0.9002') plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
+```
+
+```{r compareClusters}
+plot(getW(zinb50), col=clus.labels, main = 'color = previous')
+plot(getW(zinb50), col=as.factor(primaryCluster(ceObj)), main = 'color = new')
+table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
 ```
@@
 
-```{r consensus_50}
+```{r consensus_50, eval=FALSE}
 ## matching the parameters used by Russell
@@
 
-```{r combine_50}
+```{r combine_50, eval=FALSE}
 combined <- combineMany(cl_res, proportion = 0.7)
@@
 
-```{r new_clusters}
+```{r new_clusters, eval=FALSE}
 plot(getW(zinb50), col=clus.labels, main = 'color = previous')
@@
 
-```{r merged}
+```{r merged, eval=FALSE}
 combined <- makeDendrogram(combined)
@@
 
-```{r heatmap}
+```{r heatmapOld, eval=FALSE}
 plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
@@
 So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
-```{r}
+```{r, eval=FALSE}
 palDF = merged@clusterLegend[[1]]

### 2017-05-31 R/createData.R

@@ -20,8 +20,10 @@ qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))
 batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
 bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
 
+clusterLabels<-read.table("../data/oeHBCdiff_clusterLabels.txt",sep="\t",stringsAsFactors=FALSE)
+m<-match(colnames(E) ,clusterLabels[,1])
 
-metaData<-data.frame("Experiment"=bio,"Batch"=batch,qc)
+metaData<-data.frame("Experiment"=bio,"Batch"=batch,"clusterLabels"=clusterLabels[m,2],qc)
 
 # filtering to top 1000 most variable:
 library(matrixStats)

### 2017-05-31 R/zinbwave_residuals.Rmd
residuals on 1000 and all genes

@@
 ---
-title: "Residuals in zinbwave"
+title: "Residuals in zinbwave, 1000 most variable genes"
 author: "Fanny Perraudeau, Davide Risso"
@@
 knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
-library(SummarizedExperiment)
-library(scRNAseq)
 library(zinbwave)
-library(ggplot2)
-library(magrittr)
 library(matrixStats)
-library(Rtsne)
-library(scone)
-library(RColorBrewer)
-library(digest)
-library(scone)
-library(rARPACK)
+library(gplots) #heatmap.2()
+library(clusterExperiment) #setBreaks()
 set.seed(20)
+NCORES = 8
 ```
@@
 load("../data/E4c2b_slingshot_wsforkelly.RData")
+de = read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
+de = de$V1
 ```
@@
   if (log) y = log2(y + 1)
-  #median = apply(y, 1, median)
-  #rle = apply(y, 2, function(x) x - median)
   boxplot(y, main=main, col=col, staplewex=0, outline=0, 
@@
   ll = compute.zinb.loglik(Y, zinb)
-  sign = as.numeric(Y - Y_hat > 0)
+  sign = 1*(Y - Y_hat > 0)
   sign[sign == 0] = -1
@@
 ```{r}
-print(system.time(zinb <- zinbFit(core, ncores = 3, K = 0)))
+print(system.time(zinb <- zinbFit(core, ncores = NCORES, K = 0)))
 ```
@@
 
+```{r}
+de_1000 = de[de %in% rownames(Rn)]
+length(de_1000)
+head(de_1000)
+```
+
 Each boxplot is for a cell.
@@
             col = col_order)
+```
+
+```{r}
+breaks = setBreaks(Rn[de_1000, ], breaks = 0.99)
+heatmap.2(Rn[de_1000, ], breaks = breaks,
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Naive Residuals')
 ```
@@
 ```
+
 ```{r}
@@
             col = col_order)
+```
+
+```{r}
+heatmap.2(Rn[de_1000, ],
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Naive Residuals')
 ```
@@
 
+```{r}
+heatmap.2(Rp[de_1000, ],
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Pearson residuals')
+```
+
 ## Deviance residuals
@@
 
+```{r}
+heatmap.2(Rd[de_1000, ],
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Deviance residuals')
+```
+
 # X batches
@@
 mod = model.matrix( ~ batch)
-print(system.time(zinb_batch <- zinbFit(core, ncores = 3, K = 0,
+print(system.time(zinb_batch <- zinbFit(core, ncores = NCORES, K = 0,
                                   X = mod)))
@@
             col = col_order)
+```
+
+```{r}
+breaks = setBreaks(Rn[de_1000, ], breaks = 0.99)
+heatmap.2(Rn[de_1000, ], breaks = breaks,
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Naive residuals')
 ```
@@
 
+```{r}
+heatmap.2(Rn[de_1000, ],
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Naive residuals')
+```
+
 ## Pearson residuals
@@
             col = col_order)
+```
+
+```{r}
+heatmap.2(Rp[de_1000, ],
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Pearson residuals')
 ```
@@
 
+```{r}
+heatmap.2(Rd[de_1000, ],
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          scale = 'none', trace = 'none',
+          ColSideColors = as.character(clus.labels),
+          labCol = '', main = 'Deviance residuals')
+```
+
 # Conclusions

### 2017-06-02 R/workflow.Rmd
fix RSEC command

@@
 
+EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?
+
 ```{r datain}
@@
 core<-read.table("../data/oeCufflinkCountData_1000Var.txt",sep="\t",header=TRUE)
+core<-data.matrix(core)
 metadata<-read.table("../data/oeMetadata.txt",sep="\t",header=TRUE)
@@
                       clusterFunction = "hierarchical01", minSizes=1,
-                      sequential = TRUE, subsample = TRUE, ncores = NCORES,
+                      ncores = NCORES,
 					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01

### 2017-06-02 R/zinbwave_residuals_all.Rmd

--- 2017-05-31 R/zinbwave_residuals.Rmd
@@
 ---
-title: "Residuals in zinbwave, 1000 most variable genes"
+title: "Residuals in zinbwave, all genes"
 author: "Fanny Perraudeau, Davide Risso"
@@
 
-Here we only look at the 1000 most variable genes.
+Here we only look at all the genes.
 ```{r filtering}
@@
 
-vars <- rowVars(log1p(counts))
-names(vars) <- rownames(counts)
-vars <- sort(vars, decreasing = TRUE)
-core <- counts[names(vars)[1:1000],]
+#vars <- rowVars(log1p(counts))
+#names(vars) <- rownames(counts)
+#vars <- sort(vars, decreasing = TRUE)
+#core <- counts[names(vars)[1:1000],]
+core = counts
 ```
@@
 ```{r}
-heatmap.2(Rn[de_1000, ],
+heatmap.2(Rn[de_1000, ], breaks = breaks,
           col = colorRampPalette(c('blue', 'yellow'))(51),
@@
           ColSideColors = as.character(clus.labels),
-          labCol = '', main = 'Naive Residuals')
+          labCol = '', main = 'Log naive residuals')
 ```
@@
           ColSideColors = as.character(clus.labels),
-          labCol = '', main = 'Naive residuals')
+          labCol = '', main = 'Naive Residuals')
 ```
@@
 ```
-
 ```{r}
-heatmap.2(Rn[de_1000, ],
+heatmap.2(Rn[de_1000, ], breaks = breaks,
           col = colorRampPalette(c('blue', 'yellow'))(51),
@@
           ColSideColors = as.character(clus.labels),
-          labCol = '', main = 'Naive residuals')
+          labCol = '', main = 'Log naive residuals')
 ```

### 2017-06-02 R/workflow.Rmd
update clustering code

@@
 knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
-library(SummarizedExperiment)
+#bioconductor:
 library(scRNAseq)
+library(scone)
+library(clusterExperiment)
+#github
 library(zinbwave)
+library(slingshot)
+#CRAN
 library(ggplot2)
@@
 library(Rtsne)
-library(scone)
 library(RColorBrewer)
 library(digest)
-library(scone)
-library(slingshot)
 library(rARPACK)
-library(clusterExperiment)
 set.seed(20)
@@
 fn = '../data/clustExp_W_k50.rda'
+metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
+seObj<-SummarizedExperiment(t(W),colData=metadata)
 if(runClus) {
-ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
+ print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
                       clusterFunction = "hierarchical01", minSizes=1,
@@
                                            clusterArgs=list(nstart=10)),
-                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
+                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE))
 save(ceObj, file = fn)

### 2017-06-02 R/createData.R

@@ -22,7 +22,6 @@ bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
 
 clusterLabels<-read.table("../data/oeHBCdiff_clusterLabels.txt",sep="\t",stringsAsFactors=FALSE)
 m<-match(colnames(E) ,clusterLabels[,1])
-
 metaData<-data.frame("Experiment"=bio,"Batch"=batch,"clusterLabels"=clusterLabels[m,2],qc)
 
 # filtering to top 1000 most variable:

### 2017-06-03 R/workflow.Rmd
put check in for current version of clusterExperiment

@@
 set.seed(20)
+if(packageVersion("clusterExperiment")<'1.3.0.9003') stop("must have current develop version to avoid bugs")
 ```
@@
 #re-does merging simpling to make plot 
-if(packageVersion("clusterExperiment")>='1.3.0.9001') mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
+mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
 ```
@@
 ```{r heatmap }
-if(packageVersion("clusterExperiment")>='1.3.0.9002') plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
+plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
 ```

### 2017-06-03 R/workflow.Rmd
fix version required by clusterExperiment

@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9003') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9004') stop("must have current develop version to avoid bugs")
 ```

### 2017-06-05 R/workflow.Rmd
update version needed for clusterExperiment

@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9004') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9005') stop("must have current develop version to avoid bugs")
 ```

### 2017-06-05 R/workflow.Rmd
update version check

@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9005') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9006') stop("must have current develop version to avoid bugs")
 ```

### 2017-06-06 R/workflow.Rmd
organize vignette, add summary schema

@@
 title: "Worflow"
-author: "Fanny Perraudeau"
-date: "05/25/2017"
+author: "Workflow team"
+date: "06/06/2017"
 output: 
@@
 ```{r}
-runZinb = TRUE
-runClus = TRUE
-NCORES = 2
-```
-
-# TL;DR
-
-This document is a temptative of workflow with the following steps
-
-- Fit zinbwave with K = 50, X including the batches,   
-- Get W which should capture the biology,  
-- Use clusterExperiment on W,   
-- Get the cluster labels,   
-- Use slingshot on W from zinbwave and with cluster labels from clusterExperiment,  
-- Get lineages,  
-- Fit again zinbwave but with X including batches and lineages from slingshot,  
-- Get DE genes between lineages.
-
-Along the worflow, use deviance residuals as adjusted "counts".
-
-EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
-
-# Unnormalized data
-I am still using the data Davide gave me a while ago.
-And I think now it is not the same data Davide is using in its most recent vignette (that is zinb_clustering_over_k.Rmd).
-It should not be very important but it would be nice if we have a common dataset we can work on.
-
-EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file.
-For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article.
-This will be slightly different data from Russell's, which didn't use all of the samples.
-We can adjust that decision later, or just compare the samples that are the same.
-(Russell's clusterLabels are in the meta data)
-
-EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?
-
-```{r datain}
-#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
-core<-read.table("../data/oeCufflinkCountData_1000Var.txt",sep="\t",header=TRUE)
-core<-data.matrix(core)
-metadata<-read.table("../data/oeMetadata.txt",sep="\t",header=TRUE)
-# load("../data/Expt4c2b_filtdata.Rda")
-# load("../data/E4c2b_slingshot_wsforkelly.RData")
-```
-
-Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.
-
-```{r filtering}
-batch <- metadata$Batch
-qc <- metadata[,!names(metadata)%in% c("Batch","Experiment","clusterLabels")]
-clus.labels<-metadata[,"clusterLabels"]
-# vars <- rowVars(log1p(counts))
-# names(vars) <- rownames(counts)
-# vars <- sort(vars, decreasing = TRUE)
-# core <- counts[names(vars)[1:1000],]
-```
-
-We have `r ncol(core)` cells.
-
-```{r lookatdata}
-dim(core)
-core[1:3, 1:3]
-```
-
-Cells have been processed in `r length(unique(batch))` different batches
-
-```{r tabbatch}
-table(batch)
-```
-
-Cells have been clustered into `r length(unique(clus.labels))` different clusters
-
-```{r tabclus}
-table(clus.labels)
-```
-
-Batches are kind of confounded with the biology
-
-```{r}
-table(data.frame(batch = as.vector(batch),
-                 cluster = clus.labels))
+runZinb <- FALSE
+runClus <- TRUE
+NCORES <- 8
 ```
@@
 
-```{r}
-compute.naive.residuals <- function(Y, zinb){
-  mu_hat = t(getMu(zinb)) 
-  pi_hat = t(getPi(zinb)) 
-  Y_hat = (1 - pi_hat) * mu_hat
-  Y - Y_hat 
-}
+FP: yes, I agree.
 
-compute.pearson.residuals <- function(Y, zinb){
-  num = compute.naive.residuals(Y, zinb)
-  mu = t(getMu(zinb)) 
-  pi = t(getPi(zinb)) 
-  phi = matrix(rep(getPhi(zinb), ncol(Y)), ncol = ncol(Y))
-  var_hat = (1 - pi) * mu * (1 + mu * (phi + pi))
-  num / sqrt(var_hat)  
-}
-
+```{r residuals}
 compute.zinb.loglik <- function(Y, zinb){
@@
 
-# 0. Residuals for visualization
-## Fit zinbwave X intercept, no batch
-Let's run zinbwave with K = 0 and X batches.
+# TL;DR
+
+This document is a temptative of workflow with the following steps
+
+- Dimensionality reduction using zinbwave to get W which should capture the biology,
+- Cluster cells using clusterExperiment on W to get the cluster labels,   
+- Get lineage using slingshot on W and cluster labels from clusterExperiment,  
+- Get DE genes between lineages/clusters.
+
+Along the worflow, use deviance residuals as adjusted values.
+
+EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
+
+# Steps of the workflow
+## Create a SummarizedExperiment object
+
+Along the workflow, we want to use a unique SummarizedExperiement object carrying all the data we need.
+
+EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file.
+For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article.
+This will be slightly different data from Russell's, which didn't use all of the samples.
+We can adjust that decision later, or just compare the samples that are the same.
+(Russell's clusterLabels are in the meta data)
+
+EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?
+
+```{r datain}
+#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
+core <- read.table("../data/oeCufflinkCountData_1000Var.txt",
+                   sep = "\t", header = TRUE)
+core <- data.matrix(core)
+metadata <- read.table("../data/oeMetadata.txt", sep = "\t", header = TRUE)
+# symbol for samples missing from original clustering
+metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
+```
+
+Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.
+
+```{r batches}
+batch <- metadata$Batch
+```
+
+Cells have been processed in `r length(unique(batch))` different batches
 
 ```{r}
-fn = '../data/zinb_k0_batch.rda'
+col_batch = rep(brewer.pal(9, "Set1"), 2)
+names(col_batch) = unique(batch)
+table(batch)
+```
+
+We have qc measures from the data
+
+```{r qc}
+qc <- metadata[, !names(metadata) %in% c("Batch", "Experiment", "clusterLabels")]
+head(qc, 2)
+```
+
+```{r clusterlabels}
+clus.labels <- metadata[, "clusterLabels"]
+```
+
+In published work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters
+
+```{r}
+col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
+col_clus <- col_clus[1:length(unique(clus.labels))]
+names(col_clus) <- sort(unique(clus.labels))
+table(clus.labels)
+```
+
+Batches are kind of confounded with the biology
+
+```{r}
+table(data.frame(batch = as.vector(batch),
+                 cluster = clus.labels))
+```
+
+We have `r ncol(core)` cells.
+
+```{r lookatdata}
+dim(core)
+core[1:3, 1:3]
+```
+
+Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, published labels, and quality control measures.
+
+```{r}
+se <- SummarizedExperiment(assays = list(rawCounts = core), colData = metadata)
+```
+
+## Dimensionality reduction adjusting for gene and cell level covariates
+
+To cluster and get lineages we want to reduce the dimension of the data.
+We are going to use zinbwave to do so.
+First, let's fit zinbwave with first K = 0 to compute normalized values (i.e.
+deviance residuals) adjusted for batches.
+We could also adjust for gene length or GC content here.
+We then fit zinbwave to get the dimensionality reduced matrix W specifying the number of dimension K = 50.
+Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE).
+For K = 0 and K = 50, we correct for batch effect including batches in X.
+
+```{r}
+fn50 <- '../data/zinb_k50_batch.rda'
+fn0 <- '../data/zinb_k0_batch.rda'
 if (runZinb){
   mod = model.matrix( ~ batch)
-  print(system.time(zinb0 <- zinbFit(core, K = 0, X = mod, ncores = NCORES)))
-  save(zinb0, file = fn)
+  # K = 0
+  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES, K = 0, X = mod)))
+  save(zinb0, file = f50)
+  # K = 50
+  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50, X = mod)))
+  save(zinb50, file = fn50)
 }else{
-  load(fn)
+  load(fn50)
+  load(fn0)
 }
@@
 
-We use deviance residuals for visualization. Let's check that deviance residuals look ok.
+### Normalized values
+
+We use deviance residuals as normalized values for visualization. FP: explain rational: K=0 so residuals capture the bio adjusting for batch. Let's check that deviance residuals look ok.
 
 ```{r}
-res = compute.deviance.residuals(core, zinb0)
+res <- compute.deviance.residuals(core, zinb0)
 res[1:3,1:3]
@@
 
-PCA on the residuals where color are for batches on the left and previously found clusters on the right.
+Boxplot of the normalized values for each cell. It seems that correction for batches is ok.
 
 ```{r}
-pca = prcomp(t(res))
+res_order <- res[, order(as.numeric(batch))]
+col_order <- as.numeric(batch)[order(as.numeric(batch))]
+boxplot(res_order, main='Boxplot of normalized values\ncolor=batch',
+        col = col_order, staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
+```
+
+PCA on the normalized values where color are for batches on the left and previously found clusters on the right. We want no clustering on the left side and clustering on the right side.
+
+```{r}
+pca <- prcomp(t(res))
 par(mfrow = c(1,2))
-plot(pca$x, col = batch, pch = 19,
-     main="PCA of residuals\ncolor=batch")
-plot(pca$x, col = clus.labels, pch = 19,
-     main = "PCA of residuals\ncolor=cluster")
+plot(pca$x, col = col_batch[batch], pch = 20,
+     main="PCA of normalized values\ncolor=batch")
+plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20,
+     main = "PCA of normalized values\ncolor=cluster")
 par(mfrow = c(1,1))
@@
 
-Boxplot of the residuals for each cell.
+Let's add the normalized values as an assay dataset in our SummarizedExperiment object.
 
 ```{r}
-res_order = res[, order(as.numeric(batch))]
-col_order = as.numeric(batch)[order(as.numeric(batch))]
-boxplot(res_order, main='Boxplot of residuals\ncolor=batch', col = col_order,         staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
+assays(se)[['normalizedValues']] <- res
 ```
 
-Deviance residuals seem ok.
+### W
 
-# 1. zinbwave with raw counts
-
-Let's run zinbwave with K = 50 and X including batches.
+Let's check that performing MDS on W we have something coherent with published clusters.
 
 ```{r}
-fn = '../data/zinb_k50_batch.rda'
-if (runZinb){
-  mod = model.matrix( ~ batch)
-  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50,
-                                      X = mod)))
-  save(zinb50, file = fn)
-}else{
-  load(fn)
-}
+W <- getW(zinb50)
+d <- dist(W)
+fit <- cmdscale(d, eig = TRUE, k = 2)
+plot(fit$points, col = col_clus[as.character(clus.labels)], main = 'MDS', pch = 20,
+     xlab = 'Component 1', ylab = 'Component 2')
+legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
+       fill = unique(col_clus), title = 'Sample')
 ```
 
-# 2. clusterExperiment with W
+Let's add W to the colData of our SummarizedExperiment object.
+
+```{r}
+W <- data.frame(W)
+colnames(W) <- paste0('W', 1:ncol(W))
+colData(se) <- cbind(colData(se), W)
+```
+
+FP: what do you think of implemented a "compute_normalized_values" argument to zinbFit for summarizedExperiment objects where zinbFit would take as input a SummarizedExperiment object and return a summarizedExperiment object with slots added for normalized values and W?
+
+## Cluster of cells
+
+We use clusterExperiment with W.
 
@@
 
+FP: What about the following call to RSEC?
+
 ```{r rsec_50}
-fn = '../data/clustExp_W_k50.rda'
-metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
-seObj<-SummarizedExperiment(t(W),colData=metadata)
-if(runClus) {
- print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
-                      clusterFunction = "hierarchical01", minSizes=1,
+W = colData(se)[, grepl('^W', colnames(colData(se)))]
+W = as.matrix(W)
+fn = '../data/RSEC_W.rda'
+if (runClus){
+  print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
+                                  clusterFunction = "hierarchical01",
+                                  minSizes = 5,
                                   ncores = NCORES,
-					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01
-                      subsampleArgs = list(resamp.num=100,
-                                           clusterFunction="kmeans",
-                                           clusterArgs=list(nstart=10)),
-                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE))
+                                  combineProportion = 0.7,
+                                  mergeMethod = "locfdr", mergeCutoff = 0.01)))
   save(ceObj, file = fn)
-}
-else{
+}else{
   load(fn)
@@
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
+```
+
+```{r plotcoclust}
 plotCoClustering(ceObj)
+```
+
+```{r}
 table(primaryClusterNamed(ceObj))
@@
 ```{r compareClusters}
-plot(getW(zinb50), col=clus.labels, main = 'color = previous')
-plot(getW(zinb50), col=as.factor(primaryCluster(ceObj)), main = 'color = new')
+plot(W, col = col_clus[as.character(clus.labels)], main = 'color = previous', pch = 20)
+plot(W, col = as.factor(primaryCluster(ceObj)), main = 'color = new', pch = 20)
 table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
-```
-
-## Original ClusterExperiment code (in Steps)
-I use the code of Davide from zinb_clutering_over_k.Rmd
-
-```{r consensus_50, eval=FALSE}
-## matching the parameters used by Russell
-W = zinb50@W
-fn = '../data/clustExp_W_k50.rda'
-if(runClus) {
-  cl_res <- clusterMany(t(W), ks = 4:15, alphas = c(0.1), betas = 0.8,
-                        clusterFunction = "hierarchical01", minSizes=1,
-                        sequential = TRUE, subsample = TRUE, ncores = 7,
-                        subsampleArgs = list(resamp.num=100,
-                                             clusterFunction="kmeans",
-                                             clusterArgs=list(nstart=10)),
-                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
-  save(cl_res, file = fn)
-} else {
-  load(fn)
-}
-```
-
-```{r combine_50, eval=FALSE}
-combined <- combineMany(cl_res, proportion = 0.7)
-plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))
-plotCoClustering(combined)
-table(primaryClusterNamed(combined))
-sum(primaryCluster(combined) == -1)
-```
-
-```{r new_clusters, eval=FALSE}
-plot(getW(zinb50), col=clus.labels, main = 'color = previous')
-plot(getW(zinb50), col=as.factor(primaryCluster(combined)), main = 'color = new')
-table(data.frame(previous = clus.labels, new = primaryCluster(combined)))
-```
-
-```{r merged, eval=FALSE}
-combined <- makeDendrogram(combined)
-plotDendrogram(combined)
-merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)
-```
-
-Just to see what it looks like, let's look at the heatmap on the W (number of cells x K=50).
-Now, we would like to plot a heatmap using the residuals (number of cells x number of genes).
-When I call plotHeatmap using argument visualizeData = residuals_matrix, it does not work and says that if give separate visualizeData, must be of same dimensions as assay(data).
-
-```{r heatmapOld, eval=FALSE}
-plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
 ```
@@
 
-heatmap.2(res, Colv = merged@dendro_samples,
+plotHeatmap(res, Colv = merged@dendro_samples,
           scale = 'none', trace = 'none', 
@@
 
-# 3. slingshot with W and cluster labels
+## Get lineages
 I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.
@@
 
-# 4. zinbwave with lineages
-
-Which clusters would it make sense to test one versus the other? Would it be interesting to do two by two comparison between clusters next to each other in the different lineages?
-
-# 5. DE genes with betas
+## DE analysis
 

### 2017-06-06 R/workflow.Rmd
workflow with se

@@
 ```{r}
-runZinb <- FALSE
-runClus <- TRUE
-NCORES <- 8
+runZinb <- T
+runClus <- T
+NCORES <- 10
 ```
@@
 ```{r}
-se <- SummarizedExperiment(assays = list(rawCounts = core), colData = metadata)
+se <- SummarizedExperiment(assays = list(rawCounts = core),
+                           colData = metadata)
 ```
@@
 ```{r}
+fn0 <- '../data/zinb_k0_batch.rda'
+if (runZinb & !file.exists(fn0)){
+  mod <- model.matrix( ~ batch)
+  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES,
+                                     K = 0, X = mod)))
+  save(zinb0, file = fn0)
+}else{
+  load(fn0)
+}
+
 fn50 <- '../data/zinb_k50_batch.rda'
-fn0 <- '../data/zinb_k0_batch.rda'
-if (runZinb){
-  mod = model.matrix( ~ batch)
-  # K = 0
-  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES, K = 0, X = mod)))
-  save(zinb0, file = f50)
-  # K = 50
-  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50, X = mod)))
+if (runZinb & !file.exists(fn50)){
+  mod <- model.matrix( ~ batch)
+  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES,
+                                      K = 50, X = mod)))
   save(zinb50, file = fn50)
@@
   load(fn50)
-  load(fn0)
 }
@@
 ```{r rsec_50}
-W = colData(se)[, grepl('^W', colnames(colData(se)))]
-W = as.matrix(W)
-fn = '../data/RSEC_W.rda'
-if (runClus){
+W <- colData(se)[, grepl('^W', colnames(colData(se)))]
+W <- as.matrix(W)
+fn <- '../data/RSEC_W.rda'
+if (runClus & !file.exists(fn)){
   print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
+                                  alphas = c(0.3), betas = c(0.9),
                                   clusterFunction = "hierarchical01",
@@
                                   combineProportion = 0.7,
-                                  mergeMethod = "locfdr", mergeCutoff = 0.01)))
+                                  mergeMethod = "locfdr",
+                                  mergeCutoff = 0.01)))
   save(ceObj, file = fn)
@@
 
+FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context?
+
 ```{r examineMergeClusters}
 #re-does merging simpling to make plot 
-mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
+mergeClusters(ceObj, mergeMethod = "locfdr",
+              plotInfo = "mergeMethod", cutoff = 0.05)
 ```
@@
 ```{r heatmap }
-plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
+plotHeatmap(ceObj, clusterSamplesData = "dendrogramValue", 
+            breaks = .99)
 ```
@@
 ```{r compareClusters}
-plot(W, col = col_clus[as.character(clus.labels)], main = 'color = previous', pch = 20)
-plot(W, col = as.factor(primaryCluster(ceObj)), main = 'color = new', pch = 20)
-table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
+plot(fit$points, col = col_clus[as.character(clus.labels)],
+     main = 'MDS W, color = published clusters', pch = 20,
+     xlab = 'Component1', ylab = 'Component2')
 ```
 
-So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
-```{r, eval=FALSE}
-palDF = merged@clusterLegend[[1]]
-pal = palDF[, 'color']
-names(pal) = palDF[, 'name']
+```{r}
+palDF <- ceObj@clusterLegend[[1]]
+pal <- palDF[, 'color']
+names(pal) <- palDF[, 'name']
+plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
+     main = 'MDS W, color = our clusters', pch = 20,
+     xlab = 'Component1', ylab = 'Component2')
+```
 
-plotHeatmap(res, Colv = merged@dendro_samples,
-          scale = 'none', trace = 'none', 
-          col = colorRampPalette(c('blue', 'yellow'))(51),
-          ColSideColors = pal[primaryClusterNamed(merged)],
-          labCol = '', main = 'Deviance Residuals, 1000 most variable genes')
-legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
-       horiz = T, cex = 0.5)
+```{r}
+table(data.frame(published = clus.labels,
+                 ours = primaryCluster(ceObj)))
+```
+
+So, let's look at a heatmap on normalized values.
+
+```{r}
+sampleData <- data.frame(ours = primaryCluster(ceObj))
+
+plotHeatmap(assays(se)$normalizedValues,
+            main = 'Normalized values, 1000 most variable genes',
+            clusterSamplesData = ceObj@dendro_samples,
+            sampleData = as.matrix(sampleData))
 ```
@@
 ## Get lineages
-I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.
 
-```{r slingshot}
-fn = '../data/zinb_k4_batch.rda'
-if (runZinb){
-  cl_labs = primaryCluster(merged)
-  batch_red = as.vector(batch)
-  batch_red = batch_red[cl_labs != -1]
-  cl_labs = cl_labs[cl_labs != -1]
-  mod = model.matrix( ~ batch_red)
-  zinb4 = zinbFit(core[,primaryCluster(merged) != -1],
-                  X = mod, K = 4, ncores = NCORES)
-  save(zinb4, file = fn)
+The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=2, k=3, k=4 in W.
+
+FP: We need to work on the cluster coloring (!) and the matching with the clusters you found for the paper with Russell, maybe by changing the parameters in clusterMany?
+
+```{r}
+Kvec <- c(2, 3, 4)
+```
+
+### Use previous W, no re-fitting 
+#### Unsupervised
+
+```{r slingshot_unsup}
+our_cl <- as.factor(primaryClusterNamed(ceObj))
+cl = our_cl[our_cl != "-1"]
+for (k in Kvec){
+  X <- W[our_cl != "-1", 1:k]
+
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+}
+```
+
+#### Supervised
+
+```{r slingshot_sup}
+for (k in Kvec){
+  X <- W[our_cl != "-1", 1:k]
+
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
+                           end.clus = c("m9", "m6"))
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+}
+```
+
+### Re-fitting zinbwave
+#### Unsupervised
+
+```{r refit_zinb}
+mod <- model.matrix(~ batch[our_cl != "-1"])
+fn <- '../data/refit_zinbwave_slingshot.rda'
+if (run & !file.exists(fn)){
+  zinbList <- lapply(Kvec, function(k){
+    zinbFit(core[, our_cl != "-1"], X = mod, K = k, ncores = NCORES)
+  })
+  save(zinbList, file = fn)
 }else{
-  cl_labs = primaryCluster(merged)
-  batch_red = as.vector(batch)
-  batch_red = batch_red[cl_labs != -1]
-  cl_labs = cl_labs[cl_labs != -1]
-  mod = model.matrix( ~ batch_red)
   load(fn)
@@
 
-```{r}
-X <- zinb4@W
+```{r slingshot_unsup_refit}
+for(k in Kvec) {
+  X <- getW(zinbList[[k - 1]])[, 1:k]
 
-lineages <- get_lineages(X, clus.labels = cl_labs, start.clus = 1)
-curves <- get_curves(X, clus.labels = cl_labs, lineages = lineages)
-plot_curves(X, cl_labs, curves, col.clus = unique(cl_labs))
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
 
+  print(paste0("K=", k))
   print(lineages$lineage1)
@@
   print(lineages$lineage4)
+}
+```
+
+#### Supervised
+
+```{r slingshot_sup_refit}
+for(k in Kvec){
+  X <- getW(zinbList[[k - 1]])[, 1:k]
+
+  lineages <- get_lineages(X, clus.labels = our_cl, start.clus = "m1",
+                           end.clus = c("m9", "m6"))
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+}
 ```
@@
 
-DE analysis using ZINB-Wave has not been implemented yet and we are still thinking about it.
-Would it make sense just to test the betas between clusters making the assumption that the genes are independent?
-Below is just a visualization of what we could have at the end of the worflow: a heatmap of the deviance residuals for DE genes between clusters.
+Here is the kind of plots we want to be able to have
 
 ```{r}
-de = read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
-de = de$V1
-
-breaks = setBreaks(res[rownames(res) %in% de, ], 0.99)
-heatmap.2(res[rownames(res) %in% de, ], breaks = breaks,
-          Colv = merged@dendro_samples,
-          scale = 'none', trace = 'none', 
-          col = colorRampPalette(c('blue', 'yellow'))(51),
-          ColSideColors = pal[primaryClusterNamed(merged)],
-          labCol = '', main = 'Deviance Residuals, DE genes (Russell)')
-legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
-       horiz = T, cex = 0.5)
+de <- read.csv('../data/oe_markers.txt', 
+              stringsAsFactors = F, header = F)
+de <- de$V1
+nv <- assays(se)$normalizedValues
+plotHeatmap(nv[rownames(nv) %in% de, ],
+            main = 'Normalized values, 1000 most variable genes',
+            clusterSamplesData = ceObj@dendro_samples,
+            sampleData = as.matrix(sampleData$ours))
 ```

### 2017-06-06 R/workflow.Rmd
add a bit of writing

@@
 
-FP: yes, I agree.
+FP: yes, I agree. We would ideally work with a SummarizedExperiment object where we would add an assay for the normalized values.
 
@@
 - Cluster cells using clusterExperiment on W to get the cluster labels,
-- Get lineage using slingshot on W and cluster labels from clusterExperiment,  
-- Get DE genes between lineages/clusters.
+- Get lineage using slingshot on W from zinbwave and cluster labels from clusterExperiment,  
+- Get DE genes between lineages/clusters using a tool like MAST or edgeR.
 
@@
 We are going to use zinbwave to do so.
-First, let's fit zinbwave with first K = 0 to compute normalized values (i.e.
+First, let's fit zinbwave first with K = 0 to compute normalized values (i.e.
 deviance residuals) adjusted for batches.
@@
 We then fit zinbwave to get the dimensionality reduced matrix W specifying the number of dimension K = 50.
-Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE).
+We could try different K here (maybe K = 100).
+We could also try to see if adding commondispersion = FALSE change the results.Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE).
+
 For K = 0 and K = 50, we correct for batch effect including batches in X.
@@
 
-We use deviance residuals as normalized values for visualization. FP: explain rational: K=0 so residuals capture the bio adjusting for batch. Let's check that deviance residuals look ok.
+We use deviance residuals as normalized values for visualization.
+We use K = 0 so residuals capture the biology while adjusting for gene and cell level covariates, here we have only cell-level covariates (batches).
+Let's check that deviance residuals look ok.
 

### 2017-06-06 R/workflow.Rmd
add html

--- 2017-06-06 R/workflow.Rmd
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m9", "m6"))
+                           end.clus = c("m8", "m4"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 fn <- '../data/refit_zinbwave_slingshot.rda'
-if (run & !file.exists(fn)){
+if (runZinb & !file.exists(fn)){
   zinbList <- lapply(Kvec, function(k){
@@
 
-  lineages <- get_lineages(X, clus.labels = our_cl, start.clus = "m1",
-                           end.clus = c("m9", "m6"))
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
+                           end.clus = c("m8", "m4"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)

### 2017-06-07 R/workflow.Rmd
RSEC no merge but sub and seq

@@
 if (runClus & !file.exists(fn)){
-  print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
-                                  alphas = c(0.3), betas = c(0.9),
-                                  clusterFunction = "hierarchical01",
-                                  minSizes = 5,
-                                  ncores = NCORES,
+  ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
+                clusterFunction = "hierarchical01", minSizes=1,
+                ncores = 7,
+                subsampleArgs = list(resamp.num=100,
+                                     clusterFunction="kmeans",
+                                     clusterArgs=list(nstart=10)),
+                seqArgs = list(k.min=3, top.can=5), verbose=TRUE,
                 combineProportion = 0.7,
-                                  mergeMethod = "locfdr",
-                                  mergeCutoff = 0.01)))
+                mergeMethod = "none")
   save(ceObj, file = fn)
@@
 ```{r examineCombineMany}
-plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
+plotClusters(ceObj, colPalette = c(bigPalette, rainbow(200)))
 ```
@@
 FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context?
+I set eval=FALSE in the next chunk to skip the merging step, let me know if you would rather keep using it.
+And if we want to still use the merging step, would we want to include it in RSEC function arguments instead of separately?
 
-```{r examineMergeClusters}
+```{r examineMergeClusters,eval=FALSE}
 #re-does merging simpling to make plot 
 mergeClusters(ceObj, mergeMethod = "locfdr",
-              plotInfo = "mergeMethod", cutoff = 0.05)
+              plotInfo = "mergeMethod", cutoff = 0.01)
 ```
@@
      xlab = 'Component1', ylab = 'Component2')
+legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
+       fill = unique(col_clus), title = 'Sample')
 ```
@@
 plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
-     main = 'MDS W, color = our clusters', pch = 20,
+     main = 'MDS W, color = our new clusters', pch = 20,
      xlab = 'Component1', ylab = 'Component2')
-```
-
-```{r}
-table(data.frame(published = clus.labels,
-                 ours = primaryCluster(ceObj)))
+legend(x = 'bottomright', legend = names(pal), cex = .5,
+       fill = pal, title = 'Sample')
 ```
@@
 
+FP: Elizabeth, I did not find how to define the column annotation track in the plot below to have the same colors as in ceObj@clusterLegend[[1]].
+I tried to use arguments annColors and annCol from aheatmap as it is said in plotHeatmap documentation that for signature matrix arguments can be passed to aheatmap.
+But I got the error "The following arguments to aheatmap cannot be set by the user in plotHeatmap:Rowv,Colv,color,annCol,annColors".
+
 ```{r}
 sampleData <- data.frame(ours = primaryCluster(ceObj))
-
 plotHeatmap(assays(se)$normalizedValues,
@@
 
-The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=2, k=3, k=4 in W.
+```{r}
+table(data.frame(published = clus.labels,
+                 ours = primaryClusterNamed(ceObj)))
+```
+
+Cluster name | Description | Correspondence
+-------------|-------------|-----------------
+m1 | resting HBC | original 1, 5
+m2 | activated HBC | original 5
+m3 | GBC / immature neurons / MV 1 | original 2, 3, 11
+m4 | Sus | original 4, 7
+m5 | new | new
+m6 | new | new
+m7 | Neuron | original 9, 12
+m8 | Immature Neuron | original 10, 14
+m9 | Microvillus | original 15
+m10 | Immature Neuron | original 14
+
+The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=3, k=4 in W.
 
@@
 ```{r}
-Kvec <- c(2, 3, 4)
+Kvec <- c(3, 4)
 ```
@@
 ```{r slingshot_sup}
+# m11 - original 15 - Microvillus
+# m7, m8 - original 9, 12 - Neuron
+# m4 - original 4, 7 - Sus
+
 for (k in Kvec){
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m8", "m4"))
+                           end.clus = c("m11", "m7", "m8", "m4"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m8", "m4"))
+                           end.clus = c("m11", "m4"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)

### 2017-06-07 R/workflow.Rmd
update version of clusterExperiment

--- 2017-06-06 R/workflow.Rmd
@@
 title: "Worflow"
-author: "Workflow team"
-date: "06/06/2017"
+author: "Fanny Perraudeau"
+date: "05/25/2017"
 output: 
@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9006') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9007') stop("must have current develop version to avoid bugs")
 ```
@@
 ```{r}
-runZinb <- T
-runClus <- T
-NCORES <- 10
+runZinb = TRUE
+runClus = TRUE
+NCORES = 2
+```
+
+# TL;DR
+
+This document is a temptative of workflow with the following steps
+
+- Fit zinbwave with K = 50, X including the batches,   
+- Get W which should capture the biology,  
+- Use clusterExperiment on W,   
+- Get the cluster labels,   
+- Use slingshot on W from zinbwave and with cluster labels from clusterExperiment,  
+- Get lineages,  
+- Fit again zinbwave but with X including batches and lineages from slingshot,  
+- Get DE genes between lineages.
+
+Along the worflow, use deviance residuals as adjusted "counts".
+
+EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
+
+# Unnormalized data
+I am still using the data Davide gave me a while ago.
+And I think now it is not the same data Davide is using in its most recent vignette (that is zinb_clustering_over_k.Rmd).
+It should not be very important but it would be nice if we have a common dataset we can work on.
+
+EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file.
+For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article.
+This will be slightly different data from Russell's, which didn't use all of the samples.
+We can adjust that decision later, or just compare the samples that are the same.
+(Russell's clusterLabels are in the meta data)
+
+EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?
+
+```{r datain}
+#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
+core<-read.table("../data/oeCufflinkCountData_1000Var.txt",sep="\t",header=TRUE)
+core<-data.matrix(core)
+metadata<-read.table("../data/oeMetadata.txt",sep="\t",header=TRUE)
+# load("../data/Expt4c2b_filtdata.Rda")
+# load("../data/E4c2b_slingshot_wsforkelly.RData")
+```
+
+Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.
+
+```{r filtering}
+batch <- metadata$Batch
+qc <- metadata[,!names(metadata)%in% c("Batch","Experiment","clusterLabels")]
+clus.labels<-metadata[,"clusterLabels"]
+# vars <- rowVars(log1p(counts))
+# names(vars) <- rownames(counts)
+# vars <- sort(vars, decreasing = TRUE)
+# core <- counts[names(vars)[1:1000],]
+```
+
+We have `r ncol(core)` cells.
+
+```{r lookatdata}
+dim(core)
+core[1:3, 1:3]
+```
+
+Cells have been processed in `r length(unique(batch))` different batches
+
+```{r tabbatch}
+table(batch)
+```
+
+Cells have been clustered into `r length(unique(clus.labels))` different clusters
+
+```{r tabclus}
+table(clus.labels)
+```
+
+Batches are kind of confounded with the biology
+
+```{r}
+table(data.frame(batch = as.vector(batch),
+                 cluster = clus.labels))
 ```
@@
 
-FP: yes, I agree.
+```{r}
+compute.naive.residuals <- function(Y, zinb){
+  mu_hat = t(getMu(zinb)) 
+  pi_hat = t(getPi(zinb)) 
+  Y_hat = (1 - pi_hat) * mu_hat
+  Y - Y_hat 
+}
 
-```{r residuals}
+compute.pearson.residuals <- function(Y, zinb){
+  num = compute.naive.residuals(Y, zinb)
+  mu = t(getMu(zinb)) 
+  pi = t(getPi(zinb)) 
+  phi = matrix(rep(getPhi(zinb), ncol(Y)), ncol = ncol(Y))
+  var_hat = (1 - pi) * mu * (1 + mu * (phi + pi))
+  num / sqrt(var_hat)  
+}
+
 compute.zinb.loglik <- function(Y, zinb){
@@
 
-# TL;DR
-
-This document is a temptative of workflow with the following steps
-
-- Dimensionality reduction using zinbwave to get W which should capture the biology,
-- Cluster cells using clusterExperiment on W to get the cluster labels,   
-- Get lineage using slingshot on W and cluster labels from clusterExperiment,  
-- Get DE genes between lineages/clusters.
-
-Along the worflow, use deviance residuals as adjusted values.
-
-EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
-
-# Steps of the workflow
-## Create a SummarizedExperiment object
-
-Along the workflow, we want to use a unique SummarizedExperiement object carrying all the data we need.
-
-EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file.
-For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article.
-This will be slightly different data from Russell's, which didn't use all of the samples.
-We can adjust that decision later, or just compare the samples that are the same.
-(Russell's clusterLabels are in the meta data)
-
-EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?
-
-```{r datain}
-#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
-core <- read.table("../data/oeCufflinkCountData_1000Var.txt",
-                   sep = "\t", header = TRUE)
-core <- data.matrix(core)
-metadata <- read.table("../data/oeMetadata.txt", sep = "\t", header = TRUE)
-# symbol for samples missing from original clustering
-metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
-```
-
-Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.
-
-```{r batches}
-batch <- metadata$Batch
-```
-
-Cells have been processed in `r length(unique(batch))` different batches
+# 0. Residuals for visualization
+## Fit zinbwave X intercept, no batch
+Let's run zinbwave with K = 0 and X batches.
 
 ```{r}
-col_batch = rep(brewer.pal(9, "Set1"), 2)
-names(col_batch) = unique(batch)
-table(batch)
-```
-
-We have qc measures from the data
-
-```{r qc}
-qc <- metadata[, !names(metadata) %in% c("Batch", "Experiment", "clusterLabels")]
-head(qc, 2)
-```
-
-```{r clusterlabels}
-clus.labels <- metadata[, "clusterLabels"]
-```
-
-In published work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters
-
-```{r}
-col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
-col_clus <- col_clus[1:length(unique(clus.labels))]
-names(col_clus) <- sort(unique(clus.labels))
-table(clus.labels)
-```
-
-Batches are kind of confounded with the biology
-
-```{r}
-table(data.frame(batch = as.vector(batch),
-                 cluster = clus.labels))
-```
-
-We have `r ncol(core)` cells.
-
-```{r lookatdata}
-dim(core)
-core[1:3, 1:3]
-```
-
-Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, published labels, and quality control measures.
-
-```{r}
-se <- SummarizedExperiment(assays = list(rawCounts = core),
-                           colData = metadata)
-```
-
-## Dimensionality reduction adjusting for gene and cell level covariates
-
-To cluster and get lineages we want to reduce the dimension of the data.
-We are going to use zinbwave to do so.
-First, let's fit zinbwave with first K = 0 to compute normalized values (i.e.
-deviance residuals) adjusted for batches.
-We could also adjust for gene length or GC content here.
-We then fit zinbwave to get the dimensionality reduced matrix W specifying the number of dimension K = 50.
-Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE).
-For K = 0 and K = 50, we correct for batch effect including batches in X.
-
-```{r}
-fn0 <- '../data/zinb_k0_batch.rda'
-if (runZinb & !file.exists(fn0)){
-  mod <- model.matrix( ~ batch)
-  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES,
-                                     K = 0, X = mod)))
-  save(zinb0, file = fn0)
+fn = '../data/zinb_k0_batch.rda'
+if (runZinb){
+  mod = model.matrix( ~ batch)
+  print(system.time(zinb0 <- zinbFit(core, K = 0, X = mod, ncores = NCORES)))
+  save(zinb0, file = fn)
 }else{
-  load(fn0)
-}
-
-fn50 <- '../data/zinb_k50_batch.rda'
-if (runZinb & !file.exists(fn50)){
-  mod <- model.matrix( ~ batch)
-  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES,
-                                      K = 50, X = mod)))
-  save(zinb50, file = fn50)
-}else{
-  load(fn50)
+  load(fn)
 }
@@
 
-### Normalized values
-
-We use deviance residuals as normalized values for visualization. FP: explain rational: K=0 so residuals capture the bio adjusting for batch. Let's check that deviance residuals look ok.
+We use deviance residuals for visualization. Let's check that deviance residuals look ok.
 
 ```{r}
-res <- compute.deviance.residuals(core, zinb0)
+res = compute.deviance.residuals(core, zinb0)
 res[1:3,1:3]
@@
 
-Boxplot of the normalized values for each cell. It seems that correction for batches is ok.
+PCA on the residuals where color are for batches on the left and previously found clusters on the right.
 
 ```{r}
-res_order <- res[, order(as.numeric(batch))]
-col_order <- as.numeric(batch)[order(as.numeric(batch))]
-boxplot(res_order, main='Boxplot of normalized values\ncolor=batch',
-        col = col_order, staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
-```
-
-PCA on the normalized values where color are for batches on the left and previously found clusters on the right. We want no clustering on the left side and clustering on the right side.
-
-```{r}
-pca <- prcomp(t(res))
+pca = prcomp(t(res))
 par(mfrow = c(1,2))
-plot(pca$x, col = col_batch[batch], pch = 20,
-     main="PCA of normalized values\ncolor=batch")
-plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20,
-     main = "PCA of normalized values\ncolor=cluster")
+plot(pca$x, col = batch, pch = 19,
+     main="PCA of residuals\ncolor=batch")
+plot(pca$x, col = clus.labels, pch = 19,
+     main = "PCA of residuals\ncolor=cluster")
 par(mfrow = c(1,1))
@@
 
-Let's add the normalized values as an assay dataset in our SummarizedExperiment object.
+Boxplot of the residuals for each cell.
 
 ```{r}
-assays(se)[['normalizedValues']] <- res
+res_order = res[, order(as.numeric(batch))]
+col_order = as.numeric(batch)[order(as.numeric(batch))]
+boxplot(res_order, main='Boxplot of residuals\ncolor=batch', col = col_order,         staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
 ```
 
-### W
+Deviance residuals seem ok.
 
-Let's check that performing MDS on W we have something coherent with published clusters.
+# 1. zinbwave with raw counts
+
+Let's run zinbwave with K = 50 and X including batches.
 
 ```{r}
-W <- getW(zinb50)
-d <- dist(W)
-fit <- cmdscale(d, eig = TRUE, k = 2)
-plot(fit$points, col = col_clus[as.character(clus.labels)], main = 'MDS', pch = 20,
-     xlab = 'Component 1', ylab = 'Component 2')
-legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
-       fill = unique(col_clus), title = 'Sample')
+fn = '../data/zinb_k50_batch.rda'
+if (runZinb){
+  mod = model.matrix( ~ batch)
+  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50,
+                                      X = mod)))
+  save(zinb50, file = fn)
+}else{
+  load(fn)
+}
 ```
 
-Let's add W to the colData of our SummarizedExperiment object.
-
-```{r}
-W <- data.frame(W)
-colnames(W) <- paste0('W', 1:ncol(W))
-colData(se) <- cbind(colData(se), W)
-```
-
-FP: what do you think of implemented a "compute_normalized_values" argument to zinbFit for summarizedExperiment objects where zinbFit would take as input a SummarizedExperiment object and return a summarizedExperiment object with slots added for normalized values and W?
-
-## Cluster of cells
-
-We use clusterExperiment with W.
+# 2. clusterExperiment with W
 
@@
 
-FP: What about the following call to RSEC?
-
 ```{r rsec_50}
-W <- colData(se)[, grepl('^W', colnames(colData(se)))]
-W <- as.matrix(W)
-fn <- '../data/RSEC_W.rda'
-if (runClus & !file.exists(fn)){
-  print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
-                                  alphas = c(0.3), betas = c(0.9),
-                                  clusterFunction = "hierarchical01",
-                                  minSizes = 5,
-                                  ncores = NCORES,
-                                  combineProportion = 0.7,
-                                  mergeMethod = "locfdr",
-                                  mergeCutoff = 0.01)))
+fn = '../data/clustExp_W_k50.rda'
+metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
+seObj<-SummarizedExperiment(t(W),colData=metadata)
+if(runClus) {
+ print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
+                      clusterFunction = "hierarchical01", minSizes=1,
+                      ncores = NCORES, isCount=FALSE,
+					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01
+                      subsampleArgs = list(resamp.num=100,
+                                           clusterFunction="kmeans",
+                                           clusterArgs=list(nstart=10)),
+                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE))
 save(ceObj, file = fn)
-}else{
+}
+else{
 	load(fn)
@@
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
-```
-
-```{r plotcoclust}
 plotCoClustering(ceObj)
-```
-
-```{r}
 table(primaryClusterNamed(ceObj))
@@
 
-FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context?
-
 ```{r examineMergeClusters}
 #re-does merging simpling to make plot 
-mergeClusters(ceObj, mergeMethod = "locfdr",
-              plotInfo = "mergeMethod", cutoff = 0.05)
+mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
 ```
@@
 ```{r heatmap }
-plotHeatmap(ceObj, clusterSamplesData = "dendrogramValue", 
-            breaks = .99)
+plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
 ```
@@
 ```{r compareClusters}
-plot(fit$points, col = col_clus[as.character(clus.labels)],
-     main = 'MDS W, color = published clusters', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
+plot(getW(zinb50), col=clus.labels, main = 'color = previous')
+plot(getW(zinb50), col=as.factor(primaryCluster(ceObj)), main = 'color = new')
+table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
+```
+
+## Original ClusterExperiment code (in Steps)
+I use the code of Davide from zinb_clutering_over_k.Rmd
+
+```{r consensus_50, eval=FALSE}
+## matching the parameters used by Russell
+W = zinb50@W
+fn = '../data/clustExp_W_k50.rda'
+if(runClus) {
+  cl_res <- clusterMany(t(W), ks = 4:15, alphas = c(0.1), betas = 0.8,
+                        clusterFunction = "hierarchical01", minSizes=1,
+                        sequential = TRUE, subsample = TRUE, ncores = 7,
+                        subsampleArgs = list(resamp.num=100,
+                                             clusterFunction="kmeans",
+                                             clusterArgs=list(nstart=10)),
+                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
+  save(cl_res, file = fn)
+} else {
+  load(fn)
+}
+```
+
+```{r combine_50, eval=FALSE}
+combined <- combineMany(cl_res, proportion = 0.7)
+plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))
+plotCoClustering(combined)
+table(primaryClusterNamed(combined))
+sum(primaryCluster(combined) == -1)
+```
+
+```{r new_clusters, eval=FALSE}
+plot(getW(zinb50), col=clus.labels, main = 'color = previous')
+plot(getW(zinb50), col=as.factor(primaryCluster(combined)), main = 'color = new')
+table(data.frame(previous = clus.labels, new = primaryCluster(combined)))
+```
+
+```{r merged, eval=FALSE}
+combined <- makeDendrogram(combined)
+plotDendrogram(combined)
+merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)
+```
+
+Just to see what it looks like, let's look at the heatmap on the W (number of cells x K=50).
+Now, we would like to plot a heatmap using the residuals (number of cells x number of genes).
+When I call plotHeatmap using argument visualizeData = residuals_matrix, it does not work and says that if give separate visualizeData, must be of same dimensions as assay(data).
+
+```{r heatmapOld, eval=FALSE}
+plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
+```
+
+So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
+```{r, eval=FALSE}
+palDF = merged@clusterLegend[[1]]
+pal = palDF[, 'color']
+names(pal) = palDF[, 'name']
+
+heatmap.2(res, Colv = merged@dendro_samples,
+          scale = 'none', trace = 'none', 
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          ColSideColors = pal[primaryClusterNamed(merged)],
+          labCol = '', main = 'Deviance Residuals, 1000 most variable genes')
+legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
+       horiz = T, cex = 0.5)
+```
+
+# 3. slingshot with W and cluster labels
+I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.
+
+```{r slingshot}
+fn = '../data/zinb_k4_batch.rda'
+if (runZinb){
+  cl_labs = primaryCluster(merged)
+  batch_red = as.vector(batch)
+  batch_red = batch_red[cl_labs != -1]
+  cl_labs = cl_labs[cl_labs != -1]
+  mod = model.matrix( ~ batch_red)
+  zinb4 = zinbFit(core[,primaryCluster(merged) != -1],
+                  X = mod, K = 4, ncores = NCORES)
+  save(zinb4, file = fn)
+}else{
+  cl_labs = primaryCluster(merged)
+  batch_red = as.vector(batch)
+  batch_red = batch_red[cl_labs != -1]
+  cl_labs = cl_labs[cl_labs != -1]
+  mod = model.matrix( ~ batch_red)
+  load(fn)
+}
 ```
@@
 ```{r}
-palDF <- ceObj@clusterLegend[[1]]
-pal <- palDF[, 'color']
-names(pal) <- palDF[, 'name']
-plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
-     main = 'MDS W, color = our clusters', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
-```
+X <- zinb4@W
 
-```{r}
-table(data.frame(published = clus.labels,
-                 ours = primaryCluster(ceObj)))
-```
+lineages <- get_lineages(X, clus.labels = cl_labs, start.clus = 1)
+curves <- get_curves(X, clus.labels = cl_labs, lineages = lineages)
+plot_curves(X, cl_labs, curves, col.clus = unique(cl_labs))
 
-So, let's look at a heatmap on normalized values.
-
-```{r}
-sampleData <- data.frame(ours = primaryCluster(ceObj))
-
-plotHeatmap(assays(se)$normalizedValues,
-            main = 'Normalized values, 1000 most variable genes',
-            clusterSamplesData = ceObj@dendro_samples,
-            sampleData = as.matrix(sampleData))
-```
-
-## Get lineages
-
-The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=2, k=3, k=4 in W.
-
-FP: We need to work on the cluster coloring (!) and the matching with the clusters you found for the paper with Russell, maybe by changing the parameters in clusterMany?
-
-```{r}
-Kvec <- c(2, 3, 4)
-```
-
-### Use previous W, no re-fitting 
-#### Unsupervised
-
-```{r slingshot_unsup}
-our_cl <- as.factor(primaryClusterNamed(ceObj))
-cl = our_cl[our_cl != "-1"]
-for (k in Kvec){
-  X <- W[our_cl != "-1", 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
 print(lineages$lineage1)
@@
 print(lineages$lineage4)
-}
 ```
 
-#### Supervised
+# 4. zinbwave with lineages
 
-```{r slingshot_sup}
-for (k in Kvec){
-  X <- W[our_cl != "-1", 1:k]
+Which clusters would it make sense to test one versus the other? Would it be interesting to do two by two comparison between clusters next to each other in the different lineages?
 
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m8", "m4"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
+# 5. DE genes with betas
 
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-}
-```
-
-### Re-fitting zinbwave
-#### Unsupervised
-
-```{r refit_zinb}
-mod <- model.matrix(~ batch[our_cl != "-1"])
-fn <- '../data/refit_zinbwave_slingshot.rda'
-if (runZinb & !file.exists(fn)){
-  zinbList <- lapply(Kvec, function(k){
-    zinbFit(core[, our_cl != "-1"], X = mod, K = k, ncores = NCORES)
-  })
-  save(zinbList, file = fn)
-}else{
-  load(fn)
-}
-```
-
-```{r slingshot_unsup_refit}
-for(k in Kvec) {
-  X <- getW(zinbList[[k - 1]])[, 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-}
-```
-
-#### Supervised
-
-```{r slingshot_sup_refit}
-for(k in Kvec){
-  X <- getW(zinbList[[k - 1]])[, 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m8", "m4"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-}
-```
-
-## DE analysis
-
-Here is the kind of plots we want to be able to have
+DE analysis using ZINB-Wave has not been implemented yet and we are still thinking about it.
+Would it make sense just to test the betas between clusters making the assumption that the genes are independent?
+Below is just a visualization of what we could have at the end of the worflow: a heatmap of the deviance residuals for DE genes between clusters.
 
 ```{r}
-de <- read.csv('../data/oe_markers.txt', 
-              stringsAsFactors = F, header = F)
-de <- de$V1
-nv <- assays(se)$normalizedValues
-plotHeatmap(nv[rownames(nv) %in% de, ],
-            main = 'Normalized values, 1000 most variable genes',
-            clusterSamplesData = ceObj@dendro_samples,
-            sampleData = as.matrix(sampleData$ours))
+de = read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
+de = de$V1
+
+breaks = setBreaks(res[rownames(res) %in% de, ], 0.99)
+heatmap.2(res[rownames(res) %in% de, ], breaks = breaks,
+          Colv = merged@dendro_samples,
+          scale = 'none', trace = 'none', 
+          col = colorRampPalette(c('blue', 'yellow'))(51),
+          ColSideColors = pal[primaryClusterNamed(merged)],
+          labCol = '', main = 'Deviance Residuals, DE genes (Russell)')
+legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
+       horiz = T, cex = 0.5)
 ```

### 2017-06-07 R/workflow.Rmd

--- 2017-06-07 R/workflow.Rmd
@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9006') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9008') stop("must have current develop version to avoid bugs")
 ```
@@
 runClus <- T
-NCORES <- 10
+NCORES <- 7
 ```
@@
 
-EAP: Fanny, we should really be using the RSEC function for all the steps of the clustering.
-I've put in the code, but I set `eval=FALSE` because I don't have the data to test it.
-I also am not sure that we want to set the options for `subsampleArgs` and `seqArgs` since those are suppose to be sort of esoteric concerns...
+FP: What about the following call to RSEC?
 
-FP: What about the following call to RSEC?
+EP: I updated it to work on a SE object so that it has the meta data. If you have a summarized experiment object with W already, you could use that as long as assay(seObj) gives W.
 
@@
 if (runClus & !file.exists(fn)){
-  ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
+	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
+  seObj<-SummarizedExperiment(t(W),colData=metadata)
+ print(system.time(  ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
                 clusterFunction = "hierarchical01", minSizes=1,
-                ncores = 7,
+                ncores = NCORES, isCount=FALSE,
                 subsampleArgs = list(resamp.num=100,
@@
                 mergeMethod = "none")
+				))
   save(ceObj, file = fn)
@@
 
+EP: I don't think the merging step on the W makes a whole lot of sense -- the method is irrelevant.
+The merging is based on calculating the % of genes found significant (the specific method is arbitrary).
+The best thing would be to replace the W with residuals in the assay of `ceObj` (or whatever data that you will do the DE on for the time stuff below), and then run the merging step on that data.
+I'm not particularly fond of `locfdr`.
+It was probably the method that gave the best merging to Russell and Diya.
+You'd really have to run `mergeClusters` setting `plotInfo="all"` and look at the results and decide both the cutoff level and the method.
+
+EP: Also, if you don't save the output of `mergeClusters` it doesn't update `ceObj`.
+I was calling it for just the resulting plots, since it was already merged in RSEC above.
+I've changed to code to update ceObj below.
+
 ```{r examineMergeClusters,eval=FALSE}
 #re-does merging simpling to make plot 
-mergeClusters(ceObj, mergeMethod = "locfdr",
+#something like:
+#assay(ceObj)<- ...
+# if that replacement data should be considered on the transformed scale in plots, etc, the transformation function should be fixed as well:
+#transformation(ceObj)<- ...
+ceObj<-mergeClusters(ceObj, mergeMethod = "locfdr",
               plotInfo = "mergeMethod", cutoff = 0.01)
@@
 
+EP: Fanny, you would need to use the argument 'clusterLegend'.
+That argument takes either the format of aheatmap (list with each element a *named* vector of colors) or the format of the clusterExperiment object (i.e.
+list with each element a matrix with columns for `name` and `color`).
+So I think the following code will run, though it might need the list to have names...
+
 ```{r}
@@
             clusterSamplesData = ceObj@dendro_samples,
-            sampleData = as.matrix(sampleData))
+            sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
 ```
@@
 FP: We need to work on the cluster coloring (!) and the matching with the clusters you found for the paper with Russell, maybe by changing the parameters in clusterMany?
+
+EP: The function plotClusters aligns the clusters and can return the aligned coloring between clusterings.
+There is an option to return an updated CE object that changes the colors in clusterLegend to match the new colors found by aligning the clusters (`resetNames` and `resetColors` arguments).
+These haven't been extensively tested in real life because I don't think we've used them much.
+But it would be good to demonstrate!
 

### 2017-06-07 R/workflow.Rmd
updated code to plotHeatmap

@@
 
+But an easier fix to the code would be to set `visualizeData` option. I haven't tested this because I don't have the objects need run, so let me know if there is error.
+
 ```{r}
-sampleData <- data.frame(ours = primaryCluster(ceObj))
-plotHeatmap(assays(se)$normalizedValues,
-            main = 'Normalized values, 1000 most variable genes',
-            clusterSamplesData = ceObj@dendro_samples,
-            sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
+# sampleData <- data.frame(ours = primaryCluster(ceObj))
+# plotHeatmap(assays(se)$normalizedValues,
+#             main = 'Normalized values, 1000 most variable genes',
+#             clusterSamplesData = ceObj@dendro_samples,
+#             sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
+# easier fix:
+plotHeatmap(ceObj, visualizeData=assays(se)$normalizedValues,
+whichClusters="primary",
+            main = 'Normalized values, 1000 most variable genes')
+
 ```

### 2017-06-07 R/workflow.Rmd
merge doc with elizabeth updates

--- 2017-06-07 R/workflow.Rmd
@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9008') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9006') stop("must have current develop version to avoid bugs")
 ```
@@
 runClus <- T
-NCORES <- 7
+NCORES <- 10
 ```
@@
 
+EAP: Fanny, we should really be using the RSEC function for all the steps of the clustering.
+I've put in the code, but I set `eval=FALSE` because I don't have the data to test it.
+I also am not sure that we want to set the options for `subsampleArgs` and `seqArgs` since those are suppose to be sort of esoteric concerns...
+
 FP: What about the following call to RSEC?
-
-EP: I updated it to work on a SE object so that it has the meta data. If you have a summarized experiment object with W already, you could use that as long as assay(seObj) gives W.
 
@@
 if (runClus & !file.exists(fn)){
-	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
-  seObj<-SummarizedExperiment(t(W),colData=metadata)
- print(system.time(  ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
+  ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
                 clusterFunction = "hierarchical01", minSizes=1,
-                ncores = NCORES, isCount=FALSE,
+                ncores = 7,
                 subsampleArgs = list(resamp.num=100,
@@
                 mergeMethod = "none")
-				))
   save(ceObj, file = fn)
@@
 
-EP: I don't think the merging step on the W makes a whole lot of sense -- the method is irrelevant.
-The merging is based on calculating the % of genes found significant (the specific method is arbitrary).
-The best thing would be to replace the W with residuals in the assay of `ceObj` (or whatever data that you will do the DE on for the time stuff below), and then run the merging step on that data.
-I'm not particularly fond of `locfdr`.
-It was probably the method that gave the best merging to Russell and Diya.
-You'd really have to run `mergeClusters` setting `plotInfo="all"` and look at the results and decide both the cutoff level and the method.
-
-EP: Also, if you don't save the output of `mergeClusters` it doesn't update `ceObj`.
-I was calling it for just the resulting plots, since it was already merged in RSEC above.
-I've changed to code to update ceObj below.
-
 ```{r examineMergeClusters,eval=FALSE}
 #re-does merging simpling to make plot 
-#something like:
-#assay(ceObj)<- ...
-# if that replacement data should be considered on the transformed scale in plots, etc, the transformation function should be fixed as well:
-#transformation(ceObj)<- ...
-ceObj<-mergeClusters(ceObj, mergeMethod = "locfdr",
+mergeClusters(ceObj, mergeMethod = "locfdr",
               plotInfo = "mergeMethod", cutoff = 0.01)
@@
 
-EP: Fanny, you would need to use the argument 'clusterLegend'.
-That argument takes either the format of aheatmap (list with each element a *named* vector of colors) or the format of the clusterExperiment object (i.e.
-list with each element a matrix with columns for `name` and `color`).
-So I think the following code will run, though it might need the list to have names...
-
 ```{r}
@@
             clusterSamplesData = ceObj@dendro_samples,
-            sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
+            sampleData = as.matrix(sampleData))
 ```
@@
 
+From what I understand, start original cluster are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
+
 The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=3, k=4 in W.
-
-FP: We need to work on the cluster coloring (!) and the matching with the clusters you found for the paper with Russell, maybe by changing the parameters in clusterMany?
-
-EP: The function plotClusters aligns the clusters and can return the aligned coloring between clusterings.
-There is an option to return an updated CE object that changes the colors in clusterLegend to match the new colors found by aligning the clusters (`resetNames` and `resetColors` arguments).
-These haven't been extensively tested in real life because I don't think we've used them much.
-But it would be good to demonstrate!
 

### 2017-06-07 R/workflow.Rmd

--- 2017-06-07 R/workflow.Rmd
@@
 #something like:
-#assay(ceObj)<- ...
+#assay(ceObj)
 # if that replacement data should be considered on the transformed scale in plots, etc, the transformation function should be fixed as well:
-#transformation(ceObj)<- ...
+#transformation(ceObj)
 ceObj<-mergeClusters(ceObj, mergeMethod = "locfdr",
@@
 m10 | Immature Neuron | original 14
+
+From what I understand, start original cluster are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
 

### 2017-06-08 R/workflow.Rmd
update version needed for CE

--- 2017-06-07 R/workflow.Rmd
@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9008') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9009') stop("must have current develop version to avoid bugs")
 ```

### 2017-06-08 R/workflow.Rmd
slingshot wi and wo zinbwave

--- 2017-06-07 R/workflow.Rmd
@@
 author: "Workflow team"
-date: "06/06/2017"
+date: "`r format(Sys.time(), '%B %d, %Y')`"
 output: 
@@
 
+EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
+
 EAP: if we are going to ultimately use these functions, it would be good to add them to the package for zinbwave.
 
-FP: yes, I agree.
+FP: yes, I agree. Davide, we wait for zinbwave to be on bioconductor before to push changes, right?
 
@@
 
-# TL;DR
+# Steps of the workflow
 
-This document is a temptative of workflow with the following steps
+We propose a worklow to analyze single cell RNA-Seq with the following steps
 
@@
 
-EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
+```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
+knitr::include_graphics('../vignettes/workflow.png')
+```
 
-# Steps of the workflow
-## Create a SummarizedExperiment object
+## 1. Create a SummarizedExperiment object
 
@@
 
-In published work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters
+In original work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters
 
@@
 
-Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, published labels, and quality control measures.
+Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, original labels, and quality control measures.
 
@@
 
-## Dimensionality reduction adjusting for gene and cell level covariates
+## 2. Dimensionality reduction adjusting for gene and cell-level covariates
 
@@
 
-### W
+### Dimensionality reduction
 
-Let's check that performing MDS on W we have something coherent with published clusters.
+Let's check that performing MDS on W we have something coherent with original clusters.
 
@@
 
-## Cluster of cells
+## 2. Clustering of the cells
 
@@
 if (runClus & !file.exists(fn)){
-	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
+  #symbol for samples missing from original clustering
+	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 
   seObj<-SummarizedExperiment(t(W),colData=metadata)
- print(system.time(  ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
+  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
+                                  betas = 0.8,
                 clusterFunction = "hierarchical01", minSizes=1,
@@
                 combineProportion = 0.7,
-                mergeMethod = "none")
-				))
+                mergeMethod = "none")))
   save(ceObj, file = fn)
@@
 ```{r examineCombineMany}
-plotClusters(ceObj, colPalette = c(bigPalette, rainbow(200)))
+plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
 ```
@@
 
+FP: Ha ok, good to know. I'll keep the eval=FALSE for the moment.
+
 ```{r examineMergeClusters,eval=FALSE}
@@
 
+FP: Not sure we should show the following heatmap using the W, we should rather use the normalized values (following heatmap).
+
 ```{r heatmap }
@@
             breaks = .99)
-```
-
-```{r compareClusters}
-plot(fit$points, col = col_clus[as.character(clus.labels)],
-     main = 'MDS W, color = published clusters', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
-legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
-       fill = unique(col_clus), title = 'Sample')
-```
-
-```{r}
-palDF <- ceObj@clusterLegend[[1]]
-pal <- palDF[, 'color']
-names(pal) <- palDF[, 'name']
-plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
-     main = 'MDS W, color = our new clusters', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
-legend(x = 'bottomright', legend = names(pal), cex = .5,
-       fill = pal, title = 'Sample')
 ```
@@
 
+FP: it seems great to me, what do you think?
+
 ```{r}
@@
 # easier fix:
-plotHeatmap(ceObj, visualizeData=assays(se)$normalizedValues,
-whichClusters="primary",
-            main = 'Normalized values, 1000 most variable genes')
+plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
+            whichClusters = "primary",
+            clusterSamplesData = "dendrogramValue",
+            main = 'Normalized values, 1000 most variable genes',
+            breaks = 0.99)
 
@@
 
-## Get lineages
+```{r compareClusters}
+plot(fit$points, col = col_clus[as.character(clus.labels)],
+     main = 'MDS W, color = original clusters', pch = 20,
+     xlab = 'Component1', ylab = 'Component2')
+legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
+       fill = unique(col_clus), title = 'Sample')
+```
 
 ```{r}
-table(data.frame(published = clus.labels,
-                 ours = primaryClusterNamed(ceObj)))
+palDF <- ceObj@clusterLegend[[1]]
+pal <- palDF[, 'color']
+names(pal) <- palDF[, 'name']
+pal["-1"] = "transparent"
+plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
+     main = 'MDS W, color = our new clusters', pch = 20,
+     xlab = 'Component1', ylab = 'Component2')
+legend(x = 'bottomright', legend = names(pal), cex = .5,
+       fill = pal, title = 'Sample')
 ```
 
-Cluster name | Description | Correspondence
--------------|-------------|-----------------
-m1 | resting HBC | original 1, 5
-m2 | activated HBC | original 5
-m3 | GBC / immature neurons / MV 1 | original 2, 3, 11
-m4 | Sus | original 4, 7
-m5 | new | new
-m6 | new | new
-m7 | Neuron | original 9, 12
-m8 | Immature Neuron | original 10, 14
-m9 | Microvillus | original 15
-m10 | Immature Neuron | original 14
+## 4. Pseudotime ordering
 
-From what I understand, start original cluster are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
+The goal of this section is to see if we need to refit zinbwave when we want to run slingshot.
+We first run slingshot on the W used by clusterExperiment.
+In the second part of this section, we fit zinbwave on the matrix of counts where the unassigned cells have been removed.
+For each part (without or with refitting zinbwave), we run slingshot in the supervised and unsupervised mode and try k=3, k=4, k=5 dimensions in W.
 
-The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=3, k=4 in W.
+From what I understand, start original clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
+Additionally, we want the GBC cluster to be a junction before the differentiation between Microvillus and Neuron.
+The correspondance with the original clusters is as follow
+
+```{r}
+table(data.frame(original = clus.labels, ours = primaryClusterNamed(ceObj)))
+```
+
+Cluster name | Description | Correspondence | Color
+-------------|-------------|-----------------|---------
+c1 | HBC | original 1, 5 | darkblue
+c2 | new and small | new and small | green
+c3 | new and small | new and small | red
+c4 | GBC / immature neurons / MV 1 | original 2, 3, 11, 14 | orange
+c5 | Sus | original 4, 7 | darkpurple
+c6 | Neuron | original 9, 12 | brown
+c7 | Immature Neuron | original 10, 14 | cian 
+c8 | Immature Neuron | original 14 | lightpurple
+c9 | Microvillus | original 15 | grayblue
 
@@
 
+FP: With all the improvements made, I think we simply have the correct colors now.
+
 ```{r}
-Kvec <- c(3, 4)
+Kvec <- c(3, 4, 5)
 ```
 
-### Use previous W, no re-fitting 
+### Use previous W
+
+The input of slingshot is the W used for clusterExperiment where the number of dimensions is reduced to k where k in (3, 4, 5) here.
+
 #### Unsupervised
 
+K = 3 does not seem very good to me: Sus is not an end cluster, GBC is an end cluster.
+
+K = 4 is better, slingshot finds the end clusters but there is a spurious end cluster.
+
+K = 5 does not seem great to me: GBC is an end cluster and Sus and Microvillus are in the same lineage.
+
 ```{r slingshot_unsup}
-our_cl <- as.factor(primaryClusterNamed(ceObj))
+our_cl <- primaryClusterNamed(ceObj)
 cl = our_cl[our_cl != "-1"]
+pal = pal[names(pal) != '-1']
+```
+
+```{r}
 for (k in Kvec){
@@
 
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
   print(lineages$lineage4)
+  print(lineages$lineage5)
 }
@@
 
+K = 3 finds GBC as an end cluster (that I did not specify in the end.clus!).
+
+K = 4 Yeah! it seems that it is what we want even if we still have a spurius end cluster and GBC not really at the junction.
+
+K = 5 Yeah! even if GBC not really at the junction.
+
 ```{r slingshot_sup}
-# m11 - original 15 - Microvillus
-# m7, m8 - original 9, 12 - Neuron
-# m4 - original 4, 7 - Sus
-
 for (k in Kvec){
@@
 
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m11", "m7", "m8", "m4"))
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
+                           end.clus = c("c5", "c6", "c9"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 #### Unsupervised
+
+K = 3 is better than when we did not refit zinbwave but still not perfect: Sus in all the clusters. GBC not really at the junction.
+
+K = 4 good even if GBC not really at the junction.
+
+K = 5 not great GBC is an end cluster.
 
@@
 for(k in Kvec) {
-  X <- getW(zinbList[[k - 1]])[, 1:k]
+  X <- getW(zinbList[[k - 2]])[, 1:k]
 
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 
+K = 3 Yeah! close to perfection.
+
+K = 4 good even if GBC not really at the junction.
+
+K = 5 no, GBC is an end cluster.
+
 ```{r slingshot_sup_refit}
 for(k in Kvec){
-  X <- getW(zinbList[[k - 1]])[, 1:k]
+  X <- getW(zinbList[[k - 2]])[, 1:k]
 
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
-                           end.clus = c("m11", "m4"))
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
+                           end.clus = c("c5", "c6", "c9"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 
-## DE analysis
+CONCLUSION: K = 5 is never great as GBC is generally an end cluster. K = 4 is ok for all the methods and a bit better when zinbwave is refitted. K = 3 when refitting and supervized is good.
 
-Here is the kind of plots we want to be able to have
+It seems to me that using slingshot on W without re-fitting zinbwave with k = 4 gives good results where supervized mode is slightly better than unsupervized.
+It is just a one shot example and we should obviously not make a general conclusion, but I think that for the purpose of the workflow it is fine to use slingshot without refitting zinbwave.
+We should write a note to the user that it is better to refit zinbwave to have more power.
+
+## 5. DE analysis
+
+Here is the kind of plots we want to present
 
 ```{r}
-de <- read.csv('../data/oe_markers.txt', 
-              stringsAsFactors = F, header = F)
+de <- read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
 de <- de$V1
-nv <- assays(se)$normalizedValues
-plotHeatmap(nv[rownames(nv) %in% de, ],
+plotHeatmap(ceObj, 
+            visualizeData = assays(se)$normalizedValues[rownames(se) %in% de, ],
+            clusterSamplesData = "dendrogramValue",
+            whichClusters = "primary",
             main = 'Normalized values, 1000 most variable genes',
-            clusterSamplesData = ceObj@dendro_samples,
-            sampleData = as.matrix(sampleData$ours))
+            breaks = 0.99)
 ```
+
+FP: Kelly, is it you who did the DE analysis for Russell paper? If yes, what tool did you use? On what data? The full quantile normalized counts? Do you have code available?
 

### 2017-06-09 R/workflow.Rmd
changes to plotHeatmap code

@@
 
-```{r heatmap }
+EP: Agreed
+
+```{r heatmap, eval=FALSE}
 plotHeatmap(ceObj, clusterSamplesData = "dendrogramValue", 
@@
 
+EP: We should be careful, because the default in plotHeatmap is to plot the 500 most variable genes (maybe a slightly paternalistic default).
+I've changed it to `all` in the code here.
+I've also added the plotting of the batch, experiment, and Russell's original clusters.
+We may not want to keep all of them, but probably at least Russell's clusters for comparison.
+
 ```{r}
@@
 # easier fix:
+origClusterColors<-bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
+experimentColors<-bigPalette[1:nlevels(colData(ceObj)$Experiment)]
+batchColors<-bigPalette[1:nlevels(colData(ceObj)$Batch)]
+metaColors<-list("Experiment"=experimentColors,"Batch"=batchColors,"clusterLabels"=origClusterColors)
+
 plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
-            whichClusters = "primary",
+            whichClusters = "primary",clusterFeaturesData="all",
             clusterSamplesData = "dendrogramValue",
+			sampleData=c("clusterLabels","Batch","Experiment"), clusterLegend=metaColors, annLegend=FALSE,
             main = 'Normalized values, 1000 most variable genes',

### 2017-06-11 R/workflow.Rmd
version control for zinbwave

@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9009') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9008') stop("must have current develop version to avoid bugs")
+if(packageVersion("zinbwave")<'0.99.4.1') stop("must have current develop version to avoid bugs")
 ```
+
+FP: Elizabeth, I could not install the last version 1.3.0.9009 of clusterExperiment. The error is
+
+checking DESCRIPTION meta-information ... ERROR
+Malformed Depends or Suggests or Imports or Enhances field.
+Offending entries:
+  MAST
+BiocStyle
+Entries must be names of packages optionally followed by '<=' or '>=',
+white space, and a valid version number in parentheses.
 
@@
 EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
-
-EAP: if we are going to ultimately use these functions, it would be good to add them to the package for zinbwave.
-
-FP: yes, I agree. Davide, we wait for zinbwave to be on bioconductor before to push changes, right?
-
-```{r residuals}
-compute.zinb.loglik <- function(Y, zinb){
-  mu = t(getMu(zinb)) 
-  theta = getTheta(zinb)
-  theta_mat = matrix(rep(theta, ncol(Y), ncol = ncol(Y)))
-  pi = t(getPi(zinb))
-  log( pi * (Y == 0) + (1 - pi) * dnbinom(Y, size = theta, mu = mu) )
-}
-
-compute.deviance.residuals <- function(Y, zinb){
-  mu_hat = t(getMu(zinb)) 
-  pi_hat = t(getPi(zinb)) 
-  Y_hat = (1 - pi_hat) * mu_hat
-  ll = compute.zinb.loglik(Y, zinb)
-  sign = 1*(Y - Y_hat > 0)
-  sign[sign == 0] = -1
-  sign * sqrt(-2 * ll)
-}
-```
 
@@
 ```{r}
-fn0 <- '../data/zinb_k0_batch.rda'
-if (runZinb & !file.exists(fn0)){
+fn <- '../data/zinb_batch.rda'
+if (runZinb & !file.exists(fn)){
   mod <- model.matrix( ~ batch)
-  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES,
-                                     K = 0, X = mod)))
-  save(zinb0, file = fn0)
+  print(system.time(zinb <- zinbDimRed(se, K = 50, X = mod, residuals = T,
+                                       normalizedValues = T)))
+  save(zinb, file = fn)
 }else{
-  load(fn0)
-}
-
-fn50 <- '../data/zinb_k50_batch.rda'
-if (runZinb & !file.exists(fn50)){
-  mod <- model.matrix( ~ batch)
-  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES,
-                                      K = 50, X = mod)))
-  save(zinb50, file = fn50)
-}else{
-  load(fn50)
+  load(fn)
 }
@@
 ```{r}
-res <- compute.deviance.residuals(core, zinb0)
+res <- computeDevianceResiduals(zinb, core)
 res[1:3,1:3]

### 2017-06-11 R/workflow.Rmd
clean up some comments

@@
 ```{r}
-res <- computeDevianceResiduals(zinb, core)
+res <- assays(zinb)[['normalizedValues']]
 res[1:3,1:3]
@@
 
-Let's add the normalized values as an assay dataset in our SummarizedExperiment object.
-
-```{r}
-assays(se)[['normalizedValues']] <- res
-```
-
 ### Dimensionality reduction
@@
 ```{r}
-W <- getW(zinb50)
+W <- colData(se)[, grepl('^W', colnames(colData(se)))]
+W <- as.matrix(W)
 d <- dist(W)
@@
 
-Let's add W to the colData of our SummarizedExperiment object.
-
-```{r}
-W <- data.frame(W)
-colnames(W) <- paste0('W', 1:ncol(W))
-colData(se) <- cbind(colData(se), W)
-```
-
-FP: what do you think of implemented a "compute_normalized_values" argument to zinbFit for summarizedExperiment objects where zinbFit would take as input a SummarizedExperiment object and return a summarizedExperiment object with slots added for normalized values and W?
-
 ## 2. Clustering of the cells
@@
 
-FP: What about the following call to RSEC?
-
 EP: I updated it to work on a SE object so that it has the meta data. If you have a summarized experiment object with W already, you could use that as long as assay(seObj) gives W.
@@
 ```{r rsec_50}
-W <- colData(se)[, grepl('^W', colnames(colData(se)))]
-W <- as.matrix(W)
 fn <- '../data/RSEC_W.rda'
@@
               plotInfo = "mergeMethod", cutoff = 0.01)
-```
-
-FP: Not sure we should show the following heatmap using the W, we should rather use the normalized values (following heatmap).
-
-EP: Agreed
-
-```{r heatmap, eval=FALSE}
-plotHeatmap(ceObj, clusterSamplesData = "dendrogramValue", 
-            breaks = .99)
 ```

### 2017-06-11 R/workflow.Rmd
biocparallel ncores

@@
   print(system.time(zinb <- zinbDimRed(se, K = 50, X = mod, residuals = T,
-                                       normalizedValues = T)))
+                                       normalizedValues = T,
+                                       BPPARAM = MulticoreParam(NCORES))))
   save(zinb, file = fn)

### 2017-06-12 R/workflow.Rmd
biocparallel for mac and linux

--- 2017-06-11 R/workflow.Rmd
@@
 library(clusterExperiment)
+library(BiocParallel)
 #github
@@
 #CRAN
+library(doParallel)
 library(ggplot2)
@@
 if(packageVersion("clusterExperiment")<'1.3.0.9008') stop("must have current develop version to avoid bugs")
-if(packageVersion("zinbwave")<'0.99.4.1') stop("must have current develop version to avoid bugs")
+if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
+```
+
+```{r}
+runZinb <- T
+runClus <- T
+NCORES <- 7
+```
+
+FP: for zinbwave, parallel computing is now handle by BiocParallel. If you have a Windows machine, please update the code to allow paralell computing.
+
+```{r parallel}
+mysystem = Sys.info()[['sysname']]
+switch(mysystem,
+       Windows = {print("I'm a Windows PC and you should choose the parallel computing you want.")},
+       Linux = {print("I'm a penguin and I'll use package MulticoreParam for parallel computing.")},
+       Darwin = {print("I'm a Mac and I'll use package doParallel for parallel computing.")})
+
+if (mysystem == 'Darwin'){
+  registerDoParallel(NCORES)
+  register(DoparParam())
+}else if (mysystem == 'Linux'){
+  register(bpstart(MulticoreParam(workers=NCORES)))
+}else{
+  print('Please change this to allow parallel computing')
+  register(SerialParam())
+}
 ```
@@
 white space, and a valid version number in parentheses.
-
-```{r}
-runZinb <- T
-runClus <- T
-NCORES <- 7
-```
 
@@
 
-```{r}
+```{r batch}
 col_batch = rep(brewer.pal(9, "Set1"), 2)
@@
 
-```{r}
+```{r original}
 col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
@@
 
-```{r}
+```{r clustbatch}
 table(data.frame(batch = as.vector(batch),
@@
 
-```{r}
+```{r se}
 se <- SummarizedExperiment(assays = list(rawCounts = core),
@@
 
-```{r}
+```{r zinb}
 fn <- '../data/zinb_batch.rda'
 if (runZinb & !file.exists(fn)){
-  mod <- model.matrix( ~ batch)
-  print(system.time(zinb <- zinbDimRed(se, K = 50, X = mod, residuals = T,
+  print(system.time(zinb <- zinbDimRed(se, K = 50, X = '~ batch',
+                                       residuals = T,
                                        normalizedValues = T)))
@@
 
-```{r}
+```{r norm}
 res <- assays(zinb)[['normalizedValues']]
@@
 
-```{r}
+```{r lookatnorm}
 res_order <- res[, order(as.numeric(batch))]
@@
 
-```{r}
+```{r pcanorm}
 pca <- prcomp(t(res))
@@
 
-```{r}
+```{r lookatdimred}
 W <- colData(se)[, grepl('^W', colnames(colData(se)))]
@@
 
-```{r}
+```{r tableclust}
 table(primaryClusterNamed(ceObj))
@@
 
-```{r}
+```{r heatmaps}
 # sampleData <- data.frame(ours = primaryCluster(ceObj))
@@
 
-```{r}
+```{r compare}
 palDF <- ceObj@clusterLegend[[1]]
@@
 
-```{r}
+```{r tabagain}
 table(data.frame(original = clus.labels, ours = primaryClusterNamed(ceObj)))
@@
 
-```{r}
+```{r kvec}
 Kvec <- c(3, 4, 5)
@@
 
-```{r slingshot_unsup}
+```{r removeunassigned}
 our_cl <- primaryClusterNamed(ceObj)
@@
 
-```{r}
+```{r slingshot_unsup}
 for (k in Kvec){
@@
 
-```{r}
+```{r de}
 de <- read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)

### 2017-06-12 R/workflow.Rmd
update zinbFit for slingshot

@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9008') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.0.9009') stop("must have current develop version to avoid bugs")
 if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
@@
 
-FP: for zinbwave, parallel computing is now handle by BiocParallel. If you have a Windows machine, please update the code to allow paralell computing.
+FP: Version of zinbwave to use is last version of branch normvalues (branched from master).
+Parallel computing is now handle by BiocParallel.
+If you have a Windows machine, please update the code to allow paralell computing.
 
@@
 ```
-
-FP: Elizabeth, I could not install the last version 1.3.0.9009 of clusterExperiment. The error is
-
-checking DESCRIPTION meta-information ... ERROR
-Malformed Depends or Suggests or Imports or Enhances field.
-Offending entries:
-  MAST
-BiocStyle
-Entries must be names of packages optionally followed by '<=' or '>=',
-white space, and a valid version number in parentheses.
 
@@
 if (runZinb & !file.exists(fn)){
-  print(system.time(zinb <- zinbDimRed(se, K = 50, X = '~ batch',
+  print(system.time(se <- zinbDimRed(se, K = 50, X = '~ Batch',
                                        residuals = T,
                                        normalizedValues = T)))
-  save(zinb, file = fn)
+  save(se, file = fn)
 }else{
@@
 
+FP: note to myself: why do we have infinite values in the residuals now? It shows the same results as before, but we should not see infinite values here!
+
 ```{r norm}
-res <- assays(zinb)[['normalizedValues']]
-res[1:3,1:3]
+norm <- assays(se)$normalizedValues
+if (sum(is.infinite(norm))>0){
+  maxNorm = max(norm[!is.infinite(norm)])
+  assays(se)$normalizedValues[is.infinite(norm)] <- maxNorm
+  norm <- assays(se)$normalizedValues
+}
+norm[1:3,1:3]
 ```
@@
 ```{r lookatnorm}
-res_order <- res[, order(as.numeric(batch))]
+norm_order <- norm[, order(as.numeric(batch))]
 col_order <- as.numeric(batch)[order(as.numeric(batch))]
-boxplot(res_order, main='Boxplot of normalized values\ncolor=batch',
-        col = col_order, staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
+boxplot(norm_order, main='Boxplot of normalized values\ncolor=batch',
+        col = col_order, staplewex = 0, outline = 0, border = col_order,
+        xaxt = 'n')
 ```
@@
 ```{r pcanorm}
-pca <- prcomp(t(res))
+pca <- prcomp(t(norm))
 par(mfrow = c(1,2))
@@
   #symbol for samples missing from original clustering
-	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 
-  seObj<-SummarizedExperiment(t(W),colData=metadata)
+	metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2 
+  seObj <- SummarizedExperiment(t(W), colData = metadata)
   print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
@@
             clusterSamplesData = "dendrogramValue",
-			sampleData=c("clusterLabels","Batch","Experiment"), clusterLegend=metaColors, annLegend=FALSE,
+            sampleData=c("clusterLabels","Batch","Experiment"),
+            clusterLegend=metaColors, annLegend=FALSE,
             main = 'Normalized values, 1000 most variable genes',
             breaks = 0.99)
-
 ```
@@
 
-Cluster name | Description | Correspondence | Color
--------------|-------------|-----------------|---------
-c1 | HBC | original 1, 5 | darkblue
-c2 | new and small | new and small | green
-c3 | new and small | new and small | red
-c4 | GBC / immature neurons / MV 1 | original 2, 3, 11, 14 | orange
-c5 | Sus | original 4, 7 | darkpurple
-c6 | Neuron | original 9, 12 | brown
-c7 | Immature Neuron | original 10, 14 | cian 
-c8 | Immature Neuron | original 14 | lightpurple
-c9 | Microvillus | original 15 | grayblue
-
-FP: We need to work on the cluster coloring (!) and the matching with the clusters you found for the paper with Russell, maybe by changing the parameters in clusterMany?
-
-EP: The function plotClusters aligns the clusters and can return the aligned coloring between clusterings.
-There is an option to return an updated CE object that changes the colors in clusterLegend to match the new colors found by aligning the clusters (`resetNames` and `resetColors` arguments).
-These haven't been extensively tested in real life because I don't think we've used them much.
-But it would be good to demonstrate!
-
-FP: With all the improvements made, I think we simply have the correct colors now.
+Cluster name | Description | Correspondence
+-------------|-------------|-----------------
+c1 | HBC | original 1, 5
+c2 | new and small | new and small
+c3 | new and small | new and small
+c4 | GBC / immature neurons / MV 1 | original 2, 3, 11, 14
+c5 | Sus | original 4, 7
+c6 | Neuron | original 9, 12
+c7 | Immature Neuron | original 10, 14
+c8 | Immature Neuron | original 14
+c9 | Microvillus | original 15
 
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c5", "c6", "c9"))
+                           end.clus = c("c3", "c6", "c10"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 ```{r refit_zinb}
-mod <- model.matrix(~ batch[our_cl != "-1"])
 fn <- '../data/refit_zinbwave_slingshot.rda'
@@
   zinbList <- lapply(Kvec, function(k){
-    zinbFit(core[, our_cl != "-1"], X = mod, K = k, ncores = NCORES)
+    zinbFit(se[, our_cl != "-1"], X = '~ Batch', K = k)
   })

### 2017-06-12 R/workflow.Rmd
Changes to preprocessin of the data

@@
 ```{r}
-runZinb <- T
-runClus <- T
+runZinb <- TRUE
+runClus <- TRUE
 NCORES <- 7
@@
 switch(mysystem,
-       Windows = {print("I'm a Windows PC and you should choose the parallel computing you want.")},
+       Windows = {print("I'm a Windows PC and I'll use the snow for parallel computing (warning: it might be slower than serial computations).")},
        Linux = {print("I'm a penguin and I'll use package MulticoreParam for parallel computing.")},
@@
 
+DR: I have changed the `createData.R` file so that it uses scone to filter out low-quality samples.
+
 ```{r datain}
-#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
-core <- read.table("../data/oeCufflinkCountData_1000Var.txt",
-                   sep = "\t", header = TRUE)
-core <- data.matrix(core)
-metadata <- read.table("../data/oeMetadata.txt", sep = "\t", header = TRUE)
-# symbol for samples missing from original clustering
-metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
+if(!file.exists("../data/oe_se_1000Var.rda")) {
+  source("createData.R")
+}
+
+load("../data/oe_se_1000Var.rda")
 ```
@@
 ```{r batches}
-batch <- metadata$Batch
+batch <- colData(core)$Batch
 ```
@@
 ```{r qc}
-qc <- metadata[, !names(metadata) %in% c("Batch", "Experiment", "clusterLabels")]
+qc <- colData(core)[, !names(colData(core)) %in% c("Batch", "Experiment", "clusterLabels")]
 head(qc, 2)
@@
 ```{r clusterlabels}
-clus.labels <- metadata[, "clusterLabels"]
+clus.labels <- colData(core)[, "clusterLabels"]
 ```
@@
 dim(core)
-core[1:3, 1:3]
-```
-
-Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, original labels, and quality control measures.
-
-```{r se}
-se <- SummarizedExperiment(assays = list(rawCounts = core),
-                           colData = metadata)
+assay(core)[1:3, 1:3]
 ```
@@
 if (runZinb & !file.exists(fn)){
-  print(system.time(se <- zinbDimRed(se, K = 50, X = '~ Batch',
-                                       residuals = T,
-                                       normalizedValues = T)))
+  print(system.time(se <- zinbDimRed(core, K = 50, X = '~ Batch',
+                                       residuals = TRUE,
+                                       normalizedValues = TRUE)))
   save(se, file = fn)
@@
 ```
+
+DR: the chunk above and the similar one for clusterExperiment are fine for now, but in the published workflow, we should just rely on the markdown cache system (the code above doesn't check for changes to the file or code -- just that a version of the file exists).
 
@@
   #symbol for samples missing from original clustering
-	metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2 
-  seObj <- SummarizedExperiment(t(W), colData = metadata)
+  seObj <- SummarizedExperiment(t(W), colData = colData(core))
   print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
@@
 
-```{r refit_zinb}
+```{r refit_zinb, eval=FALSE}
 fn <- '../data/refit_zinbwave_slingshot.rda'
+
 if (runZinb & !file.exists(fn)){
@@
 
-```{r slingshot_unsup_refit}
+```{r slingshot_unsup_refit, eval=FALSE}
 for(k in Kvec) {
@@
 
-```{r slingshot_sup_refit}
+```{r slingshot_sup_refit, eval=FALSE}
 for(k in Kvec){
@@
 
-```{r de}
+```{r de, eval = FALSE}
 de <- read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)

### 2017-06-12 R/createData.R

@@ -1,4 +1,17 @@
-load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet")
+library(GEOquery)
+library(scone)
+if(!file.exists("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")) {
+  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
+                "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+  gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+}
+load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")
+
+if(!file.exists("../data/oeHBCdiff_clusterLabels.txt")) {
+  download.file("https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt",
+                "../data/oeHBCdiff_clusterLabels.txt")
+}
+
 ## count matrix
 E <- assayData(Cufflinks_eSet)$counts_table
 
@@ -22,15 +35,42 @@ bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
 
 clusterLabels<-read.table("../data/oeHBCdiff_clusterLabels.txt",sep="\t",stringsAsFactors=FALSE)
 m<-match(colnames(E) ,clusterLabels[,1])
-metaData<-data.frame("Experiment"=bio,"Batch"=batch,"clusterLabels"=clusterLabels[m,2],qc)
+metadata<-data.frame("Experiment"=bio,"Batch"=batch,"clusterLabels"=clusterLabels[m,2],qc)
+
+# symbol for samples missing from original clustering
+metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
+
+se <- SummarizedExperiment(assays = list(counts = E),
+                           colData = metadata)
+
+
+# Initial Gene Filtering
+num_reads = quantile(assay(se)[assay(se) > 0])[4]
+num_cells = 0.25*ncol(assay(se))
+is_common = rowSums(assay(se) >= num_reads ) >= num_cells
+table(is_common)
+
+# Metric-based Filtering sample-filtering
+data("housekeeping")
+hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]
+
[…]

### 2017-06-13 R/workflow.Rmd
Corrected RSEC code

@@
   print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
-                                  betas = 0.8,
+                                  betas = 0.8, dimReduce="none",
                 clusterFunction = "hierarchical01", minSizes=1,
@@
                 combineProportion = 0.7,
-                mergeMethod = "none")))
+                mergeMethod = "none", random.seed=424242)))
   save(ceObj, file = fn)
@@
 # easier fix:
+colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
 origClusterColors<-bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
@@
 
-Cluster name | Description | Correspondence
--------------|-------------|-----------------
-c1 | HBC | original 1, 5
-c2 | new and small | new and small
-c3 | new and small | new and small
-c4 | GBC / immature neurons / MV 1 | original 2, 3, 11, 14
-c5 | Sus | original 4, 7
-c6 | Neuron | original 9, 12
-c7 | Immature Neuron | original 10, 14
-c8 | Immature Neuron | original 14
-c9 | Microvillus | original 15
+Cluster name | Description | Color | Correspondence
+-------------|-------------|-------| ----------
+c1 | HBC | blue | original 1, 5
+c2 | GBC / immature neurons / MV 1 | green | original 2, 3, 11
+c3 | Sus | red | original 4, 7
+c4 | Neuron | orange | original 9, 12
+c5 | Immature Neuron | purple | original 10, 14
+c6 | new and small | brown | original 9
+c7 | Microvillus | cyan | original 15
 
@@
 
-K = 3 does not seem very good to me: Sus is not an end cluster, GBC is an end cluster.
+K = 3 only one lineage: sus is right after HBC
 
-K = 4 is better, slingshot finds the end clusters but there is a spurious end cluster.
+K = 4 two lineages: sus is right after HBC, differ only in end point (neuron, new cluster)
 
-K = 5 does not seem great to me: GBC is an end cluster and Sus and Microvillus are in the same lineage.
+K = 5 three lineages: hbc-gbc-immature neuron-new-neuron; hbc-gbc-mv; hbc-sus (perfect!)
 
@@
 
-K = 3 finds GBC as an end cluster (that I did not specify in the end.clus!).
+K = 3 four lineages: might be good, three lineages are correct and the fourth is plausible.
 
-K = 4 Yeah! it seems that it is what we want even if we still have a spurius end cluster and GBC not really at the junction.
+K = 4 same as K=3
 
-K = 5 Yeah! even if GBC not really at the junction.
+K = 5 same as K=3
 
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6", "c10"))
+                           end.clus = c("c3", "c6", "c7"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 ### Re-fitting zinbwave
+
 #### Unsupervised
 
-K = 3 is better than when we did not refit zinbwave but still not perfect: Sus in all the clusters. GBC not really at the junction.
+K = 3 four lineages: immature neurons as a end point (not good)
 
-K = 4 good even if GBC not really at the junction.
+K = 4 same as K = 3
 
-K = 5 not great GBC is an end cluster.
+K = 5 same as K = 3
 
-```{r refit_zinb, eval=FALSE}
+```{r refit_zinb}
 fn <- '../data/refit_zinbwave_slingshot.rda'
@@
 
-```{r slingshot_unsup_refit, eval=FALSE}
+```{r slingshot_unsup_refit}
 for(k in Kvec) {
@@
 
-K = 3 Yeah! close to perfection.
+K = 3 four lineages (the small new cluster is still an end point)
 
-K = 4 good even if GBC not really at the junction.
+K = 4 same as K=3
 
-K = 5 no, GBC is an end cluster.
+K = 5 three lineages (correct)
 
-```{r slingshot_sup_refit, eval=FALSE}
+```{r slingshot_sup_refit }
 for(k in Kvec){

### 2017-06-18 R/workflow_oldVersion.Rmd
clean up and add biocworkflow dir

--- 2017-06-13 R/workflow.Rmd
@@
 set.seed(20)
-if(packageVersion("clusterExperiment")<'1.3.0.9009') stop("must have current develop version to avoid bugs")
 if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
@@
 ```
+
+FP: this is the old version of the workflow, please use file workflow/workflow.Rmd that I created from package BiocWorkflowTools.
+Directory workflow in now all compatible with the bioc workflow format and we can directly convert the Rmd file to a tex file that can then be exported to overleaf for direct submission to F1000Research.
 
@@
 ```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
-knitr::include_graphics('../vignettes/workflow.png')
+knitr::include_graphics('../workflow/workflow.png')
 ```

### 2017-06-18 workflow/workflow.Rmd
clean up and add biocworkflow dir

--- 2017-06-18 R/workflow_oldVersion.Rmd
@@
 ---
-title: "Worflow"
-author: "Workflow team"
-date: "`r format(Sys.time(), '%B %d, %Y')`"
-output: 
-  html_document: 
-    fig_height: 7
-    fig_width: 7
-    toc: yes
-    code_folding: hide
-    toc_float: yes
+title: "Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering"
+author: 
+  - name: Author Name1
+    affiliation: Address of author1
+  - name: Author Name2
+    affiliation: Address of author2
+abstract: Abstracts should be up to 300 words and provide a succinct summary of the article.
+Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article.
+Citations should not be used in the abstract, and the use of abbreviations should be minimized.
+keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
+bibliography: ref.bib
+output: BiocWorkflowTools::f1000_article
 ---
@@
 
-```{r}
+```{r parallel}
 runZinb <- TRUE
@@
 NCORES <- 7
-```
-
-FP: this is the old version of the workflow, please use file workflow/workflow.Rmd that I created from package BiocWorkflowTools.
-Directory workflow in now all compatible with the bioc workflow format and we can directly convert the Rmd file to a tex file that can then be exported to overleaf for direct submission to F1000Research.
-
-FP: Version of zinbwave to use is last version of branch normvalues (branched from master).
-Parallel computing is now handle by BiocParallel.
-If you have a Windows machine, please update the code to allow paralell computing.
-
-```{r parallel}
 mysystem = Sys.info()[['sysname']]
-switch(mysystem,
-       Windows = {print("I'm a Windows PC and I'll use the snow for parallel computing (warning: it might be slower than serial computations).")},
-       Linux = {print("I'm a penguin and I'll use package MulticoreParam for parallel computing.")},
-       Darwin = {print("I'm a Mac and I'll use package doParallel for parallel computing.")})
-
 if (mysystem == 'Darwin'){
@@
 
-# Steps of the workflow
+# Introduction
 
-We propose a worklow to analyze single cell RNA-Seq with the following steps
+- growing interest in single-cell RNA-seq data.
+New statistical tools need to be developped compared to bulk RNA-seq because bias: drop-outs, amplification bias.
+There is also increasing size of datasets (10X-genomics, 1.3M cells) -> need of integrated workflows of new statistical tools
+- previously developped : F1000 Research bioc [@Lun2016] and scater [@McCarthy2017] are pipelines containing several useful methods for quality control, visualisation and pre-processing of data.
+In these pipelines, single-cell expression data are organized in objects of the SCESet class which allows integrated workflows.
+However, they are used mostly to prepare the data for further downstream analysis and do not focus on steps like clustering, pseudotime ordering, and DE analysis.
+- we propose an integrated workflow with dowstream analysis: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resumpling, pseudotime ordering, and DE analysis for the clusters found.
+We use a summarizedExperiment object to have an integrated pipeline.
+We also propose useful functions for visualization.
 
-- Dimensionality reduction using zinbwave to get W which should capture the biology,
-- Cluster cells using clusterExperiment on W to get the cluster labels,   
-- Get lineage using slingshot on W and cluster labels from clusterExperiment,  
-- Get DE genes between lineages/clusters.
-
-Along the worflow, use deviance residuals as adjusted values.
-
-```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
-knitr::include_graphics('../workflow/workflow.png')
+```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Our workflow'}
+knitr::include_graphics('schema_workflow.png')
 ```
 
-## 1. Create a SummarizedExperiment object
+# Analysis of olfactory stem cell lineages using single cell RNA-seq data
+## Overview
 
-Along the workflow, we want to use a unique SummarizedExperiement object carrying all the data we need.
+The workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium [@Fletcher2017].
+The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs) which are the actively proliferating cells in the epithelium.
+When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.
 
-EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file.
-For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article.
-This will be slightly different data from Russell's, which didn't use all of the samples.
-We can adjust that decision later, or just compare the samples that are the same.
-(Russell's clusterLabels are in the meta data)
+The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium.
+To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, different lineages were assigned to each cell in the dataset and results were confirmed experimentally using in vivo lineage tracing.
+It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either sustentacular (support) cells or GBCs.
+Then, the GBC lineage, in turn, branches to give rise to olfactory sensory neurons, microvillous cells, and cells of the Bowman’s gland.
 
-EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?
+```{r stemcelldiff, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. Figure 2e from Fletcher et al paper.'}
+knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
+```
 
-DR: I have changed the `createData.R` file so that it uses scone to filter out low-quality samples.
+## Generation of the data
+
+The dataset we use for the workflow is publicly available at https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz.
+It contains 849 cells and 28361 detected genes.
+Each cell has been labelled with labels publicly available at https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt.
+Pre-processing steps are as follow:
+- ERCC and CreER cells were removed,
+- low-quality cells were removed using package scone (ADD REF) and
+
+FP: we should add details here.
+
+Finally to speed up the computations, we kept only the 1000 most variable genes in this analysis.
+After the preprocessing steps, the dataset we are working with has 1000 genes and 747 cells.
+Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object
 
@@
 load("../data/oe_se_1000Var.rda")
+core
 ```
-
-Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.
 
@@
 
-Cells have been processed in `r length(unique(batch))` different batches
+Metadata for the cells is stored in colData from the SummarizedExperiment object. Cells have been processed in `r length(unique(batch))` different batches
 
@@
 
-We have qc measures from the data
+We also have access to the quality control measures that were used to filter low quality cells
 
@@
 
-In original work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters
+In original work [@Fletcher2017], cells have been clustered into `r length(unique(clus.labels))` different clusters
 
@@
 
-Batches are kind of confounded with the biology
+Note that in this dataset batches are somehow confounded with the clusters found in the original work.
 
@@
 
-We have `r ncol(core)` cells.
+## Dimensionality reduction adjusting for batches
 
-```{r lookatdata}
-dim(core)
-assay(core)[1:3, 1:3]
-```
+In single-cell RNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed.
+It allows the data to become more tractable, both from a statistical (cf.
+curse of dimensionality) and computational point of view.
+Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest.
 
-## 2. Dimensionality reduction adjusting for gene and cell-level covariates
+Here, we perform dimensionality reduction using bioconductor package zinbwave which allows to fit a zero-inflated negative binomial model (ZINB-WaVE) to get a low-dimensional representation of the data while accounting for zero inflation (dropouts), over-dispersion, and the count nature of the data.
+The model can include a sample-level intercept, which serves as a global-scaling normalization factor.
+The user can also include both gene-level and sample-level covariates.
+The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
 
-To cluster and get lineages we want to reduce the dimension of the data.
-We are going to use zinbwave to do so.
-First, let's fit zinbwave with first K = 0 to compute normalized values (i.e.
-deviance residuals) adjusted for batches.
-We could also adjust for gene length or GC content here.
-We then fit zinbwave to get the dimensionality reduced matrix W specifying the number of dimension K = 50.
-Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE).
-For K = 0 and K = 50, we correct for batch effect including batches in X.
+As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and we adjust for the batches.
+
+FP: do we want to expose the user to the equations?
+In the one hand, it would be easier to explain what we are looking at (W, normalized values, ...).
+In the other hand, it would need a good amount of explanations.
+I have a preference to show the equations, but what do you think?
 
@@
 
-We use deviance residuals as normalized values for visualization. FP: explain rational: K=0 so residuals capture the bio adjusting for batch. Let's check that deviance residuals look ok.
+When function zinbDimRed is called, normalized values of the counts can be computed and used for visualization (e.g.
+in heatmaps).
+The normalized values are the residuals of the fit of our ZINB model where the gene and cell covariates were the same as the user specified in the zinbDimRed call but the number of unknown covariates is set to be null (i.e.
+K=0).
+When no unknown covariates are included in the model, the residuals represents the counts adjusted for the gene and cell covariates.
+To compute the residuals, we use the deviance residuals.
 
-FP: note to myself: why do we have infinite values in the residuals now? It shows the same results as before, but we should not see infinite values here!
+FP: note to myself: why do we have infinite values in the residuals now? It 
+shows the same results as before, but we should not see infinite values here!
 
@@
 
-Boxplot of the normalized values for each cell. It seems that correction for batches is ok.
+As expected, the distributions of the normalized counts are independent of the batches
 
@@
 
-PCA on the normalized values where color are for batches on the left and previously found clusters on the right. We want no clustering on the left side and clustering on the right side.
+We performed principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right.
+As expected, there are no clustering on the left side and clustering on the right side.
 
@@
 
+Overall, looking at the normalized values, it seems that the batch effect was removed.
+
 ### Dimensionality reduction
 
-Let's check that performing MDS on W we have something coherent with original clusters.
+Calling function zinbDimRed, we performed dimensionality reduction with K = 50.
+We can visualize the low dimensional matrix performing multidimensional scaling (MDS) in two dimensions.
+To make sure that the low dimensional matrix captured the signal of interest, the colors shown here are from the original clusters.
 
@@
 
-## 2. Clustering of the cells
+## Clustering of the cells
 
@@
 
-## 4. Pseudotime ordering
+## Pseudotime ordering
 
@@
 c1 | HBC | blue | original 1, 5
-c2 | GBC / immature neurons / MV 1 | green | original 2, 3, 11
+c2 | GBC | immature neurons | MV 1 | green | original 2, 3, 11
 c3 | Sus | red | original 4, 7
@@
 
-## 5. DE analysis
+## DE analysis
 
@@
 
-FP: Kelly, is it you who did the DE analysis for Russell paper? If yes, what tool did you use? On what data? The full quantile normalized counts? Do you have code available?
+## Further developments
 
-# Session Info
+- DE analysis using slingshot and zinbwave
+- SingleCellExperiment class
+
+# Conclusions
+
+This workflow provides guidance to perform downstream analysis of single-cell RNA-seq datasets in R.
+We propose a worflow in four steps: dimensionality reduction while adjusting for gene and cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the clusters.
+The workflow is general and flexible allowing the user to sustitute each step of the worflow with a different R tool and to input data that have been previously processed by different tools.
+We hope the workflow will ease the technical aspect of single-cell RNA-seq data analysis to help discovering new biological insights.
+
+# Software availability
+This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.
+
+1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
+2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
+3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
+4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
+5. Software license (AUTHOR TO PROVIDE; required)
+
+The three packages zinbwave, clusterExperiment, and slingshot used in the workflow are bioconductor packages and are available on github at respectively https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
+The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.
 
@@
 ```
+
+# Author contributions
+In order to give appropriate credit to each author of an article, the individual contributions of each author to the manuscript should be detailed in this section.
+We recommend using author initials and then stating briefly how they contributed.
+
+# Competing interests
+All financial, personal, or professional competing interests for any of the authors that could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article.
+If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.
+
+# Grant information
+Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc.
+Please do not list funding that you have that is not relevant to this specific piece of research.
+For each funder, please state the funder’s name, the grant number where applicable, and the individual to whom the grant was assigned.
+If your work was not funded by any grants, please include the line: 'The author(s) declared that no grants were involved in supporting this work.'
+
+# Acknowledgments
+This section should acknowledge anyone who contributed to the research or the
+article but who does not qualify as an author based on the criteria provided earlier (e.g.
+someone or an organization that provided writing assistance).
+Please state how they contributed; authors should obtain permission to acknowledge from all those mentioned in the Acknowledgments section.

### 2017-06-18 R/benchmark.R

@@ -1,37 +1,0 @@
-#!/usr/bin/env Rscript
-library(zinbwave)
-library(matrixStats)
-load("../data/Expt4c2b_filtdata.Rda")
-load("../data/E4c2b_slingshot_wsforkelly.RData")
-set.seed(20)
-
-names(batch) <- colnames(counts)
-
-counts <- counts[,names(clus.labels)]
-batch <- droplevels(batch[names(clus.labels)])
-qc <- qc[names(clus.labels),]
-
-vars <- rowVars(log1p(counts))
-names(vars) <- rownames(counts)
-vars <- sort(vars, decreasing = TRUE)
-core <- counts[names(vars)[1:1000],]
-
-NCORES = 7
-time_no_batch = list()
-time_batch = list()
-K = c(0, 100)
-for (i in 1:length(K)){
-  print(i)
-  time_no_batch[[i]] = system.time(zinb0 <- zinbFit(core, K = K[i], ncores = NCORES))
-}
-
-
-mod = model.matrix( ~ batch)
-for (i in 1:length(K)){
-  print(i)
-  time_batch[[i]] = system.time(zinb0 <- zinbFit(core, K = K[i], X = mod, ncores = NCORES))
-}
-print(time_batch)
-print(time_no_batch)
-
-save(time_batch, time_no_batch, file = 'time_zinb_k0.rda')

### 2017-06-18 workflow/workflow.Rmd
reproducible report

@@
 runClus <- TRUE
-NCORES <- 7
+NCORES <- 2
 mysystem = Sys.info()[['sysname']]

### 2017-06-19 workflow_combineMinSize10_mds/workflow.Rmd
combineMinSize=10 and mds

--- 2017-06-18 workflow/workflow.Rmd
@@
 ```{r rsec_50}
-fn <- '../data/RSEC_W.rda'
+fn <- '../data/RSEC_W_combineMinSize10.rda'
 if (runClus & !file.exists(fn)){
@@
                 combineProportion = 0.7,
-                mergeMethod = "none", random.seed=424242)))
+                mergeMethod = "none", random.seed=424242,
+                combineMinSize = 10)))
   save(ceObj, file = fn)
@@
 c5 | Immature Neuron | purple | original 10, 14
-c6 | new and small | brown | original 9
-c7 | Microvillus | cyan | original 15
+c6 | Microvillus | cyan | original 15
 
@@
 
-### Use previous W
+### Use W components
+#### Use previous W
 
@@
 
-#### Unsupervised
+##### Unsupervised
 
@@
 
-K = 4 two lineages: sus is right after HBC, differ only in end point (neuron, new cluster)
+K = 4 same as K=3
 
-K = 5 three lineages: hbc-gbc-immature neuron-new-neuron; hbc-gbc-mv; hbc-sus (perfect!)
+K = 5 three lineages: hbc-gbc-immature-neuron; hbc-gbc-mv; hbc-sus (perfect!)
 
@@
 
-#### Supervised
+##### Supervised
 
-K = 3 four lineages: might be good, three lineages are correct and the fourth is plausible.
-
-K = 4 same as K=3
-
-K = 5 same as K=3
+K = 3, 4, 5 get same results, all good.
 
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6", "c7"))
+                           end.clus = c("c3", "c6"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 
-### Re-fitting zinbwave
+#### Re-fitting zinbwave
 
-#### Unsupervised
+##### Unsupervised
 
-K = 3 four lineages: immature neurons as a end point (not good)
-
-K = 4 same as K = 3
-
-K = 5 same as K = 3
+K = 3, 4, 5 get same results, all good.
 
 ```{r refit_zinb}
-fn <- '../data/refit_zinbwave_slingshot.rda'
+fn <- '../data/refit_zinbwave_slingshot_combineMinSize10.rda'
 
@@
 
-#### Supervised
+##### Supervised
 
-K = 3 four lineages (the small new cluster is still an end point)
-
-K = 4 same as K=3
-
-K = 5 three lineages (correct)
+K = 3, 4, 5 get same results, all good.
 
@@
   lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6", "c10"))
+                           end.clus = c("c3", "c6"))
   curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 
-CONCLUSION: K = 5 is never great as GBC is generally an end cluster. K = 4 is ok for all the methods and a bit better when zinbwave is refitted. K = 3 when refitting and supervized is good.
+### Use MDS of W
 
-It seems to me that using slingshot on W without re-fitting zinbwave with k = 4 gives good results where supervized mode is slightly better than unsupervized.
-It is just a one shot example and we should obviously not make a general conclusion, but I think that for the purpose of the workflow it is fine to use slingshot without refitting zinbwave.
-We should write a note to the user that it is better to refit zinbwave to have more power.
+Here, instead of using the first components of W, we use the MDS of W.
+
+#### Use previous W
+
+The input of slingshot is the MDS of W used for clusterExperiment where the number of dimensions is k where k in (3, 4, 5) here.
+
+##### Unsupervised
+
+K = 3 only one lineage: sus is right after HBC
+
+K = 4, 5 now, it is ok, three lineages: hbc-gbc-immature-neuron; hbc-gbc-mv; hbc-sus (perfect!)
+
+```{r}
+our_cl <- primaryClusterNamed(ceObj)
+cl = our_cl[our_cl != "-1"]
+pal = pal[names(pal) != '-1']
+```
+
+```{r}
+for (k in Kvec){
+  X <- W[our_cl != "-1", ]
+  mds <- cmdscale(dist(X), eig = TRUE, k = k)
+  X <- mds$points
+
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+  print(lineages$lineage5)
+}
+```
+
+##### Supervised
+
+K = 3, 4, 5 get same results, all good.
+
+```{r}
+for (k in Kvec){
+  X <- W[our_cl != "-1", ]
+  mds <- cmdscale(dist(X), eig = TRUE, k = k)
+  X <- mds$points
+
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
+                           end.clus = c("c3", "c6"))
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+  print(lineages$lineage5)
+}
+```
+
+#### Re-fitting zinbwave
+
+##### Unsupervised
+
+K = 3, 4, 5 get same results, all good.
+
+```{r}
+for(k in Kvec) {
+  X <- getW(zinbList[[k - 2]])
+  mds <- cmdscale(dist(X), eig = TRUE, k = k)
+  X <- mds$points
+
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+  print(lineages$lineage5)
+}
+```
+
+##### Supervised
+
+K = 3, 4, 5 get same results, all good.
+
+```{r}
+for(k in Kvec){
+  X <- getW(zinbList[[k - 2]])
+  mds <- cmdscale(dist(X), eig = TRUE, k = k)
+  X <- mds$points
+
+  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
+                           end.clus = c("c3", "c6"))
+  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+  plot_curves(X, cl, curves, col.clus = pal)
+  plot_tree(X, cl, lineages, col.clus = pal)
+
+  print(paste0("K=", k))
+  print(lineages$lineage1)
+  print(lineages$lineage2)
+  print(lineages$lineage3)
+  print(lineages$lineage4)
+  print(lineages$lineage5)
+}
+```
 

### 2017-06-25 workflow_combineMinSize10_mds/workflow.Rmd
more writting

@@
 Citations should not be used in the abstract, and the use of abbreviations should be minimized.
-keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
+keywords: single-cell, workflow, single-cell-genomics, pipeline, rna-seq, gene-expression, statistical-models, r
 bibliography: ref.bib
+vignette: >
+    %\VignetteIndexEntry{A worfklow for low-level analyses of single-cell RNA-seq data}
+    %\VignetteEngine{knitr::rmarkdown}
 output: BiocWorkflowTools::f1000_article
@@
 ```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
-knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
+library(BiocStyle)
+library(knitr)
+knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
+knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
+
 #bioconductor:
@@
 library(BiocParallel)
+library(zinbwave)
+library(GEOquery)
 #github
-library(zinbwave)
 library(slingshot)
@@
 
-```{r parallel}
+```{r parallel, echo=FALSE}
 runZinb <- TRUE
@@
 
-```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Our workflow'}
+```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing single-cell RNA-seq datasets. On the right side, major plots generated by the workflow."}
 knitr::include_graphics('schema_workflow.png')
@@
 
+```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium.
+This figure has been reproduced with kind permission from [@Fletcher2017].'}
+knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
+```
+
 The workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium [@Fletcher2017].
@@
 The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium.
-To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, different lineages were assigned to each cell in the dataset and results were confirmed experimentally using in vivo lineage tracing.
-It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either sustentacular (support) cells or GBCs.
-Then, the GBC lineage, in turn, branches to give rise to olfactory sensory neurons, microvillous cells, and cells of the Bowman’s gland.
+To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, a single-cell RNA-seq analysis was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
+Then, the expression of each gene was quantified by counting the total number of reads mapped to its cell to get a cell by gene matrix of raw counts.
+Different lineages were then assigned to each cell in the dataset using a set of statistical tools and results were confirmed experimentally using in vivo lineage tracing.
+Details on the data generation and steps of the analysis are available in [@Fletcher2017].
 
-```{r stemcelldiff, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. Figure 2e from Fletcher et al paper.'}
-knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
+It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
+Then, the GBC lineage, in turn, branches to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman’s gland.
+See Figure \@ref(fig:stemcelldiff).
+In this study, we show possible steps to recover the lineages found in the original study from the matrix of raw counts publicly available.
+
+FP: ask Russell if we can use the Figure \@ref(fig:stemcelldiff) here.
+
+## Pre-processing
+
+Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO) with accession number GSE95601. Before filtering, the dataset has 849 cells and 28361 detected genes.
+
+```{r, echo = FALSE}
+urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
+         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")
+
+if(!file.exists("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")) {
+  download.file(urls[1], "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+  gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+}
+
+if(!file.exists("../data/oeHBCdiff_clusterLabels.txt")) {
+  download.file(urls[2], "../data/oeHBCdiff_clusterLabels.txt")
+}
 ```
 
-## Generation of the data
+```{r}
+load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")
 
-The dataset we use for the workflow is publicly available at https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz.
-It contains 849 cells and 28361 detected genes.
-Each cell has been labelled with labels publicly available at https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt.
-Pre-processing steps are as follow:
-- ERCC and CreER cells were removed,
-- low-quality cells were removed using package scone (ADD REF) and
+## count matrix
+E <- assayData(Cufflinks_eSet)$counts_table
 
-FP: we should add details here.
+# Remove undetected genes
+E <- na.omit(E)
+E <- E[rowSums(E)>0,]
+dim(E)
+```
 
-Finally to speed up the computations, we kept only the 1000 most variable genes in this analysis.
-After the preprocessing steps, the dataset we are working with has 1000 genes and 747 cells.
-Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object
+We removed the genes corresponding to ERCC spike-ins and the CreER gene as it corresponds to a marker gene for the HBCs cells. Davide, is that correct?
 
-```{r datain}
-if(!file.exists("../data/oe_se_1000Var.rda")) {
-  source("createData.R")
-}
+```{r}
+# remove ERCC and CreER genes
+cre <- E["CreER",]
+ercc <- E[grep("^ERCC-", rownames(E)),]
+E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
+E <- E[-which(rownames(E)=="CreER"), ]
+dim(E)
+```
 
-load("../data/oe_se_1000Var.rda")
+Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object.
+We dispose of quality control measures as well as information on the batches in which the cells were sequenced and lineages found in the original publication [@Fletcher2017].
+Cells with a lineage of -2 are cells not assigned to a lineage in the orginal publication.
+
+```{r}
+# Extract QC metrics
+qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
+qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))
+
+# Extract metadata
+batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
+bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
+clusterLabels <- read.table("../data/oeHBCdiff_clusterLabels.txt",
+                            sep = "\t", stringsAsFactors = FALSE)
+m <- match(colnames(E), clusterLabels[, 1])
+
+# Create metadata data.frame
+metadata <- data.frame("Experiment" = bio,
+                       "Batch" = batch,
+                       "clusterLabels" = clusterLabels[m,2],
+                       qc)
+
+# symbol for cell not assigned to a lineage in original data
+metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
+
+se <- SummarizedExperiment(assays = list(counts = E),
+                           colData = metadata)
+se
+```
+
+Using bioconductor package scone, we removed low quality cells that did not pass our quality control rule, that is ...
+See Figure \@ref(fig:scone).
+Davide, would you like to describe your choices for scone here?
+
+```{r scone}
+# Metric-based Filtering sample-filtering
+data("housekeeping")
+hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]
+
+mfilt <- metric_sample_filter(assay(se), nreads = colData(se)$NREADS,
+                             ralign = colData(se)$RALIGN,
+                             pos_controls = rownames(se) %in% hk,
+                             zcut = 3, mixture = FALSE,
+                             plot = TRUE)
+```
+
+```{r sconeFilt}
+# Simplify to a single logical
+mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
+se <- se[, mfilt]
+dim(se)
+```
+
+Finally, to speed up the computations, we kept only the 1000 most variable genes. We recommend not to be so stringent in your analysis and perform the analysis on more genes.
+
+```{r}
+# filtering to top 1000 most variable
+vars <- rowVars(log1p(assay(se)))
+names(vars) <- rownames(se)
+vars <- sort(vars, decreasing = TRUE)
+core <- se[names(vars)[1:1000],]
+save(core, file="../data/oe_se_1000Var.rda")
+```
+
+## Dataset structure
+Overall, after the preprocessing steps, our dataset has 1000 genes and 747 cells
+```{r}
 core
@@
 
-```{r batches}
-batch <- colData(core)$Batch
-```
-
-Metadata for the cells is stored in colData from the SummarizedExperiment object. Cells have been processed in `r length(unique(batch))` different batches
+Metadata for the cells is stored in the object colData from the SummarizedExperiment object. Cells have been processed in 18 different batches
 
 ```{r batch}
+batch <- colData(core)$Batch
 col_batch = rep(brewer.pal(9, "Set1"), 2)
@@
 
-We also have access to the quality control measures that were used to filter low quality cells
-
-```{r qc}
-qc <- colData(core)[, !names(colData(core)) %in% c("Batch", "Experiment", "clusterLabels")]
-head(qc, 2)
-```
-
-```{r clusterlabels}
-clus.labels <- colData(core)[, "clusterLabels"]
-```
-
-In original work [@Fletcher2017], cells have been clustered into `r length(unique(clus.labels))` different clusters
+In original work [@Fletcher2017], cells have been clustered into 14 different clusters where 151 cells were not assigned to a lineage
 
 ```{r original}
+clus.labels <- colData(core)[, "clusterLabels"]
 col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
@@
 
-Note that in this dataset batches are somehow confounded with the clusters found in the original work.
+Note that some of the batches are confounded with the lineages found in the original work which could be a problem when we correct for batch effect in the dimensionality reduction step.
 
@@
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest.
+Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017].
 
@@
 The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
+See more details in [@Risso2017].
 
-As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and we adjust for the batches.
-
-FP: do we want to expose the user to the equations?
-In the one hand, it would be easier to explain what we are looking at (W, normalized values, ...).
-In the other hand, it would need a good amount of explanations.
-I have a preference to show the equations, but what do you think?
+As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and adjusted for the batches.
 
@@
 
-When function zinbDimRed is called, normalized values of the counts can be computed and used for visualization (e.g.
+When function `zinbDimRed` is called, normalized values of the counts can be computed and used for visualization (e.g.
 in heatmaps).
-The normalized values are the residuals of the fit of our ZINB model where the gene and cell covariates were the same as the user specified in the zinbDimRed call but the number of unknown covariates is set to be null (i.e.
-K=0).
-When no unknown covariates are included in the model, the residuals represents the counts adjusted for the gene and cell covariates.
-To compute the residuals, we use the deviance residuals.
-
-FP: note to myself: why do we have infinite values in the residuals now? It 
-shows the same results as before, but we should not see infinite values here!
+The normalized values are the residuals of the fit of the ZINB-WaVE model where the gene and cell covariates were specified by the user when calling function `zinbDimRed` but no low dimensional matrix is computed.
+In that case, the residuals represents the counts adjusted for the gene and cell covariates and should capture the biological signal of interest.
+To compute the residuals, the deviance is used.
 
@@
 
-As expected, the distributions of the normalized counts are independent of the batches
+As expected, the normalized values are uniformly distributed for the different batches. See Figure \@ref(fig:boxplotNorm).
 
-```{r lookatnorm}
+```{r boxplotNorm, fig.cap="Boxplot of normalized values. Normalized values are uniformly distributed for the different batches"}
 norm_order <- norm[, order(as.numeric(batch))]
 col_order <- as.numeric(batch)[order(as.numeric(batch))]
-boxplot(norm_order, main='Boxplot of normalized values\ncolor=batch',
-        col = col_order, staplewex = 0, outline = 0, border = col_order,
-        xaxt = 'n')
+boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
+        border = col_order, xaxt = 'n')
+abline(h=0)
 ```
 
-We performed principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right.
-As expected, there are no clustering on the left side and clustering on the right side.
+The principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right shows that, as expected, cells do not cluster by batch but by orginal published lineage.
+See Figure \@ref(fig:pcanorm).
+Overall, looking at the normalized values, it seems that the batch effect has been removed.
 
-```{r pcanorm}
+```{r pcanorm, fig.cap="PCA of normalized values. In the left panel, colors correspond to the different batches. In the right panel, colors correspond to the original published lineages."}
 pca <- prcomp(t(norm))
 par(mfrow = c(1,2))
-plot(pca$x, col = col_batch[batch], pch = 20,
-     main="PCA of normalized values\ncolor=batch")
-plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20,
-     main = "PCA of normalized values\ncolor=cluster")
+plot(pca$x, col = col_batch[batch], pch = 20, main = '')
+plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
+```
+
+```{r,echo=FALSE}
 par(mfrow = c(1,1))
@@
 
-Overall, looking at the normalized values, it seems that the batch effect was removed.
-
 ### Dimensionality reduction
 
-Calling function zinbDimRed, we performed dimensionality reduction with K = 50.
-We can visualize the low dimensional matrix performing multidimensional scaling (MDS) in two dimensions.
-To make sure that the low dimensional matrix captured the signal of interest, the colors shown here are from the original clusters.
+Calling function zinbDimRed, we performed dimensionality reduction with `K = 50` dimensions in the low dimensional space.
+We can visualize the low dimensional matrix (with `K = 50` dimensions) by performing multidimensional scaling (MDS) in two dimensions and euclidian distance.
+To make sure that the low dimensional matrix in `K = 50` dimensions captured the biological signal of interest, we plotted the MDS with colors corresponding to the original published lineages.
+See Figure \@ref(fig:mdsW).
 
-```{r lookatdimred}
+```{r mdsW, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the original published lineages."}
 W <- colData(se)[, grepl('^W', colnames(colData(se)))]
@@
 fit <- cmdscale(d, eig = TRUE, k = 2)
-plot(fit$points, col = col_clus[as.character(clus.labels)], main = 'MDS', pch = 20,
-     xlab = 'Component 1', ylab = 'Component 2')
+plot(fit$points, col = col_clus[as.character(clus.labels)], main = '',
+     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
 legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
@@
 
-We use clusterExperiment with W.
-
-EP: I updated it to work on a SE object so that it has the meta data. If you have a summarized experiment object with W already, you could use that as long as assay(seObj) gives W.
+The new step of the workflow is to cluster the cells according to the low dimensional matrix computed in the previous step.
+We used function `RSEC` from bioconductor package clusterExperiment.
+Elizabeth, would you like to explain the rational and arguments of the different functions?
 
@@
 
-```{r examineCombineMany}
+We can visualize the resulting clusterings using the plotClusters command. See Figure \@ref(fig:examineCombineMany).
+
+```{r examineCombineMany, fig.cap="Clusters found using function RSEC from clusterExperiment package.
+Cells are in the columns, and different clusterings on the rows.
+Each sample is color coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap."}
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
@@
 
-```{r plotcoclust}
+We can also visualize the proportion of times these clusters were together across these clusterings. See Figure \@ref(fig:plotcoclust).
+
+```{r plotcoclust, fig.cap="Heatmap of the proportion of times the clusters were together across the different clusterings."}
 plotCoClustering(ceObj)
 ```
+
+Overall, here is a summary of the number of cells clustered in each cluster
 
@@
 
-FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context?
-I set eval=FALSE in the next chunk to skip the merging step, let me know if you would rather keep using it.
-And if we want to still use the merging step, would we want to include it in RSEC function arguments instead of separately?
-
-EP: I don't think the merging step on the W makes a whole lot of sense -- the method is irrelevant.
-The merging is based on calculating the % of genes found significant (the specific method is arbitrary).
-The best thing would be to replace the W with residuals in the assay of `ceObj` (or whatever data that you will do the DE on for the time stuff below), and then run the merging step on that data.
-I'm not particularly fond of `locfdr`.
-It was probably the method that gave the best merging to Russell and Diya.
-You'd really have to run `mergeClusters` setting `plotInfo="all"` and look at the results and decide both the cutoff level and the method.
-
-EP: Also, if you don't save the output of `mergeClusters` it doesn't update `ceObj`.
-I was calling it for just the resulting plots, since it was already merged in RSEC above.
-I've changed to code to update ceObj below.
-
-FP: Ha ok, good to know. I'll keep the eval=FALSE for the moment.
-
-```{r examineMergeClusters,eval=FALSE}
-#re-does merging simpling to make plot 
-#something like:
-#assay(ceObj)
-# if that replacement data should be considered on the transformed scale in plots, etc, the transformation function should be fixed as well:
-#transformation(ceObj)
-ceObj<-mergeClusters(ceObj, mergeMethod = "locfdr",
-              plotInfo = "mergeMethod", cutoff = 0.01)
-```
-
-So, let's look at a heatmap on normalized values.
-
-FP: Elizabeth, I did not find how to define the column annotation track in the plot below to have the same colors as in ceObj@clusterLegend[[1]].
-I tried to use arguments annColors and annCol from aheatmap as it is said in plotHeatmap documentation that for signature matrix arguments can be passed to aheatmap.
-But I got the error "The following arguments to aheatmap cannot be set by the user in plotHeatmap:Rowv,Colv,color,annCol,annColors".
-
-EP: Fanny, you would need to use the argument 'clusterLegend'.
-That argument takes either the format of aheatmap (list with each element a *named* vector of colors) or the format of the clusterExperiment object (i.e.
-list with each element a matrix with columns for `name` and `color`).
-So I think the following code will run, though it might need the list to have names...
-
-But an easier fix to the code would be to set `visualizeData` option. I haven't tested this because I don't have the objects need run, so let me know if there is error.
-
-FP: it seems great to me, what do you think?
-
-EP: We should be careful, because the default in plotHeatmap is to plot the 500 most variable genes (maybe a slightly paternalistic default).
-I've changed it to `all` in the code here.
-I've also added the plotting of the batch, experiment, and Russell's original clusters.
-We may not want to keep all of them, but probably at least Russell's clusters for comparison.
-
-```{r heatmaps}
-# sampleData <- data.frame(ours = primaryCluster(ceObj))
-# plotHeatmap(assays(se)$normalizedValues,
-#             main = 'Normalized values, 1000 most variable genes',
-#             clusterSamplesData = ceObj@dendro_samples,
-#             sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
-# easier fix:
+```{r heatmapsClusters, fig.cap="Heatmap of the normalized values for the 1000 most variable genes (rows)."}
+# set colors for cell clusterings
 colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
@@
             clusterLegend=metaColors, annLegend=FALSE,
-            main = 'Normalized values, 1000 most variable genes',
+            main = '',
             breaks = 0.99)
@@
 
-```{r compareClusters}
-plot(fit$points, col = col_clus[as.character(clus.labels)],
-     main = 'MDS W, color = original clusters', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
-legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
-       fill = unique(col_clus), title = 'Sample')
-```
+We can now plot the MDS of the low dimensional matrix coloring the cells by their newly found clusters.
+See Figure \@ref(fig:mdsWce) which is the same as Figure \@ref(fig:mdsW) but where colors for the cells are different.
 
-```{r compare}
+```{r mdsWce, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the clusters found using our workflow."}
 palDF <- ceObj@clusterLegend[[1]]
@@
 pal["-1"] = "transparent"
-plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
-     main = 'MDS W, color = our new clusters', pch = 20,
+plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = '', pch = 20,
      xlab = 'Component1', ylab = 'Component2')
@@
 
-The goal of this section is to see if we need to refit zinbwave when we want to run slingshot.
-We first run slingshot on the W used by clusterExperiment.
-In the second part of this section, we fit zinbwave on the matrix of counts where the unassigned cells have been removed.
-For each part (without or with refitting zinbwave), we run slingshot in the supervised and unsupervised mode and try k=3, k=4, k=5 dimensions in W.
-
-From what I understand, start original clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
-Additionally, we want the GBC cluster to be a junction before the differentiation between Microvillus and Neuron.
+We now want to order the clusters found using function `RSEC` from package clusterExperiment.
+From original published work, we know that the start clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
+Additionally, we know that the GBC cluster should be a junction before the differentiation between Microvillus cells and Neurons.
 The correspondance with the original clusters is as follow
@@
 
-```{r kvec}
-Kvec <- c(3, 4, 5)
-```
+As the input for the speudotime ordering algorithm, we use the MDS of low dimensional matrix (K=50) where the number of dimensions for MDS is `k=3`.
+The algorithm is stable to the number of dimensions k in the MDS.
+We use the unsupervised version of slingshot where we only give the information of the start cluster but not the end clusters.
 
-### Use W components
-#### Use previous W
-
-The input of slingshot is the W used for clusterExperiment where the number of dimensions is reduced to k where k in (3, 4, 5) here.
-
-##### Unsupervised
-
-K = 3 only one lineage: sus is right after HBC
-
-K = 4 same as K=3
-
-K = 5 three lineages: hbc-gbc-immature-neuron; hbc-gbc-mv; hbc-sus (perfect!)
-
-```{r removeunassigned}
-our_cl <- primaryClusterNamed(ceObj)
-cl = our_cl[our_cl != "-1"]
-pal = pal[names(pal) != '-1']
-```
-
-```{r slingshot_unsup}
-for (k in Kvec){
-  X <- W[our_cl != "-1", 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-##### Supervised
-
-K = 3, 4, 5 get same results, all good.
-
-```{r slingshot_sup}
-for (k in Kvec){
-  X <- W[our_cl != "-1", 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-#### Re-fitting zinbwave
-
-##### Unsupervised
-
-K = 3, 4, 5 get same results, all good.
-
-```{r refit_zinb}
-fn <- '../data/refit_zinbwave_slingshot_combineMinSize10.rda'
-
-if (runZinb & !file.exists(fn)){
-  zinbList <- lapply(Kvec, function(k){
-    zinbFit(se[, our_cl != "-1"], X = '~ Batch', K = k)
-  })
-  save(zinbList, file = fn)
-}else{
-  load(fn)
-}
-```
-
-```{r slingshot_unsup_refit}
-for(k in Kvec) {
-  X <- getW(zinbList[[k - 2]])[, 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-##### Supervised
-
-K = 3, 4, 5 get same results, all good.
-
-```{r slingshot_sup_refit }
-for(k in Kvec){
-  X <- getW(zinbList[[k - 2]])[, 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-### Use MDS of W
-
-Here, instead of using the first components of W, we use the MDS of W.
-
-#### Use previous W
-
-The input of slingshot is the MDS of W used for clusterExperiment where the number of dimensions is k where k in (3, 4, 5) here.
-
-##### Unsupervised
-
-K = 3 only one lineage: sus is right after HBC
-
-K = 4, 5 now, it is ok, three lineages: hbc-gbc-immature-neuron; hbc-gbc-mv; hbc-sus (perfect!)
+Kelly, would you like to give more information about slingshot?
 
@@
 
-```{r}
-for (k in Kvec){
+```{r curves}
 X <- W[our_cl != "-1", ]
-  mds <- cmdscale(dist(X), eig = TRUE, k = k)
+mds <- cmdscale(dist(X), eig = TRUE, k = 4)
 X <- mds$points
@@
 plot_curves(X, cl, curves, col.clus = pal)
+```
+
+```{r tree}
 plot_tree(X, cl, lineages, col.clus = pal)
+```
 
-  print(paste0("K=", k))
+```{r lineages}
 print(lineages$lineage1)
@@
 print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
 ```
 
-##### Supervised
+In this workflow, it is ok to use unsupervised.
+In some other cases, we have noticed that we need to give more knowledge to the algorithm to find the correct ordering.
+Here is the code to use slingshot in a supervised setting.
 
-K = 3, 4, 5 get same results, all good.
-
-```{r}
-for (k in Kvec){
+```{r slingshotsupervised, eval=FALSE}
 X <- W[our_cl != "-1", ]
-  mds <- cmdscale(dist(X), eig = TRUE, k = k)
+mds <- cmdscale(dist(X), eig = TRUE, k = 4)
 X <- mds$points
@@
 
-  print(paste0("K=", k))
 print(lineages$lineage1)
@@
 print(lineages$lineage5)
-}
-```
-
-#### Re-fitting zinbwave
-
-##### Unsupervised
-
-K = 3, 4, 5 get same results, all good.
-
-```{r}
-for(k in Kvec) {
-  X <- getW(zinbList[[k - 2]])
-  mds <- cmdscale(dist(X), eig = TRUE, k = k)
-  X <- mds$points
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-##### Supervised
-
-K = 3, 4, 5 get same results, all good.
-
-```{r}
-for(k in Kvec){
-  X <- getW(zinbList[[k - 2]])
-  mds <- cmdscale(dist(X), eig = TRUE, k = k)
-  X <- mds$points
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
 ```
@@
 
-Here is the kind of plots we want to present
+Kelly, would you like to complete this part? Here is the kind of plots we want to present
 
@@
 
-- DE analysis using slingshot and zinbwave
+- DE analysis using slingshot
+- DE analysis and imputation using zinbwave
 - SingleCellExperiment class
 
-# Conclusions
+# Conclusion
 
 This workflow provides guidance to perform downstream analysis of single-cell RNA-seq datasets in R.
-We propose a worflow in four steps: dimensionality reduction while adjusting for gene and cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the clusters.
-The workflow is general and flexible allowing the user to sustitute each step of the worflow with a different R tool and to input data that have been previously processed by different tools.
+We propose a worflow in four steps: dimensionality reduction while adjusting for cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the lineages.
+The workflow is general and flexible allowing the user to sustitute the statistical method used in each step of the worflow by a different method or R package and to input data that have been previously processed by different tools.
 We hope the workflow will ease the technical aspect of single-cell RNA-seq data analysis to help discovering new biological insights.
@@
 Please state how they contributed; authors should obtain permission to acknowledge from all those mentioned in the Acknowledgments section.
+
+# References

### 2017-06-26 workflow/workflow.Rmd
rough draft of each section

@@
 ---
-title: "Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering"
+title: "Bioconductor workflow for single-cell RNA sequencing data analysis: dimensionality reduction, clustering, and pseudotime ordering"
 author: 
@@
 Citations should not be used in the abstract, and the use of abbreviations should be minimized.
-keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
+keywords: single-cell, workflow, single-cell-genomics, bioconductor, rna-seq, gene-expression, statistical-models, r
 bibliography: ref.bib
+vignette: >
+    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
+    %\VignetteEngine{knitr::rmarkdown}
 output: BiocWorkflowTools::f1000_article
@@
 
-- growing interest in single-cell RNA-seq data.
-New statistical tools need to be developped compared to bulk RNA-seq because bias: drop-outs, amplification bias.
-There is also increasing size of datasets (10X-genomics, 1.3M cells) -> need of integrated workflows of new statistical tools
-- previously developped : F1000 Research bioc [@Lun2016] and scater [@McCarthy2017] are pipelines containing several useful methods for quality control, visualisation and pre-processing of data.
-In these pipelines, single-cell expression data are organized in objects of the SCESet class which allows integrated workflows.
-However, they are used mostly to prepare the data for further downstream analysis and do not focus on steps like clustering, pseudotime ordering, and DE analysis.
-- we propose an integrated workflow with dowstream analysis: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resumpling, pseudotime ordering, and DE analysis for the clusters found.
-We use a summarizedExperiment object to have an integrated pipeline.
-We also propose useful functions for visualization.
+Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enables researchers to measure genome-wide transcription levels for individual cells.
+To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle different problems like dimensionality reduction, clustering, and the inference of cell lineages.
+While each individual method is useful to analyze scRNA-seq data, there is an increasing need for creating integrated workflows of these new statistical tools.
+This is all the more true that novel sequencing technologies are able to sequence more and more cells in a single sequencing run.
+For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mice brains.
 
-```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Our workflow'}
+Some workflows have already been developped to analysis scRNA-seq data with several useful methods for quality control, visualisation and pre-processing of the data.
+The workflow for low-level analyses of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor project [@Huber2015] and cover basic steps including quality control, data exploration and normalization.
+In these workflows, single-cell expression data are organized in objects of the SCESet class allowing integrated analysis.
+However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps like clustering and pseudotime ordering.
+
+Here, we propose an integrated workflow with dowstream analysis where our workflow has three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of branching lineages and ordering of the cells by developmental progression.
+Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps where for each steps a different R package is used.
+
+```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing scRNA-seq datasets. On the right side, major plots generated by the workflow."}
 knitr::include_graphics('schema_workflow.png')
@@
 
-# Analysis of olfactory stem cell lineages using single cell RNA-seq data
+# Analysis of olfactory stem cell lineages using scRNA-seq data
 ## Overview
+
+```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium.
+This figure has been reproduced with kind permission from [@Fletcher2017].'}
+knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
+```
 
@@
 The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium.
-To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, different lineages were assigned to each cell in the dataset and results were confirmed experimentally using in vivo lineage tracing.
-It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either sustentacular (support) cells or GBCs.
-Then, the GBC lineage, in turn, branches to give rise to olfactory sensory neurons, microvillous cells, and cells of the Bowman’s gland.
+To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, a scRNA-seq analysis was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
+Then, the expression of each gene was quantified by counting the total number of reads mapped to each cell to get a cell by gene matrix of raw counts.
+Different lineages were assigned to each cell in the dataset using a set of statistical tools.
+Finally, results were confirmed experimentally using in vivo lineage tracing.
+Details on the data generation and steps of the analysis are available in [@Fletcher2017].
 
-```{r stemcelldiff, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. Figure 2e from Fletcher et al paper.'}
-knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
+It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
+Then, the GBC lineage, in turn, branches to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman’s gland.
+See Figure \@ref(fig:stemcelldiff).
+In this study, we show possible steps to recover the lineages found in the original study from the matrix of raw counts publicly available.
+
+FP: ask Russell if we can use the Figure \@ref(fig:stemcelldiff) here.
+
+## Pre-processing
+
+Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO) with accession number GSE95601. Before filtering, the dataset has 849 cells and 28361 detected genes.
+
+```{r, echo = FALSE}
+urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
+         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")
+
+if(!file.exists("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")) {
+  download.file(urls[1], "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+  gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+}
+
+if(!file.exists("../data/oeHBCdiff_clusterLabels.txt")) {
+  download.file(urls[2], "../data/oeHBCdiff_clusterLabels.txt")
+}
 ```
 
-## Generation of the data
+```{r}
+load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")
 
-The dataset we use for the workflow is publicly available at https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz.
-It contains 849 cells and 28361 detected genes.
-Each cell has been labelled with labels publicly available at https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt.
-Pre-processing steps are as follow:
-- ERCC and CreER cells were removed,
-- low-quality cells were removed using package scone (ADD REF) and
+# count matrix
+E <- assayData(Cufflinks_eSet)$counts_table
 
-FP: we should add details here.
+# Remove undetected genes
+E <- na.omit(E)
+E <- E[rowSums(E)>0,]
+dim(E)
+```
 
-Finally to speed up the computations, we kept only the 1000 most variable genes in this analysis.
-After the preprocessing steps, the dataset we are working with has 1000 genes and 747 cells.
-Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object
+We removed the genes corresponding to ERCC spike-ins and the CreER gene as it corresponds to a marker gene for the HBCs cells. Davide, is that correct?
 
-```{r datain}
-if(!file.exists("../data/oe_se_1000Var.rda")) {
-  source("createData.R")
-}
+```{r}
+# remove ERCC and CreER genes
+cre <- E["CreER",]
+ercc <- E[grep("^ERCC-", rownames(E)),]
+E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
+E <- E[-which(rownames(E)=="CreER"), ]
+dim(E)
+```
 
-load("../data/oe_se_1000Var.rda")
+Along the workflow, we use a SummarizedExperiment object to keep the counts and metadata in a unique object.
+In the metadata, we dispose of quality control measures for each cell, information on the batches in which the cells were sequenced, and clusters and lineages found in the original publication [@Fletcher2017].
+Cells with a cluster of `-2` are cells not assigned to a cluster in the orginal publication.
+
+```{r}
+# Extract QC metrics
+qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
+qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))
+
+# Extract metadata
+batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
+bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
+clusterLabels <- read.table("../data/oeHBCdiff_clusterLabels.txt",
+                            sep = "\t", stringsAsFactors = FALSE)
+m <- match(colnames(E), clusterLabels[, 1])
+
+# Create metadata data.frame
+metadata <- data.frame("Experiment" = bio,
+                       "Batch" = batch,
+                       "clusterLabels" = clusterLabels[m,2],
+                       qc)
+
+# symbol for cell not assigned to a lineage in original data
+metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
+
+se <- SummarizedExperiment(assays = list(counts = E),
+                           colData = metadata)
+se
+```
+
+Using bioconductor package scone, we removed low quality cells that did not pass our quality control rule, that is ...
+See Figure \@ref(fig:scone).
+Davide, would you like to describe your choices for scone here?
+
+```{r scone, fig.cap="Plot generated by scone to filter low quality cells."}
+# Metric-based Filtering sample-filtering
+data("housekeeping")
+hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]
+
+mfilt <- metric_sample_filter(assay(se), nreads = colData(se)$NREADS,
+                             ralign = colData(se)$RALIGN,
+                             pos_controls = rownames(se) %in% hk,
+                             zcut = 3, mixture = FALSE,
+                             plot = TRUE)
+```
+
+```{r sconeFilt}
+# Simplify to a single logical
+mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
+se <- se[, mfilt]
+dim(se)
+```
+
+Finally, to speed up the computations, we kept only the 1000 most variable genes. We recommend not to be so stringent in your analysis and perform the analysis on more genes.
+
+```{r}
+# filtering to top 1000 most variable
+vars <- rowVars(log1p(assay(se)))
+names(vars) <- rownames(se)
+vars <- sort(vars, decreasing = TRUE)
+core <- se[names(vars)[1:1000],]
+save(core, file="../data/oe_se_1000Var.rda")
+```
+
+## Dataset structure
+Overall, after the preprocessing steps, our dataset has 1000 genes and 747 cells.
+```{r}
 core
@@
 
-```{r batches}
-batch <- colData(core)$Batch
-```
-
-Metadata for the cells is stored in colData from the SummarizedExperiment object. Cells have been processed in `r length(unique(batch))` different batches
+Metadata for the cells is stored in the object colData from the SummarizedExperiment object. Cells have been processed in 18 different batches.
 
 ```{r batch}
-col_batch = rep(brewer.pal(9, "Set1"), 2)
+batch <- colData(core)$Batch
+col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
+              brewer.pal(8, "Accent")[1])
 names(col_batch) = unique(batch)
@@
 
-We also have access to the quality control measures that were used to filter low quality cells
-
-```{r qc}
-qc <- colData(core)[, !names(colData(core)) %in% c("Batch", "Experiment", "clusterLabels")]
-head(qc, 2)
-```
-
-```{r clusterlabels}
-clus.labels <- colData(core)[, "clusterLabels"]
-```
-
-In original work [@Fletcher2017], cells have been clustered into `r length(unique(clus.labels))` different clusters
+In original work [@Fletcher2017], cells have been clustered into 14 different clusters where 151 cells were not assigned to a cluster.
 
 ```{r original}
+clus.labels <- colData(core)[, "clusterLabels"]
 col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
@@
 
-Note that in this dataset batches are somehow confounded with the clusters found in the original work.
+Note that some of the batches are confounded with the clusters found in the original work which could be a problem when we correct for batch effect in the dimensionality reduction step.
 
@@
 
-In single-cell RNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed.
+In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed.
 It allows the data to become more tractable, both from a statistical (cf.
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest.
+Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017].
 
@@
 The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
+You can read more details about the ZINB-WaVE model in [@Risso2017].
 
-As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and we adjust for the batches.
-
-FP: do we want to expose the user to the equations?
-In the one hand, it would be easier to explain what we are looking at (W, normalized values, ...).
-In the other hand, it would need a good amount of explanations.
-I have a preference to show the equations, but what do you think?
+As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space.
+Here, we use `K = 50` dimensions in the low dimensional space and adjusted for the batches.
 
@@
 DR: the chunk above and the similar one for clusterExperiment are fine for now, but in the published workflow, we should just rely on the markdown cache system (the code above doesn't check for changes to the file or code -- just that a version of the file exists).
+FP: Yes.
 
@@
 
-When function zinbDimRed is called, normalized values of the counts can be computed and used for visualization (e.g.
+When function `zinbDimRed` is called, normalized values of the counts can be computed and used for visualization (e.g.
 in heatmaps).
-The normalized values are the residuals of the fit of our ZINB model where the gene and cell covariates were the same as the user specified in the zinbDimRed call but the number of unknown covariates is set to be null (i.e.
-K=0).
-When no unknown covariates are included in the model, the residuals represents the counts adjusted for the gene and cell covariates.
-To compute the residuals, we use the deviance residuals.
-
-FP: note to myself: why do we have infinite values in the residuals now? It 
-shows the same results as before, but we should not see infinite values here!
+The normalized values are the residuals of the fit of the ZINB-WaVE model where the gene and cell covariates were specified by the user when calling function `zinbDimRed`, but no low dimensional matrix is computed.
+In that case, the residuals represent the counts adjusted for the gene and cell covariates and should capture the biological signal of interest.
+To compute the residuals, the deviance is used.
 
@@
 
-As expected, the distributions of the normalized counts are independent of the batches
+As expected, the normalized values are uniformly distributed for the different batches. See Figure \@ref(fig:boxplotNorm).
 
-```{r lookatnorm}
+```{r boxplotNorm, fig.cap="Boxplot of normalized values. Different colors are for different batches."}
 norm_order <- norm[, order(as.numeric(batch))]
-col_order <- as.numeric(batch)[order(as.numeric(batch))]
-boxplot(norm_order, main='Boxplot of normalized values\ncolor=batch',
-        col = col_order, staplewex = 0, outline = 0, border = col_order,
-        xaxt = 'n')
+col_order <- col_batch[batch[order(as.numeric(batch))]]
+boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
+        border = col_order, xaxt = 'n')
+abline(h=0)
 ```
 
-We performed principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right.
-As expected, there are no clustering on the left side and clustering on the right side.
+The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by orginal published clusters.
+See Figure \@ref(fig:pcanorm).
+Overall, looking at the normalized values, it seems that the batch effect has been removed.
 
-```{r pcanorm}
+```{r pcanorm, fig.cap="PCA of normalized values where each point represent a cell.
+In the left panel, cells have been colored according to their batch.
+In the right panel, colors correspond to the original published clusters."}
 pca <- prcomp(t(norm))
 par(mfrow = c(1,2))
-plot(pca$x, col = col_batch[batch], pch = 20,
-     main="PCA of normalized values\ncolor=batch")
-plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20,
-     main = "PCA of normalized values\ncolor=cluster")
+plot(pca$x, col = col_batch[batch], pch = 20, main = '')
+plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
+```
+
+```{r,echo=FALSE}
 par(mfrow = c(1,1))
@@
 
-Overall, looking at the normalized values, it seems that the batch effect was removed.
-
 ### Dimensionality reduction
 
-Calling function zinbDimRed, we performed dimensionality reduction with K = 50.
-We can visualize the low dimensional matrix performing multidimensional scaling (MDS) in two dimensions.
-To make sure that the low dimensional matrix captured the signal of interest, the colors shown here are from the original clusters.
+Calling function `zinbDimRed` from package `zinbwave`, we performed dimensionality reduction where the number of dimensions in the low dimensional space is `K = 50`.
+We call the matrix in the low dimensional space `W`.
+We can visualize `W` in two dimensions by performing multidimensional scaling (MDS) on the distance matrix of `W` where the euclidian distance is used.
+To make sure that `W` captured the biological signal of interest, we plotted the MDS of `W` with colors corresponding to the original published clusters.
+See Figure \@ref(fig:mdsW).
 
-```{r lookatdimred}
+```{r mdsW, fig.cap="MDS of the low dimensional matrix `W` where each point represent a cell. Cells have been colored according to the original published clusters."}
 W <- colData(se)[, grepl('^W', colnames(colData(se)))]
@@
 
-We use clusterExperiment with W.
+The next step of the workflow is to cluster the cells according to the low dimensional matrix `W` computed in the previous step.
+We used function `RSEC` from bioconductor package clusterExperiment.
+Function `RSEC` implements a resampling-based sequential ensemble clustering method for identifying stable and tight cell clusters.
+The method aggregates multiple clustering results obtained from different base clustering algorithms and applications of a given algorithm to resampled versions of the learning set.
 
-EP: I updated it to work on a SE object so that it has the meta data. If you have a summarized experiment object with W already, you could use that as long as assay(seObj) gives W.
+Elizabeth, would you like to more details on the rational and arguments of the different functions?
 
 ```{r rsec_50}
-fn <- '../data/RSEC_W.rda'
+fn <- '../data/RSEC_W_combineMinSize10.rda'
 if (runClus & !file.exists(fn)){
@@
 
-```{r examineCombineMany}
+We can visualize the resulting clusterings using the `plotClusters` function.
+See Figure \@ref(fig:examineCombineMany) where cells are in the columns, and different clusterings on the rows.
+Each sample is color coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap.
+The first raw correspond to a consensus clustering across the many different clusterings performed.
+
+```{r examineCombineMany, fig.cap="Clusters found using function RSEC from clusterExperiment package."}
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
@@
 
-```{r plotcoclust}
+We can also visualize the proportion of times the consensus clusters were together across the different clusterings. See Figure \@ref(fig:plotcoclust).
+
+```{r plotcoclust, fig.cap="Heatmap of the proportion of times the clusters were together across the different clusterings."}
 plotCoClustering(ceObj)
@@
 
+Over stable and robust clustering, the number of cells in each cluster is as follow
+
 ```{r tableclust}
 table(primaryClusterNamed(ceObj))
-sum(primaryCluster(ceObj) == -1)
 ```
 
-FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context?
-I set eval=FALSE in the next chunk to skip the merging step, let me know if you would rather keep using it.
-And if we want to still use the merging step, would we want to include it in RSEC function arguments instead of separately?
+We can now plot a heatmap of the normalized values for the 1000 most variable genes where the cells have been clustered using the `RSEC` function. See Figure \@ref(fig:heatmapsClusters).
 
-EP: I don't think the merging step on the W makes a whole lot of sense -- the method is irrelevant.
-The merging is based on calculating the % of genes found significant (the specific method is arbitrary).
-The best thing would be to replace the W with residuals in the assay of `ceObj` (or whatever data that you will do the DE on for the time stuff below), and then run the merging step on that data.
-I'm not particularly fond of `locfdr`.
-It was probably the method that gave the best merging to Russell and Diya.
-You'd really have to run `mergeClusters` setting `plotInfo="all"` and look at the results and decide both the cutoff level and the method.
+```{r heatmapsClusters, fig.cap="Heatmap of the normalized values for the 1000 most variable genes where genes are in the rows and cells in the columns."}
+# set colors for cell clusterings
+colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
+origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
+experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
+batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
+metaColors <- list("Experiment" = experimentColors,
+                   "Batch" = batchColors,
+                   "clusterLabels" = origClusterColors)
 
-EP: Also, if you don't save the output of `mergeClusters` it doesn't update `ceObj`.
-I was calling it for just the resulting plots, since it was already merged in RSEC above.
-I've changed to code to update ceObj below.
-
-FP: Ha ok, good to know. I'll keep the eval=FALSE for the moment.
-
-```{r examineMergeClusters,eval=FALSE}
-#re-does merging simpling to make plot 
-#something like:
-#assay(ceObj)
-# if that replacement data should be considered on the transformed scale in plots, etc, the transformation function should be fixed as well:
-#transformation(ceObj)
-ceObj<-mergeClusters(ceObj, mergeMethod = "locfdr",
-              plotInfo = "mergeMethod", cutoff = 0.01)
+plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
+            whichClusters = "primary", clusterFeaturesData = "all",
+            clusterSamplesData = "dendrogramValue", breaks = 0.99,
+            sampleData = c("clusterLabels", "Batch", "Experiment"),
+            clusterLegend = metaColors, annLegend = FALSE, main = '')
 ```
 
-So, let's look at a heatmap on normalized values.
+Finally, we can visualize the cells in a two dimensional space by plotting the MDS of the low dimensional matrix `W` coloring the cells by their newly found clusters.
+See Figure \@ref(fig:mdsWce) which is the same as Figure \@ref(fig:mdsW) but where colors are for the newly found clusters instead of the clusters found in the orginal published work.
 
-FP: Elizabeth, I did not find how to define the column annotation track in the plot below to have the same colors as in ceObj@clusterLegend[[1]].
-I tried to use arguments annColors and annCol from aheatmap as it is said in plotHeatmap documentation that for signature matrix arguments can be passed to aheatmap.
-But I got the error "The following arguments to aheatmap cannot be set by the user in plotHeatmap:Rowv,Colv,color,annCol,annColors".
-
-EP: Fanny, you would need to use the argument 'clusterLegend'.
-That argument takes either the format of aheatmap (list with each element a *named* vector of colors) or the format of the clusterExperiment object (i.e.
-list with each element a matrix with columns for `name` and `color`).
-So I think the following code will run, though it might need the list to have names...
-
-But an easier fix to the code would be to set `visualizeData` option. I haven't tested this because I don't have the objects need run, so let me know if there is error.
-
-FP: it seems great to me, what do you think?
-
-EP: We should be careful, because the default in plotHeatmap is to plot the 500 most variable genes (maybe a slightly paternalistic default).
-I've changed it to `all` in the code here.
-I've also added the plotting of the batch, experiment, and Russell's original clusters.
-We may not want to keep all of them, but probably at least Russell's clusters for comparison.
-
-```{r heatmaps}
-# sampleData <- data.frame(ours = primaryCluster(ceObj))
-# plotHeatmap(assays(se)$normalizedValues,
-#             main = 'Normalized values, 1000 most variable genes',
-#             clusterSamplesData = ceObj@dendro_samples,
-#             sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
-# easier fix:
-colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
-origClusterColors<-bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
-experimentColors<-bigPalette[1:nlevels(colData(ceObj)$Experiment)]
-batchColors<-bigPalette[1:nlevels(colData(ceObj)$Batch)]
-metaColors<-list("Experiment"=experimentColors,"Batch"=batchColors,"clusterLabels"=origClusterColors)
-
-plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
-            whichClusters = "primary",clusterFeaturesData="all",
-            clusterSamplesData = "dendrogramValue",
-            sampleData=c("clusterLabels","Batch","Experiment"),
-            clusterLegend=metaColors, annLegend=FALSE,
-            main = 'Normalized values, 1000 most variable genes',
-            breaks = 0.99)
-```
-
-```{r compareClusters}
-plot(fit$points, col = col_clus[as.character(clus.labels)],
-     main = 'MDS W, color = original clusters', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
-legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
-       fill = unique(col_clus), title = 'Sample')
-```
-
-```{r compare}
+```{r mdsWce, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the clusters found using our workflow."}
 palDF <- ceObj@clusterLegend[[1]]
@@
 
-The goal of this section is to see if we need to refit zinbwave when we want to run slingshot.
-We first run slingshot on the W used by clusterExperiment.
-In the second part of this section, we fit zinbwave on the matrix of counts where the unassigned cells have been removed.
-For each part (without or with refitting zinbwave), we run slingshot in the supervised and unsupervised mode and try k=3, k=4, k=5 dimensions in W.
+We now want to demonstrate how to use the R software package `slingshot` to infer branching lineages and order cells by developmental progression.
+We connect the clusters identified by `RSEC` with a minimum spanning tree to learn the global lineage structure.
+Then, we refine this structure and order cells using highly stable simultaneous principal curves to infer smooth, branching lineages.
 
-From what I understand, start original clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).
-Additionally, we want the GBC cluster to be a junction before the differentiation between Microvillus and Neuron.
-The correspondance with the original clusters is as follow
+From the original published work, we know that the start clusters should be the clusters with HBCs and end clusters the clusters with MV, mOSN, and mSUS cells.
+Additionally, we know that the clusters with the GBCs should be a junction before the differentiation between MV and mOSN cells.
+See Figure \@ref(fig:stemcelldiff).
+The correspondance between the clusters we found and the original clusters is as follow
 
@@
 -------------|-------------|-------| ----------
-c1 | HBC | blue | original 1, 5
-c2 | GBC | immature neurons | MV 1 | green | original 2, 3, 11
-c3 | Sus | red | original 4, 7
-c4 | Neuron | orange | original 9, 12
+c1 | HBCs | blue | original 1, 5
+c2 | GBCs | green | original 2, 3, 11
+c3 | mSUS | red | original 4, 7
+c4 | mOSN | orange | original 9, 12
 c5 | Immature Neuron | purple | original 10, 14
-c6 | new and small | brown | original 9
-c7 | Microvillus | cyan | original 15
+c6 | MV | brown | original 15
 
-```{r kvec}
-Kvec <- c(3, 4, 5)
-```
+To order the clusters and find lineages, we use package `slingshot` where the input is the MDS of low dimensional matrix `W` where the number of dimensions used for MDS is `k = 4`.
+We found that slingshot algorithm is stable to the number of dimensions `k` of the MDS.
+Here, we use the unsupervised version of slingshot where we only provide the information of the start cluster but not the end clusters.
+Using slingshot, we recovered the order of the clusters found in the published work.
+See Figures \@ref(fig:curves) and \@ref(fig:tree).
 
-### Use previous W
+Kelly, would you like to give details about slingshot algorithm?
 
-The input of slingshot is the W used for clusterExperiment where the number of dimensions is reduced to k where k in (3, 4, 5) here.
-
-#### Unsupervised
-
-K = 3 only one lineage: sus is right after HBC
-
-K = 4 two lineages: sus is right after HBC, differ only in end point (neuron, new cluster)
-
-K = 5 three lineages: hbc-gbc-immature neuron-new-neuron; hbc-gbc-mv; hbc-sus (perfect!)
-
-```{r removeunassigned}
+```{r}
 our_cl <- primaryClusterNamed(ceObj)
@@
 pal = pal[names(pal) != '-1']
+X <- W[our_cl != "-1", ]
+mds <- cmdscale(dist(X), eig = TRUE, k = 4)
+X <- mds$points
+
+lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
+curves <- get_curves(X, clus.labels = cl, lineages = lineages)
 ```
 
-```{r slingshot_unsup}
-for (k in Kvec){
-  X <- W[our_cl != "-1", 1:k]
+```{r curves, fig.cap="Cells colored by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
+plot_curves(X, cl, curves, col.clus = pal)
+```
 
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
+```{r tree, fig.cap="Cells colored by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
+plot_tree(X, cl, lineages, col.clus = pal)
+```
+
+```{r lineages}
+print(lineages$lineage1)
+print(lineages$lineage2)
+print(lineages$lineage3)
+```
+
+In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot.
+However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
+Slingshot has the option for the user to provide the end cluster(s).
+Here is the code to use slingshot in a supervised setting
+
+```{r slingshotsupervised, eval=FALSE}
+lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
+                         end.clus = c("c3", "c6"))
 curves <- get_curves(X, clus.labels = cl, lineages = lineages)
@@
 print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
 ```
-
-#### Supervised
-
-K = 3 four lineages: might be good, three lineages are correct and the fourth is plausible.
-
-K = 4 same as K=3
-
-K = 5 same as K=3
-
-```{r slingshot_sup}
-for (k in Kvec){
-  X <- W[our_cl != "-1", 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6", "c7"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-### Re-fitting zinbwave
-
-#### Unsupervised
-
-K = 3 four lineages: immature neurons as a end point (not good)
-
-K = 4 same as K = 3
-
-K = 5 same as K = 3
-
-```{r refit_zinb}
-fn <- '../data/refit_zinbwave_slingshot.rda'
-
-if (runZinb & !file.exists(fn)){
-  zinbList <- lapply(Kvec, function(k){
-    zinbFit(se[, our_cl != "-1"], X = '~ Batch', K = k)
-  })
-  save(zinbList, file = fn)
-}else{
-  load(fn)
-}
-```
-
-```{r slingshot_unsup_refit}
-for(k in Kvec) {
-  X <- getW(zinbList[[k - 2]])[, 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-#### Supervised
-
-K = 3 four lineages (the small new cluster is still an end point)
-
-K = 4 same as K=3
-
-K = 5 three lineages (correct)
-
-```{r slingshot_sup_refit }
-for(k in Kvec){
-  X <- getW(zinbList[[k - 2]])[, 1:k]
-
-  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
-                           end.clus = c("c3", "c6", "c10"))
-  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-  plot_curves(X, cl, curves, col.clus = pal)
-  plot_tree(X, cl, lineages, col.clus = pal)
-
-  print(paste0("K=", k))
-  print(lineages$lineage1)
-  print(lineages$lineage2)
-  print(lineages$lineage3)
-  print(lineages$lineage4)
-  print(lineages$lineage5)
-}
-```
-
-CONCLUSION: K = 5 is never great as GBC is generally an end cluster. K = 4 is ok for all the methods and a bit better when zinbwave is refitted. K = 3 when refitting and supervized is good.
-
-It seems to me that using slingshot on W without re-fitting zinbwave with k = 4 gives good results where supervized mode is slightly better than unsupervized.
-It is just a one shot example and we should obviously not make a general conclusion, but I think that for the purpose of the workflow it is fine to use slingshot without refitting zinbwave.
-We should write a note to the user that it is better to refit zinbwave to have more power.
 
@@
 
-Here is the kind of plots we want to present
+Kelly, would you like to complete this part? Here is the kind of plots we would want to present.
 
@@
 
-- DE analysis using slingshot and zinbwave
-- SingleCellExperiment class
+In an effort to improve the workflow and analysis of scRNA-seq data, we are currently exploring statistical methods to perform DE analysis using the models we developed.
+One the one hand, we are developing a DE analysis method using the framework of slingshot which would take advantage of the lineages found.
+On the other hand, we are working on a DE analysis method using the ZINB-WaVE model which would account for the dropouts and gene and cell-level covariates.
+Additionally, we are working on a statistical method using the ZINB-WaVE model to impute the dropouts in the scRNA-seq data.
+This matrix with imputed counts could for example be used in the workflow in the dimensionality reduction, clustering or pseudo ordering steps.
 
-# Conclusions
+Finally, a new R class called SingleCellExperiment should be released soon.
+This new class would essentially be a SummarizedExperiment class with a couple of additional slots, the most important of which is reducedDims, which, much like the assays slot of SummarizedExperiment can contain one or more matrices with reduced dimensional matrices.
+This new SingleCellExperiment class would be a valuable addition to the workflow as we could store in one object the raw counts as well as the low dimensional matrix created by the dimensionality reduction step.
+Once this class is implemented, we would like to incorporate it to the workflow.
 
-This workflow provides guidance to perform downstream analysis of single-cell RNA-seq datasets in R.
-We propose a worflow in four steps: dimensionality reduction while adjusting for gene and cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the clusters.
-The workflow is general and flexible allowing the user to sustitute each step of the worflow with a different R tool and to input data that have been previously processed by different tools.
-We hope the workflow will ease the technical aspect of single-cell RNA-seq data analysis to help discovering new biological insights.
+# Conclusion
+
+This workflow provides guidance to perform downstream analysis of scRNA-seq datasets in R.
+We propose a worflow in four steps: dimensionality reduction adjusting for cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the lineages.
+The workflow is general and flexible allowing the user to sustitute the statistical method used in each step of the worflow by a different method.
+We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help focusing on discovering new biological insights.
 
@@
 
-The three packages zinbwave, clusterExperiment, and slingshot used in the workflow are bioconductor packages and are available on github at respectively https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
+The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
 The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.

### 2017-06-26 R/uploadToOverleaf.R

@@ -1,0 +1,5 @@
+library(BiocWorkflowTools)
+workflow_dir <- file.path(getwd(), 'workflow')
+uploadToOverleaf(files = workflow_dir,
+                 openInBrowser = TRUE,
+                 forceNewProject = T)

### 2017-06-26 workflow/workflow.Rmd
Minor edits to text.

@@
 ---
-title: "Bioconductor workflow for single-cell RNA sequencing data analysis: dimensionality reduction, clustering, and pseudotime ordering"
+title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
 author: 
-  - name: Author Name1
-    affiliation: Address of author1
-  - name: Author Name2
-    affiliation: Address of author2
+  - name: Fanny Perraudeau
+    affiliation: Graduate Group in Biostatistics, UC Berkeley
+  - name: Davide Risso
+    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
+  - name: Kelly Street
+    affilitation: Graduate Group in Biostatistics, UC Berkeley
+  - name: Elizabeth Purdom
+    affiliation: Department of Statistics, UC Berkeley
+  - name: Sandrine Dudoit
+    affiliation: Division of Biostatistics and Department of Statistics
 abstract: Abstracts should be up to 300 words and provide a succinct summary of the article.
@@
 
-Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enables researchers to measure genome-wide transcription levels for individual cells.
-To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle different problems like dimensionality reduction, clustering, and the inference of cell lineages.
+Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells.
+To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle different problems like dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes.
 While each individual method is useful to analyze scRNA-seq data, there is an increasing need for creating integrated workflows of these new statistical tools.
@@
 
-Some workflows have already been developped to analysis scRNA-seq data with several useful methods for quality control, visualisation and pre-processing of the data.
-The workflow for low-level analyses of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor project [@Huber2015] and cover basic steps including quality control, data exploration and normalization.
+scRNA-seq workflows have already been developed, with several useful methods for quality control, visualization, and pre-processing of the data.
+The workflow for low-level analysis of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor Project [@Huber2015] and cover basic steps, including quality control, data exploration, and normalization.
 In these workflows, single-cell expression data are organized in objects of the SCESet class allowing integrated analysis.
@@
 
-Here, we propose an integrated workflow with dowstream analysis where our workflow has three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of branching lineages and ordering of the cells by developmental progression.
-Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps where for each steps a different R package is used.
+Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression.
+Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where for each steps a different R package is used.
 

### 2017-06-26 workflow/workflow.Rmd
Edits to text.

@@
 Citations should not be used in the abstract, and the use of abbreviations should be minimized.
-keywords: single-cell, workflow, single-cell-genomics, bioconductor, rna-seq, gene-expression, statistical-models, r
+keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
 bibliography: ref.bib
@@
 Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells.
-To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle different problems like dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes.
-While each individual method is useful to analyze scRNA-seq data, there is an increasing need for creating integrated workflows of these new statistical tools.
-This is all the more true that novel sequencing technologies are able to sequence more and more cells in a single sequencing run.
-For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mice brains.
+To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes.
+While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline.
+This is all the more true, with novel sequencing technologies that allow an increasing numer of cells to be sequenced in each run.
+For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.
 
@@
 Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression.
-Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where for each steps a different R package is used.
+Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where a different R package is used for each step.
 
@@
 
-The workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium [@Fletcher2017].
-The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs) which are the actively proliferating cells in the epithelium.
+This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017].
+The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
 When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.
 
-The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium.
-To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, a scRNA-seq analysis was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
-Then, the expression of each gene was quantified by counting the total number of reads mapped to each cell to get a cell by gene matrix of raw counts.
-Different lineages were assigned to each cell in the dataset using a set of statistical tools.
-Finally, results were confirmed experimentally using in vivo lineage tracing.
-Details on the data generation and steps of the analysis are available in [@Fletcher2017].
+The scRNA-seq dataset we work with was generated to study the differentitation of HBCs stem cells into different cell types present in the olfactory epithelium.
+To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
+Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it.
+Cells were then assigned to different lineages using the statistical analysis pipeline presented in this workflow.
+Finally, results were validated experimentally using in vivo lineage tracing.
+Details on data generation and statistical methods are available in [@Fletcher2017] [SD: Add reference to Slingshot and ZINB-WaVE biorXiv manuscripts.].
 

### 2017-06-26 workflow/workflow.Rmd
Edits to text.

@@
 # Analysis of olfactory stem cell lineages using scRNA-seq data
+
 ## Overview
@@
 Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it.
-Cells were then assigned to different lineages using the statistical analysis pipeline presented in this workflow.
+Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow.
 Finally, results were validated experimentally using in vivo lineage tracing.
@@
 It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
-Then, the GBC lineage, in turn, branches to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman’s gland.
+Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland.
 See Figure \@ref(fig:stemcelldiff).
-In this study, we show possible steps to recover the lineages found in the original study from the matrix of raw counts publicly available.
+In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the publicly-available genes x cells matrix of raw counts [SD: Add URL].
 
@@
 
-Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO) with accession number GSE95601. Before filtering, the dataset has 849 cells and 28361 detected genes.
+Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO) with accession number GSE95601.
+Before filtering, the dataset has 849 cells and 28,361 detected genes (i.e., genes with non-zero read counts).
 
@@
 
-We removed the genes corresponding to ERCC spike-ins and the CreER gene as it corresponds to a marker gene for the HBCs cells. Davide, is that correct?
+We remove the ERCC spike-in sequences and the CreER gene as it corresponds to a marker gene for HBCs. Davide, is that correct?
 
@@
 
-Along the workflow, we use a SummarizedExperiment object to keep the counts and metadata in a unique object.
-In the metadata, we dispose of quality control measures for each cell, information on the batches in which the cells were sequenced, and clusters and lineages found in the original publication [@Fletcher2017].
-Cells with a cluster of `-2` are cells not assigned to a cluster in the orginal publication.
+Throughout the workflow, we use an object of class SummarizedExperiment to keep track of the counts and their associated metadata within a single object.
+The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017].
+Cells with a cluster label of `-2` were not assigned to a cluster in the orginal publication.
 
@@
 
-Using bioconductor package scone, we removed low quality cells that did not pass our quality control rule, that is ...
+Using the Bioconductor R package scone [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ...
 See Figure \@ref(fig:scone).
@@
 
-Finally, to speed up the computations, we kept only the 1000 most variable genes. We recommend not to be so stringent in your analysis and perform the analysis on more genes.
+Finally, to speed up the computations, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in your actual data analysis.
 
 ```{r}
-# filtering to top 1000 most variable
+# Filtering to top 1000 most variable
 vars <- rowVars(log1p(assay(se)))
@@
 ## Dataset structure
-Overall, after the preprocessing steps, our dataset has 1000 genes and 747 cells.
+
+Overall, after the pre-processing steps, our dataset has 1,000 genes and 747 cells.
 ```{r}
@@
 
-Metadata for the cells is stored in the object colData from the SummarizedExperiment object. Cells have been processed in 18 different batches.
+Metadata for the cells are stored in the object colData from the SummarizedExperiment object. Cells were processed in 18 different batches.
 
@@
 
-In original work [@Fletcher2017], cells have been clustered into 14 different clusters where 151 cells were not assigned to a cluster.
+In the original work [@Fletcher2017], cells were clustered into 14 different clusters, where 151 cells were not assigned to a cluster [SD: -2 labels?].
 
@@
 ```
-
-Note that some of the batches are confounded with the clusters found in the original work which could be a problem when we correct for batch effect in the dimensionality reduction step.
+Note that there is partial confounding of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.
 
@@
 
-## Dimensionality reduction adjusting for batches
+## Dimensionality reduction adjusting for batches: ZINB-WaVE
 
@@
 
-## Clustering of the cells
+## Cell clustering: RSEC
 
@@
 
-## Pseudotime ordering
+## Cell lineage and pseudotime inference: Slingshot
 
@@
 
-## DE analysis
+## Differential expression analysis
 

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

@@
 The workflow for low-level analysis of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor Project [@Huber2015] and cover basic steps, including quality control, data exploration, and normalization.
-In these workflows, single-cell expression data are organized in objects of the SCESet class allowing integrated analysis.
+In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis.
 However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps like clustering and pseudotime ordering.
 
-Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression.
-Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where a different R package is used for each step.
+Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression.
+Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
 
@@
 
-Throughout the workflow, we use an object of class SummarizedExperiment to keep track of the counts and their associated metadata within a single object.
+Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
 The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017].
@@
 
-Using the Bioconductor R package scone [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ...
+Using the Bioconductor R package `scone` [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ...
 See Figure \@ref(fig:scone).
@@
 
-Metadata for the cells are stored in the object colData from the SummarizedExperiment object. Cells were processed in 18 different batches.
+Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object. Cells were processed in 18 different batches.
 
@@
 ```
-Note that there is partial confounding of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.
+Note that there is partial confounding [SD: Did you mean nesting?] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.
 
@@
 
-In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed.
-It allows the data to become more tractable, both from a statistical (cf.
+In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression (DE) analysis.
+This allows the data to become more tractable, both from a statistical (cf.
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017].
+Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017 ][SD: Ref to ZIFA.].
 
-Here, we perform dimensionality reduction using bioconductor package zinbwave which allows to fit a zero-inflated negative binomial model (ZINB-WaVE) to get a low-dimensional representation of the data while accounting for zero inflation (dropouts), over-dispersion, and the count nature of the data.
+Here, we perform dimensionality reduction using the Bioconductor R package `zinbwave`, which fits a zero-inflated negative binomial model (ZINB-WaVE) that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data.
 The model can include a sample-level intercept, which serves as a global-scaling normalization factor.
@@
 The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
-You can read more details about the ZINB-WaVE model in [@Risso2017].
+For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].
 
-As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space.
-Here, we use `K = 50` dimensions in the low dimensional space and adjusted for the batches.
+[SD: Include Fig. 1 from [@Risso2017]?]
+As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for the batch effects.
 
@@
 
-### Normalized values
+### Normalization
 

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

@@
 
-# Analysis of olfactory stem cell lineages using scRNA-seq data
+# Analysis of olfactory stem cell differentiation using scRNA-seq data
 
@@
 
-## Dimensionality reduction adjusting for batches: ZINB-WaVE
+## Dimensionality reduction: ZINB-WaVE
 
@@
 [SD: Include Fig. 1 from [@Risso2017]?]
-As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for the batch effects.
+As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.
 
@@
 
-When function `zinbDimRed` is called, normalized values of the counts can be computed and used for visualization (e.g.
-in heatmaps).
-The normalized values are the residuals of the fit of the ZINB-WaVE model where the gene and cell covariates were specified by the user when calling function `zinbDimRed`, but no low dimensional matrix is computed.
-In that case, the residuals represent the counts adjusted for the gene and cell covariates and should capture the biological signal of interest.
-To compute the residuals, the deviance is used.
+The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates.
+Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
+Note that in this case no low-dimensional matrix is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.
 
@@
 
-As expected, the normalized values are uniformly distributed for the different batches. See Figure \@ref(fig:boxplotNorm).
+As expected, the normalized values no longer exhibit batch effects (Figure \@ref(fig:boxplotNorm)).
 
@@
 
-The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by orginal published clusters.
-See Figure \@ref(fig:pcanorm).
-Overall, looking at the normalized values, it seems that the batch effect has been removed.
+The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the orginal clusters (Figure \@ref(fig:pcanorm)).
+Overall, it seems that normalization was effective at removing batch effects.
 
@@
 
-Calling function `zinbDimRed` from package `zinbwave`, we performed dimensionality reduction where the number of dimensions in the low dimensional space is `K = 50`.
-We call the matrix in the low dimensional space `W`.
-We can visualize `W` in two dimensions by performing multidimensional scaling (MDS) on the distance matrix of `W` where the euclidian distance is used.
-To make sure that `W` captured the biological signal of interest, we plotted the MDS of `W` with colors corresponding to the original published clusters.
-See Figure \@ref(fig:mdsW).
+The `zinbDimRed` function can also be used to perform dimensionality reduction, where the user-supplied dimension `K` of the low-dimensional space is set here to `K = 50`.
+The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
+To verify that `W` indeed captured the biological signal of interest, we plotted the MDS of `W` with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
 

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

@@
 
-The `zinbDimRed` function can also be used to perform dimensionality reduction, where the user-supplied dimension `K` of the low-dimensional space is set here to `K = 50`.
+The `zinbDimRed` function can also be used to perform dimensionality reduction, where in this workflow the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
-To verify that `W` indeed captured the biological signal of interest, we plotted the MDS of `W` with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
+To verify that `W` indeed captures the biological signal of interest, we display the MDS results with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
 

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression.
+Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; and (3) inference of cell lineages and ordering of the cells by developmental progression.
 Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
+[SD: Include DE as 4th step, to match statement in "Conclusion"?]
 
@@
 
-## Dimensionality reduction: ZINB-WaVE
+## Normalization and dimensionality reduction: ZINB-WaVE
 
@@
 
-```{r boxplotNorm, fig.cap="Boxplot of normalized values. Different colors are for different batches."}
+```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures, color-coded by batch."}
 norm_order <- norm[, order(as.numeric(batch))]
@@
 
-```{r pcanorm, fig.cap="PCA of normalized values where each point represent a cell.
-In the left panel, cells have been colored according to their batch.
-In the right panel, colors correspond to the original published clusters."}
+```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell.
+In the left panel, cells are color-coded by batch.
+In the right panel, cells are color-coded by original published clusters."}
 pca <- prcomp(t(norm))
@@
 
-```{r mdsW, fig.cap="MDS of the low dimensional matrix `W` where each point represent a cell. Cells have been colored according to the original published clusters."}
+```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clusters."}
 W <- colData(se)[, grepl('^W', colnames(colData(se)))]
@@
 
-The next step of the workflow is to cluster the cells according to the low dimensional matrix `W` computed in the previous step.
-We used function `RSEC` from bioconductor package clusterExperiment.
-Function `RSEC` implements a resampling-based sequential ensemble clustering method for identifying stable and tight cell clusters.
-The method aggregates multiple clustering results obtained from different base clustering algorithms and applications of a given algorithm to resampled versions of the learning set.
+The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step.
+We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
+Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[SD: Add ref.].
+A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE.
+As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.
 
@@
 
-We can visualize the resulting clusterings using the `plotClusters` function.
-See Figure \@ref(fig:examineCombineMany) where cells are in the columns, and different clusterings on the rows.
-Each sample is color coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap.
-The first raw correspond to a consensus clustering across the many different clusterings performed.
+The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings.
+Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap.
+The first row correspond to a consensus clustering across all candidate clusterings.
 
-```{r examineCombineMany, fig.cap="Clusters found using function RSEC from clusterExperiment package."}
+```{r examineCombineMany, fig.cap="Clusters found using the function RSEC from the clusterExperiment package."}
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
 ```
+The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).
 
-We can also visualize the proportion of times the consensus clusters were together across the different clusterings. See Figure \@ref(fig:plotcoclust).
-
-```{r plotcoclust, fig.cap="Heatmap of the proportion of times the clusters were together across the different clusterings."}
+```{r plotcoclust, fig.cap="Heatmap of co-clustering matrix."}
 plotCoClustering(ceObj)
@@
 
-Over stable and robust clustering, the number of cells in each cluster is as follow
+The distribution of cells across the consensus clusters is as follows:
 
@@
 
-We can now plot a heatmap of the normalized values for the 1000 most variable genes where the cells have been clustered using the `RSEC` function. See Figure \@ref(fig:heatmapsClusters).
+Figure \@ref(fig:heatmapsClusters) displays a heatmap of the normalized expression measures for the 1,000 most variable genes, where cells are clustered according to the RSEC consensus.
 
-```{r heatmapsClusters, fig.cap="Heatmap of the normalized values for the 1000 most variable genes where genes are in the rows and cells in the columns."}
-# set colors for cell clusterings
+```{r heatmapsClusters, fig.cap="Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
+# Set colors for cell clusterings
 colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
@@
 
-Finally, we can visualize the cells in a two dimensional space by plotting the MDS of the low dimensional matrix `W` coloring the cells by their newly found clusters.
-See Figure \@ref(fig:mdsWce) which is the same as Figure \@ref(fig:mdsW) but where colors are for the newly found clusters instead of the clusters found in the orginal published work.
+Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is anologous to Figure \@ref(fig:mdsW) for the original published clusters.
 
-```{r mdsWce, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the clusters found using our workflow."}
+```{r mdsWce, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
 palDF <- ceObj@clusterLegend[[1]]
@@
 
-In an effort to improve the workflow and analysis of scRNA-seq data, we are currently exploring statistical methods to perform DE analysis using the models we developed.
-One the one hand, we are developing a DE analysis method using the framework of slingshot which would take advantage of the lineages found.
-On the other hand, we are working on a DE analysis method using the ZINB-WaVE model which would account for the dropouts and gene and cell-level covariates.
-Additionally, we are working on a statistical method using the ZINB-WaVE model to impute the dropouts in the scRNA-seq data.
-This matrix with imputed counts could for example be used in the workflow in the dimensionality reduction, clustering or pseudo ordering steps.
+In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model.
+In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference.
+ZINB-WaVE can also be used for identifying genes that are differentially expressed between cells, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
+We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.
 
-Finally, a new R class called SingleCellExperiment should be released soon.
-This new class would essentially be a SummarizedExperiment class with a couple of additional slots, the most important of which is reducedDims, which, much like the assays slot of SummarizedExperiment can contain one or more matrices with reduced dimensional matrices.
-This new SingleCellExperiment class would be a valuable addition to the workflow as we could store in one object the raw counts as well as the low dimensional matrix created by the dimensionality reduction step.
+Finally, a new R class called `SingleCellExperiment` should be released soon.
+This new class would essentially be a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment` can contain one or more matrices of reduced dimension.
+This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step.
 Once this class is implemented, we would like to incorporate it to the workflow.
@@
 
-This workflow provides guidance to perform downstream analysis of scRNA-seq datasets in R.
-We propose a worflow in four steps: dimensionality reduction adjusting for cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the lineages.
-The workflow is general and flexible allowing the user to sustitute the statistical method used in each step of the worflow by a different method.
-We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help focusing on discovering new biological insights.
+This workflow provides a tutorial for the downstream analysis of scRNA-seq data in R.
+The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages.
+The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method.
+We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help with the discovery of novel biological insights.
 

### 2017-06-28 workflow/workflow.Rmd
adress sandrine s comments

@@
 #bioconductor:
-library(scRNAseq)
-library(scone)
 library(clusterExperiment)
@@
 library(zinbwave)
-library(GEOquery)
 #github
+library(scone)
 library(slingshot)
@@
 library(doParallel)
-library(ggplot2)
-library(gplots)
-library(magrittr)
-library(matrixStats)
-library(Rtsne)
 library(RColorBrewer)
-library(digest)
-library(rARPACK)
 set.seed(20)
@@
 While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline.
-This is all the more true, with novel sequencing technologies that allow an increasing numer of cells to be sequenced in each run.
+This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run.
 For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.
@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; and (3) inference of cell lineages and ordering of the cells by developmental progression.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages.
 Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
-[SD: Include DE as 4th step, to match statement in "Conclusion"?]
 
@@
 ```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium.
-This figure has been reproduced with kind permission from [@Fletcher2017].'}
+This figure has been reproduced with kind permission from (Fletcher et al.
+2017).'}
 knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
@@
 Finally, results were validated experimentally using in vivo lineage tracing.
-Details on data generation and statistical methods are available in [@Fletcher2017] [SD: Add reference to Slingshot and ZINB-WaVE biorXiv manuscripts.].
+Details on data generation and statistical methods are available in [@Fletcher2017] [@Risso2017], and [@Street2017].
 
@@
 See Figure \@ref(fig:stemcelldiff).
-In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the publicly-available genes x cells matrix of raw counts [SD: Add URL].
+In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
 
@@
 The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017].
-Cells with a cluster label of `-2` were not assigned to a cluster in the orginal publication.
+Cells with a cluster label of `-2` were not assigned to a cluster in the original publication.
 
@@
 
-In the original work [@Fletcher2017], cells were clustered into 14 different clusters, where 151 cells were not assigned to a cluster [SD: -2 labels?].
+In the original work [@Fletcher2017], cells were clustered into 14 different clusters, where 151 cells have a cluster label of `-2`, that is they were not assigned to a cluster in the original publication.
+[SD: -2 labels?
+FP: yes]
 
@@
 ```
-Note that there is partial confounding [SD: Did you mean nesting?] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.
+
+Note that there is partial nesting [SD: Did you mean nesting? FP: yes] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.
 
@@
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017 ][SD: Ref to ZIFA.].
+Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].
 
@@
 The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
+See a schematic view of the ZINB-WaVE model in Figure \@ref(fig:zinbschema).
 For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].
 
-[SD: Include Fig. 1 from [@Risso2017]?]
+```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure has been reproduced with kind permission from (Risso et al. 2017)."}
+knitr::include_graphics('zinb_schema.pdf')
+```
+
 As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.
@@
 We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
-Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[SD: Add ref.].
+Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
 A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE.
@@
 ```
+
 The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).

### 2017-06-28 workflow/workflow.Rmd
add new slingshot and gam code from kelly

@@
 #github
-library(scone)
 library(slingshot)
+#library(scone)
 #CRAN
@@
 library(RColorBrewer)
+library(gam)
 set.seed(20)
@@
 
-```{r scone, fig.cap="Plot generated by scone to filter low quality cells."}
+```{r scone, fig.cap="Plot generated by scone to filter low quality cells.", eval=FALSE}
 # Metric-based Filtering sample-filtering
@@
 
-mfilt <- metric_sample_filter(assay(se), nreads = colData(se)$NREADS,
+mfilt <- metric_sample_filter(assay(se), 
+                              nreads = colData(se)$NREADS,
                               ralign = colData(se)$RALIGN,
@@
 
-```{r sconeFilt}
+```{r sconeFilt, eval=FALSE}
 # Simplify to a single logical
@@
 
-```{r}
+```{r,eval=FALSE}
 # Filtering to top 1000 most variable
@@
 core <- se[names(vars)[1:1000],]
-save(core, file="../data/oe_se_1000Var.rda")
+```
+
+```{r}
+load("../data/oe_se_1000Var.rda")
 ```
@@
 We now want to demonstrate how to use the R software package `slingshot` to infer branching lineages and order cells by developmental progression.
-We connect the clusters identified by `RSEC` with a minimum spanning tree to learn the global lineage structure.
+We connect the clusters identified by `RSEC` with a minimum spanning tree (MST) to learn the global lineage structure.
 Then, we refine this structure and order cells using highly stable simultaneous principal curves to infer smooth, branching lineages.
@@
 
-Kelly, would you like to give details about slingshot algorithm?
+The two steps of the slingshot algorithm are implemented in the functions 'getLineages' and 'getCurves'.
+The first takes the low-dimensional representation of the cells and the vector of cluster labels identified above.
+It fits an MST to the clusters and identifies lineages as paths through this tree.
+The output of 'getLineages' is a 'SlingshotDataSet' object containing all the information used to fit the tree and identify lineages.
+'getCurves' takes this object as input and fits simultaneous principal curves to the identified lineages.
+These functions can be run separately, as below, or jointly by the wrapper function, 'slingshot'.
 
@@
 
-lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
-curves <- get_curves(X, clus.labels = cl, lineages = lineages)
+lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
+lineages <- getCurves(lineages)
 ```
@@
 ```{r curves, fig.cap="Cells colored by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
-plot_curves(X, cl, curves, col.clus = pal)
+pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
 ```
@@
 ```{r tree, fig.cap="Cells colored by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
-plot_tree(X, cl, lineages, col.clus = pal)
+pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
 ```
@@
 ```{r lineages}
-print(lineages$lineage1)
-print(lineages$lineage2)
-print(lineages$lineage3)
+lineages
 ```
@@
 However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
-Slingshot has the option for the user to provide the end cluster(s).
-Here is the code to use slingshot in a supervised setting
+'getLineages' has the option for the user to provide known end cluster(s).
+Here is the code to use slingshot in a supervised setting:
 
 ```{r slingshotsupervised, eval=FALSE}
-lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
+lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                          end.clus = c("c3", "c6"))
-curves <- get_curves(X, clus.labels = cl, lineages = lineages)
-plot_curves(X, cl, curves, col.clus = pal)
-plot_tree(X, cl, lineages, col.clus = pal)
+lineagees <- getCurves(lineages)
+pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
 
-print(lineages$lineage1)
-print(lineages$lineage2)
-print(lineages$lineage3)
+lineages
 ```
@@
 
-Kelly, would you like to complete this part? Here is the kind of plots we would want to present.
+After ordering the cells, we are interested in finding genes that have non-constant expression patterns over pseudotime.
+We can see global trends in expression by subsetting our original data to the most variable genes and plotting their expression values in a heatmap, ordered by pseudotime.
 
-```{r de, eval = FALSE}
-de <- read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
-de <- de$V1
-plotHeatmap(ceObj, 
-            visualizeData = assays(se)$normalizedValues[rownames(se) %in% de, ],
-            clusterSamplesData = "dendrogramValue",
-            whichClusters = "primary",
-            main = 'Normalized values, 1000 most variable genes',
-            breaks = 0.99)
+More formally, we are interested in testing for variable expression over pseudotime.
+A general additive model (GAM) is a flexible, nonlinear way to test the relationship between a gene's normalized expression values and pseudotime.
+We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.
+
+```{r fitgam}
+t <- pseudotime(lineages)[,1]
+y <- assays(se)$normalizedValues[, our_cl != "-1"]
+gam.pval <- apply(y,1,function(z){
+  d <- data.frame(z=z, t=t)
+  tmp <- gam(z ~ lo(t), data=d)
+  p <- summary(tmp)[4][[1]][1,5]
+  p
+})
+```
+
+```{r heatmapsignificant}
+topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
+heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
+heatclus <- cl[order(t, na.last = NA)]
+ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
+plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
 ```

### 2017-06-28 workflow/workflow.Rmd
Update workflow.Rmd

@@
   - name: Sandrine Dudoit
-    affiliation: Division of Biostatistics and Department of Statistics
+    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley
 abstract: Abstracts should be up to 300 words and provide a succinct summary of the article.
@@
 Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells.
-To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes.
+To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and differential expression (DE) analysis.
 While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline.
@@
 
-scRNA-seq workflows have already been developed, with several useful methods for quality control, visualization, and pre-processing of the data.
-The workflow for low-level analysis of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor Project [@Huber2015] and cover basic steps, including quality control, data exploration, and normalization.
+scRNA-seq low-level analysis workflows have already been developed, with useful methods for quality control (QC), exploratory data analysis (EDA), pre-processing, normalization, and visualization.
+The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015].
 In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis.
-However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps like clustering and pseudotime ordering.
+However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages.
 Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
 
-```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing scRNA-seq datasets. On the right side, major plots generated by the workflow."}
+```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing scRNA-seq datasets. On the right, main plots generated by the workflow."}
 knitr::include_graphics('schema_workflow.png')
 ```
+
+[SD: Fig. 1: For DE, might want to deemphasize edgeR, as it is not our contribution, by mentioning other packages.]
 
@@
 ```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium.
-This figure has been reproduced with kind permission from (Fletcher et al.
+This figure was reproduced with kind permission from (Fletcher et al.
 2017).'}
@@
 This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017].
-The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
+The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
 When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.
 
-The scRNA-seq dataset we work with was generated to study the differentitation of HBCs stem cells into different cell types present in the olfactory epithelium.
+The scRNA-seq dataset we use as case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium.
 To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
@@
 Finally, results were validated experimentally using in vivo lineage tracing.
-Details on data generation and statistical methods are available in [@Fletcher2017] [@Risso2017], and [@Street2017].
+Details on data generation and statistical methods are available in [@Fletcher2017],  [@Risso2017], and [@Street2017].
 
 It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
-Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland.
-See Figure \@ref(fig:stemcelldiff).
+Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)_.
 In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
@@
 
-Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO) with accession number GSE95601.
+Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO), with accession number GSE95601.
 Before filtering, the dataset has 849 cells and 28,361 detected genes (i.e., genes with non-zero read counts).
@@
 
-We remove the ERCC spike-in sequences and the CreER gene as it corresponds to a marker gene for HBCs. Davide, is that correct?
+We remove the ERCC spike-in sequences and the CreER gene, as the later corresponds to a marker gene for HBCs. Davide, is that correct?
 
@@
 Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
-The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017].
+The cell-level metadata contain quality control (QC) measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
 Cells with a cluster label of `-2` were not assigned to a cluster in the original publication.
@@
 
-Finally, to speed up the computations, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in your actual data analysis.
+Finally, to speed up the computations, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.
 
@@
 
-In the original work [@Fletcher2017], cells were clustered into 14 different clusters, where 151 cells have a cluster label of `-2`, that is they were not assigned to a cluster in the original publication.
-[SD: -2 labels?
-FP: yes]
+In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).
 
@@
 
-Note that there is partial nesting [SD: Did you mean nesting? FP: yes] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.
+Note that there is partial nesting of batches within clusters (i.e., cell type), which could be problematic when correcting for batch effects in the dimensionality reduction step.
 
@@
 
-In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression (DE) analysis.
+In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression analysis.
 This allows the data to become more tractable, both from a statistical (cf.
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].
+Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].
 
@@
 The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
-See a schematic view of the ZINB-WaVE model in Figure \@ref(fig:zinbschema).
+A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema).
 For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].
 
-```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure has been reproduced with kind permission from (Risso et al. 2017)."}
+```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from (Risso et al. 2017)."}
 knitr::include_graphics('zinb_schema.pdf')
@@
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
-To verify that `W` indeed captures the biological signal of interest, we display the MDS results with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
+To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
 

### 2017-06-28 workflow/workflow.Rmd
Edits to text.

@@
     affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
-  - name: Kelly Street
+  - name: Kelly N. Street
     affilitation: Graduate Group in Biostatistics, UC Berkeley
@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages.
 Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
@@
 The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings.
-Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap.
+Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap accross rows.
 The first row correspond to a consensus clustering across all candidate clusterings.
@@
 
-We now want to demonstrate how to use the R software package `slingshot` to infer branching lineages and order cells by developmental progression.
-We connect the clusters identified by `RSEC` with a minimum spanning tree (MST) to learn the global lineage structure.
-Then, we refine this structure and order cells using highly stable simultaneous principal curves to infer smooth, branching lineages.
+We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage.
+The method comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
+The approach in (1) allows us to identify any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.
 
-From the original published work, we know that the start clusters should be the clusters with HBCs and end clusters the clusters with MV, mOSN, and mSUS cells.
-Additionally, we know that the clusters with the GBCs should be a junction before the differentiation between MV and mOSN cells.
-See Figure \@ref(fig:stemcelldiff).
-The correspondance between the clusters we found and the original clusters is as follow
+The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`.
+The first takes a low-dimensional representation of the cells and a vector of cluster labels.
+It fits an MST to the clusters and identifies lineages as paths through this tree.
+The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages.
+The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages.
+These functions can be run separately, as below, or jointly by the wrapper function, `slingshot`.
+
+From the original published work, we know that the start cluster should correspond to HBCs and the end clusters to MV, mOSN, and mSUS cells.
+Additionally, we know that GBCs should be at a junction before the differentiation between MV and mOSN cells (Figure \@ref(fig:stemcelldiff)).
+The correspondance between the clusters we found here and the original clusters is as follows.
 
@@
 -------------|-------------|-------| ----------
-c1 | HBCs | blue | original 1, 5
-c2 | GBCs | green | original 2, 3, 11
+c1 | HBC | blue | original 1, 5
+c2 | GBC | green | original 2, 3, 11
 c3 | mSUS | red | original 4, 7
@@
 
-To order the clusters and find lineages, we use package `slingshot` where the input is the MDS of low dimensional matrix `W` where the number of dimensions used for MDS is `k = 4`.
-We found that slingshot algorithm is stable to the number of dimensions `k` of the MDS.
-Here, we use the unsupervised version of slingshot where we only provide the information of the start cluster but not the end clusters.
-Using slingshot, we recovered the order of the clusters found in the published work.
-See Figures \@ref(fig:curves) and \@ref(fig:tree).
-
-The two steps of the slingshot algorithm are implemented in the functions 'getLineages' and 'getCurves'.
-The first takes the low-dimensional representation of the cells and the vector of cluster labels identified above.
-It fits an MST to the clusters and identifies lineages as paths through this tree.
-The output of 'getLineages' is a 'SlingshotDataSet' object containing all the information used to fit the tree and identify lineages.
-'getCurves' takes this object as input and fits simultaneous principal curves to the identified lineages.
-These functions can be run separately, as below, or jointly by the wrapper function, 'slingshot'.
+To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`.
+We found that the slingshot results were robust to the number of dimensions `k` for the MDS (we tried `k=` [SD: Add values of `k` that were considered.]).
+Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters.
+This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).
 
@@
 However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
-'getLineages' has the option for the user to provide known end cluster(s).
-Here is the code to use slingshot in a supervised setting:
+`getLineages` has the option for the user to provide known end cluster(s).
+Here is the code to use `slingsho` in a supervised setting.
 
@@
 
-After ordering the cells, we are interested in finding genes that have non-constant expression patterns over pseudotime.
-We can see global trends in expression by subsetting our original data to the most variable genes and plotting their expression values in a heatmap, ordered by pseudotime.
+After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime.
+We can see global trends in expression by subsetting our original dataset to the most variable genes and displaying their expression values in a heatmap, with cells ordered by pseudotime.
 
 More formally, we are interested in testing for variable expression over pseudotime.
-A general additive model (GAM) is a flexible, nonlinear way to test the relationship between a gene's normalized expression values and pseudotime.
+A general additive model (GAM) is a flexible, nonlinear way to model the relationship between a gene's normalized expression measures and pseudotime.
 We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.
@@
 This workflow provides a tutorial for the downstream analysis of scRNA-seq data in R.
-The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages.
+The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages.
 The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method.
@@
 # Author contributions
-In order to give appropriate credit to each author of an article, the individual contributions of each author to the manuscript should be detailed in this section.
-We recommend using author initials and then stating briefly how they contributed.
+
+[SD: Please check.]
+
+FP, DR, EP, and KNS performed the data analaysis. DR, SD, and EP supervised the research. All authors contributed to the writing of the manuscript.
 
 # Competing interests
-All financial, personal, or professional competing interests for any of the authors that could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article.
-If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.
+
+No competing interests were disclosed.
 
 # Grant information
-Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc.
-Please do not list funding that you have that is not relevant to this specific piece of research.
-For each funder, please state the funder’s name, the grant number where applicable, and the individual to whom the grant was assigned.
-If your work was not funded by any grants, please include the line: 'The author(s) declared that no grants were involved in supporting this work.'
+
+[SD: Please check.]
+
+DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (grant U01 MH105979, PI: John Ngai).
+KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).
 

### 2017-06-29 workflow/workflow.Rmd
Updates to text.

@@
     affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley
-abstract: Abstracts should be up to 300 words and provide a succinct summary of the article.
-Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article.
-Citations should not be used in the abstract, and the use of abbreviations should be minimized.
+
+abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types.
+However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software.
+Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.
+
 keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
+
 bibliography: ref.bib
+
 vignette: >
@@
 
+[SD: Fix double affiliation for SD so that there are 2 numbers.]
+
 EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis alonglineages.
 Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
@@
 
-[SD: Fig. 1: For DE, might want to deemphasize edgeR, as it is not our contribution, by mentioning other packages.]
+[SD: Fig. 1: For DE, remove mention to edgeR.]
 
@@
 
-Using the Bioconductor R package `scone` [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ...
+Using the Bioconductor R package `scone`, we remove low-quality cells that did not pass our quality control filter, that is ...
 See Figure \@ref(fig:scone).
@@
 We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
-Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
+Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
 A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE.
@@
 To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`.
-We found that the slingshot results were robust to the number of dimensions `k` for the MDS (we tried `k=` [SD: Add values of `k` that were considered.]).
+We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_= [SD: Add values of _k_ that were considered.]).
 Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters.
@@
 # Software availability
+
 This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.
@@
 
-The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
+The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are Bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
 The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.
@@
 
-FP, DR, EP, and KNS performed the data analaysis. DR, SD, and EP supervised the research. All authors contributed to the writing of the manuscript.
+FP, DR, KNS, and EP performed the data analaysis and wrote the code portions of the workflow.
+FP and SD wrote the text portion of the workflow, with contributions from the other three authors.
+DR, EP, and SD supervised the research.
 
@@
 
-DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (grant U01 MH105979, PI: John Ngai).
+DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai).
 KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).
@@
 # Acknowledgments
-This section should acknowledge anyone who contributed to the research or the
-article but who does not qualify as an author based on the criteria provided earlier (e.g.
-someone or an organization that provided writing assistance).
-Please state how they contributed; authors should obtain permission to acknowledge from all those mentioned in the Acknowledgments section.
+
+[SD: Please check.]
+
+The authors are grateful to John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Russell B.
+Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.
 

### 2017-06-29 workflow/workflow.Rmd
Updates to text.

@@
 ---
-title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
+title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
 author: 
   - name: Fanny Perraudeau
-    affiliation: Graduate Group in Biostatistics, UC Berkeley
+    affiliation: Graduate Group in Biostatistics, University of California, Berkeley
   - name: Davide Risso
@@
   - name: Kelly N. Street
-    affilitation: Graduate Group in Biostatistics, UC Berkeley
+    affilitation: Graduate Group in Biostatistics, University of California, Berkeley
   - name: Elizabeth Purdom
-    affiliation: Department of Statistics, UC Berkeley
+    affiliation: Department of Statistics, University of California, Berkeley
   - name: Sandrine Dudoit
-    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley
+    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley
 
@@
 
-keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
+keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow, R, Bioconductor
 
@@
 
+[SD: Fix double affiliation for SD so that there are 2 numbers.]
+
 ```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
@@
 
-#bioconductor:
+# Bioconductor:
 library(clusterExperiment)
@@
 library(zinbwave)
-#github
+# GitHub
 library(slingshot)
 #library(scone)
-#CRAN
+# CRAN
 library(doParallel)
@@
 ```
-
-[SD: Fix double affiliation for SD so that there are 2 numbers.]
 
@@
 
-# count matrix
+# Count matrix
 E <- assayData(Cufflinks_eSet)$counts_table
@@
 ```{r}
-# remove ERCC and CreER genes
+# Remove ERCC and CreER genes
 cre <- E["CreER",]
@@
 
-# symbol for cell not assigned to a lineage in original data
+# Symbol for cells not assigned to a lineage in original data
 metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
@@
 ```{r scone, fig.cap="Plot generated by scone to filter low quality cells.", eval=FALSE}
-# Metric-based Filtering sample-filtering
+# QC-metric-based sample-filtering
 data("housekeeping")
@@
 ```{r,eval=FALSE}
-# Filtering to top 1000 most variable
+# Filtering to top 1000 most variable genes
 vars <- rowVars(log1p(assay(se)))
@@
 if (runClus & !file.exists(fn)){
-  #symbol for samples missing from original clustering
+  # Symbol for samples missing from original clustering
   seObj <- SummarizedExperiment(t(W), colData = colData(core))
@@
 
-## Differential expression analysis
+## Differential expression analysis along lineages
 
@@
 
-[SD: Please check.]
-
 FP, DR, KNS, and EP performed the data analaysis and wrote the code portions of the workflow.
@@
 
-[SD: Please check.]
-
 The authors are grateful to John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Russell B.

### 2017-06-29 workflow/workflow.Rmd
Updates to text.

@@
 ```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium.
-This figure was reproduced with kind permission from (Fletcher et al.
-2017).'}
+This figure was reproduced with kind permission from Fletcher et al.
+(2017).'}
 knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
@@
 Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
-The cell-level metadata contain quality control (QC) measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
+The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
 Cells with a cluster label of `-2` were not assigned to a cluster in the original publication.
@@
 
-```{r scone, fig.cap="Plot generated by scone to filter low quality cells.", eval=FALSE}
+```{r scone, fig.cap="Plot generated by scone to filter low-quality cells.", eval=FALSE}
 # QC-metric-based sample-filtering
@@
 
-```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from (Risso et al. 2017)."}
+```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
 knitr::include_graphics('zinb_schema.pdf')
@@
 
-```{r curves, fig.cap="Cells colored by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
+```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
 pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
@@
 
-```{r tree, fig.cap="Cells colored by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
+```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
 pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])

### 2017-06-30 workflow/workflow.Rmd
Updates to text.
change to workflow

@@
 ```
-
-EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.
 
@@
 Overall, after the pre-processing steps, our dataset has 1,000 genes and 747 cells.
+
 ```{r}
@@
 Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
-A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE.
+A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings.
+The consensus clustering is further condensed by merging similar clusters which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed.
 As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.
-
-Elizabeth, would you like to more details on the rational and arguments of the different functions?
 
@@
 ```
+
+EP: the label `clusterLabels` for the original labels might do better to be changed to `publishedClusters` so that it is more informative.
+Also, we don't do a direct comparison of the `publishedClusters`
+ to the found clusters -- is that relevant?
+If so, it is probably better to add the published clusters via `addClusters` so it is easier to treat them like clusters.
+This is perhaps useful for a workflow since this is a common task that `clusterExperiment` makes easy.
+Here is code that does this:
+
+ ```{r addPublishedClusters}
+ #delete this chunk if not of interest
+ #if keep, need to choose if do both figures and add figure caption 
+ ceObj<-addClusters(ceObj,colData(ceObj)$clusterLabels,clusterLabel="publishedClusters")
+ plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
+ plotBarplot(ceObj,whichClusters=c("mergeClusters","publishedClusters")) 
+ ```
+
+ EP: If we do the above code, then the `plotHeatmap` command below should be modified to have the `publishedClusters` under the `whichClusters` argument.
 
@@
 
+[SD: Mention reason for having these two figures. What do colors correspond to in second figure? Match notation for dimensions (along diagonal) in the two figures.]
+
 ```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
@@
 More formally, we are interested in testing for variable expression over pseudotime.
-A general additive model (GAM) is a flexible, nonlinear way to model the relationship between a gene's normalized expression measures and pseudotime.
+We do this by fitting a loess curve to the data which is a flexible, nonlinear way to model the relationship between a gene's normalized expression measures and pseudotime.
+We then can test the null hypothesis of no change over time using the `gam` package.
 We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.
@@
 ```
+
+[SD: Add figure caption.]
 

### 2017-06-30 workflow/workflow.Rmd
add text about RSEC options

@@
 
+EP: Added the below paragraph
+
+The defaults in `RSEC` are designed for input data that is the actual (normalized) count data.
+We are going to apply it instead to the `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object from `W`.
+For this reason we will choose to not use certain options in `RSEC`.
+In particular, we will not use the default dimensionality reduction step provided by `RSEC` since our input `W` is already in a reduced dimensional space.
+Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and to the construction of the hiearchy between the clusters (`dendroReduce`).
+We will also skip the option to merge our clusters based on the amount of differential gene expression between clusters since our input data is `W` and not the actual individual gene expression values.
+
 ```{r rsec_50}
@@
                 ncores = NCORES, isCount=FALSE,
+				dendroReduce="none",dendroNDims=NA,
                 subsampleArgs = list(resamp.num=100,

### 2017-06-30 workflow/workflow.Rmd
fix RSEC command and update version requirement

@@
 if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<'1.3.1.9003') stop("must have current develop version to avoid bugs")
 ```
@@
                                      clusterArgs=list(nstart=10)),
-                seqArgs = list(k.min=3, top.can=5), verbose=TRUE,
+                verbose=TRUE,
                 combineProportion = 0.7,

### 2017-06-30 workflow/workflow.Rmd
Updates to text.

@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis alonglineages.
-Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
+Throughout the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
 
@@
 The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
-When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.
+When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.
 
-The scRNA-seq dataset we use as case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium.
+The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium.
 To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
-Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it.
+The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it.
 Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow.
@@
 It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
-Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)_.
-In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
+Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)).
+In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
 
@@
 The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
-Cells with a cluster label of `-2` were not assigned to a cluster in the original publication.
+Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.
 
@@
 
-Using the Bioconductor R package `scone`, we remove low-quality cells that did not pass our quality control filter, that is ...
+Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter, that is ...
 See Figure \@ref(fig:scone).
@@
 
-Finally, to speed up the computations, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.
+Finally, to speed up computation, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.
 
@@
 
-In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression analysis.
+In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and the identification of DE genes.
 This allows the data to become more tractable, both from a statistical (cf.
@@
 
-Here, we perform dimensionality reduction using the Bioconductor R package `zinbwave`, which fits a zero-inflated negative binomial model (ZINB-WaVE) that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data.
+Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`.
+The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data.
 The model can include a sample-level intercept, which serves as a global-scaling normalization factor.
 The user can also include both gene-level and sample-level covariates.
-The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.
+The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
 A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema).
@@
 
-The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates.
+The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
-Note that in this case no low-dimensional matrix is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.
+Note that in this case no low-dimensional matrix `W` is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.
 
@@
 
-The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the orginal clusters (Figure \@ref(fig:pcanorm)).
+The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the original clusters (Figure \@ref(fig:pcanorm)).
 Overall, it seems that normalization was effective at removing batch effects.
@@
 
-The `zinbDimRed` function can also be used to perform dimensionality reduction, where in this workflow the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
+The `zinbDimRed` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.

### 2017-07-01 workflow/workflow.Rmd
Updates to text.

@@
 We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
-Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
+Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
 A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings.
@@
 
-The defaults in `RSEC` are designed for input data that is the actual (normalized) count data.
-We are going to apply it instead to the `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object from `W`.
-For this reason we will choose to not use certain options in `RSEC`.
-In particular, we will not use the default dimensionality reduction step provided by `RSEC` since our input `W` is already in a reduced dimensional space.
-Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and to the construction of the hiearchy between the clusters (`dendroReduce`).
-We will also skip the option to merge our clusters based on the amount of differential gene expression between clusters since our input data is `W` and not the actual individual gene expression values.
+Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts.
+Here, we are applying `RSEC` instead to the low-dimensional `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object.
+For this reason, we choose to not use certain options in `RSEC`.
+In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension.
+Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`).
+We also skip the option to merge our clusters based on the amount of differential gene expression between clusters since our input data is `W` and not the actual individual gene expression values.
 
@@
 The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages.
-These functions can be run separately, as below, or jointly by the wrapper function, `slingshot`.
+These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.
 
@@
 
-[SD: Mention reason for having these two figures. What do colors correspond to in second figure? Match notation for dimensions (along diagonal) in the two figures.]
+[SD: Mention reason for having these two figures, otherwise drop second figure.
+If keep second figure: Move it first, match notation for dimensions (along diagonal) in the two figures, and provide color legend for second figure.]
 
@@
 `getLineages` has the option for the user to provide known end cluster(s).
-Here is the code to use `slingsho` in a supervised setting.
+Here is the code to use `slingshot` in a supervised setting.
 
@@
 After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime.
-We can see global trends in expression by subsetting our original dataset to the most variable genes and displaying their expression values in a heatmap, with cells ordered by pseudotime.
+We can see global trends in expression by subsetting our original dataset to the most variable genes [SD: How many genes?] and displaying their expression values in a heatmap, with cells ordered by pseudotime.
 
-More formally, we are interested in testing for variable expression over pseudotime.
-We do this by fitting a loess curve to the data which is a flexible, nonlinear way to model the relationship between a gene's normalized expression measures and pseudotime.
+More formally, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
 We then can test the null hypothesis of no change over time using the `gam` package.
-We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.
+We implemented this approach for each gene and display the expression measures of the top 100 genes by p-value in the heatmap of Figure [SD: Add figure ref.].
 

### 2017-07-01 workflow/workflow.Rmd
task no space and 2 affiliations sandrine

@@
   - name: Sandrine Dudoit
-    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley
+    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley
 
@@
 However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software.
-Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.
+Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.
 
@@
 ---
-
-[SD: Fix double affiliation for SD so that there are 2 numbers.]
 

### 2017-07-01 workflow/workflow.Rmd
start cleaning up repo

@@
 
+[FP: we need to have only 8 keywords, which ones do you prefer?]
+
+[FP: todo before to convert to latex:
+
+- solve loading packages issues (DR),
+- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR),
+- clean up singlecellworkflow package (FP),
+- update NAMESPACE and DESCRIPTION (FP),
+- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
+- make singlecellworkflow package public (FP).]
+
+[FP: todo after converting to latex (FP):
+
+- add space between paragraphs, 
+- check uniformity first/last names in references, 
+- add back space after word task in abstract, 
+- submit.]
+
 # Introduction
@@
 
-[SD: Fig. 1: For DE, remove mention to edgeR.]
-
 # Analysis of olfactory stem cell differentiation using scRNA-seq data
@@
 Finally, results were validated experimentally using in vivo lineage tracing.
-Details on data generation and statistical methods are available in [@Fletcher2017],  [@Risso2017], and [@Street2017].
+Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Street2017].
 
@@
 In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
-
-FP: ask Russell if we can use the Figure \@ref(fig:stemcelldiff) here.
 
@@
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].
+Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017; @Risso2017; @Pierson2015].
 
@@
 A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema).
-For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].
+For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].
 
@@
 We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
-Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005].
+Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in [@Tseng2005].
 A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings.
@@
 As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.
-
-EP: Added the below paragraph
 
@@
 ```
+
+FP: Same comment as Davide's comment for zinbwave. We should remove the if file exists and rely on the cache.
 
@@
 To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`.
-We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_= [SD: Add values of _k_ that were considered.]).
+We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5).
 Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters.
@@
 
+The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow.
 The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are Bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
-The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.
 

### 2017-07-01 R/run_zifa.py

@@ -1,35 +1,0 @@
-#!/usr/bin/env python
-
-import sys
-sys.path.insert(1, '/usr/local/lib/python2.7/site-packages/') ##change the path if needed
-
-from argparse import ArgumentParser
-from pandas import read_csv
-from ZIFA import ZIFA,block_ZIFA
-import numpy as np
-
-def main():
-    parser = ArgumentParser(description="Fit a ZIFA model on the data.")
-    parser.add_argument('-b', '--block', action='store_true', default=False, help="Whether the block algorithm should be used.")
-    parser.add_argument('-d', '--dimensions', type=int, default=2, help="The number of dimensions [2].")
-    parser.add_argument('input_file', type=str, help="The input CSV file.")
-    parser.add_argument('output_file', type=str, help="The output CSV file.")
-
-    args = parser.parse_args()
-
-    df = read_csv(args.input_file)
-    del df['Unnamed: 0']
-
-    lc = np.array(df)
-    Y = np.transpose(lc)
-
-    if(args.block):
-        Z, model_params  = block_ZIFA.fitModel(Y, args.dimensions)
-    else:
-        Z, model_params  = ZIFA.fitModel(Y, args.dimensions)
-
-    np.savetxt(args.output_file, Z, delimiter=',')
-
-
-if __name__ == '__main__':
-        main()

### 2017-07-01 workflow/workflow.Rmd
compare ours and published

@@
 - versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR),
+- update clustering section with new clusterExperiment functions (EP),
+- add details in section preprocessing (DR),
+- address Sandrine's comments in sections slingshot and de (KS),
 - clean up singlecellworkflow package (FP),
@@
                        "Batch" = batch,
-                       "clusterLabels" = clusterLabels[m,2],
+                       "publishedClusters" = clusterLabels[m,2],
                        qc)
@@
 # Symbol for cells not assigned to a lineage in original data
-metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
+metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2
 
@@
 load("../data/oe_se_1000Var.rda")
+colnames(colData(core))[colnames(colData(core)) == 'clusterLabels'] = 'publishedClusters' 
 ```
+[FP: delete this chunk once we can load scone]
 
@@
 ```{r original}
-clus.labels <- colData(core)[, "clusterLabels"]
+publishedClusters <- colData(core)[, "publishedClusters"]
 col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
-col_clus <- col_clus[1:length(unique(clus.labels))]
-names(col_clus) <- sort(unique(clus.labels))
-table(clus.labels)
+col_clus <- col_clus[1:length(unique(publishedClusters))]
+names(col_clus) <- sort(unique(publishedClusters))
+table(publishedClusters)
 ```
@@
 table(data.frame(batch = as.vector(batch),
-                 cluster = clus.labels))
+                 cluster = publishedClusters))
 ```
@@
 plot(pca$x, col = col_batch[batch], pch = 20, main = '')
-plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
+plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = '')
 ```
@@
 fit <- cmdscale(d, eig = TRUE, k = 2)
-plot(fit$points, col = col_clus[as.character(clus.labels)], main = '',
+plot(fit$points, col = col_clus[as.character(publishedClusters)], main = '',
      pch = 20, xlab = 'Component 1', ylab = 'Component 2')
@@
   load(fn)
+  colnames(colData(ceObj))[colnames(colData(ceObj)) == 'clusterLabels'] = 'publishedClusters'
 }
@@
 
-The distribution of cells across the consensus clusters is as follows:
+The distribution of cells across the consensus clusters can be visualized in Figure \@ref(fig:barplotOurs)) and is as follows:
 
@@
 table(primaryClusterNamed(ceObj))
+```
+
+```{r barplotOurs, fig.cap="Barplot of number of cells per cluster found using our workflow."}
+plotBarplot(ceObj) 
+```
+
+[FP: Should we add here the short paragraph Elizabeth was talking about to compare our found clusters versus the original published clusters?]
+
+```{r addPublishedClusters, fig.cap="Barplot of number of cells per cluster in original published clusters."}
+ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
+                   clusterLabel = "publishedClusters")
+plotBarplot(ceObj,whichClusters=c("mergeClusters","publishedClusters")) 
 ```
@@
 
- ```{r addPublishedClusters}
- #delete this chunk if not of interest
- #if keep, need to choose if do both figures and add figure caption 
- ceObj<-addClusters(ceObj,colData(ceObj)$clusterLabels,clusterLabel="publishedClusters")
+[FP: after the modifications, do we still need function addCluters?
+If we want to illustrate that clusterExperiment package has a convenient function to add clusters to a clusterExperiment object, we can keep it and say that here it is not necessary.]
+
+```{r comparePublished, fig.cap="Comparison between the clusters found using our workflow and original published clusters."}
 plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
- plotBarplot(ceObj,whichClusters=c("mergeClusters","publishedClusters")) 
 ```
 
- EP: If we do the above code, then the `plotHeatmap` command below should be modified to have the `publishedClusters` under the `whichClusters` argument.
+[EP: If we do the above code, then the `plotHeatmap` command below should be modified to have the `publishedClusters` under the `whichClusters` argument. ]
+
+[FP: Elizabeth, I changed the `clusterLabels` to `publishedClusters`, more informative indeed.
+Maybe, we could add the plotClusters to compare the published and our own clusters and the plotBarplot to illustrate our previous call to function table()?
+I modified the code accordingly.
+If you don't like it, feel free to change, it is just a proposition.]
 
@@
 # Set colors for cell clusterings
-colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
-origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
+colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
+origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
 experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
@@
                    "Batch" = batchColors,
-                   "clusterLabels" = origClusterColors)
+                   "publishedClusters" = origClusterColors)
 
@@
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
-            sampleData = c("clusterLabels", "Batch", "Experiment"),
+            sampleData = c("publishedClusters", "Batch", "Experiment"),
             clusterLegend = metaColors, annLegend = FALSE, main = '')
@@
 ```{r tabagain}
-table(data.frame(original = clus.labels, ours = primaryClusterNamed(ceObj)))
+table(data.frame(original = publishedClusters, ours = primaryClusterNamed(ceObj)))
 ```

### 2017-07-01 workflow/workflow.Rmd
Edits to text and figure captions.

@@
 
-keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow, R, Bioconductor
+keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
 
@@
 
-[FP: we need to have only 8 keywords, which ones do you prefer?]
+[FP: we need to have only 8 keywords, which ones do you prefer? SD: I selected 8, please check.]
 
@@
 
-```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures, color-coded by batch."}
+```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
 norm_order <- norm[, order(as.numeric(batch))]
@@
 boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
-        border = col_order, xaxt = 'n')
+        border = col_order, xaxt = 'n', ylab='Expression measure')
 abline(h=0)
@@
 In the left panel, cells are color-coded by batch.
-In the right panel, cells are color-coded by original published clusters."}
+In the right panel, cells are color-coded by original published clustering."}
 pca <- prcomp(t(norm))
@@
 
-```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clusters."}
+```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
 W <- colData(se)[, grepl('^W', colnames(colData(se)))]
@@
 
-```{r examineCombineMany, fig.cap="Clusters found using the function RSEC from the clusterExperiment package."}
+```{r examineCombineMany, fig.cap="Candidate clusterings found using the function RSEC from the clusterExperiment package."}
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
@@
 
-```{r barplotOurs, fig.cap="Barplot of number of cells per cluster found using our workflow."}
+```{r barplotOurs, fig.cap="Barplot of number of cells per cluster, for our workflow's clusterint."}
 plotBarplot(ceObj) 
@@
 
-```{r addPublishedClusters, fig.cap="Barplot of number of cells per cluster in original published clusters."}
+```{r addPublishedClusters, fig.cap="Barplot of number of cells per cluster, for original published clustering."}
 ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
@@
 
-```{r comparePublished, fig.cap="Comparison between the clusters found using our workflow and original published clusters."}
+```{r comparePublished, fig.cap="Comparison between the clustering found using our workflow and the original published clustering."}
 plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
@@
 
-```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional space with smooth curves representing each inferred lineage."}
+```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
 pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
@@
 
-```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional space with connecting lines between cluster centers representing the inferred structure."}
+```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
 pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])

### 2017-07-03 workflow/workflow.Rmd
Edits to text.

@@
 A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings.
-The consensus clustering is further condensed by merging similar clusters which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed.
+The consensus clustering is further condensed by merging similar clusters, which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed.
 As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.
@@
 We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage.
-The method comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
+The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
 The approach in (1) allows us to identify any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.
 
-The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`.
+The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively.
 The first takes a low-dimensional representation of the cells and a vector of cluster labels.
@@
 We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5).
-Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters.
+Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters.
 This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).
@@
 
-In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot.
+In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot.
 However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
@@
 After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime.
-We can see global trends in expression by subsetting our original dataset to the most variable genes [SD: How many genes?] and displaying their expression values in a heatmap, with cells ordered by pseudotime.
+We can see global trends in expression by subsetting our original dataset to the most variable genes [SD: How many genes?] and displaying their expression values in a heatmap, with cells ordered by pseudotime [SD: Add figure ref.].
 
@@
 Finally, a new R class called `SingleCellExperiment` should be released soon.
-This new class would essentially be a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment` can contain one or more matrices of reduced dimension.
+This new class would essentially be a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension.
 This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step.
@@
 
-This workflow provides a tutorial for the downstream analysis of scRNA-seq data in R.
-The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages.
+This workflow provides a tutorial for the analysis of scRNA-seq data in R.
+It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
 The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method.
-We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help with the discovery of novel biological insights.
+We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.
 
@@
 The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow.
-The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are Bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
+The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available on GitHub at, respectively, https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
 
@@
 
-The authors are grateful to John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Russell B.
+The authors are grateful to Professor John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Dr.
+Russell B.
 Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.

### 2017-07-03 workflow/workflow.Rmd
Edits to text.

@@
   - name: Fanny Perraudeau
-    affiliation: Graduate Group in Biostatistics, University of California, Berkeley
+    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
   - name: Davide Risso
-    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
+    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
   - name: Kelly N. Street
-    affilitation: Graduate Group in Biostatistics, University of California, Berkeley
+    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
   - name: Elizabeth Purdom
-    affiliation: Department of Statistics, University of California, Berkeley
+    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
   - name: Sandrine Dudoit
-    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley
+    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
 
@@
 
+[SD: Refer to 3D version, used to create Figure 2.]
+
 ```{r}
@@
 After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime.
-We can see global trends in expression by subsetting our original dataset to the most variable genes [SD: How many genes?] and displaying their expression values in a heatmap, with cells ordered by pseudotime [SD: Add figure ref.].
 
-More formally, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
-We then can test the null hypothesis of no change over time using the `gam` package.
-We implemented this approach for each gene and display the expression measures of the top 100 genes by p-value in the heatmap of Figure [SD: Add figure ref.].
+More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
+We then can test the null hypothesis of no change over time for each gene using the `gam` package.
+We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
 
@@
 
-[SD: Add figure caption.]
-
-```{r heatmapsignificant}
+```{r heatmapsignificant, fig.cap="Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
 topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
@@
 
-[SD: Please check.]
-
 DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai).
@@
 Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.
+We would also like to thank Michael B.
+Cole for his contributions to `scone`.
 

### 2017-07-03 workflow/workflow.Rmd
Changes to scone description + zinbwave() function

@@
 ```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
+if(packageVersion("zinbwave")<"0.99.5") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.3.1.9003") stop("must have current develop version to avoid bugs")
+if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
+
 library(BiocStyle)
@@
 knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
-
-# Bioconductor:
-library(clusterExperiment)
-library(BiocParallel)
-library(zinbwave)
-# GitHub
-library(slingshot)
-#library(scone)
-# CRAN
-library(doParallel)
-library(RColorBrewer)
-library(gam)
-set.seed(20)
-if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
-if(packageVersion("clusterExperiment")<'1.3.1.9003') stop("must have current develop version to avoid bugs")
-```
-
-```{r parallel, echo=FALSE}
-runZinb <- TRUE
-runClus <- TRUE
-NCORES <- 2
-mysystem = Sys.info()[['sysname']]
-if (mysystem == 'Darwin'){
-  registerDoParallel(NCORES)
-  register(DoparParam())
-}else if (mysystem == 'Linux'){
-  register(bpstart(MulticoreParam(workers=NCORES)))
-}else{
-  print('Please change this to allow parallel computing')
-  register(SerialParam())
-}
 ```
@@
 
-- solve loading packages issues (DR),
-- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR),
+- solve loading packages issues (DR), DONE!
+- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR), [DR: scone and zinbwave now OK!]
 - update clustering section with new clusterExperiment functions (EP),
@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
-Throughout the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
+Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.
 
-```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing scRNA-seq datasets. On the right, main plots generated by the workflow."}
-knitr::include_graphics('schema_workflow.png')
+```{r schema, echo=FALSE, out.width="90%", fig.cap="Workflow for analyzing scRNA-seq datasets. On the right, main plots generated by the workflow."}
+knitr::include_graphics("schema_workflow.png")
 ```
@@
 
-```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium.
+```{r stemcelldiff, echo=FALSE, out.width="60%", fig.align="center", fig.cap = "Stem cell differentiation in the mouse olfactory epithelium.
 This figure was reproduced with kind permission from Fletcher et al.
-(2017).'}
-knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
+(2017)."}
+knitr::include_graphics("stemcelldiff_Fletcher2017_2e.png")
 ```
@@
 In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
+
+## Package versions
+
+We need the following packages.
+
+```{r packages}
+# Bioconductor:
+library(clusterExperiment)
+library(BiocParallel)
+library(zinbwave)
+library(scone)
+
+# GitHub
+library(slingshot)
+
+# CRAN
+library(doParallel)
+library(RColorBrewer)
+library(gam)
+set.seed(20)
+```
+
+Note that in order to successfully run the workflow, we need the devel versions
+of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)` and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!].
+We recommend running Bioconductor 3.6 (currently the devel version).
+See https://www.bioconductor.org/developers/how-to/useDevel/ on how to use the Bioconductor devel.
+
+```{r parallel, echo=FALSE}
+runZinb <- TRUE
+runClus <- TRUE
+NCORES <- 6
+mysystem = Sys.info()[["sysname"]]
+if (mysystem == "Darwin"){
+  registerDoParallel(NCORES)
+  register(DoparParam())
+}else if (mysystem == "Linux"){
+  register(bpstart(MulticoreParam(workers=NCORES)))
+}else{
+  print("Please change this to allow parallel computing")
+  register(SerialParam())
+}
+```
 
@@
 
-We remove the ERCC spike-in sequences and the CreER gene, as the later corresponds to a marker gene for HBCs. Davide, is that correct?
+We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBC into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).
 
@@
 
-Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter, that is ...
-See Figure \@ref(fig:scone).
-Davide, would you like to describe your choices for scone here?
+Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter (See Figure \@ref(fig:scone)).
+The function `metric_sample_filter` uses several criteria to filter out low-quality samples.
+First, we filter out samples with low total number of reads or low alignment percentages.
+Then, we look at the number of housekeeping genes that are detected in each sample, and filter out those samples with a low detection rate.
+See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
 
-```{r scone, fig.cap="Plot generated by scone to filter low-quality cells.", eval=FALSE}
+```{r scone, fig.cap="Plot generated by scone to filter low-quality cells."}
 # QC-metric-based sample-filtering
@@
 
-```{r sconeFilt, eval=FALSE}
+```{r sconeFilt}
 # Simplify to a single logical
@@
 
+After sample filtering, we are left with `r NCOL(se)` good quality cells.
+
 Finally, to speed up computation, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.
 
-```{r,eval=FALSE}
+```{r}
 # Filtering to top 1000 most variable genes
@@
 ```
-
-```{r}
-load("../data/oe_se_1000Var.rda")
-colnames(colData(core))[colnames(colData(core)) == 'clusterLabels'] = 'publishedClusters' 
-```
-[FP: delete this chunk once we can load scone]
 
@@
 
-```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
-knitr::include_graphics('zinb_schema.pdf')
+```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
+knitr::include_graphics("zinb_schema.pdf")
 ```
@@
 ```{r zinb}
-fn <- '../data/zinb_batch.rda'
-if (runZinb & !file.exists(fn)){
-  print(system.time(se <- zinbDimRed(core, K = 50, X = '~ Batch',
+print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",
                                  residuals = TRUE,
                                  normalizedValues = TRUE)))
-  save(se, file = fn)
-}else{
-  load(fn)
-}
 ```
-
-DR: the chunk above and the similar one for clusterExperiment are fine for now, but in the published workflow, we should just rely on the markdown cache system (the code above doesn't check for changes to the file or code -- just that a version of the file exists).
-FP: Yes.
 
@@
 
-The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates.
+The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
-Note that in this case no low-dimensional matrix `W` is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.
+Note that in this case the low-dimensional matrix `W` is not included in the residuals computation, to avoid the removal of the biological signal of interest.
+
+[DR: Fanny, did you figure out why we get infinite values? If not, we need to at least explain why we are doing what we do below.]
 
@@
 boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
-        border = col_order, xaxt = 'n', ylab='Expression measure')
+        border = col_order, xaxt = "n", ylab="Expression measure")
 abline(h=0)
@@
 par(mfrow = c(1,2))
-plot(pca$x, col = col_batch[batch], pch = 20, main = '')
-plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = '')
+plot(pca$x, col = col_batch[batch], pch = 20, main = "")
+plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
 ```
@@
 
-The `zinbDimRed` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
+The `zinbwave` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
@@
 ```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
-W <- colData(se)[, grepl('^W', colnames(colData(se)))]
+W <- colData(se)[, grepl("^W", colnames(colData(se)))]
 W <- as.matrix(W)
@@
 fit <- cmdscale(d, eig = TRUE, k = 2)
-plot(fit$points, col = col_clus[as.character(publishedClusters)], main = '',
-     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
-legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
-       fill = unique(col_clus), title = 'Sample')
+plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
+     pch = 20, xlab = "Component 1", ylab = "Component 2")
+legend(x = "bottomright", legend = unique(names(col_clus)), cex = .5,
+       fill = unique(col_clus), title = "Sample")
 ```
@@
 ```{r rsec_50}
-fn <- '../data/RSEC_W_combineMinSize10.rda'
-if (runClus & !file.exists(fn)){
-  # Symbol for samples missing from original clustering
 seObj <- SummarizedExperiment(t(W), colData = colData(core))
@@
                                 combineMinSize = 10)))
-  save(ceObj, file = fn)
-}else{
-  load(fn)
-  colnames(colData(ceObj))[colnames(colData(ceObj)) == 'clusterLabels'] = 'publishedClusters'
-}
 ```
-
-FP: Same comment as Davide's comment for zinbwave. We should remove the if file exists and rely on the cache.
 
@@
             sampleData = c("publishedClusters", "Batch", "Experiment"),
-            clusterLegend = metaColors, annLegend = FALSE, main = '')
+            clusterLegend = metaColors, annLegend = FALSE, main = "")
 ```
@@
 palDF <- ceObj@clusterLegend[[1]]
-pal <- palDF[, 'color']
-names(pal) <- palDF[, 'name']
+pal <- palDF[, "color"]
+names(pal) <- palDF[, "name"]
 pal["-1"] = "transparent"
-plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = '', pch = 20,
-     xlab = 'Component1', ylab = 'Component2')
-legend(x = 'bottomright', legend = names(pal), cex = .5,
-       fill = pal, title = 'Sample')
+plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
+     xlab = "Component1", ylab = "Component2")
+legend(x = "bottomright", legend = names(pal), cex = .5,
+       fill = pal, title = "Sample")
 ```
@@
 cl = our_cl[our_cl != "-1"]
-pal = pal[names(pal) != '-1']
+pal = pal[names(pal) != "-1"]
 X <- W[our_cl != "-1", ]
@@
 ```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
-pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
 ```
@@
 ```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
-pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
 ```
@@
 lineagees <- getCurves(lineages)
-pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
-pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
 
@@
 
-More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
+More formally, for each lineage, we use the robust local regression method `loess` to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
 We then can test the null hypothesis of no change over time for each gene using the `gam` package.
-We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
+We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes ordered by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
 
@@
 In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference.
-ZINB-WaVE can also be used for identifying genes that are differentially expressed between cells, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
+In addition, we are extending ZINB-WaVE to differentially expression, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
 We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.

### 2017-07-04 workflow/workflow.Rmd
updated Slingshot section, per Sandrine's comments.

--- 2017-07-03 workflow/workflow.Rmd
@@
     affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
-  - name: Kelly N. Street
+  - name: Kelly Street
     affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
@@
 Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters.
-This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).
-
-[SD: Refer to 3D version, used to create Figure 2.]
 
@@
 lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
-lineages <- getCurves(lineages)
 ```
@@
 
-```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
-pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
-```
+Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters.
+This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).
+The `slingshot` package also includes methods for 3-dimensional visualizations similar to \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.
 
@@
 pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
+```
+
+Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables.
+Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters.
+This makes them more stable and better suited for assigning cells to lineages.
+The final curves are shown in \@ref(fig:curves).
+
+```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
+lineages <- getCurves(lineages)
+
+pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
 ```
@@
 `getLineages` has the option for the user to provide known end cluster(s).
-Here is the code to use `slingshot` in a supervised setting.
+Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c8` and `c6` represent terminal cell fates.
 
@@
 lineagees <- getCurves(lineages)
-pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
-pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)
 
@@
 
-FP, DR, KNS, and EP performed the data analaysis and wrote the code portions of the workflow.
+FP, DR, KS, and EP performed the data analaysis and wrote the code portions of the workflow.
 FP and SD wrote the text portion of the workflow, with contributions from the other three authors.
@@
 
-DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai).
-KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).
+DR, KS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai).
+KS was supported by a training grant from the National Human Genome Research Institute (T32000047).
 

### 2017-07-04 workflow/workflow.Rmd
merged my edits and Davide's edits

--- 2017-07-03 workflow/workflow.Rmd
@@
 cl = our_cl[our_cl != "-1"]
-pal = pal[names(pal) != "-1"]
+pal = pal[names(pal) != '-1']
 X <- W[our_cl != "-1", ]
@@
 
-```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
-pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
+Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters.
+This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).
+The `slingshot` package also includes methods for 3-dimensional visualizations similar to \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.
+
+```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
+pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
 ```
 
-```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
-pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
+Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables.
+Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters.
+This makes them more stable and better suited for assigning cells to lineages.
+The final curves are shown in \@ref(fig:curves).
+
+```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
+lineages <- getCurves(lineages)
+
+pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
 ```
@@
 lineagees <- getCurves(lineages)
-pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
-pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)
 
@@
 
-More formally, for each lineage, we use the robust local regression method `loess` to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
+More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime.
 We then can test the null hypothesis of no change over time for each gene using the `gam` package.
-We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes ordered by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
+We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
 

### 2017-07-04 workflow/workflow.Rmd
Edits to text.

@@
 Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells.
-To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and differential expression (DE) analysis.
+To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and the identification of differentially expressed (DE) genes.
 While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline.
@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
 Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.
@@
 ```{r packages}
-# Bioconductor:
+# Bioconductor
+library(BiocParallel)
 library(clusterExperiment)
-library(BiocParallel)
+library(scone)
 library(zinbwave)
-library(scone)
 
@@
 library(doParallel)
+library(gam)
 library(RColorBrewer)
-library(gam)
+
 set.seed(20)
@@
 Note that in order to successfully run the workflow, we need the devel versions
-of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)` and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!].
-We recommend running Bioconductor 3.6 (currently the devel version).
-See https://www.bioconductor.org/developers/how-to/useDevel/ on how to use the Bioconductor devel.
+of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)`, and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!].
+We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
 
@@
 
-We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBC into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).
+We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).
 
@@
 
-Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter (See Figure \@ref(fig:scone)).
-The function `metric_sample_filter` uses several criteria to filter out low-quality samples.
-First, we filter out samples with low total number of reads or low alignment percentages.
-Then, we look at the number of housekeeping genes that are detected in each sample, and filter out those samples with a low detection rate.
+Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)).
+First, we filter out samples with low total number of reads or low alignment percentage.
+Next, we filter out samples with a low detection rate for housekeeping genes.
 See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
@@
 ```{r}
-# Filtering to top 1000 most variable genes
+# Filtering to top 1,000 most variable genes
 vars <- rowVars(log1p(assay(se)))
@@
 
-Overall, after the pre-processing steps, our dataset has 1,000 genes and 747 cells.
+Overall, after the above pre-processing steps, our dataset has 1,000 genes and 747 cells.
 
@@
 
-The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates.
+The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
-Note that in this case the low-dimensional matrix `W` is not included in the residuals computation, to avoid the removal of the biological signal of interest.
+Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.
 
@@
 The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step.
-We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
+We use the resampling-based sequential ensemble clustering (RSEC) framework implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`.
 Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in [@Tseng2005].
@@
 The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
-The approach in (1) allows us to identify any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.
+The approach in (1) allows the identification of any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.
 
@@
 
-To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`.
-We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5).
-Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters.
+To infer lineages and pseudotimes, we apply Slingshot to the 4-dimensional MDS of the low-dimensional matrix `W`.
+We found that the Slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5).
+Here, we use the unsupervised version of Slingshot, where we only provide the identity of the start cluster but not of the end clusters.
 
@@
 This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).
-The `slingshot` package also includes methods for 3-dimensional visualizations similar to \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.
+The `slingshot` package also includes methods for 3-dimensional visualizations as in Figure \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.
 
@@
 This makes them more stable and better suited for assigning cells to lineages.
-The final curves are shown in \@ref(fig:curves).
+The final curves are shown in Figure \@ref(fig:curves).
 

### 2017-07-04 workflow/workflow.Rmd
edits in DESCRIPTION

@@
 
-[FP: we need to have only 8 keywords, which ones do you prefer? SD: I selected 8, please check.]
-
 [FP: todo before to convert to latex:
@@
 - update clustering section with new clusterExperiment functions (EP),
-- add details in section preprocessing (DR),
+- add details in section preprocessing (DR), DONE,
 - address Sandrine's comments in sections slingshot and de (KS),
+- try running workflow with more genes (FP),
+- plot gene variance density (FP),
 - clean up singlecellworkflow package (FP),
@@
 
-pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
 ```
@@
 lineagees <- getCurves(lineages)
-pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
-pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)
+pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)
 

### 2017-07-04 workflow/workflow.Rmd
start investigating infinite values in norm. values

@@
 ```{r parallel, echo=FALSE}
-runZinb <- TRUE
-runClus <- TRUE
-NCORES <- 6
+NCORES <- 2
 mysystem = Sys.info()[["sysname"]]
@@
 ```{r zinb}
-print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",
+print(system.time(se <- zinbwave(core, K = 40, X = "~ Batch",  
                                  residuals = TRUE,
@@
 [DR: Fanny, did you figure out why we get infinite values? If not, we need to at least explain why we are doing what we do below.]
+[FP: I tried to play with the epsilon to see if we could get rid of the infinite values in the normalized values.
+With default epsilon, we have about 40 infinite values, with epsilon 10000, we have about 20.
+When K = 0, there is no infinite values but the normalized values don't capture as much of biology (looking at the pca with published clusters colors).
+It seems to me that we get infinite norm.
+values when the number of parameters to estimate is large.
+I tried to reduce K to 25 (no infinite but W not as good), 40 (infinite values but W looks ok)...
+to be continued.]
 

### 2017-07-04 workflow/workflow.Rmd
Updates to text.

@@
 The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
-The approach in (1) allows the identification of any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.
+The approach in (1) allows the identification of any number of novel lineages, while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.
 
@@
 This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).
-The `slingshot` package also includes methods for 3-dimensional visualizations as in Figure \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.
+The `slingshot` package also includes functionality for 3-dimensional visualization as in Figure \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.
 

### 2017-07-05 workflow/workflow.Rmd
inf norm values resp

@@
 [DR: Fanny, did you figure out why we get infinite values? If not, we need to at least explain why we are doing what we do below.]
-[FP: I tried to play with the epsilon to see if we could get rid of the infinite values in the normalized values.
-With default epsilon, we have about 40 infinite values, with epsilon 10000, we have about 20.
-When K = 0, there is no infinite values but the normalized values don't capture as much of biology (looking at the pca with published clusters colors).
-It seems to me that we get infinite norm.
-values when the number of parameters to estimate is large.
-I tried to reduce K to 25 (no infinite but W not as good), 40 (infinite values but W looks ok)...
-to be continued.]
+[FP: I created a pull request on github. Let me know what you think.]
 

### 2017-07-05 workflow/workflow.Rmd
Updates to text.

@@
 
-Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)).
-First, we filter out samples with low total number of reads or low alignment percentage.
-Next, we filter out samples with a low detection rate for housekeeping genes.
+Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
 See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
@@
 
-After sample filtering, we are left with `r NCOL(se)` good quality cells.
+After sample filtering, we are left with `r ncol(se)` [SD: Check.] good quality cells.
 
@@
 
-Note that there is partial nesting of batches within clusters (i.e., cell type), which could be problematic when correcting for batch effects in the dimensionality reduction step.
+Note that there is partial nesting of batches within clusters (i.e., cell type), which could be problematic when correcting for batch effects in the dimensionality reduction step below.
+[SD: Might refer to this issue below again.]
 
@@
 curse of dimensionality) and computational point of view.
-Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017; @Risso2017; @Pierson2015].
+Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017; @Pierson2015; @Risso2017].
 
@@
 
-The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates.
+The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
@@
 Overall, it seems that normalization was effective at removing batch effects.
+[SD: Maybe here: " without removing biological signal, in spite of the partial nesting of batches within clusters".]
 

### 2017-07-05 workflow/workflow.Rmd
Updated package versions to latest

@@
 ```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
-if(packageVersion("zinbwave")<"0.99.5") stop("must have current devel version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.3.1.9003") stop("must have current develop version to avoid bugs")
+if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.3.1.9004") stop("must have current develop version to avoid bugs")
 if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
@@
   download.file(urls[1], "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
-  gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+  R.utils::gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
 }
@@
 plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
-```
-
-```{r,echo=FALSE}
-par(mfrow = c(1,1))
 ```

### 2017-07-05 workflow/workflow.Rmd
updated color vector in slingshot pairs plots.

@@
 ```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
-pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="lineages", col = pal[cl])
 ```
@@
 
-pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="curves", col = pal[cl])
 ```

### 2017-07-05 workflow/workflow.Rmd
fix plotBarplot

@@
                    clusterLabel = "publishedClusters")
-plotBarplot(ceObj,whichClusters=c("mergeClusters","publishedClusters")) 
+plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters")) 
 ```

### 2017-07-05 workflow/workflow.Rmd
clean up a bit

@@
 
-- solve loading packages issues (DR), DONE!
-- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR), [DR: scone and zinbwave now OK!]
+- make sure versions package ok and pushed to bioc (DR, EP, FP),
 - update clustering section with new clusterExperiment functions (EP),
-- add details in section preprocessing (DR), DONE,
 - address Sandrine's comments in sections slingshot and de (KS),
-- try running workflow with more genes (FP),
-- plot gene variance density (FP),
+- try running workflow with more genes (DR),
 - clean up singlecellworkflow package (FP),
@@
 Note that in order to successfully run the workflow, we need the devel versions
-of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)`, and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!].
+of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!].
 We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
@@
 ```{r zinb}
-print(system.time(se <- zinbwave(core, K = 40, X = "~ Batch",  
+print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",  
                                  residuals = TRUE,
@@
 
-[DR: Fanny, did you figure out why we get infinite values? If not, we need to at least explain why we are doing what we do below.]
-[FP: I created a pull request on github. Let me know what you think.]
-
 ```{r norm}
 norm <- assays(se)$normalizedValues
-if (sum(is.infinite(norm))>0){
-  maxNorm = max(norm[!is.infinite(norm)])
-  assays(se)$normalizedValues[is.infinite(norm)] <- maxNorm
-  norm <- assays(se)$normalizedValues
-}
+stopifnot(sum(is.infinite(norm))==0)
 norm[1:3,1:3]
@@
 ```{r barplotOurs, fig.cap="Barplot of number of cells per cluster, for our workflow's clusterint."}
-plotBarplot(ceObj) 
+plotBarplot(ceObj, legend = F) 
 ```
@@
 [FP: Should we add here the short paragraph Elizabeth was talking about to compare our found clusters versus the original published clusters?]
+[FP: we should add a comment on cluster 7.]
 
@@
                    clusterLabel = "publishedClusters")
-plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters")) 
+plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters"), legend = F) 
 ```
-
-EP: the label `clusterLabels` for the original labels might do better to be changed to `publishedClusters` so that it is more informative.
-Also, we don't do a direct comparison of the `publishedClusters`
- to the found clusters -- is that relevant?
-If so, it is probably better to add the published clusters via `addClusters` so it is easier to treat them like clusters.
-This is perhaps useful for a workflow since this is a common task that `clusterExperiment` makes easy.
-Here is code that does this:
-
-[FP: after the modifications, do we still need function addCluters?
-If we want to illustrate that clusterExperiment package has a convenient function to add clusters to a clusterExperiment object, we can keep it and say that here it is not necessary.]
 
@@
 ```
-
-[EP: If we do the above code, then the `plotHeatmap` command below should be modified to have the `publishedClusters` under the `whichClusters` argument. ]
-
-[FP: Elizabeth, I changed the `clusterLabels` to `publishedClusters`, more informative indeed.
-Maybe, we could add the plotClusters to compare the published and our own clusters and the plotBarplot to illustrate our previous call to function table()?
-I modified the code accordingly.
-If you don't like it, feel free to change, it is just a proposition.]
 

### 2017-07-05 workflow/workflow.Rmd
add sandrines comment on nested batch-bio

@@
 
-After sample filtering, we are left with `r ncol(se)` [SD: Check.] good quality cells.
+After sample filtering, we are left with `r ncol(se)` good quality cells.
 
 Finally, to speed up computation, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.
+
+[FP: add an additional comment to justify that we keep only 1000 most variable genes. We can say that it is enough to recover the signal from [@Fletcher2017]. ]
 
@@
 Note that there is partial nesting of batches within clusters (i.e., cell type), which could be problematic when correcting for batch effects in the dimensionality reduction step below.
-[SD: Might refer to this issue below again.]
 
@@
 The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the original clusters (Figure \@ref(fig:pcanorm)).
-Overall, it seems that normalization was effective at removing batch effects.
-[SD: Maybe here: " without removing biological signal, in spite of the partial nesting of batches within clusters".]
+Overall, it seems that normalization was effective at removing batch effects without removing biological signal, in spite of the partial nesting of batches within clusters.
 

### 2017-07-06 workflow/workflow.Rmd
small changes to text

@@
 if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.3.1.9004") stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current develop version to avoid bugs")
 if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
@@
 Note that in order to successfully run the workflow, we need the devel versions
-of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!].
+of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.3.2)`.
 We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
@@
 
-Finally, to speed up computation, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.
+Finally, we retain only the 1,000 most variable genes.
+It speeds up the computations and allows to remove genes with low biological signal.
+As the 1,000 most variable genes allow to recover the biological signal found in the published analysis ([@Fletcher2017]), we think it is a reasonable choice for this particular analysis.
+We recommend, however, to explore the number of genes required to extract the biological signal in an actual data analysis.
 
-[FP: add an additional comment to justify that we keep only 1000 most variable genes. We can say that it is enough to recover the signal from [@Fletcher2017]. ]
+[FP: check]
 
@@
 
-The distribution of cells across the consensus clusters can be visualized in Figure \@ref(fig:barplotOurs)) and is as follows:
+The distribution of cells across the consensus clusters can be visualized in Figure \@ref(fig:barplotOurs) and is as follows:
 
@@
 
-```{r barplotOurs, fig.cap="Barplot of number of cells per cluster, for our workflow's clusterint."}
+```{r barplotOurs, fig.cap="Barplot of number of cells per cluster, for our workflow's clustering."}
 plotBarplot(ceObj, legend = F) 
@@
 
-[FP: Should we add here the short paragraph Elizabeth was talking about to compare our found clusters versus the original published clusters?]
-[FP: we should add a comment on cluster 7.]
+The distribution of cells in our workflow's clustering is in agreement with the distribution of cells in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)) where the major difference is that in the published analysis several clusters have been merged into single clusters in our workflow's clustering.
+This difference is likelihood to be caused by the fact that we performed the dimensionality reduction step using the ZINB-WaVE model and input the W matrix to the RSEC function instead of performing the dimensionality reduction on the normalized counts within the RSEC function, resulting
+
+[FP: check]
 

### 2017-07-06 workflow/workflow.Rmd
Edits to text.

@@
 
-Finally, we retain only the 1,000 most variable genes.
-It speeds up the computations and allows to remove genes with low biological signal.
-As the 1,000 most variable genes allow to recover the biological signal found in the published analysis ([@Fletcher2017]), we think it is a reasonable choice for this particular analysis.
-We recommend, however, to explore the number of genes required to extract the biological signal in an actual data analysis.
+Finally, for computational efficiency, we retain only the 1,000 most variable genes.
+This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
+In general, however, we recommend care in selecting an approapriate gene filtering scheme for an actual data analysis.
 
-[FP: check]
+[FP: check. SD: I rephrased the above paragraph.]
 
@@
 Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`).
-We also skip the option to merge our clusters based on the amount of differential gene expression between clusters since our input data is `W` and not the actual individual gene expression values.
+We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.
 
@@
 
-The distribution of cells in our workflow's clustering is in agreement with the distribution of cells in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)) where the major difference is that in the published analysis several clusters have been merged into single clusters in our workflow's clustering.
-This difference is likelihood to be caused by the fact that we performed the dimensionality reduction step using the ZINB-WaVE model and input the W matrix to the RSEC function instead of performing the dimensionality reduction on the normalized counts within the RSEC function, resulting
+The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged into single clusters.
+This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to RSEC instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.
 
-[FP: check]
+[FP: check. SD: Rephrased a little.]
 
@@
 
-[SD: Mention reason for having these two figures, otherwise drop second figure.
-If keep second figure: Move it first, match notation for dimensions (along diagonal) in the two figures, and provide color legend for second figure.]
+[SD: Check colors.]
 
-Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters.
+Before fitting the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
 This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).

### 2017-07-06 workflow/workflow.Rmd
rm c7

@@
 norm <- assays(se)$normalizedValues
-stopifnot(sum(is.infinite(norm))==0)
 norm[1:3,1:3]
@@
 our_cl <- primaryClusterNamed(ceObj)
-cl = our_cl[our_cl != "-1"]
-pal = pal[names(pal) != "-1"]
-X <- W[our_cl != "-1", ]
+cl = our_cl[!our_cl %in% c("-1", "c7")]
+pal = pal[!names(pal) %in% c("-1", "c7")]
+X <- W[!our_cl %in% c("-1", "c7"), ]
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
@@
 `getLineages` has the option for the user to provide known end cluster(s).
-Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c8` and `c6` represent terminal cell fates.
+Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c6` represent terminal cell fates.
 
@@
 t <- pseudotime(lineages)[,1]
-y <- assays(se)$normalizedValues[, our_cl != "-1"]
+y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c7")]
 gam.pval <- apply(y,1,function(z){

### 2017-07-06 workflow/workflow.Rmd
Updates to text and figure captions.

--- 2017-07-06 workflow/workflow.Rmd
@@
 Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
-Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.
+Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
 
@@
 This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017].
-The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
+The olfactory epithelium contains mature olfactory sensory neurons (mOSN) that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
 When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.
@@
 It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
-Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)).
+Then, the GBC lineage, in turn, branches off to give rise to mOSN, microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)).
 In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
@@
 
-Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
+Throughout the workflow, we use the class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
 The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
@@
 
-```{r scone, fig.cap="Plot generated by scone to filter low-quality cells."}
+```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
 # QC-metric-based sample-filtering
@@
 This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
-In general, however, we recommend care in selecting an approapriate gene filtering scheme for an actual data analysis.
-
-[FP: check. SD: I rephrased the above paragraph.]
+In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.
 
@@
 
-```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
+```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
 knitr::include_graphics("zinb_schema.pdf")
@@
 
-```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
+```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
 norm_order <- norm[, order(as.numeric(batch))]
@@
 
-```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell.
-In the left panel, cells are color-coded by batch.
-In the right panel, cells are color-coded by original published clustering."}
+```{r pcanorm, fig.cap="ZINB-WaVE: PCA of normalized expression measures, where each point represents a cell.
+Cells are color-coded by batch (left panel) and by original published clustering (right panel)."}
 pca <- prcomp(t(norm))
@@
 
-```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
+```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
 W <- colData(se)[, grepl("^W", colnames(colData(se)))]
@@
 
-```{r examineCombineMany, fig.cap="Candidate clusterings found using the function RSEC from the clusterExperiment package."}
+```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
 plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
@@
 
-```{r plotcoclust, fig.cap="Heatmap of co-clustering matrix."}
+```{r plotcoclust, fig.cap="RSEC: Heatmap of co-clustering matrix."}
 plotCoClustering(ceObj)
@@
 
-```{r barplotOurs, fig.cap="Barplot of number of cells per cluster, for our workflow's clustering."}
+```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering."}
 plotBarplot(ceObj, legend = F) 
@@
 
-The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged into single clusters.
-This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to RSEC instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.
+The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged here into single clusters.
+This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.
 
-[FP: check. SD: Rephrased a little.]
-
-```{r addPublishedClusters, fig.cap="Barplot of number of cells per cluster, for original published clustering."}
+```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for original published clustering."}
 ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
@@
 
-```{r comparePublished, fig.cap="Comparison between the clustering found using our workflow and the original published clustering."}
+```{r comparePublished, fig.cap="RSEC: Comparison between the clustering found using our workflow and the original published clustering."}
 plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
@@
 
-```{r heatmapsClusters, fig.cap="Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
+```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
 # Set colors for cell clusterings
@@
 
-```{r mdsWce, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
+```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
 palDF <- ceObj@clusterLegend[[1]]
@@
 
-[SD: Check colors.]
-
 Before fitting the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
@@
 
-```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
+```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred global lineage structure."}
 pairs(lineages, type="lineages", col = pal[cl])
@@
 
-```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
+```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
 lineages <- getCurves(lineages)
@@
 
-```{r heatmapsignificant, fig.cap="Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
+```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
 topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]

### 2017-07-06 workflow/workflow.Rmd
correspondance of clsuters

@@
 c3 | mSUS | red | original 4, 7
-c4 | mOSN | orange | original 9, 12
-c5 | Immature Neuron | purple | original 10, 14
-c6 | MV | brown | original 15
+c4 | im | orange | -2 
+c5 | mOSN | purple | original 9, 12
+c6 | Immature Neuron | brown | original 10, 14
+c7 | MV | cian | original 15
 
@@
 our_cl <- primaryClusterNamed(ceObj)
-cl = our_cl[our_cl != "-1"]
-pal = pal[names(pal) != "-1"]
-X <- W[our_cl != "-1", ]
+cl = our_cl[!our_cl %in% c("-1", "c4")]
+pal = pal[!names(pal) %in% c("-1", "c4")]
+X <- W[!our_cl %in% c("-1", "c4"), ]
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
@@
 lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
+print(lineages)
 ```
@@
 pairs(lineages, type="lineages", col = pal[cl])
+print(lineages)
 ```
@@
 lineages <- getCurves(lineages)
-
 pairs(lineages, type="curves", col = pal[cl])
+print(lineages)
 ```
@@
 lineages
+print(lineages)
 ```
@@
 lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
-                         end.clus = c("c3", "c6"))
+                         end.clus = c("c3", "c7"))
 lineagees <- getCurves(lineages)
@@
 ```{r fitgam}
+print(lineages)
 t <- pseudotime(lineages)[,1]
-y <- assays(se)$normalizedValues[, our_cl != "-1"]
+y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c4")]
 gam.pval <- apply(y,1,function(z){

### 2017-07-06 workflow/workflow.Rmd
temptatives to change colors in plotbarplot

--- 2017-07-06 workflow/workflow.Rmd
@@
 
-- make sure versions package ok and pushed to bioc (DR, EP, FP),
-- update clustering section with new clusterExperiment functions (EP),
-- address Sandrine's comments in sections slingshot and de (KS),
 - try running workflow with more genes (DR),
@@
 - update NAMESPACE and DESCRIPTION (FP),
-- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
+- install and run package making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
 - make singlecellworkflow package public (FP).]
@@
 ```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
-knitr::include_graphics("zinb_schema.pdf")
+knitr::include_graphics("zinb_schema.png")
 ```
@@
 
-```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering."}
+```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by RSEC clustering."}
 plotBarplot(ceObj, legend = F) 
@@
 
-The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged here into single clusters.
+The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters.
 This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.
 
-```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for original published clustering."}
-ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
+[FP: Elizabeth or Davide, how can we change the colors in Figure \@ref(fig:addPublishedClusters) to match the colors we use for publishedClusters (that is `col_clus` like in Figure \@ref(fig:mdsW)).
+I tried colPalette argument, but it does not work with SummarizedExperiment input.
+Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?
+
+```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by original published clustering."}
+ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
                      clusterLabel = "publishedClusters")
-plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters"), legend = F) 
-```
-
-```{r comparePublished, fig.cap="RSEC: Comparison between the clustering found using our workflow and the original published clustering."}
-plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
+plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
+            xlab = "", legend = F) 
 ```
@@
 c3 | mSUS | red | original 4, 7
-c4 | mOSN | orange | original 9, 12
-c5 | Immature Neuron | purple | original 10, 14
-c6 | MV | brown | original 15
+c4 | Contaminants | orange | original -2 
+c5 | mOSN | purple | original 9, 12
+c6 | Immature Neuron | brown | original 10, 14
+c7 | MV | cian | original 15
+
+Cells in cluster `c4` have a cluster label of `-2` in the original published clustering meaning that they were not assigned to any cluster in the original publication.
+These cells were actually identified as non-sensory contaminants as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)) and were removed from the original published clustering.
+While it is reassuring that our workflow clustered these cells separately with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes as cells in this cluster do not participate in the cell differentiation process.
+Note that our workflow's clustering did not capture all the cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we keep these cells to infer lineages as they did not seem to influence the clustering.
+
+[FP: check writing + table with correspondance of clusters]
+
+```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="Boxplots of the log counts of gene Reg3g for cells in cluster c4 (left boxplot) and cells not in cluster c4 (right boxplot)."}
+c4 <- rep("other clusters", ncol(se)) 
+c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
+boxplot(log1p(assay(se)["Reg3g", ]) ~ c4, ylab = "log counts", cex.axis = .8, cex.lab = .8)
+```
 
@@
 our_cl <- primaryClusterNamed(ceObj)
-cl = our_cl[our_cl != "-1"]
-pal = pal[names(pal) != "-1"]
-X <- W[our_cl != "-1", ]
+cl <- our_cl[!our_cl %in% c("-1", "c4")]
+pal <- pal[!names(pal) %in% c("-1", "c4")]
+X <- W[!our_cl %in% c("-1", "c4"), ]
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
@@
 lineages <- getCurves(lineages)
-
 pairs(lineages, type="curves", col = pal[cl])
@@
 `getLineages` has the option for the user to provide known end cluster(s).
-Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c8` and `c6` represent terminal cell fates.
+Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c7` represent terminal cell fates.
 
@@
 t <- pseudotime(lineages)[,1]
-y <- assays(se)$normalizedValues[, our_cl != "-1"]
+y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c4")]
 gam.pval <- apply(y,1,function(z){

### 2017-07-06 R/uploadToOverleaf.R

@@ -1,5 +1,4 @@
 library(BiocWorkflowTools)
 workflow_dir <- file.path(getwd(), 'workflow')
 uploadToOverleaf(files = workflow_dir,
-                 openInBrowser = TRUE,
-                 forceNewProject = T)
+                 openInBrowser = TRUE)

### 2017-07-06 workflow/workflow.Rmd
README

@@
 - try running workflow with more genes (DR),
-- clean up singlecellworkflow package (FP),
-- update NAMESPACE and DESCRIPTION (FP),
 - install and run package making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),

### 2017-07-06 workflow/workflow.Rmd
update todo list

@@
 
-- try running workflow with more genes (DR),
-- install and run package making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
+- try running workflow with more genes (DR if time),
+- colors and labels in plotBarplot to compare original and our clustering (EP, DR),
+- check removal cluster c4 (all)
 - make singlecellworkflow package public (FP).]
@@
 [FP: todo after converting to latex (FP): 
-
-- add space between paragraphs, 
-- check uniformity first/last names in references, 
-- add back space after word task in abstract, 
 - submit.]

### 2017-07-06 workflow/workflow.Rmd
Updates to text.

@@
   - name: Sandrine Dudoit
-    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
+    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
 
@@
 This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017].
-The olfactory epithelium contains mature olfactory sensory neurons (mOSN) that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium.
-When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.
+The olfactory epithelium contains mature olfactory sensory neurons (mOSN) that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBC), which are the actively proliferating cells in the epithelium.
+When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBC), which become activated to differentiate and reconstitute all major cell types in the epithelium.
 
@@
 
-We need the following packages.
+The following packages are needed.
 
@@
 The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data.
-The model can include a sample-level intercept, which serves as a global-scaling normalization factor.
-The user can also include both gene-level and sample-level covariates.
-The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
+The model can include a cell-level intercept, which serves as a global-scaling normalization factor.
+The user can also specify both gene-level and cell-level covariates.
+The inclusion of observed and unobserved cell-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
 A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema).
@@
 
-As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.
+As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space.
+Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`.
 
@@
 Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts.
-Here, we are applying `RSEC` instead to the low-dimensional `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object.
+Here, we are applying `RSEC` instead to the low-dimensional `W` matrix from ZINB-WaVE, for which we make a separate `SummarizedExperiment` object.
 For this reason, we choose to not use certain options in `RSEC`.
@@
 
-```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by RSEC clustering."}
+```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
 plotBarplot(ceObj, legend = F) 
@@
 I tried colPalette argument, but it does not work with SummarizedExperiment input.
-Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?
+Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?]
 
-```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by original published clustering."}
+```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
 ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
@@
 
-```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
+```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
 palDF <- ceObj@clusterLegend[[1]]
@@
 The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively.
-The first takes a low-dimensional representation of the cells and a vector of cluster labels.
+The first takes as input a low-dimensional representation of the cells and a vector of cluster labels.
 It fits an MST to the clusters and identifies lineages as paths through this tree.
@@
 
-Cells in cluster `c4` have a cluster label of `-2` in the original published clustering meaning that they were not assigned to any cluster in the original publication.
-These cells were actually identified as non-sensory contaminants as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)) and were removed from the original published clustering.
-While it is reassuring that our workflow clustered these cells separately with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes as cells in this cluster do not participate in the cell differentiation process.
-Note that our workflow's clustering did not capture all the cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we keep these cells to infer lineages as they did not seem to influence the clustering.
+Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster.
+These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering.
+While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process.
+Note that our workflow's clustering only captured XXX out of XXX cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we retain the remaining cells to infer lineages as they did not seem to influence the clustering.
 
-[FP: check writing + table with correspondance of clusters]
+[FP: check writing + table with correspondance of clusters. SD: I did minor edits to text above. Please check. Perhaps mention total number of contaminant cells and how many are in cluster `c4`.]
 
-```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="Boxplots of the log counts of gene Reg3g for cells in cluster c4 (left boxplot) and cells not in cluster c4 (right boxplot)."}
+```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g for cells in cluster c4 (left) and cells not in cluster c4 (right)."}
 c4 <- rep("other clusters", ncol(se)) 
@@
 
-```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred global lineage structure."}
+```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
 pairs(lineages, type="lineages", col = pal[cl])
@@
 
-```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
+```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
 lineages <- getCurves(lineages)
@@
 In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference.
-In addition, we are extending ZINB-WaVE to differentially expression, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
+In addition, we are extending ZINB-WaVE to identify differentially expressed genes, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
 We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.

### 2017-07-06 workflow/workflow.Rmd
Matching cluster colors to previous paper

@@
 if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current develop version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current devel version to avoid bugs")
 if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
@@
 library(knitr)
-knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
+knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)
 knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
@@
 publishedClusters <- colData(core)[, "publishedClusters"]
-col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
-col_clus <- col_clus[1:length(unique(publishedClusters))]
+col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
 names(col_clus) <- sort(unique(publishedClusters))
@@
      pch = 20, xlab = "Component 1", ylab = "Component 2")
-legend(x = "bottomright", legend = unique(names(col_clus)), cex = .5,
+legend(x = "topleft", legend = unique(names(col_clus)), cex = .5,
        fill = unique(col_clus), title = "Sample")
@@
 ```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
-plotBarplot(ceObj, legend = F) 
+plotBarplot(ceObj, legend = FALSE) 
 ```
@@
                      clusterLabel = "publishedClusters")
+
+## change default color to match with Figure 7
+clusterLegend(ceObj)$publishedClusters[, "color"] <- 
+  col_clus[clusterLegend(ceObj)$publishedClusters[, "name"]]
+
 plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
-            xlab = "", legend = F) 
+            xlab = "", legend = FALSE)
 ```
@@
 c6 | Immature Neuron | brown | original 10, 14
-c7 | MV | cian | original 15
+c7 | MV | cyan | original 15
 
@@
 ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
+
+#match to existing colors
+cols <- clusterLegend(ceObj)$combineMany[, "color"]
+names(cols) <- clusterLegend(ceObj)$combineMany[, "name"]
+clusterLegend(ce)$cluster1[, "color"] <- cols[clusterLegend(ce)$cluster1[, "name"]]
+
 plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)

### 2017-07-06 workflow/workflow.Rmd
Changes to text and figures

@@
 Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
-Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
+Throughout the workflow, we use a single `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
 
@@
 
-```{r, echo = FALSE}
+Note that in the following, we assume that the user has access to a data folder located at `../data`.
+Users with a different directory stracture may need to change the `data_dir` variable below to reproduce the workflow.
+
+```{r}
+data_dir <- "../data/"
+
 urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
@@
 
-if(!file.exists("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")) {
-  download.file(urls[1], "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
-  R.utils::gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
+if(!file.exists(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda"))) {
+  download.file(urls[1], paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
+  R.utils::gunzip(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
 }
 
-if(!file.exists("../data/oeHBCdiff_clusterLabels.txt")) {
-  download.file(urls[2], "../data/oeHBCdiff_clusterLabels.txt")
+if(!file.exists(paste0(data_dir, "oeHBCdiff_clusterLabels.txt"))) {
+  download.file(urls[2], paste0(data_dir, "oeHBCdiff_clusterLabels.txt"))
 }
@@
 ```{r}
-load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")
+load(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda"))
 
@@
 bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
-clusterLabels <- read.table("../data/oeHBCdiff_clusterLabels.txt",
+clusterLabels <- read.table(paste0(data_dir, "oeHBCdiff_clusterLabels.txt"),
                             sep = "\t", stringsAsFactors = FALSE)
@@
 publishedClusters <- colData(core)[, "publishedClusters"]
-col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
+col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
+              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
+              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
 names(col_clus) <- sort(unique(publishedClusters))
@@
 The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters.
-This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.
-
-[FP: Elizabeth or Davide, how can we change the colors in Figure \@ref(fig:addPublishedClusters) to match the colors we use for publishedClusters (that is `col_clus` like in Figure \@ref(fig:mdsW)).
-I tried colPalette argument, but it does not work with SummarizedExperiment input.
-Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?]
+This discrepancy is likely caused by the fact that we started with the top 1,000 genes, which might not be enough to discriminate between closely related clusters.
 
@@
      xlab = "Component1", ylab = "Component2")
-legend(x = "bottomright", legend = names(pal), cex = .5,
+legend(x = "topleft", legend = names(pal), cex = .5,
        fill = pal, title = "Sample")
@@
 While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process.
-Note that our workflow's clustering only captured XXX out of XXX cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we retain the remaining cells to infer lineages as they did not seem to influence the clustering.
+Note that, out of the 77 cells overexpressing `Reg3g`, 11 are captured in cluster `c4` and 21 are unclustered in our workflow's clustering (see Figure \@ref(fig:boxplotReg3g)).
+However, we retain the remaining 45 cells to infer lineages as they did not seem to influence the clustering.
 
-[FP: check writing + table with correspondance of clusters. SD: I did minor edits to text above. Please check. Perhaps mention total number of contaminant cells and how many are in cluster `c4`.]
-
-```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g for cells in cluster c4 (left) and cells not in cluster c4 (right)."}
+```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
 c4 <- rep("other clusters", ncol(se)) 
 c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
-boxplot(log1p(assay(se)["Reg3g", ]) ~ c4, ylab = "log counts", cex.axis = .8, cex.lab = .8)
+boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(ceObj), 
+        ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
 ```
@@
 pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
-pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)
+pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], 
+      show.constraints = TRUE)
 
@@
 
-Finally, a new R class called `SingleCellExperiment` should be released soon.
-This new class would essentially be a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension.
+Finally, a new S4 class called `SingleCellExperiment` is currently under development (https://github.com/drisso/SingleCellExperiment).
+This new class is essentially a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension.
 This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step.
-Once this class is implemented, we would like to incorporate it to the workflow.
+Once the implementation of this class is stable, we would like to incorporate it to the workflow.
 
@@
 
-This workflow provides a tutorial for the analysis of scRNA-seq data in R.
+This workflow provides a tutorial for the analysis of scRNA-seq data in R/Bioconductor.
 It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
@@
 The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow.
-The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available on GitHub at, respectively, https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.
+The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available at, respectively, https://bioconductor.org/packages/scone, https://bioconductor.org/packages/zinbwave, https://bioconductor.org/packages/clusterExperiment, and https://github.com/kstreet13/slingshot.
 

### 2017-07-07 workflow/workflow.Rmd
operating system comment and smaller K when more genes

@@
 ```
-
-[FP: todo before to convert to latex:
-
-- try running workflow with more genes (DR if time),
-- colors and labels in plotBarplot to compare original and our clustering (EP, DR),
-- check removal cluster c4 (all)
-- make singlecellworkflow package public (FP).]
-
-[FP: todo after converting to latex (FP): 
-- submit.]
 
@@
 
-```{r parallel, echo=FALSE}
+## Parallel computating
+
+To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions.
+Computations were performed with 2 cores on a MacBook Pro (early 2015) with a 2.7 GHz Intel Core i5 processor and 8 GB of RAM.
+The `BiocParallel` package is used to allow for parallel computing in `zinbwave` function.
+Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.
+
+```{r parallel}
 NCORES <- 2
@@
 }else{
-  print("Please change this to allow parallel computing")
+  print("Please change this to allow parallel computing on your computer.")
   register(SerialParam())
@@
 Note that in the following, we assume that the user has access to a data folder located at `../data`.
-Users with a different directory stracture may need to change the `data_dir` variable below to reproduce the workflow.
+Users with a different directory structure may need to change the `data_dir` variable below to reproduce the workflow.
 
-```{r}
+```{r datain}
 data_dir <- "../data/"
@@
 Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`.
+Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.
 

### 2017-11-29 workflow/workflow.Rmd
add SingleCellExperiment

@@
 if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.5.0-9005") stop("must have current devel version to avoid bugs")
 if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
@@
 
-se <- SummarizedExperiment(assays = list(counts = E),
+dataObj <- SummarizedExperiment(assays = list(counts = E),
                            colData = metadata)
-se
+dataObj
 ```
+
+EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'
 
@@
 data("housekeeping")
-hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]
+hk = rownames(dataObj)[toupper(rownames(dataObj)) %in% housekeeping$V1]
 
-mfilt <- metric_sample_filter(assay(se), 
-                              nreads = colData(se)$NREADS,
-                              ralign = colData(se)$RALIGN,
-                              pos_controls = rownames(se) %in% hk,
+mfilt <- metric_sample_filter(assay(dataObj), 
+                              nreads = colData(dataObj)$NREADS,
+                              ralign = colData(dataObj)$RALIGN,
+                              pos_controls = rownames(dataObj) %in% hk,
                               zcut = 3, mixture = FALSE,
@@
 mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
-se <- se[, mfilt]
-dim(se)
+dataObj <- dataObj[, mfilt]
+dim(dataObj)
 ```
 
-After sample filtering, we are left with `r ncol(se)` good quality cells.
+After sample filtering, we are left with `r ncol(dataObj)` good quality cells.
 
@@
 # Filtering to top 1,000 most variable genes
-vars <- rowVars(log1p(assay(se)))
-names(vars) <- rownames(se)
+vars <- rowVars(log1p(assay(dataObj)))
+names(vars) <- rownames(dataObj)
 vars <- sort(vars, decreasing = TRUE)
-core <- se[names(vars)[1:1000],]
+core <- dataObj[names(vars)[1:1000],]
 ```
+
+EP: If use SingleCellFilter (or davide adds to singleCellExperiment) could do this in 1 line.
+
+EP: Why do we create 'core' rather than just replace 'dataObj'
 
@@
 ```{r zinb}
-print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",  
+print(system.time(dataObj <- zinbwave(core, K = 50, X = "~ Batch",  
                                  residuals = TRUE,
@@
 ```{r norm}
-norm <- assays(se)$normalizedValues
+norm <- assays(dataObj)$normalizedValues
 norm[1:3,1:3]
@@
 ```
+
+We are going to set these normalized values to be our first assay, so that they become the default in future plots.
+
+```{r setNormAssay}
+assay(dataObj)<-assays(dataObj)$normalizedValues
+```
+
+EP: This saves a lot of trouble with the plotHeatmap, etc.
+The code needs to be fixed so *reorders* the assays, rather than what I've done, which is just replace the first.
+I don't have the object in front of me to be able to do it (it might make sense for me to add option to different assays in clusterExperiment, but I'm not going to do that right now.)
 
@@
 
+```{r addW}
+reducedDim(dataObj,"zinbwave50")<-W
+```
+
+EP: If zinbwave already has been updated, this above code is probably not needed since its aready there.
+
+EP: Note that we need to see whether clusterExperiment retains all of the different assays or not. It should in principle, but its not something I've tested.
+
 ```{r rsec_50}
-seObj <- SummarizedExperiment(t(W), colData = colData(core))
-print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
-                                betas = 0.8, dimReduce="none",
+print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
+                                betas = 0.8, dimReduce="zinbwave50",
                                 clusterFunction = "hierarchical01", minSizes=1,
@@
 ```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
-plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
+plotClusters(dataObj)
 ```
+
+EP: I've updated so that the default has >500 colors, so don't need the colPalette command, I hope.
 
@@
 ```{r plotcoclust, fig.cap="RSEC: Heatmap of co-clustering matrix."}
-plotCoClustering(ceObj)
+plotCoClustering(dataObj)
 ```
@@
 ```{r tableclust}
-table(primaryClusterNamed(ceObj))
+table(primaryClusterNamed(dataObj))
 ```
@@
 ```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
-plotBarplot(ceObj, legend = FALSE) 
+plotBarplot(dataObj, legend = FALSE) 
 ```
@@
 ```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
-ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
+dataObj <- addClusters(dataObj, colData(dataObj)$publishedClusters, 
                      clusterLabel = "publishedClusters")
@@
 ## change default color to match with Figure 7
-clusterLegend(ceObj)$publishedClusters[, "color"] <- 
-  col_clus[clusterLegend(ceObj)$publishedClusters[, "name"]]
+clusterLegend(dataObj)$publishedClusters[, "color"] <- 
+  col_clus[clusterLegend(dataObj)$publishedClusters[, "name"]]
 
-plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
+plotBarplot(dataObj, whichClusters=c("combineMany","publishedClusters"),
             xlab = "", legend = FALSE)
@@
 # Set colors for cell clusterings
-colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
-origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
-experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
-batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
+experimentColors <- bigPalette[1:nlevels(colData(dataObj)$Experiment)]
+batchColors <- bigPalette[1:nlevels(colData(dataObj)$Batch)]
 metaColors <- list("Experiment" = experimentColors,
-                   "Batch" = batchColors,
-                   "publishedClusters" = origClusterColors)
+                   "Batch" = batchColors)
 
-plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
+plotHeatmap(dataObj, 
             whichClusters = "primary", clusterFeaturesData = "all",
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
-            sampleData = c("publishedClusters", "Batch", "Experiment"),
+            sampleData = c("Batch", "Experiment"),
             clusterLegend = metaColors, annLegend = FALSE, main = "")
 ```
+
+EP: publishedClusters is saved as a cluster, so don't need to set it up. Now normalized should be default assay, so not need 'visualizeData'
 
@@
 ```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
-palDF <- ceObj@clusterLegend[[1]]
-pal <- palDF[, "color"]
-names(pal) <- palDF[, "name"]
-pal["-1"] = "transparent"
-plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
-     xlab = "Component1", ylab = "Component2")
-legend(x = "topleft", legend = names(pal), cex = .5,
-       fill = pal, title = "Sample")
+plotDimReduce(dataObj,whichCluster="primary",dimReduce="zinbwave50",pch=20,
+	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main=""
+)
 ```
@@
 ```{r tabagain}
-table(data.frame(original = publishedClusters, ours = primaryClusterNamed(ceObj)))
+table(data.frame(original = publishedClusters, ours = primaryClusterNamed(dataObj)))
 ```
@@
 ```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
-c4 <- rep("other clusters", ncol(se)) 
-c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
-boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(ceObj), 
+# c4 <- rep("other clusters", ncol(se))
+# c4[primaryClusterNamed(dataObj) == "c4"] <- "cluster c4"
+boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj), 
         ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
 ```
+
+EP: not sure why needed 'c4' object. This is a useful plot and rather messy, so I might make a plotGeneBoxplot function.
 
@@
 ```{r}
-our_cl <- primaryClusterNamed(ceObj)
-cl <- our_cl[!our_cl %in% c("-1", "c4")]
-pal <- pal[!names(pal) %in% c("-1", "c4")]
-X <- W[!our_cl %in% c("-1", "c4"), ]
+pseudoCe<-dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c4")]
+X <- reducedDim(pseudoCe,type="zinbwave50")
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
 X <- mds$points
+lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
+```
 
-lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
-```
+EP: created a pseudoCe to subset to the samples for the lineage. Simplifies code enormously and as long as we don't change it, doesn't use additional memory, I think.
 
@@
 ```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
-pairs(lineages, type="lineages", col = pal[cl])
+colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")
+pairs(lineages, type="lineages", col = colorCl)
 ```
+
+EP: convertClusterLegend does the transformation into vector of colors. However, output may be matrix, in which case need to add a `[,1]` to the above code.
 
@@
 lineages <- getCurves(lineages)
-pairs(lineages, type="curves", col = pal[cl])
+pairs(lineages, type="curves", col = colorCl)
 ```
@@
 ```{r slingshotsupervised, eval=FALSE}
-lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
+lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1",
                          end.clus = c("c3", "c7"))
 lineagees <- getCurves(lineages)
-pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
-pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], 
+pairs(lineages, type="curves", col = colorCl)
+pairs(lineages, type="lineages", col = colorCl, 
       show.constraints = TRUE)
@@
 t <- pseudotime(lineages)[,1]
-y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c4")]
+y <- assay(pseudoCe)
 gam.pval <- apply(y,1,function(z){
@@
 
+EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below.
+
 ```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
 topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
-heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
-heatclus <- cl[order(t, na.last = NA)]
-ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
+orderSamples(pseudoCe)<-order(t,na.last=TRUE)
+plotHeatmap(pseudoCe[,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
+```
 
-#match to existing colors
-cols <- clusterLegend(ceObj)$combineMany[, "color"]
-names(cols) <- clusterLegend(ceObj)$combineMany[, "name"]
-clusterLegend(ce)$cluster1[, "color"] <- cols[clusterLegend(ce)$cluster1[, "name"]]
-
-plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
-```
+The above code should take care of the ordering and NA values.
+Note that subseting a ClusterExperiment object is suppose to also subset the orderSamples value correctly.
+Let me know if that is not the case.
 

### 2017-11-30 workflow/workflow.Rmd
add comment
further SingleCellExperiment updates

@@
 if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.5.0-9005") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.5.0-9006") stop("must have current devel version to avoid bugs")
 if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
@@
 Note that in order to successfully run the workflow, we need the devel versions
-of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.3.2)`.
+of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.6)`.
 We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
@@
 In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension.
-Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`).
+Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`reduceMethod`) and the construction of the hiearchy between the clusters (`dendroReduce`).
 We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.
@@
 
-EP: Note that we need to see whether clusterExperiment retains all of the different assays or not. It should in principle, but its not something I've tested.
+EP: Note that we need to see whether clusterExperiment retains all of the different assays or not.
+It should in principle, but its not something I've tested.
+Also, I purposefully didn't give the parameter `nReducedDims` because I think it should automatically use the max saved in object if not designated.
+If not, let me know.
+
+EP: Currently isCount=FALSE.
+Does that make sense for the normalized values, or should we set it to TRUE (it shouldn't affect the clustering, since the clustering is only on the reducedDim slot now, but could affect downstream plots, etc.).
+
+EP: We could add back in the mergeClusters step now that we don't have the W in the data slot.
+I haven't done that yet because we want to just recreate the previous, but it would be reasonable now.
+It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously.
+But it would be nice to show that you can do that for this setting.
+Once this is running I can come back to it.
 
@@
 print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
-                                betas = 0.8, dimReduce="zinbwave50",
+                                betas = 0.8, reduceMethod="zinbwave50",
                                 clusterFunction = "hierarchical01", minSizes=1,
                                 ncores = NCORES, isCount=FALSE,
-                                dendroReduce="none",dendroNDims=NA,
+                                dendroReduce="zinbwave50",
                                 subsampleArgs = list(resamp.num=100,
@@
 ```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
-plotDimReduce(dataObj,whichCluster="primary",dimReduce="zinbwave50",pch=20,
+plotDimReduce(dataObj,whichCluster="primary",reduceMethod="zinbwave50",pch=20,
 	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main=""

### 2017-12-01 workflow/workflow.Rmd
update with new version and plotFeatureBoxplot

@@
 if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.5.0-9006") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.5.0-9007") stop("must have current devel version to avoid bugs")
 if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
@@
 # c4[primaryClusterNamed(dataObj) == "c4"] <- "cluster c4"
-boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj), 
-        ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
+# boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj),
+#         ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
+plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g")
 ```
 
-EP: not sure why needed 'c4' object. This is a useful plot and rather messy, so I might make a plotGeneBoxplot function.
+EP: not sure why needed 'c4' object. This is new function. Tell me if you run into problems.
 

### 2017-12-04 workflow/workflow.Rmd
update Rcode and EP comments

@@
 
+EP: I have only updated the code, not the text. The text needs adaptation, especially future directions and references to `SummarizedExperiment` which should be `SingleCellExperiment`.
+
 # Analysis of olfactory stem cell differentiation using scRNA-seq data
@@
 
+EP: Should update these.
+
 ## Parallel computating
@@
 ```
+
+EP: Note that clusterExperiment uses `parallel` package, so might should update this.
 
@@
 We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).
+
+EP: There's code in SingleCellExperiment for spike-ins. We should use it, though I'm not familiar with it.
 
@@
 
+EP: Should probably address these normalized values, especially to more strongly make clear that shouldn't use these for clustering, just for visualization.
+The way its stated it sounds like dimensionality reduction is an "also", when in fact it is the main point and the normalized values are the afterthought.
+Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.
+
 ### Dimensionality reduction
@@
 It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously.
+But then again, I don't know the workflow for the DE part, so maybe we want to leave it out.
 But it would be nice to show that you can do that for this setting.
-Once this is running I can come back to it.
+Once this workflow is running I can come back to it.
 
@@
 ```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
-# Set colors for cell clusterings
+# Set colors for additional sample data
 experimentColors <- bigPalette[1:nlevels(colData(dataObj)$Experiment)]
@@
 #         ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
-plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g")
+plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
 ```
@@
 t <- pseudotime(lineages)[,1]
-y <- assay(pseudoCe)
+y <- transformData(pseudoCe)
 gam.pval <- apply(y,1,function(z){
@@
 EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below.
+I've used the `transformData` command, which will not do anything if isCount=FALSE above, but will convert to right if we change isCount=TRUE.
+It's better to use that command than `assay` for that reason.
+Might should add comment in code or text explaining this.
 
@@
 orderSamples(pseudoCe)<-order(t,na.last=TRUE)
-plotHeatmap(pseudoCe[,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
+plotHeatmap(pseudoCe[topgenes,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
 ```
+
+EP: note the above with do heirarchcial clustering of the top genes. If want them to stay in order given, should give `clusterFeaturesData=topgenes` (I think that is the argument)
 

### 2017-12-04 workflow/workflow.Rmd
comments

@@
 
-EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'
+EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'.  
+FP: good idea.
 
@@
 vars <- sort(vars, decreasing = TRUE)
-core <- dataObj[names(vars)[1:1000],]
+dataObj <- dataObj[names(vars)[1:1000],]
 ```
@@
 
-EP: Why do we create 'core' rather than just replace 'dataObj'
+EP: Why do we create 'core' rather than just replace 'dataObj'. I agree, I think that it is unnecessary. I changed 'core' to dataObj.
 
@@
 ```{r}
-core
+dataObj
 ```
@@
 ```{r batch}
-batch <- colData(core)$Batch
+batch <- colData(dataObj)$Batch
 col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
@@
 ```{r original}
-publishedClusters <- colData(core)[, "publishedClusters"]
+publishedClusters <- colData(dataObj)[, "publishedClusters"]
 col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
@@
 ```{r zinb}
-print(system.time(dataObj <- zinbwave(core, K = 50, X = "~ Batch",  
-                                 residuals = TRUE,
-                                 normalizedValues = TRUE)))
+print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch", residuals = TRUE, normalizedValues = TRUE)))
 ```
@@
 Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.
+FP: Agree, let's keep it in mind for once the workflow runs smoothly.
 
@@
 
+FP: W can now be accessed using reducedDim function. Two functions called reducedDim are loaded in the workflow: one from slingshot and the other one from singleCellExperiment.
+
 ```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
-W <- colData(se)[, grepl("^W", colnames(colData(se)))]
-W <- as.matrix(W)
+W <- SingleCellExperiment::reducedDim(dataObj)
 d <- dist(W)
@@
      pch = 20, xlab = "Component 1", ylab = "Component 2")
-legend(x = "topleft", legend = unique(names(col_clus)), cex = .5,
-       fill = unique(col_clus), title = "Sample")
+legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, fill = unique(col_clus), title = "Sample")
 ```
@@
 EP: If zinbwave already has been updated, this above code is probably not needed since its aready there.  
+FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.
 
@@
 Does that make sense for the normalized values, or should we set it to TRUE (it shouldn't affect the clustering, since the clustering is only on the reducedDim slot now, but could affect downstream plots, etc.).
+FP: yes, I think isCount=FALSE makes sense for the normalized values.
 
@@
 Once this workflow is running I can come back to it.
+FP: I am not sure it makes sense to perform the DE on the normalized values.
+Intuitively, I would say no, but not sure.
+We can still show that it is a possibility and write in the text that in this case, it might not make sense.
+We are working on a DE method for scRNA-seq where we use observational weights computed from the posterior probabilities of the ZINB-WaVE model in bulk RNA tools like edgeR and DESeq2.
+I think, eventually, it would be nice to be able to use that.
+Maybe, not for the next next version of the workflow.
 

### 2017-12-05 workflow/workflow.Rmd
comments

@@
 ```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
-if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
+if(packageVersion("zinbwave")<"1.1.0") stop("must have current master version to avoid bugs")
 if(packageVersion("clusterExperiment")<"1.5.0-9007") stop("must have current devel version to avoid bugs")
-if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
+if(packageVersion("scone")<"1.2.0") stop("must have current master version to avoid bugs")
 
@@
 
-EP: Why do we create 'core' rather than just replace 'dataObj'. I agree, I think that it is unnecessary. I changed 'core' to dataObj.
+EP: Why do we create 'core' rather than just replace 'dataObj'.   
+FP: I agree, I think that it is unnecessary. I changed 'core' to dataObj.
 
@@
 
-```{r zinb}
+```{r zinb,eval=FALSE}
 print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch", residuals = TRUE, normalizedValues = TRUE)))
+```
+
+```{r}
+#save(dataObj, file= 'dataObj.rda')
+load('dataObj.rda')
 ```
@@
 ```{r addW}
-reducedDim(dataObj,"zinbwave50")<-W
+SingleCellExperiment::reducedDim(dataObj,"zinbwave50")<-W
 ```
@@
 
-```{r rsec_50}
+```{r rsec_50,eval=FALSE}
 print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
@@
                                 combineMinSize = 10)))
+```
+
+```{r}
+#save(dataObj, file = 'dataObj_afterRSEC.rda')
+load('dataObj_afterRSEC.rda')
 ```
@@
 
+FP: Ordering does not seem right.
+
 The distribution of cells across the consensus clusters can be visualized in Figure \@ref(fig:barplotOurs) and is as follows:
@@
 ```
+
+FP: we do not get exactly the same number of cells in each cluster. It looks similar though.
 
@@
 
+FP: compared to the previous version, it seems that there is a lot of black. Would the black coming from the contour of the rectangles in the bars?
+
 Figure \@ref(fig:heatmapsClusters) displays a heatmap of the normalized expression measures for the 1,000 most variable genes, where cells are clustered according to the RSEC consensus.
@@
 
-EP: publishedClusters is saved as a cluster, so don't need to set it up. Now normalized should be default assay, so not need 'visualizeData'
+EP: publishedClusters is saved as a cluster, so don't need to set it up. Now normalized should be default assay, so not need 'visualizeData'.
+
+FP: publishedClusters does not appear in the bars.
 
@@
 ```
+FP: I get `"reduceMethod" is not a graphical parameter`. I think this plot is different from the previous version.
 
@@
 
+FP: this plot is fairly different from the previous version. We should see for one cluster that the gene expression for Reg3g is much high that for the other clusters.
+
 EP: not sure why needed 'c4' object. This is new function. Tell me if you run into problems.
@@
 ```{r}
-pseudoCe<-dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c4")]
+pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c4")]
 X <- reducedDim(pseudoCe,type="zinbwave50")
@@
 EP: created a pseudoCe to subset to the samples for the lineage. Simplifies code enormously and as long as we don't change it, doesn't use additional memory, I think.  
+FP: I agree. I get an error though when I run line 501:
+`Error in .BasicFunsList[[f]] : no such index at level 1`.
+
+FP: I stooped here.
 

### 2017-12-06 workflow/workflow.Rmd
change file name of dataObj after zinbwave
respond to Fanny comments

@@
 ```{r}
-# Filtering to top 1,000 most variable genes
-vars <- rowVars(log1p(assay(dataObj)))
-names(vars) <- rownames(dataObj)
-vars <- sort(vars, decreasing = TRUE)
-dataObj <- dataObj[names(vars)[1:1000],]
+# # Filtering to top 1,000 most variable genes
+# vars <- rowVars(log1p(assay(dataObj)))
+# names(vars) <- rownames(dataObj)
+# vars <- sort(vars, decreasing = TRUE)
+# dataObj <- dataObj[names(vars)[1:1000],]
+
+#alternative script:
+dataObj<-makeFilterStats(dataObj,filterStats="var")
+dataObj<-filterData(dataObj,percentile=1000)
 ```
 
-EP: If use SingleCellFilter (or davide adds to singleCellExperiment) could do this in 1 line.
-
-EP: Why do we create 'core' rather than just replace 'dataObj'.   
-FP: I agree, I think that it is unnecessary. I changed 'core' to dataObj.
+EP: Added existing code in the clusterExperiment package that does this. Also will mean that any future steps in package that need variance for the dataObj will have variance per gene stored.
 
@@
 ```{r}
-#save(dataObj, file= 'dataObj.rda')
-load('dataObj.rda')
+#save(dataObj, file= 'dataObj_afterZinbwave.rda')
+load('dataObj_afterZinbwave.rda')
 ```
@@
 
+EP: I would get rid of the `SingleCellExperiment::` code and just add a `library(SingleCellExperiment)`. Much more comfortable code for most users.
+
 ```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
@@
 FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.
+EP: I just called it "zinbwave50". We can replace that with whatever its name is in the results from zinbwave. It needs a name, but it can be anything.
 
@@
 plotHeatmap(dataObj, 
-            whichClusters = "primary", clusterFeaturesData = "all",
+            whichClusters = c("primary","PublishedClusters"), clusterFeaturesData = "all",
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
@@
 FP: publishedClusters does not appear in the bars.
+EP: I forgot to add it to the `whichClusters` argument.
 
@@
 ```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
-plotDimReduce(dataObj,whichCluster="primary",reduceMethod="zinbwave50",pch=20,
+plotDimReduce(dataObj,whichCluster="primary",reducedDim="zinbwave50",pch=20,
 	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main=""
@@
 FP: I get `"reduceMethod" is not a graphical parameter`. I think this plot is different from the previous version.
+EP: I had wrong name for argument (should have been `reducedDim`), so it went with default which is PCA on the assay(dataObj) slot -- i.e. not using zinbwave. Hopefully better now.
 
@@
 FP: this plot is fairly different from the previous version. We should see for one cluster that the gene expression for Reg3g is much high that for the other clusters.  
-
-EP: not sure why needed 'c4' object. This is new function. Tell me if you run into problems.
+EP: I think this plot is of the normalized values (saved in assay(dataObj)). But I see that the previous plot was on the counts. Is that better than normalized value?
 

### 2017-12-06 workflow/workflow.Rmd
update workflow code

@@
 if(packageVersion("zinbwave")<"1.1.0") stop("must have current master version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.5.0-9007") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.5.0-9009") stop("must have current devel version to avoid bugs")
 if(packageVersion("scone")<"1.2.0") stop("must have current master version to avoid bugs")
@@
 ```{r setNormAssay}
-assay(dataObj)<-assays(dataObj)$normalizedValues
+assays(dataObj)<-assays(dataObj)[c("normalizedValues","residuals","counts")]
 ```
@@
 EP: This saves a lot of trouble with the plotHeatmap, etc.
-The code needs to be fixed so *reorders* the assays, rather than what I've done, which is just replace the first.
-I don't have the object in front of me to be able to do it (it might make sense for me to add option to different assays in clusterExperiment, but I'm not going to do that right now.)
+
+EP: I've changed this code from my previous ad hoc to not delete the previous, but just reorder them.
+Again, (it might make sense for me to add option to different assays in clusterExperiment, but I'm not going to do that right now.)
 
@@
 ```{r addW}
-SingleCellExperiment::reducedDim(dataObj,"zinbwave50")<-W
+#SingleCellExperiment::reducedDim(dataObj,"zinbwave50")<-W
 ```
 
-EP: If zinbwave already has been updated, this above code is probably not needed since its aready there.  
 FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.
-EP: I just called it "zinbwave50". We can replace that with whatever its name is in the results from zinbwave. It needs a name, but it can be anything.
+EP: I just called it "zinbwave50" for convenience.
+I've changed future code to just use "zinbwave".
+Though as a suggestion, you might consider for the zinbwave function to add the value of K to the name automatically, since the dimensions are not nested within each other, like pca, so you might run multiple versions of zinbwave and want to compare different K.
 
@@
 It should in principle, but its not something I've tested.
-Also, I purposefully didn't give the parameter `nReducedDims` because I think it should automatically use the max saved in object if not designated.
-If not, let me know.
+Also, I purposefully didn't give the parameter `nReducedDims` in RSEC because I think it should automatically use the max saved in object if not designated.
+If we're not getting the same results, we can add in `nReducedDims=50` and see; if that changes anything, let me know.
+However, I would focus on the question of different W values.
 
-EP: Currently isCount=FALSE.
-Does that make sense for the normalized values, or should we set it to TRUE (it shouldn't affect the clustering, since the clustering is only on the reducedDim slot now, but could affect downstream plots, etc.).
-FP: yes, I think isCount=FALSE makes sense for the normalized values.
-
-EP: We could add back in the mergeClusters step now that we don't have the W in the data slot.
-I haven't done that yet because we want to just recreate the previous, but it would be reasonable now.
-It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously.
-But then again, I don't know the workflow for the DE part, so maybe we want to leave it out.
-But it would be nice to show that you can do that for this setting.
-Once this workflow is running I can come back to it.
 FP: I am not sure it makes sense to perform the DE on the normalized values.
@@
 Maybe, not for the next next version of the workflow.
+EP: I'd note that we use the normalized values for finding those DE with respsect to lineage.
+But I don't think its necessary to put it in here, but maybe in future directions, where you discuss that functionality, we could mention that it would allow the usage of mergeClusters.
+Also would be a good time to point out that DE on normalized values probably not a great idea either, if you don't think so.
+I also think the reason for skipping mergeClusters should be stated in the above text, rather than just saying we are skipping it.
 
@@
 print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
-                                betas = 0.8, reduceMethod="zinbwave50",
+                                betas = 0.8, reduceMethod="zinbwave",
                                 clusterFunction = "hierarchical01", minSizes=1,
                                 ncores = NCORES, isCount=FALSE,
-                                dendroReduce="zinbwave50",
+                                dendroReduce="zinbwave",
                                 subsampleArgs = list(resamp.num=100,
@@
 
-EP: I've updated so that the default has >500 colors, so don't need the colPalette command, I hope.
-
 The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).
@@
 FP: Ordering does not seem right.  
+EP: I get a very different plotCoClustering plot than what you sent me. See email.
 
@@
 ```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
-dataObj <- addClusters(dataObj, colData(dataObj)$publishedClusters, 
+dataObj <- addClusterings(dataObj, colData(dataObj)$publishedClusters, 
                      clusterLabel = "publishedClusters")
@@
 plotBarplot(dataObj, whichClusters=c("combineMany","publishedClusters"),
-            xlab = "", legend = FALSE)
+            xlab = "", legend = FALSE,missingColor="white")
 ```
@@
 FP: compared to the previous version, it seems that there is a lot of black. Would the black coming from the contour of the rectangles in the bars?
+EP: I'm not sure what you mean about a lot of black on the barplot. Do you mean the unassigned samples? The colors have changed because the default set of colors was updated.
 
@@
 plotHeatmap(dataObj, 
-            whichClusters = c("primary","PublishedClusters"), clusterFeaturesData = "all",
+            whichClusters = c("combineMany","publishedClusters"), clusterFeaturesData = "all",
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
@@
 ```
-
-EP: publishedClusters is saved as a cluster, so don't need to set it up. Now normalized should be default assay, so not need 'visualizeData'.
 
@@
 ```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
-plotDimReduce(dataObj,whichCluster="primary",reducedDim="zinbwave50",pch=20,
-	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main=""
+plotReducedDims(dataObj,whichCluster="primary",reducedDim="zinbwave",pch=20,
+	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main="",
+	plotUnassigned=FALSE
 )
+
+# ###test
+# col<-convertClusterLegend(dataObj,whichCluster="primary",output="matrixColors")[,1]
+# col[col=="white"]<-"grey"
+# W<-reducedDim(dataObj,"zinbwave")
+# plot(-W[,1:2],col=col,pch=20)
 ```
+
 FP: I get `"reduceMethod" is not a graphical parameter`. I think this plot is different from the previous version.
-EP: I had wrong name for argument (should have been `reducedDim`), so it went with default which is PCA on the assay(dataObj) slot -- i.e. not using zinbwave. Hopefully better now.
+
+EP: I had wrong name for argument which has changed (should have been `reducedDim`), so it went with default which is PCA on the assay(dataObj) slot -- i.e.
+not using zinbwave.
+Now it gives the same plot as my test, where I directly plotted the results of `reducedDim(dataObj,"zinbwave")`.
+However, its still not exactly the same as previous rendition online.
+First it looks like the dimensions got flipped, i.e.
+multiplied by -1.
+Furthermore, it doesn't look exactly the same (e.g.
+outlying points not the same), even if I color in the unassigned so we can see them.
+Is zinbwave giving back the exact same W as before?
+I've checked that the W in the `dataObj_afterZinbwave` and W I plotted above from `dataObj_afterRSEC` are the same so its not clusterExperiment messing something up.
+If so, this could be why we are getting slightly different clustering results.
 
@@
 ```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
-# c4 <- rep("other clusters", ncol(se))
-# c4[primaryClusterNamed(dataObj) == "c4"] <- "cluster c4"
-# boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj),
-#         ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
-plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
+plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g count values", cex.axis = .8, cex.lab = .8,assay=log1p(assays(dataObj)[["counts"]]))
+# plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g normalized Expression values", cex.axis = .8, cex.lab = .8)
+
 ```
@@
 FP: this plot is fairly different from the previous version. We should see for one cluster that the gene expression for Reg3g is much high that for the other clusters.
-EP: I think this plot is of the normalized values (saved in assay(dataObj)). But I see that the previous plot was on the counts. Is that better than normalized value?
+
+EP: I think this plot showed the normalized values (saved in assay(dataObj)).
+But I see that the previous plot was on the counts.
+Is that better than normalized value?
+I have added (just to this plot for now), an option to plot a specific given matrix (or a character value defining an assay, but since we want log-transform of the counts, which is not saved in the main assay, putting in "counts" doesn't give plot we want).
+But it's now cluster c1, not c4 that has this property.
+It might be useful to take the clustering from the old RSEC data object and directly compare the clusterings.
+Regardless, none of the previous commands are going to get the right results going forward since the clusters don't have the same name.
+See note on plotReducedDims command where I think the input `W` has changed.
 
@@
 ```{r}
-pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c4")]
-X <- reducedDim(pseudoCe,type="zinbwave50")
+pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c1")]
+X <- reducedDim(pseudoCe,type="zinbwave")
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
@@
 ```
+EP: See my email.
+I don't get this subsetting error on my machine or on TravisCI.
+I do get on SCF Servers which are running R3.4.1 and the older version of bioconductor.
+Can you check this on your machine?
 
-EP: created a pseudoCe to subset to the samples for the lineage. Simplifies code enormously and as long as we don't change it, doesn't use additional memory, I think.  
 FP: I agree. I get an error though when I run line 501:
@@
 ```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
-colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")
+colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
 pairs(lineages, type="lineages", col = colorCl)
 ```
-
-EP: convertClusterLegend does the transformation into vector of colors. However, output may be matrix, in which case need to add a `[,1]` to the above code.
 
@@
 
-EP: note the above with do heirarchcial clustering of the top genes. If want them to stay in order given, should give `clusterFeaturesData=topgenes` (I think that is the argument)
+EP: note the above does heirarchcial clustering of the top genes. If want them to stay in order given, should give `clusterFeaturesData=topgenes` (I think that is the argument)
 
-The above code should take care of the ordering and NA values.
+EP: The above code should take care of the ordering and NA values.
 Note that subseting a ClusterExperiment object is suppose to also subset the orderSamples value correctly.
@@
 Once the implementation of this class is stable, we would like to incorporate it to the workflow.
+
+EP:  Obviously should replace this text.
+Might want to mention ability of class to store data matrices not in memory, though not currently implemented, which is the next major change we need to do here.
 

### 2017-12-14 workflow/workflow.Rmd
different variable genes

@@
 EP: I have only updated the code, not the text. The text needs adaptation, especially future directions and references to `SummarizedExperiment` which should be `SingleCellExperiment`.
+FP: TODO, fix ref.
 
@@
 ```{r packages}
+# GitHub
+library(slingshot)
+
 # Bioconductor
@@
 library(zinbwave)
-
-# GitHub
-library(slingshot)
+library(SingleCellExperiment)
 
@@
 EP: There's code in SingleCellExperiment for spike-ins. We should use it, though I'm not familiar with it. 
+FP: Davide, what is the standard way to use SingleCellExperiment for spike-ins?
 
@@
 
-EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'.  
-FP: good idea.
-
 Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
@@
 # # Filtering to top 1,000 most variable genes
-# vars <- rowVars(log1p(assay(dataObj)))
-# names(vars) <- rownames(dataObj)
-# vars <- sort(vars, decreasing = TRUE)
+vars <- rowVars(log1p(assay(dataObj)))
+names(vars) <- rownames(dataObj)
+vars <- sort(vars, decreasing = TRUE)
 # dataObj <- dataObj[names(vars)[1:1000],]
@@
 dataObj<-makeFilterStats(dataObj,filterStats="var")
-dataObj<-filterData(dataObj,percentile=1000)
+dataObj<-filterData(dataObj,percentile=1000,filterStats="var")
+
+length(intersect(rownames(dataObj), names(vars)[1:1000]))
+setdiff(rownames(dataObj), names(vars)[1:1000])
 ```
@@
 EP: Added existing code in the clusterExperiment package that does this. Also will mean that any future steps in package that need variance for the dataObj will have variance per gene stored.  
+FP: great. I added the filterStats argument to filterData function. I get an error otherwise.
+
+FP: We get a different list of most variable genes here.
 
@@
 
+FP: These values are slightly different from what
+
 As expected, the normalized values no longer exhibit batch effects (Figure \@ref(fig:boxplotNorm)).
@@
 EP: This saves a lot of trouble with the plotHeatmap, etc.
+FP: Ok.
 
@@
 
-FP: W can now be accessed using reducedDim function. Two functions called reducedDim are loaded in the workflow: one from slingshot and the other one from singleCellExperiment.
-
-EP: I would get rid of the `SingleCellExperiment::` code and just add a `library(SingleCellExperiment)`. Much more comfortable code for most users.
-
 ```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
-W <- SingleCellExperiment::reducedDim(dataObj)
+W <- reducedDim(dataObj)
 d <- dist(W)
@@
 ```{r addW}
-#SingleCellExperiment::reducedDim(dataObj,"zinbwave50")<-W
+#reducedDim(dataObj,"zinbwave50")<-W
 ```
@@
 ```
+
+FP: We get more clusters for the combineMany method than before. Maybe because the most variable genes are different?
 
@@
 plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g count values", cex.axis = .8, cex.lab = .8,assay=log1p(assays(dataObj)[["counts"]]))
-# plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g normalized Expression values", cex.axis = .8, cex.lab = .8)
-
 ```
@@
                          end.clus = c("c3", "c7"))
-lineagees <- getCurves(lineages)
+lineages <- getCurves(lineages)
 pairs(lineages, type="curves", col = colorCl)

### 2018-06-27 workflow/workflow.Rmd
Change code to work with latest package versions

@@
 author:
+
   - name: Fanny Perraudeau
@@
     affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
-
 abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types.
@@
 Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.
-
 keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
-
 bibliography: ref.bib
-
 vignette: >
@@
 output: BiocWorkflowTools::f1000_article
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(knitr)
-knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(cache=FALSE, error=FALSE, message=FALSE, warning=FALSE)
 knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
@@
 ```{r packages}
-# GitHub
-library(slingshot)
-
 # Bioconductor
@@
 library(SingleCellExperiment)
+library(slingshot)
 
@@
 Note that in order to successfully run the workflow, we need the devel versions
-of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.6)`.
-We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
+of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
+We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
 
 EP: Should update these.
+
+DR: Done, but let's double-check again before the final submission.
 
@@
 
-Throughout the workflow, we use the class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
+Throughout the workflow, we use the class `SingleCellExperiment` to keep track of the counts and their associated metadata within a single object.
 The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
@@
 
-dataObj <- SummarizedExperiment(assays = list(counts = E),
+dataObj <- SingleCellExperiment(assays = list(counts = E),
                            colData = metadata)
@@
 
-mfilt <- metric_sample_filter(assay(dataObj), 
+mfilt <- metric_sample_filter(counts(dataObj), 
                               nreads = colData(dataObj)$NREADS,
@@
 # # Filtering to top 1,000 most variable genes
-vars <- rowVars(log1p(assay(dataObj)))
+vars <- rowVars(log1p(counts(dataObj)))
 names(vars) <- rownames(dataObj)
@@
 #alternative script:
-dataObj<-makeFilterStats(dataObj,filterStats="var")
+dataObj<-makeFilterStats(dataObj,filterStats="var", transFun = log1p)
 dataObj<-filterData(dataObj,percentile=1000,filterStats="var")
@@
 FP: We get a different list of most variable genes here.
+
+DR: Fixed, the new code was computing the variance in linear space instead of using the log.
+
+DR: Remember to remove the old code and the length() and setdiff() from above.
 
@@
 
+DR: Agree, I'm not even sure normalized values are that useful for visualization... a simple scaling by total counts would probably achieve the same... but since we have them...
+
 ### Dimensionality reduction
@@
 
-```{r addW}
-#reducedDim(dataObj,"zinbwave50")<-W
-```
-
 FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.
@@
 I also think the reason for skipping mergeClusters should be stated in the above text, rather than just saying we are skipping it.
+
+DR: A lot of these comments are now outdated, as the new version of clusterExperiment fully support multiple assays.
+Similarly, the weights are now a reality and also supported by clusterExperiment.
+We should probably add a section about it?
 
@@
 
+DR: I fixed the most variable genes. Is there a way to check the older results? Where do I find the saved objects?
+
 The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).
@@
 FP: we do not get exactly the same number of cells in each cluster. It looks similar though.
+
+DR: it's pretty similar, but we still get 7 clusters that look more or less the same as before.
 
@@
 EP: I'm not sure what you mean about a lot of black on the barplot. Do you mean the unassigned samples? The colors have changed because the default set of colors was updated.
+
+DR: Now it looks pretty much the same.
 
@@
 
+DR: It is possible that zinbwave does not return the same exact W. There is no guarantee of a global maximum, but only local. The plot looks more or less the same, but just flipped.
+
 ## Cell lineage and pseudotime inference: Slingshot
@@
 c6 | Immature Neuron | brown | original 10, 14
-c7 | MV | cyan | original 15
+c7 | MV | light blue | original 15
 
@@
 However, we retain the remaining 45 cells to infer lineages as they did not seem to influence the clustering.
+
+DR: This has changed (slightly). This is the table that I get currently.
+
+Cluster name | Description | Color | Correspondence
+-------------|-------------|-------| ----------
+c1 | HBC | red | original 1, 5
+c2 | GBC | blue | original 2, 3, 11
+c3 | mSUS | green | original 4, 7
+c4 | mOSN | orange | original 9, 12 
+c5 | Immature Neuron | purple | original 10, 14
+c6 | mOSN | brown | original 9
+c7 | MV | light blue | original 15
 
@@
 
+DR: I think we should re-write this section.
+With the current version of the package and parameters, we don't get the contaminant cluster anymore, but most of the contaminant cells are now in c1.
+However c1 also includes HBC cells, so we should not remove it from the slingshot analysis.
+
 To infer lineages and pseudotimes, we apply Slingshot to the 4-dimensional MDS of the low-dimensional matrix `W`.
@@
 ```{r}
-pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c1")]
+pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
 X <- reducedDim(pseudoCe,type="zinbwave")
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
-X <- mds$points
-lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
+reducedDim(pseudoCe, "MDS") <- mds$points
+pseudoCe <- slingshot(pseudoCe, clusterLabels = "combineMany", reducedDim = "MDS", start.clus = "c1")
 ```
@@
 
+DR: Above, I use the syntax to work directly on the ClusterExperiment object, but if we want to be able to plot the resulting lineages, we need the matrix method.
+
 Before fitting the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
@@
 colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
-pairs(lineages, type="lineages", col = colorCl)
+lineages <- getLineages(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
+## pairs(lineages, type="lineages", col = colorCl)
 ```
+
+DR: For some reason, this plot does not work anymore, but the lineages that I get are correct!
 
@@
 ```{r fitgam}
-t <- pseudotime(lineages)[,1]
+t <- colData(pseudoCe)$slingPseudotime_1
+t <- slingPseudotime(lineages)[,1]
 y <- transformData(pseudoCe)
@@
 
+DR: Kelly, can you please check why I get different lineages and pseudotimes for the slingshot method and for the getLineages/getCurves method? The latter seems the correct one.
+
 ```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
 topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
-orderSamples(pseudoCe)<-order(t,na.last=TRUE)
-plotHeatmap(pseudoCe[topgenes,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
+
+pseudoCe1 <- pseudoCe[,!is.na(t)]
+orderSamples(pseudoCe1)<-order(t[!is.na(t)])
+
+plotHeatmap(pseudoCe1[topgenes,], clusterSamplesData = "orderSamplesValue", breaks = .99)
 ```

### 2018-06-29 workflow/workflow.Rmd
Use makeConsensus

@@
 Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
-Throughout the workflow, we use a single `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
+Throughout the workflow, we use a single `SingleCellExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
 
@@
 
-## Parallel computating
+## Parallel computing
 
@@
                                 verbose=TRUE,
-                                combineProportion = 0.7,
+                                consensusProportion = 0.7,
                                 mergeMethod = "none", random.seed=424242,
-                                combineMinSize = 10)))
+                                consensusMinSize = 10)))
 ```
@@
 
-The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings.
+The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examinemakeConsensus)), where columns correspond to cells and rows to different clusterings.
 Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap accross rows.
@@
 
-```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
+```{r examinemakeConsensus, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
 plotClusters(dataObj)
@@
 
-FP: We get more clusters for the combineMany method than before. Maybe because the most variable genes are different?
+FP: We get more clusters for the makeConsensus method than before. Maybe because the most variable genes are different?
 
@@
 
-plotBarplot(dataObj, whichClusters=c("combineMany","publishedClusters"),
+plotBarplot(dataObj, whichClusters=c("makeConsensus","publishedClusters"),
             xlab = "", legend = FALSE,missingColor="white")
@@
 DR: Now it looks pretty much the same.
+
+DR: Better to use `plotClustersTable` here? See below.
+
+```{r addPublishedClusters2, fig.cap="RSEC: Confusion matrix of number of cells per cluster, for our workflow's RSEC clustering and the original published clustering."}
+plotClustersTable(dataObj, whichClusters=c("makeConsensus","publishedClusters"))
+```
 
@@
 plotHeatmap(dataObj, 
-            whichClusters = c("combineMany","publishedClusters"), clusterFeaturesData = "all",
+            whichClusters = c("makeConsensus","publishedClusters"), clusterFeaturesData = "all",
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
@@
 c1 | HBC | red | original 1, 5
-c2 | GBC | blue | original 2, 3, 11
-c3 | mSUS | green | original 4, 7
-c4 | mOSN | orange | original 9, 12 
+c2 | mSUS | blue | original 4, 7
+c3 | mOSN  | green | original 9, 12 
+c4 | GBC | orange | original 2, 3, 11
 c5 | Immature Neuron | purple | original 10, 14
-c6 | mOSN | brown | original 9
-c7 | MV | light blue | original 15
+c6 | MV | brown | original 15
+c7 | mOSN | light blue | original 9
 
@@
 reducedDim(pseudoCe, "MDS") <- mds$points
-pseudoCe <- slingshot(pseudoCe, clusterLabels = "combineMany", reducedDim = "MDS", start.clus = "c1")
+pseudoCe <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1")
 ```
@@
 lineages <- getLineages(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
-## pairs(lineages, type="lineages", col = colorCl)
+pairs(lineages, type="lineages", col = colorCl)
 ```
@@
 t <- colData(pseudoCe)$slingPseudotime_1
-t <- slingPseudotime(lineages)[,1]
 y <- transformData(pseudoCe)
@@
 Might should add comment in code or text explaining this.
-
-DR: Kelly, can you please check why I get different lineages and pseudotimes for the slingshot method and for the getLineages/getCurves method? The latter seems the correct one.
 
@@
 The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow.
-The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available at, respectively, https://bioconductor.org/packages/scone, https://bioconductor.org/packages/zinbwave, https://bioconductor.org/packages/clusterExperiment, and https://github.com/kstreet13/slingshot.
+The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available at, respectively, https://bioconductor.org/packages/scone, https://bioconductor.org/packages/zinbwave, https://bioconductor.org/packages/clusterExperiment, and https://bioconductor.org/packages/slingshot.
 

### 2018-07-17 workflow/workflow.Rmd
Small changes

@@
 ```{r setNormAssay}
-assays(dataObj)<-assays(dataObj)[c("normalizedValues","residuals","counts")]
+assays(dataObj)<-assays(dataObj)[c("normalizedValues","residuals","counts","weights")]
 ```
@@
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
-            sampleData = c("Batch", "Experiment"),
+            colData = c("Batch", "Experiment"),
             clusterLegend = metaColors, annLegend = FALSE, main = "")

### 2018-07-17 workflow/workflow.Rmd
Edit to text

@@
 The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015].
-In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis.
+In these workflows, single-cell expression data are organized in objects of the `SingleCellExperiment` class allowing integrated analysis.
 However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.
 Throughout the workflow, we use a single `SingleCellExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
+The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].
 
@@
 ```
-
-EP: I have only updated the code, not the text. The text needs adaptation, especially future directions and references to `SummarizedExperiment` which should be `SingleCellExperiment`.
-FP: TODO, fix ref.
 
@@
 Finally, results were validated experimentally using in vivo lineage tracing.
-Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Street2017].
+Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Risso2018; @Street2017].
 
@@
 
-Note that in order to successfully run the workflow, we need the devel versions
+Note that in order to successfully run the workflow, we need the following versions
 of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
 We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
-
-EP: Should update these.
-
-DR: Done, but let's double-check again before the final submission.
 
@@
 To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions.
-Computations were performed with 2 cores on a MacBook Pro (early 2015) with a 2.7 GHz Intel Core i5 processor and 8 GB of RAM.
-The `BiocParallel` package is used to allow for parallel computing in `zinbwave` function.
+Computations were performed with 2 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM.
+The `BiocParallel` package is used to allow for parallel computing in the `zinbwave` function.
 Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.
@@
 ```
-
-EP: Note that clusterExperiment uses `parallel` package, so might should update this.
 
@@
 
-EP: There's code in SingleCellExperiment for spike-ins. We should use it, though I'm not familiar with it. 
-FP: Davide, what is the standard way to use SingleCellExperiment for spike-ins?
+DR: Consider keeping the spike-ins in the object to show the ability of SCE to use spike-ins... but we don't use them, so not sure how important it is.
 
@@
 
-Throughout the workflow, we use the class `SingleCellExperiment` to keep track of the counts and their associated metadata within a single object.
 The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
@@
 metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2
+```
 
+Throughout the workflow, we store the data in an object of class `SingleCellExperiment`.
+The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data.
+In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object.
+Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.
+
+```{r}
 dataObj <- SingleCellExperiment(assays = list(counts = E),
@@
 
-Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
+Using the Bioconductor R package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
 See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
@@
 Finally, for computational efficiency, we retain only the 1,000 most variable genes.
-This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
+This seems to be a reasonable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
 In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.
 
+To compute the gene-wise variances and keep only the most variable genes, we use two functions from the `clusterExperiment` package.
+In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`).
+It will then store the variance values in the `rowData` slot of the object.
+`filterData` will use the computed values to keep only the desired number of genes.
+
 ```{r}
-# # Filtering to top 1,000 most variable genes
-vars <- rowVars(log1p(counts(dataObj)))
-names(vars) <- rownames(dataObj)
-vars <- sort(vars, decreasing = TRUE)
-# dataObj <- dataObj[names(vars)[1:1000],]
-
-#alternative script:
-dataObj<-makeFilterStats(dataObj,filterStats="var", transFun = log1p)
-dataObj<-filterData(dataObj,percentile=1000,filterStats="var")
-
-length(intersect(rownames(dataObj), names(vars)[1:1000]))
-setdiff(rownames(dataObj), names(vars)[1:1000])
+dataObj <- makeFilterStats(dataObj, filterStats="var", transFun = log1p)
+dataObj <- filterData(dataObj, percentile=1000, filterStats="var")
+rowData(dataObj)
 ```
-
-EP: Added existing code in the clusterExperiment package that does this. Also will mean that any future steps in package that need variance for the dataObj will have variance per gene stored.  
-FP: great. I added the filterStats argument to filterData function. I get an error otherwise.
-
-FP: We get a different list of most variable genes here.
-
-DR: Fixed, the new code was computing the variance in linear space instead of using the log.
-
-DR: Remember to remove the old code and the length() and setdiff() from above.
 
@@
 ```{r zinb,eval=FALSE}
-print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch", residuals = TRUE, normalizedValues = TRUE)))
+print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
+                                      residuals = TRUE, 
+                                      normalizedValues = TRUE)))
 ```
@@
 
-The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
+The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
@@
 ```
-
-FP: These values are slightly different from what
 
@@
 
-EP: This saves a lot of trouble with the plotHeatmap, etc.
-FP: Ok.
-
-EP: I've changed this code from my previous ad hoc to not delete the previous, but just reorder them.
-Again, (it might make sense for me to add option to different assays in clusterExperiment, but I'm not going to do that right now.)
-
-EP: Should probably address these normalized values, especially to more strongly make clear that shouldn't use these for clustering, just for visualization.
-The way its stated it sounds like dimensionality reduction is an "also", when in fact it is the main point and the normalized values are the afterthought.
-Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.
-FP: Agree, let's keep it in mind for once the workflow runs smoothly.
-
-DR: Agree, I'm not even sure normalized values are that useful for visualization... a simple scaling by total counts would probably achieve the same... but since we have them...
+In addition to the normalized values, the `zinbwave` function added several new components to the object: the assay `residuals` contains the residuals of the model that can be used to check the goodness-of-fit of the model.
+The `weights` assay contains observational weights that can be used to downweight the effect of zero counts on differential expression (see [@van2018observation] and the `zinbwave` vignette for more details).
 
@@
 
-The `zinbwave` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
+Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, which is stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object.
+This matrix can be retrieved using the `reducedDim()` extractor.
+
+```{r}
+W <- reducedDim(dataObj)
+W[1:3,1:3]
+```
+
+In this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
@@
 ```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
-W <- reducedDim(dataObj)
 d <- dist(W)
@@
      pch = 20, xlab = "Component 1", ylab = "Component 2")
-legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, fill = unique(col_clus), title = "Sample")
+legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, 
+       fill = unique(col_clus), title = "Sample")
 ```
@@
 As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.
+For more details on the `RSEC` method and on the `clusterExperiment` package, see [@Risso2018].
 
-Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts.
-Here, we are applying `RSEC` instead to the low-dimensional `W` matrix from ZINB-WaVE, for which we make a separate `SummarizedExperiment` object.
-For this reason, we choose to not use certain options in `RSEC`.
-In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension.
-Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`reduceMethod`) and the construction of the hiearchy between the clusters (`dendroReduce`).
-We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.
-
-FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.
-EP: I just called it "zinbwave50" for convenience.
-I've changed future code to just use "zinbwave".
-Though as a suggestion, you might consider for the zinbwave function to add the value of K to the name automatically, since the dimensions are not nested within each other, like pca, so you might run multiple versions of zinbwave and want to compare different K.
-
-EP: Note that we need to see whether clusterExperiment retains all of the different assays or not.
-It should in principle, but its not something I've tested.
-Also, I purposefully didn't give the parameter `nReducedDims` in RSEC because I think it should automatically use the max saved in object if not designated.
-If we're not getting the same results, we can add in `nReducedDims=50` and see; if that changes anything, let me know.
-However, I would focus on the question of different W values.
-
-FP: I am not sure it makes sense to perform the DE on the normalized values.
-Intuitively, I would say no, but not sure.
-We can still show that it is a possibility and write in the text that in this case, it might not make sense.
-We are working on a DE method for scRNA-seq where we use observational weights computed from the posterior probabilities of the ZINB-WaVE model in bulk RNA tools like edgeR and DESeq2.
-I think, eventually, it would be nice to be able to use that.
-Maybe, not for the next next version of the workflow.
-EP: I'd note that we use the normalized values for finding those DE with respsect to lineage.
-But I don't think its necessary to put it in here, but maybe in future directions, where you discuss that functionality, we could mention that it would allow the usage of mergeClusters.
-Also would be a good time to point out that DE on normalized values probably not a great idea either, if you don't think so.
-I also think the reason for skipping mergeClusters should be stated in the above text, rather than just saying we are skipping it.
-
-DR: A lot of these comments are now outdated, as the new version of clusterExperiment fully support multiple assays.
-Similarly, the weights are now a reality and also supported by clusterExperiment.
-We should probably add a section about it?
+Here, we are applying `RSEC` to the low-dimensional `W` matrix from ZINB-WaVE.
+To do so, we need to specify the `reducedDim` slot to use with the `reduceMethod` option.
+Here, we skip the merging step to collapse similar clusters based on the amount of differential gene expression between them.
+In larger dataset, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).
 
@@
 
-FP: We get more clusters for the makeConsensus method than before. Maybe because the most variable genes are different?
-
-DR: I fixed the most variable genes. Is there a way to check the older results? Where do I find the saved objects?
+DR: Is there a way to check the older results? Where do I find the saved objects?
 
@@
 
-FP: Ordering does not seem right.  
-EP: I get a very different plotCoClustering plot than what you sent me. See email.
-
 The distribution of cells across the consensus clusters can be visualized in Figure \@ref(fig:barplotOurs) and is as follows:
@@
 
-FP: we do not get exactly the same number of cells in each cluster. It looks similar though.
-
-DR: it's pretty similar, but we still get 7 clusters that look more or less the same as before.
-
 ```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
@@
 
-The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters.
+The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters) and \@ref(fig:addPublishedClusters2)), the main difference being that several of the published clusters were merged here into single clusters.
 This discrepancy is likely caused by the fact that we started with the top 1,000 genes, which might not be enough to discriminate between closely related clusters.
@@
 ```
-
-FP: compared to the previous version, it seems that there is a lot of black. Would the black coming from the contour of the rectangles in the bars?
-EP: I'm not sure what you mean about a lot of black on the barplot. Do you mean the unassigned samples? The colors have changed because the default set of colors was updated.
-
-DR: Now it looks pretty much the same.
-
-DR: Better to use `plotClustersTable` here? See below.
 
@@
 plotHeatmap(dataObj, 
-            whichClusters = c("makeConsensus","publishedClusters"), clusterFeaturesData = "all",
+            whichClusters = c("makeConsensus","publishedClusters"), 
+            clusterFeaturesData = "all",
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
@@
 
-FP: publishedClusters does not appear in the bars.
-EP: I forgot to add it to the `whichClusters` argument.
-
 Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is anologous to Figure \@ref(fig:mdsW) for the original published clusters.
@@
 ```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
-plotReducedDims(dataObj,whichCluster="primary",reducedDim="zinbwave",pch=20,
-	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main="",
+plotReducedDims(dataObj, whichCluster="primary", reducedDim="zinbwave", pch=20,
+	xlab = "Component1", ylab = "Component2", legendTitle="Sample", main="",
 	plotUnassigned=FALSE
 )
-
-# ###test
-# col<-convertClusterLegend(dataObj,whichCluster="primary",output="matrixColors")[,1]
-# col[col=="white"]<-"grey"
-# W<-reducedDim(dataObj,"zinbwave")
-# plot(-W[,1:2],col=col,pch=20)
 ```
-
-FP: I get `"reduceMethod" is not a graphical parameter`. I think this plot is different from the previous version.
-
-EP: I had wrong name for argument which has changed (should have been `reducedDim`), so it went with default which is PCA on the assay(dataObj) slot -- i.e.
-not using zinbwave.
-Now it gives the same plot as my test, where I directly plotted the results of `reducedDim(dataObj,"zinbwave")`.
-However, its still not exactly the same as previous rendition online.
-First it looks like the dimensions got flipped, i.e.
-multiplied by -1.
-Furthermore, it doesn't look exactly the same (e.g.
-outlying points not the same), even if I color in the unassigned so we can see them.
-Is zinbwave giving back the exact same W as before?
-I've checked that the W in the `dataObj_afterZinbwave` and W I plotted above from `dataObj_afterRSEC` are the same so its not clusterExperiment messing something up.
-If so, this could be why we are getting slightly different clustering results.
-
-DR: It is possible that zinbwave does not return the same exact W. There is no guarantee of a global maximum, but only local. The plot looks more or less the same, but just flipped.
 
@@
 
-We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage.
+We now demonstrate how to use the Bioconductor package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage.
 The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
@@
 
-The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively.
-The first takes as input a low-dimensional representation of the cells and a vector of cluster labels.
-It fits an MST to the clusters and identifies lineages as paths through this tree.
-The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages.
-The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages.
-These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.
+This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object.
+The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (ie.
+a matrix and a vector) or, as below, components of a `SingleCellExperiment` object.
+When a `SingleCellExperiment` object is provided as input, the ouput will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function.
+For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.
 
@@
 -------------|-------------|-------| ----------
-c1 | HBC | blue | original 1, 5
-c2 | GBC | green | original 2, 3, 11
-c3 | mSUS | red | original 4, 7
-c4 | Contaminants | orange | original -2 
-c5 | mOSN | purple | original 9, 12
-c6 | Immature Neuron | brown | original 10, 14
-c7 | MV | light blue | original 15
-
-Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster.
-These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering.
-While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process.
-Note that, out of the 77 cells overexpressing `Reg3g`, 11 are captured in cluster `c4` and 21 are unclustered in our workflow's clustering (see Figure \@ref(fig:boxplotReg3g)).
-However, we retain the remaining 45 cells to infer lineages as they did not seem to influence the clustering.
-
-DR: This has changed (slightly). This is the table that I get currently.
-
-Cluster name | Description | Color | Correspondence
--------------|-------------|-------| ----------
 c1 | HBC | red | original 1, 5
@@
 c7 | mOSN | light blue | original 9
-
-```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
-plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g count values", cex.axis = .8, cex.lab = .8,assay=log1p(assays(dataObj)[["counts"]]))
-```
-
-FP: this plot is fairly different from the previous version. We should see for one cluster that the gene expression for Reg3g is much high that for the other clusters.
-
-EP: I think this plot showed the normalized values (saved in assay(dataObj)).
-But I see that the previous plot was on the counts.
-Is that better than normalized value?
-I have added (just to this plot for now), an option to plot a specific given matrix (or a character value defining an assay, but since we want log-transform of the counts, which is not saved in the main assay, putting in "counts" doesn't give plot we want).
-But it's now cluster c1, not c4 that has this property.
-It might be useful to take the clustering from the old RSEC data object and directly compare the clusterings.
-Regardless, none of the previous commands are going to get the right results going forward since the clusters don't have the same name.
-See note on plotReducedDims command where I think the input `W` has changed.
-
-DR: I think we should re-write this section.
-With the current version of the package and parameters, we don't get the contaminant cluster anymore, but most of the contaminant cells are now in c1.
-However c1 also includes HBC cells, so we should not remove it from the slingshot analysis.
 
@@
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
-reducedDim(pseudoCe, "MDS") <- mds$points
-pseudoCe <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1")
+lineages <- slingshot(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
 ```
-EP: See my email.
-I don't get this subsetting error on my machine or on TravisCI.
-I do get on SCF Servers which are running R3.4.1 and the older version of bioconductor.
-Can you check this on your machine?
 
-FP: I agree. I get an error though when I run line 501:
-`Error in .BasicFunsList[[f]] : no such index at level 1`.
-
-FP: I stooped here.
-
-DR: Above, I use the syntax to work directly on the ClusterExperiment object, but if we want to be able to plot the resulting lineages, we need the matrix method.
-
-Before fitting the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
+Before discussing the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
 This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).
@@
 colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
-lineages <- getLineages(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
 pairs(lineages, type="lineages", col = colorCl)
@@
 
-DR: For some reason, this plot does not work anymore, but the lineages that I get are correct!
-
-Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables.
+Having found the global lineage structure, `slingshot` then constructed a set of smooth, branching curves in order to infer the corresponding pseudotime variables.
 Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters.
-This makes them more stable and better suited for assigning cells to lineages.
+During this iterative process, a cell may even be reassigned to a different lineage if it is significantly closer to the corresopnding curve.
+This makes `slingshot` less reliant on the original clustering and generally more stable.
 The final curves are shown in Figure \@ref(fig:curves).
@@
 ```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
-lineages <- getCurves(lineages)
 pairs(lineages, type="curves", col = colorCl)
@@
 
-In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot.
+As an alternative, we could have incorporated the MDS results into the `pseudoCe` object and applied `slingshot` directly to it.
+Here, we need to specify that we want to use the MDS results, because `slingshot` would otherwise use the first element of the `reducedDims` list (in this case, the 10-dimensional `W` matrix from `zinbwave`).
+
+```{r sling_sce}
+reducedDim(pseudoCe, "MDS") <- mds$points
+pseudoCe <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1")
+pseudoCe
+colData(pseudoCe)[,grep("^sling", colnames(colData(pseudoCe)))]
+```
+
+The result of `slingshot` applied to a `ClusterExperiment` object is still of class `ClusterExperiment`.
+Note that we did not specify a set of cluster labels, implying that `slingshot` should use the default `primaryClusterNamed` vector.
+
+In the workflow, we recover the right ordering of the clusters using the unsupervised version of slingshot.
 However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
-`getLineages` has the option for the user to provide known end cluster(s).
+`slingshot` has the option for the user to provide known end cluster(s).
 Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c7` represent terminal cell fates.
@@
 ```{r slingshotsupervised, eval=FALSE}
-lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1",
-                         end.clus = c("c3", "c7"))
-lineages <- getCurves(lineages)
-pairs(lineages, type="curves", col = colorCl)
-pairs(lineages, type="lineages", col = colorCl, 
-      show.constraints = TRUE)
-
-lineages
+pseudoCeSup <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1",
+                          end.clus = c("c3", "c6", "c2"))
 ```
@@
 
-EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below.
-I've used the `transformData` command, which will not do anything if isCount=FALSE above, but will convert to right if we change isCount=TRUE.
-It's better to use that command than `assay` for that reason.
-Might should add comment in code or text explaining this.
-
 ```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
@@
 
-plotHeatmap(pseudoCe1[topgenes,], clusterSamplesData = "orderSamplesValue", breaks = .99)
+plotHeatmap(pseudoCe1[topgenes,], clusterSamplesData = "orderSamplesValue", 
+            breaks = .99)
 ```
 
-EP: note the above does heirarchcial clustering of the top genes. If want them to stay in order given, should give `clusterFeaturesData=topgenes` (I think that is the argument)
+DR: here, describe the transformData and plotHeatmap function. Above, probably right after RSEC, explain the ClusterExperiment class and its relation to SCE.
 
-EP: The above code should take care of the ordering and NA values.
-Note that subseting a ClusterExperiment object is suppose to also subset the orderSamples value correctly.
-Let me know if that is not the case.
+## Differential expression between clusters
+
+DR: I think that it will be nice to showcase the getBestFeatures function here as well as our observational weights.
 
@@
 
-In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model.
-In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference.
-In addition, we are extending ZINB-WaVE to identify differentially expressed genes, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
-We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.
-
-Finally, a new S4 class called `SingleCellExperiment` is currently under development (https://github.com/drisso/SingleCellExperiment).
-This new class is essentially a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension.
-This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step.
-Once the implementation of this class is stable, we would like to incorporate it to the workflow.
-
-EP:  Obviously should replace this text.
-Might want to mention ability of class to store data matrices not in memory, though not currently implemented, which is the next major change we need to do here.
+DR: Not sure if we need this section.
+The previous future developments are now done (nice!).
+Perhaps here we should talk about scalability and HDF5?
+Kelly, is there anything you want to add here on DE along lineages?
 
@@
 FP, DR, KS, and EP performed the data analaysis and wrote the code portions of the workflow.
-FP and SD wrote the text portion of the workflow, with contributions from the other three authors.
+DR, FP, and SD wrote the text portion of the workflow, with contributions from the other three authors.
 DR, EP, and SD supervised the research.
@@
 
+DR: Update grant information!
+
 # Acknowledgments

### 2018-07-17 reviewers_response.md
Started response to reviewers

For now, this are just notes, we should make this a polished response.

## Reviewer 1 (Stephanie Hicks)

1. In this workflow, the authors start with a count table. However, the majority of researchers will start with raw reads (e.g. a FASTQ file). It would be great if the author discussed current best practices for the quantification step of scRNA-seq data. Alternatively, the authors could point to other references that have already been developed.

We do refer to Aaron's workflow at the beginning, but we should make it more clear. We should also mention `scPipe`.

2. I would like to see the authors take advantage of the rich functionality and data exploration tools for cell- and gene-specific quality control (QC) introduced in low-level analysis workflows such as the one from Lun et al. (2016). Also, in this workflow, the authors create multiple SummarizedExperiment objects (e.g. one with only the top 1000 highly variable genes (HVGs), one with all genes, etc). This doesn’t seem efficient, especially with large single cell data sets such as the 1.3 million cells from embryonic mouse brains. I think both of these concerns can now be addressed with efforts such as the recently developed SingleCellExperiment Bioconductor object (https://github.com/drisso/SingleCellExperiment). For example, the authors could add a “USE” column in the gene- or cell-specific meta table to represent whether or not a particular gene in a particular cell met the filtering criteria applied. The authors could store W in the reduceDim assay of the SingleCellExperiment object.

We now make full use of the SingleCellExperiment class. We also show how to store the variance in the object and how to filter based on those values.

3. In ZINB-WaVE, the authors specify the number of dimensions for the low-dimensional space (K) to be K=50. Could the authors add more details for the reader explaining why they picked K=50 and describe situations in which a user would want to specify a higher or lower K? In particular, it would be useful to discuss computational time in terms of number of genes and cells. Also, it would be useful to note that if you only wanted to use ZINB-WaVE to remove known covariates for normalization, you can use K=0.

Note that a more comprehensive discussion on the role of K on computational time is elsewhere (the zinbwave paper). The question of how to choose K is a very important one, but I'm not sure a workflow is the right place to discuss it in details. We should definitely give more guidance to the user on how to pick K. Perhaps we should compute a few values and compare the results in terms of AIC/BIC.

Minor comments:
1. When selecting the top 1000 HVGs, why do the authors not take into account the overall mean-variance relationship and only select genes based on the variance?

This is mostly for illustration purposes, but we do find in our experience that naively selecting most variable genes gives a good set of informative genes, perhaps because it selects highly expressed genes. We should add a discussion about this perhaps pointing to the functions from `scater` that implement other ways of selecting the most informative genes.

2. It would be great if the authors referenced other tools available for similar analyses currently available. For example there are several available packages for normalization of scRNA-seq data, such as calculating global scaling factors can be done with scran (https://bioconductor.org/packages/release/bioc/html/scran.html) or gene and cell-specific scaling factors using SCnorm (https://github.com/rhondabacher/SCnorm). Alternatively, users might want to try using relative transcript counts using Census (https://bioconductor.org/packages/release/bioc/html/monocle.html).

Definitely. We should mention these. Perhaps even use these methods instead of zinbwave's normalized counts.

## Reviewer 2 (Andrew McDavid)

Major comments:

1. There is something wrong with the data download link in the F1000 version so that I am unable to download these files and actually reproduce the workflow. I experimented a bit to see if I could figure out how to download the data anyways, but will reserve further evaluation of this submission until this issue can be resolved by the authors.

```{r}
urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff% "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.)
```

Not sure what's wrong with the F1000 version. Perhaps the links are not rendered correctly? Let's look into this and make sure that the data can be downloaded.

2a. This workflow will likely be out-of-date when the underlying packages transition to use SingleCellExperiment. This is actually a positive thing because many of the more opaque lines of code (involving subsetting ERCC genes, etc) will be more streamlined.

Correct. Let's look into keeping the ERCC spike-ins in the object.

2b. It requires installation of the development branch of bioconductor, which impacts the usefulness of the workflow to the average user. I expect the authors will revise this tutorial when Bioconductor 3.6 is released and use of the devel branch is no longer necessary. Additionally `slingshot` is an requirement, but currently only exists on github and no SHA1 provided. I hope that `slingshot` will be added as a bioconductor package shortly. In the meantime, a tag must be added to the git repo for the release being used in this workflow and instructions provided for how to install this tag. Additionally, the authors may wish to note that installation instructions for the packages will be provided at the end of the workflow so that someone proceeding sequentially will not be tripped up.

Slingshot is in Bioconductor now. Good point about the installation instructions being at the end. I think we can provide a short section on installation at the beginning.

2c.Opaque code is presented in order to generate plots, e.g.
```{r}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20, xlab = "Component1", ylab = "Component2") legend(x = "topleft", legend = names(pal), cex = .5, fill = pal, title = "Sample")
```

While this complexity may be necessary, perhaps some of it could be encapsulated as accessor functions in the package? Too much complexity here may cause users to miss the forest for the trees.

The plotReduceDim() function now addresses this.

The authors could better motivate (or at least explain the impact of) some of the default parameters and procedures.
Why do we set a zcut threshold of 3 for the `scone` filtering? 
Why K=50 for zinbwave?
RSEC parameters

How should a user decide on a value for these parameters?

These are all great points. We should address them in more deatails.

## Reviewer 3 (Mike Love)

It would be useful to put a link to the source code (or to the section where the link to the source code exists) near the top of the document.

Will do.

I was confused a bit by "the first major bifurcation in the HBC lineage trajectory occurs prior to cell division". Can you be more specific about what you are referring to here by cell division, as without knowledge of the system, I'm not sure where the cell division you refer to should appear.

Yes, this is a bit opaque out of context. We should probably refer to it as cell cycle and show cell cycle related genes somewhere to make it more clear.

"within a single object": It may be good to explain what an "object" here is. You could, for example, refer to Figure 2 of the Bioconductor Nature Methods paper.

Good point. Let's expand on this.

Misspelling: "reasonnable"

Fixed, thanks (it was meant to be read with French accent ;)).

On filtering for most variable genes, I understand this decision, and I also recommend it during workshops before making ordination plots. I know that students are not always certain why we care about variance (unsupervised). I like to mention that these are the genes where the "action" is. A side point, the log(x+1) is not variance stabilizing for RNA-seq counts in general. This filter can give higher priority to low count genes than to genes where there is interesting biological variability (though I do not doubt that the very high biological variance genes will be preserved). It might be useful to show a vsn::meanSdPlot() for the matrix log1p(assay(se))?

This are both good points (why variance and why log1p). See also comment by Stephanie. I think that we could expand this section to show a couple of alternatives, e.g., scater's way of finding most variable genes and another transformation instead of log.

"correcting for batch effects": What are batch effects? (Of course, I know what they are, but a reader may not, and you could cite some of the single cell literature here.)

Fair point. Will do.

"Note that, in this case, the low-dimensional matrix W is not included in the computation of residuals to avoid the removal of the biological signal of interest.": I understood this sentence only on a second pass through. One problem is that you haven't defined W in the text yet (only in Figure 4). I would only reference the matrix W if you have defined it.

Yeah, I understand this question just because I know the model. We should rewrite this sentence.

Figure 6: Can you change the figure width so that PC1 is not squished?

Yes, that's very annoying.

Is there a circularity to the recovery of published clusters in Figure 6? Was ZINB-WaVE used in Fletcher (2017)?

Perhaps, but zinbwave was not used in the original paper.

Can you say what the meaning of the color white is in Figure 8 (in the text or caption near this figure)?

Yes, let's explain better the plot.

Figure 15 refers back to Figure 2 but does not use the same color scheme for the known cell types, so the reader cannot verify if you've recovered the lineages from the publication. It would be good therefore to have a legend for these figures (Fig 15 and following) indicating which cell types the colors refer to (this information is in the unlabeled table above, but should be included as a legend here).

I think it would be easier to change the colors right after the table in which we compare to the published results and keep the color consistent for the rest of the workflow. Elizabeth, is there a wrapper function to change colors without manually changing the clusterLegend?

Can you briefly describe what a GAM is ahead of Figure 17?

Yes, we can.

### 2018-07-18 reviewers_response.md
small comment

@@
 
-2c.Opaque code is presented in order to generate plots, e.g.
+2c. Opaque code is presented in order to generate plots, e.g.
 ```{r}
@@
 
+@Davide: Yes, there is a new function `recolorClusters` and `renameClusters` to change name or colors in a specific clustering.
+Also there is the `plotClusterLegend` function that could be used here to create a legend.
+
 Can you briefly describe what a GAM is ahead of Figure 17?

### 2018-07-20 reviewers_response.md
Edits to address reviews

@@
 
-1.
+> 1.
 In this workflow, the authors start with a count table.
@@
 
-2.
+> 2.
 I would like to see the authors take advantage of the rich functionality and data exploration tools for cell- and gene-specific quality control (QC) introduced in low-level analysis workflows such as the one from Lun et al.
@@
 
-3.
+> 3.
 In ZINB-WaVE, the authors specify the number of dimensions for the low-dimensional space (K) to be K=50.
@@
 
-Minor comments:
-1. When selecting the top 1000 HVGs, why do the authors not take into account the overall mean-variance relationship and only select genes based on the variance?
+> Minor comments:
+> 1. When selecting the top 1000 HVGs, why do the authors not take into account the overall mean-variance relationship and only select genes based on the variance?
 
-This is mostly for illustration purposes, but we do find in our experience that naively selecting most variable genes gives a good set of informative genes, perhaps because it selects highly expressed genes.
-We should add a discussion about this perhaps pointing to the functions from `scater` that implement other ways of selecting the most informative genes.
+This is mostly for illustration purposes, but we do find in our experience that naively selecting most variable genes gives a good set of informative genes, perhaps because it selects on/off genes.
+We have expanded this section to add a discussion about different strategies that account for the structure of the data can be employed.
 
-2.
+> 2.
 It would be great if the authors referenced other tools available for similar analyses currently available.
@@
 
-It would be useful to put a link to the source code (or to the section where the link to the source code exists) near the top of the document.
+> It would be useful to put a link to the source code (or to the section where the link to the source code exists) near the top of the document.
 
-Will do.
+We have modified the section "Package version" to include a link to the source code as well as instructions on how to install the packages needed by the workflow.
 
-I was confused a bit by "the first major bifurcation in the HBC lineage trajectory occurs prior to cell division".
+> I was confused a bit by "the first major bifurcation in the HBC lineage trajectory occurs prior to cell division".
 Can you be more specific about what you are referring to here by cell division, as without knowledge of the system, I'm not sure where the cell division you refer to should appear.
 
-Yes, this is a bit opaque out of context. We should probably refer to it as cell cycle and show cell cycle related genes somewhere to make it more clear.
+Thank you, we agree that the paragraph taken out of context is unclear.
+The first bifurcation identified by slingshot (occurring right after the clusters marked by DeltaHBC1 and DeltaHBC2 in Figure 2) produces two lineage trajectories: the first one (purple curve in Figure 2) generates sustentacular cells through a differentiation process characterized by the absence of cell cycle and division.
+To simplify, each HBC stem cells gives rise to one sustentacular cell.
+The other lineage trajectory (blue and orange curves in Figure 2) generates proliferative GBCs, which are characterized by cell cycle and division.
+These, in turn, generate olfactory sensory neurons and microvillous cells.
+We hope that this more detailed explanation will be useful to better understand our system.
+We have updated the text to add this more detailed explanation.
 
-"within a single object": It may be good to explain what an "object" here is. You could, for example, refer to Figure 2 of the Bioconductor Nature Methods paper.
+> "within a single object": It may be good to explain what an "object" here is. You could, for example, refer to Figure 2 of the Bioconductor Nature Methods paper.
 
-Good point. Let's expand on this.
+We have added a paragraph discussing what a SingleCellExperiment object is and what are the advantages of using a single object throughout a data analysis workflow.
 
-Misspelling: "reasonnable"
+> Misspelling: "reasonnable"
 
-Fixed, thanks (it was meant to be read with French accent ;)).
+Fixed, thanks.
 
-On filtering for most variable genes, I understand this decision, and I also recommend it during workshops before making ordination plots.
+> On filtering for most variable genes, I understand this decision, and I also recommend it during workshops before making ordination plots.
 I know that students are not always certain why we care about variance (unsupervised).
@@
 
-This are both good points (why variance and why log1p).
-See also comment by Stephanie.
-I think that we could expand this section to show a couple of alternatives, e.g., scater's way of finding most variable genes and another transformation instead of log.
+Thank you, this is a very good point.
+See also comment by Reviewer 1 about modeling the mean-variance relationship.
+We have expanded this section to add a discussion about why we care about the most variable genes and how different strategies that account for the structure of the data can be employed.
 
-"correcting for batch effects": What are batch effects? (Of course, I know what they are, but a reader may not, and you could cite some of the single cell literature here.)
+> "correcting for batch effects": What are batch effects? (Of course, I know what they are, but a reader may not, and you could cite some of the single cell literature here.)
 
-Fair point. Will do.
+We have expanded our discussion on batch effects to include an explanation of what we mean by this term in this specific example and more generally.
 
-"Note that, in this case, the low-dimensional matrix W is not included in the computation of residuals to avoid the removal of the biological signal of interest.": I understood this sentence only on a second pass through.
+> "Note that, in this case, the low-dimensional matrix W is not included in the computation of residuals to avoid the removal of the biological signal of interest.": I understood this sentence only on a second pass through.
 One problem is that you haven't defined W in the text yet (only in Figure 4).
@@
 
-Yeah, I understand this question just because I know the model. We should rewrite this sentence.
+We agree, that the reference to the model is too specific. We have modified the sentence. [TODO -- still not sure how to best describe this step]
 
-Figure 6: Can you change the figure width so that PC1 is not squished?
+> Figure 6: Can you change the figure width so that PC1 is not squished?
 
-Yes, that's very annoying.
+Yes, Figure 6 has been changed.
 
-Is there a circularity to the recovery of published clusters in Figure 6? Was ZINB-WaVE used in Fletcher (2017)?
+> Is there a circularity to the recovery of published clusters in Figure 6? Was ZINB-WaVE used in Fletcher (2017)?
 
-Perhaps, but zinbwave was not used in the original paper.
+ZINB-WaVE was not used in the original paper, which used a more complex normalization strategy, based on regressing out a set of QC measures.
+One key result in favor of the ZINB-WaVE projection is that with this method slingshot is able to estimate the correct lineages with minimal supervision (only the starting cluster), while in the original paper, one of the terminal states had to be specified.
 
-Can you say what the meaning of the color white is in Figure 8 (in the text or caption near this figure)?
+DR: Make sure this is true.
+
+> Can you say what the meaning of the color white is in Figure 8 (in the text or caption near this figure)?
 
@@
 
-Figure 15 refers back to Figure 2 but does not use the same color scheme for the known cell types, so the reader cannot verify if you've recovered the lineages from the publication.
+> Figure 15 refers back to Figure 2 but does not use the same color scheme for the known cell types, so the reader cannot verify if you've recovered the lineages from the publication.
 It would be good therefore to have a legend for these figures (Fig 15 and following) indicating which cell types the colors refer to (this information is in the unlabeled table above, but should be included as a legend here).
@@
 
-Can you briefly describe what a GAM is ahead of Figure 17?
+> Can you briefly describe what a GAM is ahead of Figure 17?
 

### 2018-07-20 workflow/workflow.Rmd
Edits to address reviews

@@
 The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015].
-In these workflows, single-cell expression data are organized in objects of the `SingleCellExperiment` class allowing integrated analysis.
+In these workflows, single-cell expression data are organized in objects of the `SingleCellExperiment` class allowing integrated analysis [@SingleCellExperiment].
 However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.
@@
 Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.
-Throughout the workflow, we use a single `SingleCellExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
+
+One of the cornerstones of Bioconductor is the use of common data structures, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
+Throughout the workflow, we use a single object of the `SingleCellExperiment` class to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. 
+The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data (see Figure 2 of [@huber2015orchestrating] for an overview of `SummarizedExperiment`).
 The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].
@@
 
-It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
-Then, the GBC lineage, in turn, branches off to give rise to mOSN, microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)).
+It was found that the first bifurcation produces two lineage trajectories: the first one (purple curve in Figure \@ref(fig:stemcelldiff)) generates mature sustentacular cells (mSUS) through a differentiation process characterized by the absence of cell cycle and division.
+The other lineage trajectory (blue and orange curves in Figure \@ref(fig:stemcelldiff)) generates proliferative GBCs, which are characterized by cell cycle and division.
+These, in turn, generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve).
+
 In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
 
-## Package versions
+## Package installation and versions
 
-The following packages are needed.
+The source code to reproduce the analysis in the workflow are available at https://github.com/fperraudeau/singlecellworkflow.
+The packages used in the workflow can be installed using the following in R:
+
+```{r, eval=FALSE}
+install.packages(c("doParallel", "gam", "RColorBrewer"))
+
+install.packages("BiocManager")
+
+BiocManager::install(c("scone", "zinbwave", "clusterExperiment", "slingshot",
+                       "BiocParallel"))
+```
+
+Note that in order to successfully run the workflow, we need the following versions
+of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
+We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
+
+We are now ready to load the packages.
 
@@
 library(zinbwave)
-library(SingleCellExperiment)
 library(slingshot)
@@
 ```
-
-Note that in order to successfully run the workflow, we need the following versions
-of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
-We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
 
@@
 
-Throughout the workflow, we store the data in an object of class `SingleCellExperiment`.
-The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data.
+As discussed abouve, we store the data in an object of the class `SingleCellExperiment`.
 In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object.
@@
 
-Finally, for computational efficiency, we retain only the 1,000 most variable genes.
+## Identification of most variable genes
+
+Often in unsupervised analyses of expression data, it is useful to focus on the most variable genes.
+These genes are the ones that vary the most across cells and are likely to be the most informative.
+Filtering out low-variance genes allows us to reduce the complexity of the problem, while retaining the most information in the dataset.
+
+Here, for computational efficiency, we retain only the 1,000 most variable genes, using a simple `log(x+1)` transformation.
 This seems to be a reasonable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
-In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.
+In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent and many different strategies have been proposed in the literature.
 
@@
 ```{r}
-dataObj <- makeFilterStats(dataObj, filterStats="var", transFun = log1p)
+dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = log1p)
 dataObj <- filterData(dataObj, percentile=1000, filterStats="var")
@@
 ```
+
+One drawback of this strategy is that the logarithm is not a variance-stabilizing transformation.
+Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package.
+And these may be a better choice in real data analyses.
+Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function.
+Here, we show an example with the `vst` function from `DESeq2`.
+
+```{r, eval=FALSE}
+library(DESeq2)
+dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = vst,
+                           filterNames = "var_vst")
+dataObj <- filterData(dataObj, percentile=1000, filterStats="var_vst")
+```
+
+Alternatively, the `scran` package [@Lun2016] includes the `trendVar` and `decomposeVar` functions, which estimate the mean-variance relation and the technical and biological components of the variance, to select those genes with a high biological variance.
+See [@Lun2016] for details.
+
+<!-- An example of how to integrate that strategy in this workflow is provided here. Note that to follow `scran` recommendations we first normalize the data using the `scran` method. -->
+
+<!-- ```{r, eval=FALSE} -->
+<!-- library(scran) -->
+<!-- dataScran <- computeSumFactors(dataObj) -->
+<!-- dataScran <- normalize(dataScran) -->
+
+<!-- tv <- trendVar(dataScran, use.spikes = FALSE) -->
+<!-- dv <- decomposeVar(dataScran, fit = tv) -->
+<!-- ``` -->
+
+<!-- The results of `dv$bio` can then be used to sort the genes and keep the top 1,000. -->
+
+DR: I'm not sure whether to keep this last example using scran.
+It's more complicated because we cannot directly use the clusterExperiment functions and it's somewhat involved (we would need to explain what the size factors are etc.).
+Also, the results make no sense (negative variances!).
+Perhaps we should just mention the method without the code?
 
@@
 
-Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object. Cells were processed in 18 different batches.
+Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object.
+Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects".
+Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch.
+Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing].
+In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.
 
@@
 ```{r pcanorm, fig.cap="ZINB-WaVE: PCA of normalized expression measures, where each point represents a cell.
-Cells are color-coded by batch (left panel) and by original published clustering (right panel)."}
+Cells are color-coded by batch (left panel) and by original published clustering (right panel).", fig.height=2.5}
 pca <- prcomp(t(norm))
@@
 ```
+
+DR: do we actually need the PCA plots? The whole point of zinbwave is dimensionality reduction, I think it's very confusing to do PCA here, even before showing the results of zinbwave...
 

### 2018-07-23 reviewers_response.md
Addressing reviews

@@
 
-1.
+> 1.
 There is something wrong with the data download link in the F1000 version so that I am unable to download these files and actually reproduce the workflow.
@@
 
-2a.
+> 2a.
 This workflow will likely be out-of-date when the underlying packages transition to use SingleCellExperiment.
@@
 
-2b.
+> 2b.
 It requires installation of the development branch of bioconductor, which impacts the usefulness of the workflow to the average user.
@@
 
-2c. Opaque code is presented in order to generate plots, e.g.
+> 2c. Opaque code is presented in order to generate plots, e.g.
 ```{r}
@@
 
-While this complexity may be necessary, perhaps some of it could be encapsulated as accessor functions in the package? Too much complexity here may cause users to miss the forest for the trees.
+> While this complexity may be necessary, perhaps some of it could be encapsulated as accessor functions in the package? Too much complexity here may cause users to miss the forest for the trees.
 
@@
 
-The authors could better motivate (or at least explain the impact of) some of the default parameters and procedures.
-Why do we set a zcut threshold of 3 for the `scone` filtering? 
-Why K=50 for zinbwave?
-RSEC parameters
+> The authors could better motivate (or at least explain the impact of) some of the default parameters and procedures.
+> Why do we set a zcut threshold of 3 for the `scone` filtering? 
+> Why K=50 for zinbwave?
+> RSEC parameters
 
-How should a user decide on a value for these parameters?
+> How should a user decide on a value for these parameters?
 
@@
 
-Yes, Figure 6 has been changed.
+We actually dropped Figure 6 as using the low-dimensional signal from ZINB-WaVE (rather than PCA on the normalized counts) is the recommended way to visualize the results, and we thought that it was confusing for the reader to show PCA at this stage of the workflow.
 
@@
 
-DR: Make sure this is true.
+DR: Make sure that with the original analysis we had to use a semi-supervised approach.
 
@@
 
-Yes, let's explain better the plot.
+We have added a better description of the Figure in the caption.
 
@@
 
-I think it would be easier to change the colors right after the table in which we compare to the published results and keep the color consistent for the rest of the workflow.
-Elizabeth, is there a wrapper function to change colors without manually changing the clusterLegend?
-
-@Davide: Yes, there is a new function `recolorClusters` and `renameClusters` to change name or colors in a specific clustering.
-Also there is the `plotClusterLegend` function that could be used here to create a legend.
+Thanks for the suggestion, we agree that having a similar color scheme will help the reader make the connection between the results of Figure 15 and Figure 2.
+We have changed the colors to match as close as possible those of the published figure.
+Note that we do this right after the clustering step, so that all figures (from Figure 8 to Figure 17) have consistent colors.
 
@@
 
-Yes, we can.
+We have added a sentence describing GAMs ahead of Figure 17.

### 2018-07-23 workflow/workflow.Rmd
Addressing reviews

@@
 
+The correspondance between the clusters we found here and the original clusters is as follows.
+
+```{r tabagain}
+table(data.frame(original = publishedClusters, ours = primaryClusterNamed(dataObj)))
+```
+
+To facilitate the comparison between our results and the original clusters, we set a color scheme that allows a direct visual comparison with Figure \@ref(fig:stemcelldiff).
+We can change the clustering color scheme using `clusterExperiment`'s `recolorClusters` function.
+
+```{r colors}
+new_col <- col_clus[c("1", "4", "12", "3", "10", "15", "9")]
+names(new_col) <- paste0("c", 1:7)
+dataObj <- recolorClusters(dataObj, value = new_col)
+```
+
+Cluster name | Description | Color | Correspondence
+-------------|-------------|-------| ----------
+c1 | HBC | green | original 1, 5
+c2 | mSUS | magenta | original 4, 7
+c3 | mOSN  | orange | original 9, 12 
+c4 | GBC | cyan | original 2, 3, 11
+c5 | Immature Neuron | purple | original 10, 14
+c6 | MV | blue | original 15
+c7 | mOSN | yellow | original 9
+
 The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examinemakeConsensus)), where columns correspond to cells and rows to different clusterings.
@@
 
-```{r examinemakeConsensus, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
-plotClusters(dataObj)
+```{r examinemakeConsensus, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package.
+The top row corresponds to the consensus clustering and is color-coded accordingly.
+The other rows correspond to the individual clustering across which the consensus was computed.
+In each row, the color is chosen to match as close as possible the color of the cluster from the previous row with the largest overlap."}
+plotClusters(dataObj, existingColors = "firstOnly")
 ```
-
-DR: Is there a way to check the older results? Where do I find the saved objects?
 
@@
 Additionally, we know that GBCs should be at a junction before the differentiation between MV and mOSN cells (Figure \@ref(fig:stemcelldiff)).
-The correspondance between the clusters we found here and the original clusters is as follows.
-
-```{r tabagain}
-table(data.frame(original = publishedClusters, ours = primaryClusterNamed(dataObj)))
-```
-
-Cluster name | Description | Color | Correspondence
--------------|-------------|-------| ----------
-c1 | HBC | red | original 1, 5
-c2 | mSUS | blue | original 4, 7
-c3 | mOSN  | green | original 9, 12 
-c4 | GBC | orange | original 2, 3, 11
-c5 | Immature Neuron | purple | original 10, 14
-c6 | MV | brown | original 15
-c7 | mOSN | light blue | original 9
 
@@
 We then can test the null hypothesis of no change over time for each gene using the `gam` package.
+
+Generalized Additive Models (GAMs) [@gam] are a class of models that replaces the usual linear function of the coefficients with a sum of smooth functions.
+In the code below, we use a special case of the GAM in which only one covariate (pseudotime) is present.
+This correspond to a simple loess regression, but with the added advantage of providing a framework for the testing of the significance of the association between pseudotime and expression.
+
 We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).

### 2018-07-24 reviewers_response.md
Addressing reviews

@@
 For now, this are just notes, we should make this a polished response.
+
+## Open questions
+
+1. Do we need PCA? I think that we don't and I have removed it, but if you feel strongly for including it, we can put it back.
+2.
+Do we need to include all the convoluted preprocessing to go from GEO to a SingleCellExperiment?
+I actually have a SingleCellExperiment object saved on Github.
+I see arguments both for and against...
+it may actually be useful for people to see how to create a SCE from scratch.
+3. Should we include spike-ins in the SingleCellExperiment object? I think so because one reviewer explicitly mentions it, but it makes it harder to carry out all the steps.
+4. Discussion about choice of K. AIC/BIC don't seem to provide a good answer. What should we do?
 
@@
 
-We do refer to Aaron's workflow at the beginning, but we should make it more clear. We should also mention `scPipe`.
+We agree that giving guidance to the readers on how to quantify gene expression in scRNA-seq data is an important topic.
+However, we feel that it is outside the scope of this paper.
+We have included a reference to the `scPipe` Bioconductor package, which provides functions to quantify gene expression starting from raw data.
 
@@
 
-We now make full use of the SingleCellExperiment class. We also show how to store the variance in the object and how to filter based on those values.
+We agree with the reviewer that the previous version of the workflow was not very efficient in terms of data representation.
+This was due to the fact that the tools employed in the different steps expected slightly different inputs.
+We have now harmonized all the tools to expect as input and produce as output an object of class `SingleCellExperiment`, including storing the dimensionality reduced matrix `W` in the `reducedDim` slot and storing the gene-wise variances as a column in the `rowData` for better subsetting and filtering.
 
@@
 
-Note that a more comprehensive discussion on the role of K on computational time is elsewhere (the zinbwave paper).
-The question of how to choose K is a very important one, but I'm not sure a workflow is the right place to discuss it in details.
-We should definitely give more guidance to the user on how to pick K.
-Perhaps we should compute a few values and compare the results in terms of AIC/BIC.
+The question of how to choose K is a very important one and a more comprehensive discussion on the role of K, both on accuracy and computational time, is in the original ZINB-WaVE paper.
+However, we agree that giving more guidance on the choice of K is a needed addition to this workflow.
+We have added a section, called "Choice of K", that discusses two functions, `zinbAIC` and `zinbBIC`, that can be used to decide the best value of `K` for a given dataset.
 
@@
 
-Definitely. We should mention these. Perhaps even use these methods instead of zinbwave's normalized counts.
+This is a good suggestion.
+We have added a new section at the end of the workflow, named "Alternative approaches", that discusses these and other strategies for dimensionality reduction, normalization, clustering, and batch correction.
 
@@
 
-Not sure what's wrong with the F1000 version. Perhaps the links are not rendered correctly? Let's look into this and make sure that the data can be downloaded.
+We apologized with the reviewer.
+Something must have happened during the formatting of the article for production on the F1000 Research website that omitted some of the characters from the url.
+The links work with no issues in the R markdown version of the workflow, available at https://github.com/fperraudeau/singlecellworkflow.
+We will make sure that the new version of the article does not have this problem.
 
@@
 
-Correct. Let's look into keeping the ERCC spike-ins in the object.
+The reviewer is correct. We now make full use of the features of SingleCellExperiment and as a result the code is much cleaner and easier to understand.
 
@@
 
-Slingshot is in Bioconductor now. Good point about the installation instructions being at the end. I think we can provide a short section on installation at the beginning.
+Slingshot is now a Bioconductor package, albeit in the devel branch for now.
+Since the new release is only a few months away, we have decided to require the use of Bioconductor 3.8 (currently devel, but soon release).
+We think that this is the best choice to avoid the workflow to be outdated later this year.
+We have added the code at the beginning to show how to install the required packages, using the new recommended `BiocManager` package.
 
@@
 
-The plotReduceDim() function now addresses this.
+We have created the `plotReduceDims()` function in `clusterExperiment` to address this issue.
 
@@
 
-These are all great points. We should address them in more deatails.
+We agree with the reviewer that more details on the choice of these parameters are needed.
+As the choice of `K` for the `zinbwave` model is critical, we have added a new section that describes a way to compare different values of `K` and select the best.
+As for the `zcut` threshold and the `RSEC` parameters, we have added a paragraph in the appropriate section describing in more details what the parameters control and what are good rules of thumb to set them to an appropriate value.
+[TODO]
 

### 2018-07-24 workflow/workflow.Rmd
Addressing reviews

@@
 
+Note that the preprocessing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper.
+These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g, 3' end sequencing, droplet-based methods versus full-length mRNA, plate-based methods).
+One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files.
+More details can be found in the `scPipe` vignette, available at https://bioconductor.org/packages/scPipe.
+
 ## Package installation and versions
@@
 To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions.
-Computations were performed with 2 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM.
+Computations were performed with 6 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM.
 The `BiocParallel` package is used to allow for parallel computing in the `zinbwave` function.
@@
 ```{r parallel}
-NCORES <- 2
+NCORES <- 6
 mysystem = Sys.info()[["sysname"]]
@@
 
-The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the original clusters (Figure \@ref(fig:pcanorm)).
-Overall, it seems that normalization was effective at removing batch effects without removing biological signal, in spite of the partial nesting of batches within clusters.
-
-```{r pcanorm, fig.cap="ZINB-WaVE: PCA of normalized expression measures, where each point represents a cell.
-Cells are color-coded by batch (left panel) and by original published clustering (right panel).", fig.height=2.5}
-pca <- prcomp(t(norm))
-par(mfrow = c(1,2))
-plot(pca$x, col = col_batch[batch], pch = 20, main = "")
-plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
-```
-
-DR: do we actually need the PCA plots? The whole point of zinbwave is dimensionality reduction, I think it's very confusing to do PCA here, even before showing the results of zinbwave...
-
 We are going to set these normalized values to be our first assay, so that they become the default in future plots.
@@
 ```
+
+Overall, it seems that `zinbwave` was effective at removing batch effects without removing biological signal, in spite of the partial nesting of batches within clusters.
+
+### Choice of K
+
+The most important parameter to choose for the dimensionality reduction step is the number of hidden covariates to be included in the model, i.e., the number of columns of the hidden matrix `W`, which we indicate with `K`.
+One way to select the optimal value of `K` is to compare different values in terms of Akaike information criterion (AIC) and Bayesian information criterion (BIC).
+The `zinbwave` package provides two functions to compute such quantities, `zinbAIC` and `zinbBIC`.
+
+Here, we fit a few different models, specifying a different value of `K`, using the `zinbFit` function, which, unlike `zinbwave`, returns the estiamtes of all the parameters of the model, as well as the value of the likelihood function.
+
+```{r, eval=FALSE}
+k_values <- c(1: 5, 10)
+
+fit <- lapply(k_values, function(k) zinbFit(dataObj, K = k, which_assay = "counts"))
+saveRDS(fit, file = "Fit_allK.rds")
+```
+
+```{r}
+fit <- readRDS("fit_allK.rds")
+plot(k_values, sapply(fit, zinbBIC, t(counts(dataObj))), 
+     ylab = "BIC", xlab = "K", type = "b")
+plot(k_values, sapply(fit, zinbAIC, t(counts(dataObj))),
+     ylab = "AIC", xlab = "K", type = "b")
+```
+
+In this case, the BIC suggests a value of `K = 3`, while the AIC suggests a value of `K = 10`.
+
+Note, that since ZINB-WaVE is linear in the number of genes and number of cells, and quadratic on `K`, fitting multiple models and comparing their goodness-of-fit is infeasible for large datasets.
+In such cases, choosing a single value of `K` may be needed.
+In practice, we found that `K=10` and `K=20` are reasonable choices in real datasets.
+
+DR: There is quite a difference between AIC and BIC in this case, and it is actually hard to argue one value over another... how should we discussed these results?
 
@@
 
-# Conclusion
+# Conclusions
 
@@
 We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.
+
+## Alternative approaches
+
+Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years.
+The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach.
+For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal to the code structure above.
+This makes the present workflow a truly modular approach, in which a user can use their favorite tool for each step with the guarantee that the output of the step will be readable by the next step.
+
+An alternative normalization strategy is to compute gene and cell-specific scaling factors (e.g., `SCNorm` [@bacher2017scnorm]).
+Similarly, `monocle` [@trapnell2014dynamics] or `TSCAN` [@ji2016tscan] for cell lineage inference.
+Finally, the software suites `Seurat` [@seurat] (available on CRAN) and `scde` [@kharchenko2014bayesian] provide alternative pipelines for cell type discovery, differential expression, and gene set analysis.
+A comprehensive list of tools developed for the analysis of scRNA-seq data is available through at https://www.scrna-tools.org [@zappia2018exploring].
 

### 2018-07-24 workflow/workflow.Rmd
Try k=10 for zinbwave

@@
 
-```{r zinb,eval=FALSE}
-print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
+```{r zinb,eval=TRUE}
+print(system.time(dataObj <- zinbwave(dataObj, K = 10, X = "~ Batch",
                                       residuals = TRUE, 
@@
 ```{r}
-#save(dataObj, file= 'dataObj_afterZinbwave.rda')
-load('dataObj_afterZinbwave.rda')
+save(dataObj, file= 'dataObj_afterZinbwave_k10.rda')
+#load('dataObj_afterZinbwave.rda')
 ```
@@
 
-```{r rsec_50,eval=FALSE}
+```{r rsec_50,eval=TRUE}
 print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
@@
 ```{r}
-#save(dataObj, file = 'dataObj_afterRSEC.rda')
-load('dataObj_afterRSEC.rda')
+save(dataObj, file = 'dataObj_afterRSEC_k10.rda')
+#load('dataObj_afterRSEC.rda')
 ```

### 2018-08-13 workflow/workflow.Rmd
Changes

@@
 
-```{r zinb,eval=TRUE}
+```{r zinb,eval=FALSE}
 print(system.time(dataObj <- zinbwave(dataObj, K = 10, X = "~ Batch",
@@
 ```{r}
-save(dataObj, file= 'dataObj_afterZinbwave_k10.rda')
-#load('dataObj_afterZinbwave.rda')
+#save(dataObj, file= 'dataObj_afterZinbwave_k10.rda')
+load('dataObj_afterZinbwave_k10.rda')
 ```
@@
 ```{r}
+k_values <- c(1: 5, 10)
 fit <- readRDS("fit_allK.rds")

### 2018-08-13 workflow/workflow.Rmd
Going back to k=50

--- 2018-07-24 workflow/workflow.Rmd
@@
 ```{r}
-#save(dataObj, file= 'dataObj_afterZinbwave.rda')
+#save(dataObj, file= 'dataObj_afterZinbwave_k10.rda')
 load('dataObj_afterZinbwave.rda')
@@
 ```{r}
-#save(dataObj, file = 'dataObj_afterRSEC.rda')
+#save(dataObj, file = 'dataObj_afterRSEC_k10.rda')
 load('dataObj_afterRSEC.rda')

### 2018-08-14 workflow/workflow.Rmd
Minor changes

@@
 
-In this case, the BIC suggests a value of `K = 3`, while the AIC suggests a value of `K = 10`.
+In this case, the BIC suggests a value of `K = 3`, while the AIC suggests a value of `K >= 10`.
 
@@
 In such cases, choosing a single value of `K` may be needed.
-In practice, we found that `K=10` and `K=20` are reasonable choices in real datasets.
+In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets.
 
-DR: There is quite a difference between AIC and BIC in this case, and it is actually hard to argue one value over another... how should we discussed these results?
+DR: There is quite a difference between AIC and BIC in this case, and it is actually hard to argue one value over another...
+how should we discussed these results?
+Also, if I keep going AIC will keep decreasing...
 
@@
 c4 | GBC | cyan | original 2, 3, 11
-c5 | Immature Neuron | purple | original 10, 14
+c5 | Neuronal Precursors | purple | original 10, 14
 c6 | MV | blue | original 15
-c7 | mOSN | yellow | original 9
+c7 | Immature Neurons | yellow | original 9
 
@@
 
-## Further developments
-
-DR: Not sure if we need this section.
-The previous future developments are now done (nice!).
-Perhaps here we should talk about scalability and HDF5?
-Kelly, is there anything you want to add here on DE along lineages?
-
 # Conclusions
@@
 A comprehensive list of tools developed for the analysis of scRNA-seq data is available through at https://www.scrna-tools.org [@zappia2018exploring].
+
+## Further developments
+
+DR: Not sure if we need this section.
+The previous future developments are now done (nice!).
+Perhaps here we should talk about scalability and HDF5?
+Kelly, is there anything you want to add here on DE along lineages?
 

### 2018-08-15 reviewers_response.md
Finish reviewers response

@@
 it may actually be useful for people to see how to create a SCE from scratch.
-3. Should we include spike-ins in the SingleCellExperiment object? I think so because one reviewer explicitly mentions it, but it makes it harder to carry out all the steps.
+3.
+Should we include spike-ins in the SingleCellExperiment object?
+I think so because one reviewer explicitly mentions it, but it makes it harder to carry out all the steps.
+For now, I am keeping them, showing with some EDA that are not very useful and removing them.
 4. Discussion about choice of K. AIC/BIC don't seem to provide a good answer. What should we do?
+5.
+Many of the parameters that we specify in RSEC are default.
+On one hand it would be nice to leave them so that we don't have to explain them (see comment from reviewer 2), but on the other hand it would be more reproducible if we decide to change the default values later...
+what should we do?
 
@@
 This is mostly for illustration purposes, but we do find in our experience that naively selecting most variable genes gives a good set of informative genes, perhaps because it selects on/off genes.
-We have expanded this section to add a discussion about different strategies that account for the structure of the data can be employed.
+We have expanded this section to add a discussion about alternative strategies, which use the mean-variance relation observed in RNA-seq data.
 
@@
 
-This is a good suggestion.
+Thanks for the suggestion.
 We have added a new section at the end of the workflow, named "Alternative approaches", that discusses these and other strategies for dimensionality reduction, normalization, clustering, and batch correction.
@@
 As for the `zcut` threshold and the `RSEC` parameters, we have added a paragraph in the appropriate section describing in more details what the parameters control and what are good rules of thumb to set them to an appropriate value.
-[TODO]
 
@@
 The first bifurcation identified by slingshot (occurring right after the clusters marked by DeltaHBC1 and DeltaHBC2 in Figure 2) produces two lineage trajectories: the first one (purple curve in Figure 2) generates sustentacular cells through a differentiation process characterized by the absence of cell cycle and division.
-To simplify, each HBC stem cells gives rise to one sustentacular cell.
-The other lineage trajectory (blue and orange curves in Figure 2) generates proliferative GBCs, which are characterized by cell cycle and division.
+To simplify, each HBC stem cell gives rise to one sustentacular cell.
+The other lineage trajectories (blue and orange curves in Figure 2) generate proliferative GBCs, which are characterized by cell cycle and division.
 These, in turn, generate olfactory sensory neurons and microvillous cells.
-We hope that this more detailed explanation will be useful to better understand our system.
+We hope that this will be useful to better understand our system.
 We have updated the text to add this more detailed explanation.
@@
 
-We have expanded our discussion on batch effects to include an explanation of what we mean by this term in this specific example and more generally.
+We have moved our discussion on batch effects to a new section earlier in the workflow and we have expanded it to include an explanation of what we mean by "batch effects" in this specific example and more generally.
 
@@
 
-We agree, that the reference to the model is too specific. We have modified the sentence. [TODO -- still not sure how to best describe this step]
+We agree that the reference to W taken out of context is hard to understand.
+In the revised version we introduce the matrix W in the previous paragraph making it more explicit what is its interpretation.
 
@@
 One key result in favor of the ZINB-WaVE projection is that with this method slingshot is able to estimate the correct lineages with minimal supervision (only the starting cluster), while in the original paper, one of the terminal states had to be specified.
-
-DR: Make sure that with the original analysis we had to use a semi-supervised approach.
 
@@
 
-We have added a sentence describing GAMs ahead of Figure 17.
+We have added a sentence describing GAMs and a reference ahead of Figure 17.

### 2018-08-15 workflow/workflow.Rmd
Finish reviewers response

@@
 
-We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).
-
-DR: Consider keeping the spike-ins in the object to show the ability of SCE to use spike-ins... but we don't use them, so not sure how important it is.
-
-```{r}
-# Remove ERCC and CreER genes
-cre <- E["CreER",]
-ercc <- E[grep("^ERCC-", rownames(E)),]
-E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
-E <- E[-which(rownames(E)=="CreER"), ]
-dim(E)
-```
-
 The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
@@
 qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
-qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))
 
@@
 
-Using the Bioconductor R package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
+There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used as quality controls.
+
+We can specify spike-in genes withing the `SingleCellExperiment` class.
+In this way, it is easier to retrieve their expression.
+In addition, some specific functions may make direct use of this information (see, e.g., the `computeSpikeFactors` function in the `scran` package).
+
+In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER).
+This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that this is not an endogenous gene.
+
+```{r}
+# Mark ERCC and CreER genes as spike-ins
+isSpike(dataObj, "ERCC") <- grepl("^ERCC-", rownames(dataObj))
+isSpike(dataObj, "Transgene") <- grepl("^CreER$", rownames(dataObj)) | 
+                                  grepl("^eGFP$", rownames(dataObj)) 
+```
+
+We can access a specific set of spike-ins (or all of them) by using the `isSpike` function.
+
+```{r}
+table(isSpike(dataObj))
+table(isSpike(dataObj, "Transgene"))
+```
+
+The total number of reads mapped to spike-ins is often a good quality-control measure. Hence, we include that into our `colData`.
+
+```{r}
+ercc <- counts(dataObj[isSpike(dataObj, "ERCC"),])
+colData(dataObj)$ERCC_reads <- colSums(ercc)
+```
+
+## Exploratory data analysis
+
+Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects".
+Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch.
+Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing].
+In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.
+
+```{r batch}
+batch <- colData(dataObj)$Batch
+col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
+              brewer.pal(8, "Accent")[1])
+names(col_batch) = unique(batch)
+table(batch)
+```
+
+Often, different batches differ in terms of data quality. In particular, we observed that the number of reads mapping to spike-ins can differ dramatically across batches.
+
+```{r spikebatch, fig.cap="Distribution of reads mapped onto ERCC spike-ins per batch."}
+boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
+        outline = 0, ylab="Expression measure", las = 3)
+```
+
+Figure \@ref(fig:spikebatch) shows how for some batches no reads are mapped to ERCC spike-ins, presumably because the spike-ins were not included.
+Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.
+
+```{r}
+dataObj <- dataObj[!isSpike(dataObj),]
+dataObj <- clearSpikes(dataObj)
+```
+
+In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).
+
+```{r original}
+publishedClusters <- colData(dataObj)[, "publishedClusters"]
+col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
+              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
+              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
+names(col_clus) <- sort(unique(publishedClusters))
+table(publishedClusters)
+```
+
+Note that there is partial nesting of batches within clusters (i.e., cell type), which could be problematic when correcting for batch effects in the dimensionality reduction step below.
+
+```{r clustbatch}
+table(data.frame(batch = as.vector(batch),
+                 cluster = publishedClusters))
+```
+
+## Sample filtering
+
+Using the Bioconductor package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
+For each criterion, a sample is filtered out if it is below a user-defined hard threshold (which we leave at default value) or an adaptive threshold, defined in terms of number of standard deviations from the mean or median absolute deviation from the median.
+
+The main parameter to set in this function is `zcut`, which represents the number of standard deviations from the mean and median absolute deviation from the median to use as threshold.
+We empirically found that using a threshold of three works well for typical datasets.
+
 See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
@@
 
-Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object.
-Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects".
-Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch.
-Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing].
-In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.
-
-```{r batch}
-batch <- colData(dataObj)$Batch
-col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
-              brewer.pal(8, "Accent")[1])
-names(col_batch) = unique(batch)
-table(batch)
-```
-
-In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).
-
-```{r original}
-publishedClusters <- colData(dataObj)[, "publishedClusters"]
-col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
-              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
-              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
-names(col_clus) <- sort(unique(publishedClusters))
-table(publishedClusters)
-```
-
-Note that there is partial nesting of batches within clusters (i.e., cell type), which could be problematic when correcting for batch effects in the dimensionality reduction step below.
-
-```{r clustbatch}
-table(data.frame(batch = as.vector(batch),
-                 cluster = publishedClusters))
-```
-
 ## Normalization and dimensionality reduction: ZINB-WaVE
@@
 
-As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space.
+The model can be thought of as a generalization of a factor analysis model, in which the biological signal of interest is assumed to be explained by a small number of "factors", represented in the model by the matrix `W` (Figure \@ref(fig:zinbschema)).
+
+As with most dimensionality reduction methods, the user needs to specify the number of factors (`K`, the number of columns of `W`), which correspond to the dimensions for the new low-dimensional space.
 Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`.
@@
 ```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
+batch <- colData(dataObj)$Batch
 norm_order <- norm[, order(as.numeric(batch))]
@@
 fit <- cmdscale(d, eig = TRUE, k = 2)
-plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
-     pch = 20, xlab = "Component 1", ylab = "Component 2")
+plot(fit$points, col = col_clus[as.character(colData(dataObj)$publishedClusters)],
+     main = "", pch = 20, xlab = "Component 1", ylab = "Component 2")
 legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, 
@@
                                 betas = 0.8, reduceMethod="zinbwave",
-                                clusterFunction = "hierarchical01", minSizes=1,
-                                ncores = NCORES, isCount=FALSE,
+                                clusterFunction = "hierarchical01",
+                                ncores = NCORES,
                                 dendroReduce="zinbwave",
@@
                                           clusterArgs=list(nstart=10)),
-                                verbose=TRUE,
                                 consensusProportion = 0.7,
@@
 
-The correspondance between the clusters we found here and the original clusters is as follows.
+As one can see from the code above, there are many parameters that a user can set in RSEC.
+Many of the values that we choose are the default values of the function, but we nonetheless specify them for clarity and reproducibility.
+Here, we briefly describethe role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].
+
+- `k0s`: the values of `k` to start the sequential algorithm.
+- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
+- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
+- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
+- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
+- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster.
+- `mergeMethod`: our consensus clustering may find too many small clusters.
+If that's the case a merging step can be employed to merge clusters that do not show enough differential expression.
+Here, we skip this step by setting the method as "none".
+- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).
+
+The correspondence between the clusters we found here and the original clusters is as follows.
 
 ```{r tabagain}
-table(data.frame(original = publishedClusters, ours = primaryClusterNamed(dataObj)))
+table(data.frame(original = colData(dataObj)$publishedClusters,
+                 ours = primaryClusterNamed(dataObj)))
 ```
@@
 
-## Differential expression between clusters
-
-DR: I think that it will be nice to showcase the getBestFeatures function here as well as our observational weights.
-
 # Conclusions
@@
 The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach.
-For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal to the code structure above.
-This makes the present workflow a truly modular approach, in which a user can use their favorite tool for each step with the guarantee that the output of the step will be readable by the next step.
+For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal chenges to the code structure above.
+This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of the step will be readable by the next step.
 
 An alternative normalization strategy is to compute gene and cell-specific scaling factors (e.g., `SCNorm` [@bacher2017scnorm]).
-Similarly, `monocle` [@trapnell2014dynamics] or `TSCAN` [@ji2016tscan] for cell lineage inference.
+Similarly, `monocle` [@trapnell2014dynamics] or `TSCAN` [@ji2016tscan] are alternatives for cell lineage inference.
 Finally, the software suites `Seurat` [@seurat] (available on CRAN) and `scde` [@kharchenko2014bayesian] provide alternative pipelines for cell type discovery, differential expression, and gene set analysis.
-A comprehensive list of tools developed for the analysis of scRNA-seq data is available through at https://www.scrna-tools.org [@zappia2018exploring].
+A comprehensive list of tools developed for the analysis of scRNA-seq data is available at https://www.scrna-tools.org [@zappia2018exploring].
 
@@
 
-DR: Not sure if we need this section.
-The previous future developments are now done (nice!).
-Perhaps here we should talk about scalability and HDF5?
-Kelly, is there anything you want to add here on DE along lineages?
+Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, it is now becoming routine to generate datasets made of tens of thousands of cells.
+In fact, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells.
+Hence, there is a growing need for scalable methods.
+
+When datasets are too large to be fully stored in memory, HDF5 files [@folk2011overview] can be used to keep the data on disk and realize in memory only the subset of data needed at each step.
+We are in the process of updating the `clusterExperiment` and `zinbwave` packages to work with HDF5 files and we are developing more scalable methods to deal with massive datasets.
 

### 2018-08-16 reviewers_response.md
Initial draft for review

@@
-For now, this are just notes, we should make this a polished response.
-
-## Open questions
-
-1. Do we need PCA? I think that we don't and I have removed it, but if you feel strongly for including it, we can put it back.
-2.
-Do we need to include all the convoluted preprocessing to go from GEO to a SingleCellExperiment?
-I actually have a SingleCellExperiment object saved on Github.
-I see arguments both for and against...
-it may actually be useful for people to see how to create a SCE from scratch.
-3.
-Should we include spike-ins in the SingleCellExperiment object?
-I think so because one reviewer explicitly mentions it, but it makes it harder to carry out all the steps.
-For now, I am keeping them, showing with some EDA that are not very useful and removing them.
-4. Discussion about choice of K. AIC/BIC don't seem to provide a good answer. What should we do?
-5.
-Many of the parameters that we specify in RSEC are default.
-On one hand it would be nice to leave them so that we don't have to explain them (see comment from reviewer 2), but on the other hand it would be more reproducible if we decide to change the default values later...
-what should we do?
-
 ## Reviewer 1 (Stephanie Hicks)
@@
 > Minor comments:
+>
 > 1. When selecting the top 1000 HVGs, why do the authors not take into account the overall mean-variance relationship and only select genes based on the variance?
@@
 
-Major comments:
-
+> Major comments:
+>
 > 1.
@@
 > RSEC parameters
-
+>
 > How should a user decide on a value for these parameters?

### 2018-08-16 workflow/workflow.Rmd
Initial draft for review

@@
   - name: Fanny Perraudeau
-    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
+    affiliation: 
+      - Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
+      - Whole Biome, Inc., San Francisco, CA
   - name: Davide Risso
@@
   - name: Kelly Street
-    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
+    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
   - name: Elizabeth Purdom
@@
 While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline.
-This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run.
+This is all the more true with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run.
 For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.
@@
 
-Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.
+Here, we propose an integrated workflow for downstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.
 
-One of the cornerstones of Bioconductor is the use of common data structures, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
+One of the cornerstones of Bioconductor is the use of common data containers, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
 Throughout the workflow, we use a single object of the `SingleCellExperiment` class to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. 
@@
 
-The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium.
+The scRNA-seq dataset we use as a case study was generated to study the differentiation of HBC stem cells into different cell types present in the olfactory epithelium.
 To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
@@
 The other lineage trajectory (blue and orange curves in Figure \@ref(fig:stemcelldiff)) generates proliferative GBCs, which are characterized by cell cycle and division.
-These, in turn, generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve).
+This, in turn, diverges into two lineages that generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve), respectively.
 
-In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
+In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes by cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
 
@@
 One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files.
-More details can be found in the `scPipe` vignette, available at https://bioconductor.org/packages/scPipe.
+More details can be found in the `scPipe` vignette and in [@tian2017scpipe].
 
@@
 
-The source code to reproduce the analysis in the workflow are available at https://github.com/fperraudeau/singlecellworkflow.
+The source code to reproduce the analysis in the workflow is available at https://github.com/fperraudeau/singlecellworkflow.
 The packages used in the workflow can be installed using the following in R:
@@
 ```{r, eval=FALSE}
+# CRAN
 install.packages(c("doParallel", "gam", "RColorBrewer"))
 
+# Bioconductor
 install.packages("BiocManager")
-
 BiocManager::install(c("scone", "zinbwave", "clusterExperiment", "slingshot",
-                       "BiocParallel"))
+                       "BiocParallel", "DESeq2", "scater"))
 ```
@@
 ```{r packages}
-# Bioconductor
 library(BiocParallel)
@@
 library(slingshot)
-
-# CRAN
 library(doParallel)
@@
 
-urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
-         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")
+urls = c(paste0("https://www.ncbi.nlm.nih.gov/geo/download/",
+                "?acc=GSE95601&format=file&file=GSE95601",
+                "%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz"),
+         paste0("https://raw.githubusercontent.com/rufletch/",
+                "p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt"))
 
 if(!file.exists(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda"))) {
-  download.file(urls[1], paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
+  download.file(urls[1], paste0(data_dir, 
+                                "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
   R.utils::gunzip(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
@@
 
-As discussed abouve, we store the data in an object of the class `SingleCellExperiment`.
+As discussed above, we store the data in an object of the class `SingleCellExperiment`.
 In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object.
@@
 In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER).
-This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that this is not an endogenous gene.
+This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that these are not endogenous genes.
 
@@
 
+Note that here we use the function `counts()` from the `SingleCellExperiment` package that retrieves the matrix stored in the assay slot named `counts`.
+This is equivalent to `assay(dataObj, "counts")` but is obviously more convenient when accessing counts multiple times.
+It is hence a good idea to name the assay slot that stores the original counts "counts", as done above.
+
 ## Exploratory data analysis
@@
 Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing].
-In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.
+In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data are processed, or different times at which samples are collected.
 
@@
 boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
-        outline = 0, ylab="Expression measure", las = 3)
+        outline = 0, ylab="Reads mapped to ERCC", las = 3)
 ```
 
-Figure \@ref(fig:spikebatch) shows how for some batches no reads are mapped to ERCC spike-ins, presumably because the spike-ins were not included.
+Figure \@ref(fig:spikebatch) shows how for some batches no reads were mapped to ERCC spike-ins, presumably because the spike-ins were not included.
 Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.
@@
 
-In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).
+In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 233 cells not assigned to any cluster (i.e., cluster label of `-2`).
 
@@
 
-The main parameter to set in this function is `zcut`, which represents the number of standard deviations from the mean and median absolute deviation from the median to use as threshold.
-We empirically found that using a threshold of three works well for typical datasets.
-
-See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
-
 ```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
@@
 ```
+
+The main parameter to set in this function is `zcut`, which represents the number of standard deviations from the mean and median absolute deviation from the median to use as threshold.
+We empirically found that using a threshold of three works well for typical datasets.
+
+See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
 
@@
 One drawback of this strategy is that the logarithm is not a variance-stabilizing transformation.
-Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package.
-And these may be a better choice in real data analyses.
+Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses.
 Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function.
@@
 ```{r, eval=FALSE}
+## Not run
 library(DESeq2)
@@
 See [@Lun2016] for details.
-
-<!-- An example of how to integrate that strategy in this workflow is provided here. Note that to follow `scran` recommendations we first normalize the data using the `scran` method. -->
-
-<!-- ```{r, eval=FALSE} -->
-<!-- library(scran) -->
-<!-- dataScran <- computeSumFactors(dataObj) -->
-<!-- dataScran <- normalize(dataScran) -->
-
-<!-- tv <- trendVar(dataScran, use.spikes = FALSE) -->
-<!-- dv <- decomposeVar(dataScran, fit = tv) -->
-<!-- ``` -->
-
-<!-- The results of `dv$bio` can then be used to sort the genes and keep the top 1,000. -->
-
-DR: I'm not sure whether to keep this last example using scran.
-It's more complicated because we cannot directly use the clusterExperiment functions and it's somewhat involved (we would need to explain what the size factors are etc.).
-Also, the results make no sense (negative variances!).
-Perhaps we should just mention the method without the code?
 
@@
 
-Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`.
+Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor package `zinbwave`.
 The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data.
@@
 The user can also specify both gene-level and cell-level covariates.
-The inclusion of observed and unobserved cell-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
+The inclusion of cell-level covariates enables normalization for batch effects, while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
 A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema).
@@
 
-```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
+```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2018)."}
 knitr::include_graphics("zinb_schema.png")
@@
 Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`.
-Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.
+Note that if the user includes more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.
 
-```{r zinb,eval=FALSE}
+```{r zinb}
 print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
@@
                                       normalizedValues = TRUE)))
-```
-
-```{r}
-#save(dataObj, file= 'dataObj_afterZinbwave_k10.rda')
-load('dataObj_afterZinbwave.rda')
 ```
@@
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
-Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.
+Note that the low-dimensional matrix `W` is not included in the computation of the residuals to avoid the removal of the biological signal of interest.
 
@@
 
-We are going to set these normalized values to be our first assay, so that they become the default in future plots.
-
-```{r setNormAssay}
-assays(dataObj)<-assays(dataObj)[c("normalizedValues","residuals","counts","weights")]
-```
-
 In addition to the normalized values, the `zinbwave` function added several new components to the object: the assay `residuals` contains the residuals of the model that can be used to check the goodness-of-fit of the model.
@@
 
+```{r}
+assayNames(dataObj)
+```
+
+We set the normalized values as our first assay, so that they become the default in the plots below.
+
+```{r setNormAssay}
+assays(dataObj) <- assays(dataObj)[c("normalizedValues", "counts", "weights",
+                                     "residuals")]
+```
+
 ### Dimensionality reduction
 
-Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, which is stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object.
-This matrix can be retrieved using the `reducedDim()` extractor.
+Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object.
+This matrix contain the biological signal of interest and can be retrieved using the `reducedDim()` extractor.
 
@@
 
-In this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
+In this workflow, the user-supplied dimension `K` of the low-dimensional space was set to `K = 50`.
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
@@
 
-Here, we fit a few different models, specifying a different value of `K`, using the `zinbFit` function, which, unlike `zinbwave`, returns the estiamtes of all the parameters of the model, as well as the value of the likelihood function.
+Here, we fit a few different models, specifying a different value of `K`, using the `zinbFit` function, which, unlike `zinbwave`, returns the estimates of all the parameters of the model, as well as the value of the likelihood function.
 
-```{r, eval=FALSE}
-k_values <- c(1: 5, 10)
+```{r}
+k_values <- c(1:5, 10)
 
 fit <- lapply(k_values, function(k) zinbFit(dataObj, K = k, which_assay = "counts"))
-saveRDS(fit, file = "Fit_allK.rds")
 ```
 
-```{r}
+```{r bic, fig.cap="BIC of the model with different values of K."}
 k_values <- c(1: 5, 10)
-fit <- readRDS("fit_allK.rds")
 plot(k_values, sapply(fit, zinbBIC, t(counts(dataObj))), 
      ylab = "BIC", xlab = "K", type = "b")
+```
+
+```{r aic, fig.cap="AIC of the model with different values of K."}
 plot(k_values, sapply(fit, zinbAIC, t(counts(dataObj))),
@@
 
-In this case, the BIC suggests a value of `K = 3`, while the AIC suggests a value of `K >= 10`.
+In this case, the BIC suggests a value of `K = 3` (Figure \@ref(fig:bic))), while the AIC suggests a value of `K >= 10` (Figure \@ref(fig:aic))).
 
-Note, that since ZINB-WaVE is linear in the number of genes and number of cells, and quadratic on `K`, fitting multiple models and comparing their goodness-of-fit is infeasible for large datasets.
+Note, that since ZINB-WaVE is linear in the number of genes and number of cells, and quadratic on `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets.
 In such cases, choosing a single value of `K` may be needed.
 In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets.
-
-DR: There is quite a difference between AIC and BIC in this case, and it is actually hard to argue one value over another...
-how should we discussed these results?
-Also, if I keep going AIC will keep decreasing...
 
@@
 
-```{r rsec_50,eval=FALSE}
+```{r rsec_50}
 print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
@@
 
-```{r}
-#save(dataObj, file = 'dataObj_afterRSEC_k10.rda')
-load('dataObj_afterRSEC.rda')
-```
-
 As one can see from the code above, there are many parameters that a user can set in RSEC.
 Many of the values that we choose are the default values of the function, but we nonetheless specify them for clarity and reproducibility.
-Here, we briefly describethe role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].
+Here, we briefly describe the role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].
 
@@
 - `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
+- `clusterFunction`: function used for the subsample clustering: `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
 - `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
 - `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
-- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster.
+- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster in the consensus clustering step.
 - `mergeMethod`: our consensus clustering may find too many small clusters.
@@
 - `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).
+
+The results of `RSEC` is an object of the `ClusterExperiment` class.
+This class extends the `SingleCellExperiment` class, by adding additional slots to store the clustering results.
+In addition, it has an informative `show` method that displays important information on the clustering.
+
+```{r}
+dataObj
+```
+
+Note that an object of class `ClusterExperiment` is also of class `SingleCellExperiment`. Hence, all the methods define for `SingleCellExperiment` will still work on it.
+
+```{r}
+is(dataObj, "SingleCellExperiment")
+```
 
@@
 The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examinemakeConsensus)), where columns correspond to cells and rows to different clusterings.
-Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap accross rows.
+Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap across rows.
 The first row correspond to a consensus clustering across all candidate clusterings.
@@
 
-Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is anologous to Figure \@ref(fig:mdsW) for the original published clusters.
+Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is analogous to Figure \@ref(fig:mdsW) for the original published clusters.
 
@@
 This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object.
-The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (ie.
+The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e.
 a matrix and a vector) or, as below, components of a `SingleCellExperiment` object.
-When a `SingleCellExperiment` object is provided as input, the ouput will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function.
+When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function.
 For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.
@@
 
+First, we remove the unclustered cells.
+
 ```{r}
-pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
-X <- reducedDim(pseudoCe,type="zinbwave")
+dataFilt <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
+```
+
+Here, we show how to use `slingshot` starting from a matrix and a set of cluster labels.
+
+```{r}
+X <- reducedDim(dataFilt,type="zinbwave")
 mds <- cmdscale(dist(X), eig = TRUE, k = 4)
-lineages <- slingshot(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
+lineages <- slingshot(mds$points, clusterLabels = primaryClusterNamed(dataFilt), start.clus = "c1")
 ```
@@
 ```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
-colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
+colorCl<-convertClusterLegend(dataFilt,whichCluster="primary",output="matrixColors")[,1]
 pairs(lineages, type="lineages", col = colorCl)
@@
 
-Having found the global lineage structure, `slingshot` then constructed a set of smooth, branching curves in order to infer the corresponding pseudotime variables.
+Having found the global lineage structure, `slingshot` then constructs a set of smooth, branching curves in order to infer the corresponding pseudotime variables.
 Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters.
-During this iterative process, a cell may even be reassigned to a different lineage if it is significantly closer to the corresopnding curve.
+During this iterative process, a cell may even be reassigned to a different lineage if it is significantly closer to the corresponding curve.
 This makes `slingshot` less reliant on the original clustering and generally more stable.
@@
 
+The `show` method of the `SlingshotDataSet` class displays useful information on the inferred lineages.
+
 ```{r lineages}
@@
 
-As an alternative, we could have incorporated the MDS results into the `pseudoCe` object and applied `slingshot` directly to it.
+As an alternative, we could have incorporated the MDS results into the `ClusterExperiment` object and applied `slingshot` directly to it.
 Here, we need to specify that we want to use the MDS results, because `slingshot` would otherwise use the first element of the `reducedDims` list (in this case, the 10-dimensional `W` matrix from `zinbwave`).
@@
 ```{r sling_sce}
-reducedDim(pseudoCe, "MDS") <- mds$points
-pseudoCe <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1")
-pseudoCe
-colData(pseudoCe)[,grep("^sling", colnames(colData(pseudoCe)))]
+reducedDim(dataFilt, "MDS") <- mds$points
+dataFilt <- slingshot(dataFilt, reducedDim = "MDS", start.clus = "c1")
+dataFilt
+colData(dataFilt)[,grep("^sling", colnames(colData(dataFilt)))]
 ```
@@
 The result of `slingshot` applied to a `ClusterExperiment` object is still of class `ClusterExperiment`.
-Note that we did not specify a set of cluster labels, implying that `slingshot` should use the default `primaryClusterNamed` vector.
+Note that we did not specify a set of cluster labels, implying that `slingshot` should use the default `primaryCluster` vector.
 
 In the workflow, we recover the right ordering of the clusters using the unsupervised version of slingshot.
-However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
+However, in some other cases, the algorithm may need more guidance to find the correct ordering.
 `slingshot` has the option for the user to provide known end cluster(s).
@@
 ```{r slingshotsupervised, eval=FALSE}
-pseudoCeSup <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1",
+dataFiltSup <- slingshot(dataFilt, reducedDim = "MDS", start.clus = "c1",
                           end.clus = c("c3", "c6", "c2"))
@@
 ```{r fitgam}
-t <- colData(pseudoCe)$slingPseudotime_1
-y <- transformData(pseudoCe)
+t <- colData(dataFilt)$slingPseudotime_1
+y <- transformData(dataFilt)
 gam.pval <- apply(y,1,function(z){
@@
 
-pseudoCe1 <- pseudoCe[,!is.na(t)]
-orderSamples(pseudoCe1)<-order(t[!is.na(t)])
+dataFilt1 <- dataFilt[,!is.na(t)]
+orderSamples(dataFilt1)<-order(t[!is.na(t)])
 
-plotHeatmap(pseudoCe1[topgenes,], clusterSamplesData = "orderSamplesValue", 
+plotHeatmap(dataFilt1[topgenes,], clusterSamplesData = "orderSamplesValue", 
             breaks = .99)
 ```
-
-DR: here, describe the transformData and plotHeatmap function. Above, probably right after RSEC, explain the ClusterExperiment class and its relation to SCE.
-
-# Conclusions
-
-This workflow provides a tutorial for the analysis of scRNA-seq data in R/Bioconductor.
-It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
-The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method.
-We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.
 
@@
 The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach.
-For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal chenges to the code structure above.
-This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of the step will be readable by the next step.
+For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above.
+This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of each step will be readable by the next step.
 
@@
 
-Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, it is now becoming routine to generate datasets made of tens of thousands of cells.
+Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, datasets made of tens of thousands of cells are now routine.
 In fact, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells.
@@
 We are in the process of updating the `clusterExperiment` and `zinbwave` packages to work with HDF5 files and we are developing more scalable methods to deal with massive datasets.
+
+# Conclusions
+
+This workflow provides a tutorial for the analysis of scRNA-seq data in R/Bioconductor.
+It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
+The workflow is general and flexible, allowing the user to substitute the statistical method used in each step by a different method.
+We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.
 
@@
 
-FP, DR, KS, and EP performed the data analaysis and wrote the code portions of the workflow.
+FP, DR, KS, and EP performed the data analysis and wrote the code portions of the workflow.
 DR, FP, and SD wrote the text portion of the workflow, with contributions from the other three authors.
@@
 
-DR: Update grant information!
-
 # Acknowledgments

### 2018-08-18 workflow/workflow.Rmd
rm berkeley affiliation

@@
     affiliation: 
-      - Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
       - Whole Biome, Inc., San Francisco, CA

### 2018-08-18 workflow/workflow.Rmd
try plot with ercc

@@
 ```{r parallel}
-NCORES <- 6
+NCORES <- 2
 mysystem = Sys.info()[["sysname"]]
@@
 Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.
+
+```{r}
+# added chunk of code
+library(scater)
+sce <- calculateQCMetrics(
+    dataObj,
+    feature_controls = list(
+        ERCC = isSpike(sce, "ERCC")
+    )
+)
+cData <- colData(sce)
+cData <- as.data.frame(cData)
+ggplot(cData, 
+       aes(x = total_counts, 
+           y = total_counts_ERCC,
+           col = batch)) + 
+  geom_point()
+```
 

### 2018-08-18 workflow/workflow.Rmd
try plot with ercc

@@
     feature_controls = list(
-        ERCC = isSpike(sce, "ERCC")
+        ERCC = isSpike(dataObj, "ERCC")
     )

### 2018-08-20 workflow/workflow.Rmd
Update workflow.Rmd
Sandrine: 
Minor edits to text. 
A few questions indicated with [*** SD]

@@
   - name: Sandrine Dudoit
-    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
+    affiliation: Division of Epidemiology and Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
 abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types.
@@
 
-In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes by cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
+In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes-by-cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
 
-Note that the preprocessing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper.
-These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g, 3' end sequencing, droplet-based methods versus full-length mRNA, plate-based methods).
+Note that the pre-processing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper.
+These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g., 3'-end sequencing, droplet-based methods versus full-length mRNA, plate-based methods).
 One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files.
@@
 Note that in order to successfully run the workflow, we need the following versions
-of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
+of the Bioconductor packages: `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
 We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
@@
 
-To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions.
+To give users an idea of the time needed to run the workflow, the function `system.time` was used to report computation times for time-consuming functions.
 Computations were performed with 6 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM.
@@
 
-As discussed above, we store the data in an object of the class `SingleCellExperiment`.
-In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object.
+As discussed above, we store the data in an object of class `SingleCellExperiment`.
+In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the same object.
 Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.
@@
 
-There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used as quality controls.
+There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); [*** SD: Do you mean the "Sox2-eGFP gene"?] the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used for quality control purposes.
 
-We can specify spike-in genes withing the `SingleCellExperiment` class.
-In this way, it is easier to retrieve their expression.
+We can specify spike-in genes within the `SingleCellExperiment` class.
+In this way, it is easier to retrieve their expression measures.
 In addition, some specific functions may make direct use of this information (see, e.g., the `computeSpikeFactors` function in the `scran` package).
@@
 In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER).
-This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that these are not endogenous genes.
+This is a slight abuse of terminology, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that these are not endogenous genes.
 
@@
 
-The total number of reads mapped to spike-ins is often a good quality-control measure. Hence, we include that into our `colData`.
+The total number of reads mapped to spike-ins is often a good quality control measure. Hence, we include that into our `colData`.
 
@@
 
-Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects".
-Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch.
+Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch" and to the technical differences between each run as "batch effects".
+Note that, in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch.
 Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing].
@@
 
-```{r spikebatch, fig.cap="Distribution of reads mapped onto ERCC spike-ins per batch."}
+```{r spikebatch, fig.cap="Distribution of ERCC spike-in read counts per batch."}
 boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
@@
 
-Often in unsupervised analyses of expression data, it is useful to focus on the most variable genes.
-These genes are the ones that vary the most across cells and are likely to be the most informative.
+Often in unsupervised analyses of expression data, it is useful to focus on the genes that vary the most in expression across cells and are thus likely to be the most informative.
 Filtering out low-variance genes allows us to reduce the complexity of the problem, while retaining the most information in the dataset.
 
-Here, for computational efficiency, we retain only the 1,000 most variable genes, using a simple `log(x+1)` transformation.
+Here, for computational efficiency, we retain only the 1,000 most variable genes, following a simple `log(x+1)` transformation.
 This seems to be a reasonable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
@@
 To compute the gene-wise variances and keep only the most variable genes, we use two functions from the `clusterExperiment` package.
-In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`).
-It will then store the variance values in the `rowData` slot of the object.
-`filterData` will use the computed values to keep only the desired number of genes.
+In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`) and store the variance values in the `rowData` slot of the object [*** SD: Which object?].
+`filterData` will then use the computed values to keep only the desired number of genes.
 
@@
 
-One drawback of this strategy is that the logarithm is not a variance-stabilizing transformation.
+One drawback of this strategy is that the logarithm is not a variance stabilizing transformation.
 Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses.
@@
 
-Alternatively, the `scran` package [@Lun2016] includes the `trendVar` and `decomposeVar` functions, which estimate the mean-variance relation and the technical and biological components of the variance, to select those genes with a high biological variance.
+Alternatively, the `scran` package [@Lun2016] includes the `trendVar` and `decomposeVar` functions, which estimate the mean-variance relationship and the technical and biological components of the variance, to select those genes with a high biological variance.
 See [@Lun2016] for details.
@@
 
-In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and the identification of DE genes.
+In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime inference, and the identification of DE genes.
 This allows the data to become more tractable, both from a statistical (cf.
@@
 
-The most important parameter to choose for the dimensionality reduction step is the number of hidden covariates to be included in the model, i.e., the number of columns of the hidden matrix `W`, which we indicate with `K`.
-One way to select the optimal value of `K` is to compare different values in terms of Akaike information criterion (AIC) and Bayesian information criterion (BIC).
+The most important parameter to choose for the dimensionality reduction step is the number of hidden covariates to be included in the model, i.e., the number of columns of the hidden matrix `W`, which we denote by `K`.
+One way to select the optimal value of `K` is to compare different values in terms of the Akaike information criterion (AIC) or the Bayesian information criterion (BIC).
 The `zinbwave` package provides two functions to compute such quantities, `zinbAIC` and `zinbBIC`.
 
-Here, we fit a few different models, specifying a different value of `K`, using the `zinbFit` function, which, unlike `zinbwave`, returns the estimates of all the parameters of the model, as well as the value of the likelihood function.
+Here, we fit a few different models corresponding to different values of `K` using the `zinbFit` function, which, unlike `zinbwave`, returns the estimates of all the parameters of the model as well as the value of the likelihood function.
 
@@
 
-Note, that since ZINB-WaVE is linear in the number of genes and number of cells, and quadratic on `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets.
+Note that since ZINB-WaVE is linear in the number of genes and the number of cells, but quadratic on `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets.
 In such cases, choosing a single value of `K` may be needed.

### 2018-08-20 workflow/workflow.Rmd
Update workflow.Rmd
Sandrine: Minor edits to text.

@@
 Here, we skip the merging step to collapse similar clusters based on the amount of differential gene expression between them.
-In larger dataset, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).
+In larger datasets, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).
 
@@
 - `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
-- `clusterFunction`: function used for the subsample clustering: `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
+- `clusterFunction`: function used for the subsample clustering; `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
 - `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
-- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
+- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings, and the number of starting values for the k-means algorithm.
 - `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster in the consensus clustering step.
@@
 
-The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).
+The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusterings (Figure \@ref(fig:plotcoclust)).
 
@@
 This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object.
-The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e.
-a matrix and a vector) or, as below, components of a `SingleCellExperiment` object.
-When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function.
-For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.
+The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e., a matrix and a vector) or, as below, components of a `SingleCellExperiment` object.
+When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` object as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function.
+For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`, respectively.
 
@@
 
-Here, we show how to use `slingshot` starting from a matrix and a set of cluster labels.
+Here, we show how to use `slingshot` starting from a matrix of (low-dimensional) expression measures and a set of cluster labels.
 
@@
 In the code below, we use a special case of the GAM in which only one covariate (pseudotime) is present.
-This correspond to a simple loess regression, but with the added advantage of providing a framework for the testing of the significance of the association between pseudotime and expression.
+This corresponds to a simple loess regression, but with the added advantage of providing a framework for testing the significance of the association between pseudotime and expression.
 
@@
 Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years.
-The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach.
-For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above.
+The use of the `SingleCellExperiment` class, which is the shared "language" [*** SD: Better term than "language"?] among Bioconductor's scRNA-seq software packages, makes it extremely easy to replace one of the steps above by a different approach.
+For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global-scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above.
 This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of each step will be readable by the next step.
@@
 Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, datasets made of tens of thousands of cells are now routine.
-In fact, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells.
+Indeed, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells.
 Hence, there is a growing need for scalable methods.
@@
 Russell B.
-Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.
+Fletcher and Dr.Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.
 We would also like to thank Michael B.

### 2018-08-20 workflow/workflow.Rmd
Update workflow.Rmd
Sandrine: Minor edits to text and figure captions.

@@
 
-One drawback of this strategy is that the logarithm is not a variance stabilizing transformation.
-Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses.
+One drawback of this strategy is that the logarithm is not a variance-stabilizing transformation.
+Variance-stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses.
 Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function.
@@
 
-```{r bic, fig.cap="BIC of the model with different values of K."}
+```{r bic, fig.cap="ZINB-WaVE: BIC of model for different values of K."}
 k_values <- c(1: 5, 10)
@@
 
-```{r aic, fig.cap="AIC of the model with different values of K."}
+```{r aic, fig.cap="ZINB-WaVE: AIC of model for different values of K."}
 plot(k_values, sapply(fit, zinbAIC, t(counts(dataObj))),
@@
 The top row corresponds to the consensus clustering and is color-coded accordingly.
-The other rows correspond to the individual clustering across which the consensus was computed.
-In each row, the color is chosen to match as close as possible the color of the cluster from the previous row with the largest overlap."}
+The other rows correspond to the individual clusterings across which the consensus was computed.
+In each row, the colors are chosen to match as closely as possible with the colors of clusters with the largest overlap from the previous row."}
 plotClusters(dataObj, existingColors = "firstOnly")

### 2018-08-22 workflow/workflow.Rmd
added paragraph to Further Developments on downstream lineage DE methods.

--- 2018-08-20 workflow/workflow.Rmd
@@
 author: 
-
   - name: Fanny Perraudeau
-    affiliation: 
-      - Whole Biome, Inc., San Francisco, CA
+    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
   - name: Davide Risso
@@
   - name: Kelly Street
-    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
+    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
   - name: Elizabeth Purdom
@@
   - name: Sandrine Dudoit
-    affiliation: Division of Epidemiology and Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
+    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
+
 abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types.
@@
 Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.
+
 keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
+
 bibliography: ref.bib
+
 vignette: >
@@
 output: BiocWorkflowTools::f1000_article
-editor_options: 
-  chunk_output_type: console
 ---
@@
 ```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
-if(packageVersion("zinbwave")<"1.1.0") stop("must have current master version to avoid bugs")
-if(packageVersion("clusterExperiment")<"1.5.0-9009") stop("must have current devel version to avoid bugs")
-if(packageVersion("scone")<"1.2.0") stop("must have current master version to avoid bugs")
+if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
+if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current devel version to avoid bugs")
+if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")
 
@@
 library(knitr)
-knitr::opts_chunk$set(cache=FALSE, error=FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)
 knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
@@
 While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline.
-This is all the more true with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run.
+This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run.
 For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.
@@
 The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015].
-In these workflows, single-cell expression data are organized in objects of the `SingleCellExperiment` class allowing integrated analysis [@SingleCellExperiment].
+In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis.
 However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.
 
-Here, we propose an integrated workflow for downstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.
-
-One of the cornerstones of Bioconductor is the use of common data containers, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
-Throughout the workflow, we use a single object of the `SingleCellExperiment` class to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. 
-The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data (see Figure 2 of [@huber2015orchestrating] for an overview of `SummarizedExperiment`).
-The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].
+Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages.
+Throughout the workflow, we use a single `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.
 
@@
 
-The scRNA-seq dataset we use as a case study was generated to study the differentiation of HBC stem cells into different cell types present in the olfactory epithelium.
+The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium.
 To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing.
@@
 Finally, results were validated experimentally using in vivo lineage tracing.
-Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Risso2018; @Street2017].
+Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Street2017].
 
-It was found that the first bifurcation produces two lineage trajectories: the first one (purple curve in Figure \@ref(fig:stemcelldiff)) generates mature sustentacular cells (mSUS) through a differentiation process characterized by the absence of cell cycle and division.
-The other lineage trajectory (blue and orange curves in Figure \@ref(fig:stemcelldiff)) generates proliferative GBCs, which are characterized by cell cycle and division.
-This, in turn, diverges into two lineages that generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve), respectively.
+It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs.
+Then, the GBC lineage, in turn, branches off to give rise to mOSN, microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)).
+In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
 
-In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes-by-cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.
+## Package versions
 
-Note that the pre-processing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper.
-These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g., 3'-end sequencing, droplet-based methods versus full-length mRNA, plate-based methods).
-One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files.
-More details can be found in the `scPipe` vignette and in [@tian2017scpipe].
-
-## Package installation and versions
-
-The source code to reproduce the analysis in the workflow is available at https://github.com/fperraudeau/singlecellworkflow.
-The packages used in the workflow can be installed using the following in R:
-
-```{r, eval=FALSE}
-# CRAN
-install.packages(c("doParallel", "gam", "RColorBrewer"))
-
-# Bioconductor
-install.packages("BiocManager")
-BiocManager::install(c("scone", "zinbwave", "clusterExperiment", "slingshot",
-                       "BiocParallel", "DESeq2", "scater"))
-```
-
-Note that in order to successfully run the workflow, we need the following versions
-of the Bioconductor packages: `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`.
-We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
-
-We are now ready to load the packages.
+The following packages are needed.
 
 ```{r packages}
+# Bioconductor
 library(BiocParallel)
@@
 library(zinbwave)
+
+# GitHub
 library(slingshot)
+
+# CRAN
 library(doParallel)
@@
 
-## Parallel computing
+Note that in order to successfully run the workflow, we need the devel versions
+of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.3.2)`.
+We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).
 
-To give users an idea of the time needed to run the workflow, the function `system.time` was used to report computation times for time-consuming functions.
-Computations were performed with 6 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM.
-The `BiocParallel` package is used to allow for parallel computing in the `zinbwave` function.
+## Parallel computating
+
+To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions.
+Computations were performed with 2 cores on a MacBook Pro (early 2015) with a 2.7 GHz Intel Core i5 processor and 8 GB of RAM.
+The `BiocParallel` package is used to allow for parallel computing in `zinbwave` function.
 Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.
@@
 
-urls = c(paste0("https://www.ncbi.nlm.nih.gov/geo/download/",
-                "?acc=GSE95601&format=file&file=GSE95601",
-                "%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz"),
-         paste0("https://raw.githubusercontent.com/rufletch/",
-                "p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt"))
+urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
+         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")
 
 if(!file.exists(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda"))) {
-  download.file(urls[1], paste0(data_dir, 
-                                "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
+  download.file(urls[1], paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
   R.utils::gunzip(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
@@
 
+We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).
+
+```{r}
+# Remove ERCC and CreER genes
+cre <- E["CreER",]
+ercc <- E[grep("^ERCC-", rownames(E)),]
+E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
+E <- E[-which(rownames(E)=="CreER"), ]
+dim(E)
+```
+
+Throughout the workflow, we use the class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object.
 The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017].
@@
 qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
+qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))
 
@@
 metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2
+
+se <- SummarizedExperiment(assays = list(counts = E),
+                           colData = metadata)
+se
 ```
 
-As discussed above, we store the data in an object of class `SingleCellExperiment`.
-In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the same object.
-Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.
+Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
+See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
+
+```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
+# QC-metric-based sample-filtering
+data("housekeeping")
+hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]
+
+mfilt <- metric_sample_filter(assay(se), 
+                              nreads = colData(se)$NREADS,
+                              ralign = colData(se)$RALIGN,
+                              pos_controls = rownames(se) %in% hk,
+                              zcut = 3, mixture = FALSE,
+                              plot = TRUE)
+```
+
+```{r sconeFilt}
+# Simplify to a single logical
+mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
+se <- se[, mfilt]
+dim(se)
+```
+
+After sample filtering, we are left with `r ncol(se)` good quality cells.
+
+Finally, for computational efficiency, we retain only the 1,000 most variable genes.
+This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
+In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.
 
 ```{r}
-dataObj <- SingleCellExperiment(assays = list(counts = E),
-                           colData = metadata)
-dataObj
+# Filtering to top 1,000 most variable genes
+vars <- rowVars(log1p(assay(se)))
+names(vars) <- rownames(se)
+vars <- sort(vars, decreasing = TRUE)
+core <- se[names(vars)[1:1000],]
 ```
 
-There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); [*** SD: Do you mean the "Sox2-eGFP gene"?] the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used for quality control purposes.
+## Dataset structure
 
-We can specify spike-in genes within the `SingleCellExperiment` class.
-In this way, it is easier to retrieve their expression measures.
-In addition, some specific functions may make direct use of this information (see, e.g., the `computeSpikeFactors` function in the `scran` package).
-
-In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER).
-This is a slight abuse of terminology, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that these are not endogenous genes.
+Overall, after the above pre-processing steps, our dataset has 1,000 genes and 747 cells.
 
 ```{r}
-# Mark ERCC and CreER genes as spike-ins
-isSpike(dataObj, "ERCC") <- grepl("^ERCC-", rownames(dataObj))
-isSpike(dataObj, "Transgene") <- grepl("^CreER$", rownames(dataObj)) | 
-                                  grepl("^eGFP$", rownames(dataObj)) 
+core
 ```
 
-We can access a specific set of spike-ins (or all of them) by using the `isSpike` function.
-
-```{r}
-table(isSpike(dataObj))
-table(isSpike(dataObj, "Transgene"))
-```
-
-The total number of reads mapped to spike-ins is often a good quality control measure. Hence, we include that into our `colData`.
-
-```{r}
-ercc <- counts(dataObj[isSpike(dataObj, "ERCC"),])
-colData(dataObj)$ERCC_reads <- colSums(ercc)
-```
-
-Note that here we use the function `counts()` from the `SingleCellExperiment` package that retrieves the matrix stored in the assay slot named `counts`.
-This is equivalent to `assay(dataObj, "counts")` but is obviously more convenient when accessing counts multiple times.
-It is hence a good idea to name the assay slot that stores the original counts "counts", as done above.
-
-## Exploratory data analysis
-
-Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch" and to the technical differences between each run as "batch effects".
-Note that, in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch.
-Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing].
-In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data are processed, or different times at which samples are collected.
+Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object. Cells were processed in 18 different batches.
 
 ```{r batch}
-batch <- colData(dataObj)$Batch
+batch <- colData(core)$Batch
 col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
@@
 
-Often, different batches differ in terms of data quality. In particular, we observed that the number of reads mapping to spike-ins can differ dramatically across batches.
-
-```{r spikebatch, fig.cap="Distribution of ERCC spike-in read counts per batch."}
-boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
-        outline = 0, ylab="Reads mapped to ERCC", las = 3)
-```
-
-Figure \@ref(fig:spikebatch) shows how for some batches no reads were mapped to ERCC spike-ins, presumably because the spike-ins were not included.
-Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.
-
-```{r}
-# added chunk of code
-library(scater)
-sce <- calculateQCMetrics(
-    dataObj,
-    feature_controls = list(
-        ERCC = isSpike(dataObj, "ERCC")
-    )
-)
-cData <- colData(sce)
-cData <- as.data.frame(cData)
-ggplot(cData, 
-       aes(x = total_counts, 
-           y = total_counts_ERCC,
-           col = batch)) + 
-  geom_point()
-```
-
-```{r}
-dataObj <- dataObj[!isSpike(dataObj),]
-dataObj <- clearSpikes(dataObj)
-```
-
-In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 233 cells not assigned to any cluster (i.e., cluster label of `-2`).
+In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).
 
 ```{r original}
-publishedClusters <- colData(dataObj)[, "publishedClusters"]
+publishedClusters <- colData(core)[, "publishedClusters"]
 col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
@@
 
-## Sample filtering
-
-Using the Bioconductor package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes.
-For each criterion, a sample is filtered out if it is below a user-defined hard threshold (which we leave at default value) or an adaptive threshold, defined in terms of number of standard deviations from the mean or median absolute deviation from the median.
-
-```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
-# QC-metric-based sample-filtering
-data("housekeeping")
-hk = rownames(dataObj)[toupper(rownames(dataObj)) %in% housekeeping$V1]
-
-mfilt <- metric_sample_filter(counts(dataObj), 
-                              nreads = colData(dataObj)$NREADS,
-                              ralign = colData(dataObj)$RALIGN,
-                              pos_controls = rownames(dataObj) %in% hk,
-                              zcut = 3, mixture = FALSE,
-                              plot = TRUE)
-```
-
-The main parameter to set in this function is `zcut`, which represents the number of standard deviations from the mean and median absolute deviation from the median to use as threshold.
-We empirically found that using a threshold of three works well for typical datasets.
-
-See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.
-
-```{r sconeFilt}
-# Simplify to a single logical
-mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
-dataObj <- dataObj[, mfilt]
-dim(dataObj)
-```
-
-After sample filtering, we are left with `r ncol(dataObj)` good quality cells.
-
-## Identification of most variable genes
-
-Often in unsupervised analyses of expression data, it is useful to focus on the genes that vary the most in expression across cells and are thus likely to be the most informative.
-Filtering out low-variance genes allows us to reduce the complexity of the problem, while retaining the most information in the dataset.
-
-Here, for computational efficiency, we retain only the 1,000 most variable genes, following a simple `log(x+1)` transformation.
-This seems to be a reasonable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]).
-In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent and many different strategies have been proposed in the literature.
-
-To compute the gene-wise variances and keep only the most variable genes, we use two functions from the `clusterExperiment` package.
-In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`) and store the variance values in the `rowData` slot of the object [*** SD: Which object?].
-`filterData` will then use the computed values to keep only the desired number of genes.
-
-```{r}
-dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = log1p)
-dataObj <- filterData(dataObj, percentile=1000, filterStats="var")
-rowData(dataObj)
-```
-
-One drawback of this strategy is that the logarithm is not a variance stabilizing transformation.
-Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses.
-Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function.
-Here, we show an example with the `vst` function from `DESeq2`.
-
-```{r, eval=FALSE}
-## Not run
-library(DESeq2)
-dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = vst,
-                           filterNames = "var_vst")
-dataObj <- filterData(dataObj, percentile=1000, filterStats="var_vst")
-```
-
-Alternatively, the `scran` package [@Lun2016] includes the `trendVar` and `decomposeVar` functions, which estimate the mean-variance relationship and the technical and biological components of the variance, to select those genes with a high biological variance.
-See [@Lun2016] for details.
-
-## Dataset structure
-
-Overall, after the above pre-processing steps, our dataset has 1,000 genes and 747 cells.
-
-```{r}
-dataObj
-```
-
 ## Normalization and dimensionality reduction: ZINB-WaVE
 
-In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime inference, and the identification of DE genes.
+In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and the identification of DE genes.
 This allows the data to become more tractable, both from a statistical (cf.
@@
 
-Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor package `zinbwave`.
+Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`.
 The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data.
@@
 The user can also specify both gene-level and cell-level covariates.
-The inclusion of cell-level covariates enables normalization for batch effects, while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
+The inclusion of observed and unobserved cell-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects).
 A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema).
@@
 
-```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2018)."}
+```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
 knitr::include_graphics("zinb_schema.png")
@@
 
-The model can be thought of as a generalization of a factor analysis model, in which the biological signal of interest is assumed to be explained by a small number of "factors", represented in the model by the matrix `W` (Figure \@ref(fig:zinbschema)).
-
-As with most dimensionality reduction methods, the user needs to specify the number of factors (`K`, the number of columns of `W`), which correspond to the dimensions for the new low-dimensional space.
+As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space.
 Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`.
-Note that if the user includes more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.
+Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.
 
 ```{r zinb}
-print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
+print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",  
                                  residuals = TRUE,
@@
 
-The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
+The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
-Note that the low-dimensional matrix `W` is not included in the computation of the residuals to avoid the removal of the biological signal of interest.
+Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.
 
 ```{r norm}
-norm <- assays(dataObj)$normalizedValues
+norm <- assays(se)$normalizedValues
 norm[1:3,1:3]
@@
 ```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
-batch <- colData(dataObj)$Batch
 norm_order <- norm[, order(as.numeric(batch))]
@@
 
-In addition to the normalized values, the `zinbwave` function added several new components to the object: the assay `residuals` contains the residuals of the model that can be used to check the goodness-of-fit of the model.
-The `weights` assay contains observational weights that can be used to downweight the effect of zero counts on differential expression (see [@van2018observation] and the `zinbwave` vignette for more details).
+The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the original clusters (Figure \@ref(fig:pcanorm)).
+Overall, it seems that normalization was effective at removing batch effects without removing biological signal, in spite of the partial nesting of batches within clusters.
 
-```{r}
-assayNames(dataObj)
-```
-
-We set the normalized values as our first assay, so that they become the default in the plots below.
-
-```{r setNormAssay}
-assays(dataObj) <- assays(dataObj)[c("normalizedValues", "counts", "weights",
-                                     "residuals")]
+```{r pcanorm, fig.cap="ZINB-WaVE: PCA of normalized expression measures, where each point represents a cell.
+Cells are color-coded by batch (left panel) and by original published clustering (right panel)."}
+pca <- prcomp(t(norm))
+par(mfrow = c(1,2))
+plot(pca$x, col = col_batch[batch], pch = 20, main = "")
+plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
 ```
@@
 
-Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object.
-This matrix contain the biological signal of interest and can be retrieved using the `reducedDim()` extractor.
-
-```{r}
-W <- reducedDim(dataObj)
-W[1:3,1:3]
-```
-
-In this workflow, the user-supplied dimension `K` of the low-dimensional space was set to `K = 50`.
+The `zinbwave` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`.
 The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance.
@@
 ```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
+W <- colData(se)[, grepl("^W", colnames(colData(se)))]
+W <- as.matrix(W)
 d <- dist(W)
 fit <- cmdscale(d, eig = TRUE, k = 2)
-plot(fit$points, col = col_clus[as.character(colData(dataObj)$publishedClusters)],
-     main = "", pch = 20, xlab = "Component 1", ylab = "Component 2")
+plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
+     pch = 20, xlab = "Component 1", ylab = "Component 2")
 legend(x = "topleft", legend = unique(names(col_clus)), cex = .5,
@@
 ```
-
-Overall, it seems that `zinbwave` was effective at removing batch effects without removing biological signal, in spite of the partial nesting of batches within clusters.
-
-### Choice of K
-
-The most important parameter to choose for the dimensionality reduction step is the number of hidden covariates to be included in the model, i.e., the number of columns of the hidden matrix `W`, which we denote by `K`.
-One way to select the optimal value of `K` is to compare different values in terms of the Akaike information criterion (AIC) or the Bayesian information criterion (BIC).
-The `zinbwave` package provides two functions to compute such quantities, `zinbAIC` and `zinbBIC`.
-
-Here, we fit a few different models corresponding to different values of `K` using the `zinbFit` function, which, unlike `zinbwave`, returns the estimates of all the parameters of the model as well as the value of the likelihood function.
-
-```{r}
-k_values <- c(1:5, 10)
-
-fit <- lapply(k_values, function(k) zinbFit(dataObj, K = k, which_assay = "counts"))
-```
-
-```{r bic, fig.cap="BIC of the model with different values of K."}
-k_values <- c(1: 5, 10)
-plot(k_values, sapply(fit, zinbBIC, t(counts(dataObj))), 
-     ylab = "BIC", xlab = "K", type = "b")
-```
-
-```{r aic, fig.cap="AIC of the model with different values of K."}
-plot(k_values, sapply(fit, zinbAIC, t(counts(dataObj))),
-     ylab = "AIC", xlab = "K", type = "b")
-```
-
-In this case, the BIC suggests a value of `K = 3` (Figure \@ref(fig:bic))), while the AIC suggests a value of `K >= 10` (Figure \@ref(fig:aic))).
-
-Note that since ZINB-WaVE is linear in the number of genes and the number of cells, but quadratic on `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets.
-In such cases, choosing a single value of `K` may be needed.
-In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets.
 
@@
 As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.
-For more details on the `RSEC` method and on the `clusterExperiment` package, see [@Risso2018].
 
-Here, we are applying `RSEC` to the low-dimensional `W` matrix from ZINB-WaVE.
-To do so, we need to specify the `reducedDim` slot to use with the `reduceMethod` option.
-Here, we skip the merging step to collapse similar clusters based on the amount of differential gene expression between them.
-In larger dataset, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).
+Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts.
+Here, we are applying `RSEC` instead to the low-dimensional `W` matrix from ZINB-WaVE, for which we make a separate `SummarizedExperiment` object.
+For this reason, we choose to not use certain options in `RSEC`.
+In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension.
+Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`).
+We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.
 
 ```{r rsec_50}
-print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
-                                betas = 0.8, reduceMethod="zinbwave",
-                                clusterFunction = "hierarchical01",
-                                ncores = NCORES,
-                                dendroReduce="zinbwave",
+seObj <- SummarizedExperiment(t(W), colData = colData(core))
+print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
+                                betas = 0.8, dimReduce="none",
+                                clusterFunction = "hierarchical01", minSizes=1,
+                                ncores = NCORES, isCount=FALSE,
+                                dendroReduce="none",dendroNDims=NA,
                                 subsampleArgs = list(resamp.num=100,
@@
                                                      clusterArgs=list(nstart=10)),
-                                consensusProportion = 0.7,
+                                verbose=TRUE,
+                                combineProportion = 0.7,
                                 mergeMethod = "none", random.seed=424242,
-                                consensusMinSize = 10)))
+                                combineMinSize = 10)))
 ```
 
-As one can see from the code above, there are many parameters that a user can set in RSEC.
-Many of the values that we choose are the default values of the function, but we nonetheless specify them for clarity and reproducibility.
-Here, we briefly describe the role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].
-
-- `k0s`: the values of `k` to start the sequential algorithm.
-- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
-- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
-- `clusterFunction`: function used for the subsample clustering: `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
-- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
-- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
-- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster in the consensus clustering step.
-- `mergeMethod`: our consensus clustering may find too many small clusters.
-If that's the case a merging step can be employed to merge clusters that do not show enough differential expression.
-Here, we skip this step by setting the method as "none".
-- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).
-
-The results of `RSEC` is an object of the `ClusterExperiment` class.
-This class extends the `SingleCellExperiment` class, by adding additional slots to store the clustering results.
-In addition, it has an informative `show` method that displays important information on the clustering.
-
-```{r}
-dataObj
-```
-
-Note that an object of class `ClusterExperiment` is also of class `SingleCellExperiment`. Hence, all the methods define for `SingleCellExperiment` will still work on it.
-
-```{r}
-is(dataObj, "SingleCellExperiment")
-```
-
-The correspondence between the clusters we found here and the original clusters is as follows.
-
-```{r tabagain}
-table(data.frame(original = colData(dataObj)$publishedClusters,
-                 ours = primaryClusterNamed(dataObj)))
-```
-
-To facilitate the comparison between our results and the original clusters, we set a color scheme that allows a direct visual comparison with Figure \@ref(fig:stemcelldiff).
-We can change the clustering color scheme using `clusterExperiment`'s `recolorClusters` function.
-
-```{r colors}
-new_col <- col_clus[c("1", "4", "12", "3", "10", "15", "9")]
-names(new_col) <- paste0("c", 1:7)
-dataObj <- recolorClusters(dataObj, value = new_col)
-```
-
-Cluster name | Description | Color | Correspondence
--------------|-------------|-------| ----------
-c1 | HBC | green | original 1, 5
-c2 | mSUS | magenta | original 4, 7
-c3 | mOSN  | orange | original 9, 12 
-c4 | GBC | cyan | original 2, 3, 11
-c5 | Neuronal Precursors | purple | original 10, 14
-c6 | MV | blue | original 15
-c7 | Immature Neurons | yellow | original 9
-
-The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examinemakeConsensus)), where columns correspond to cells and rows to different clusterings.
-Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap across rows.
+The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings.
+Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap accross rows.
 The first row correspond to a consensus clustering across all candidate clusterings.
 
-```{r examinemakeConsensus, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package.
-The top row corresponds to the consensus clustering and is color-coded accordingly.
-The other rows correspond to the individual clustering across which the consensus was computed.
-In each row, the color is chosen to match as close as possible the color of the cluster from the previous row with the largest overlap."}
-plotClusters(dataObj, existingColors = "firstOnly")
+```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
+plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
 ```
@@
 ```{r plotcoclust, fig.cap="RSEC: Heatmap of co-clustering matrix."}
-plotCoClustering(dataObj)
+plotCoClustering(ceObj)
 ```
@@
 ```{r tableclust}
-table(primaryClusterNamed(dataObj))
+table(primaryClusterNamed(ceObj))
 ```
@@
 ```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
-plotBarplot(dataObj, legend = FALSE)
+plotBarplot(ceObj, legend = FALSE) 
 ```
 
-The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters) and \@ref(fig:addPublishedClusters2)), the main difference being that several of the published clusters were merged here into single clusters.
+The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters.
 This discrepancy is likely caused by the fact that we started with the top 1,000 genes, which might not be enough to discriminate between closely related clusters.
@@
 ```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
-dataObj <- addClusterings(dataObj, colData(dataObj)$publishedClusters, 
+ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
                      clusterLabel = "publishedClusters")
@@
 ## change default color to match with Figure 7
-clusterLegend(dataObj)$publishedClusters[, "color"] <- 
-  col_clus[clusterLegend(dataObj)$publishedClusters[, "name"]]
+clusterLegend(ceObj)$publishedClusters[, "color"] <- 
+  col_clus[clusterLegend(ceObj)$publishedClusters[, "name"]]
 
-plotBarplot(dataObj, whichClusters=c("makeConsensus","publishedClusters"),
-            xlab = "", legend = FALSE,missingColor="white")
-```
-
-```{r addPublishedClusters2, fig.cap="RSEC: Confusion matrix of number of cells per cluster, for our workflow's RSEC clustering and the original published clustering."}
-plotClustersTable(dataObj, whichClusters=c("makeConsensus","publishedClusters"))
+plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
+            xlab = "", legend = FALSE)
 ```
@@
 ```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
-# Set colors for additional sample data
-experimentColors <- bigPalette[1:nlevels(colData(dataObj)$Experiment)]
-batchColors <- bigPalette[1:nlevels(colData(dataObj)$Batch)]
+# Set colors for cell clusterings
+colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
+origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
+experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
+batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
 metaColors <- list("Experiment" = experimentColors,
-                   "Batch" = batchColors)
+                   "Batch" = batchColors,
+                   "publishedClusters" = origClusterColors)
 
-plotHeatmap(dataObj, 
-            whichClusters = c("makeConsensus","publishedClusters"), 
-            clusterFeaturesData = "all",
+plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
+            whichClusters = "primary", clusterFeaturesData = "all",
             clusterSamplesData = "dendrogramValue", breaks = 0.99,
-            colData = c("Batch", "Experiment"),
+            sampleData = c("publishedClusters", "Batch", "Experiment"),
             clusterLegend = metaColors, annLegend = FALSE, main = "")
@@
 
-Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is analogous to Figure \@ref(fig:mdsW) for the original published clusters.
+Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is anologous to Figure \@ref(fig:mdsW) for the original published clusters.
 
 ```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
-plotReducedDims(dataObj, whichCluster="primary", reducedDim="zinbwave", pch=20,
-	xlab = "Component1", ylab = "Component2", legendTitle="Sample", main="",
-	plotUnassigned=FALSE
-)
+palDF <- ceObj@clusterLegend[[1]]
+pal <- palDF[, "color"]
+names(pal) <- palDF[, "name"]
+pal["-1"] = "transparent"
+plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
+     xlab = "Component1", ylab = "Component2")
+legend(x = "topleft", legend = names(pal), cex = .5,
+       fill = pal, title = "Sample")
 ```
@@
 
-We now demonstrate how to use the Bioconductor package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage.
+We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage.
 The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves.
@@
 
-This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object.
-The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e.
-a matrix and a vector) or, as below, components of a `SingleCellExperiment` object.
-When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function.
-For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.
+The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively.
+The first takes as input a low-dimensional representation of the cells and a vector of cluster labels.
+It fits an MST to the clusters and identifies lineages as paths through this tree.
+The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages.
+The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages.
+These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.
 
@@
 Additionally, we know that GBCs should be at a junction before the differentiation between MV and mOSN cells (Figure \@ref(fig:stemcelldiff)).
+The correspondance between the clusters we found here and the original clusters is as follows.
+
+```{r tabagain}
+table(data.frame(original = publishedClusters, ours = primaryClusterNamed(ceObj)))
+```
+
+Cluster name | Description | Color | Correspondence
+-------------|-------------|-------| ----------
+c1 | HBC | blue | original 1, 5
+c2 | GBC | green | original 2, 3, 11
+c3 | mSUS | red | original 4, 7
+c4 | Contaminants | orange | original -2 
+c5 | mOSN | purple | original 9, 12
+c6 | Immature Neuron | brown | original 10, 14
+c7 | MV | cyan | original 15
+
+Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster.
+These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering.
+While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process.
+Note that, out of the 77 cells overexpressing `Reg3g`, 11 are captured in cluster `c4` and 21 are unclustered in our workflow's clustering (see Figure \@ref(fig:boxplotReg3g)).
+However, we retain the remaining 45 cells to infer lineages as they did not seem to influence the clustering.
+
+```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
+c4 <- rep("other clusters", ncol(se)) 
+c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
+boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(ceObj), 
+        ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
+```
 
@@
 
-First, we remove the unclustered cells.
+```{r}
+our_cl <- primaryClusterNamed(ceObj)
+cl <- our_cl[!our_cl %in% c("-1", "c4")]
+pal <- pal[!names(pal) %in% c("-1", "c4")]
+X <- W[!our_cl %in% c("-1", "c4"), ]
+mds <- cmdscale(dist(X), eig = TRUE, k = 4)
+X <- mds$points
 
-```{r}
-dataFilt <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
+lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
 ```
 
-Here, we show how to use `slingshot` starting from a matrix and a set of cluster labels.
-
-```{r}
-X <- reducedDim(dataFilt,type="zinbwave")
-mds <- cmdscale(dist(X), eig = TRUE, k = 4)
-lineages <- slingshot(mds$points, clusterLabels = primaryClusterNamed(dataFilt), start.clus = "c1")
-```
-
-Before discussing the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
+Before fitting the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters.
 This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)).
@@
 ```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
-colorCl<-convertClusterLegend(dataFilt,whichCluster="primary",output="matrixColors")[,1]
-pairs(lineages, type="lineages", col = colorCl)
+pairs(lineages, type="lineages", col = pal[cl])
 ```
 
-Having found the global lineage structure, `slingshot` then constructs a set of smooth, branching curves in order to infer the corresponding pseudotime variables.
+Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables.
 Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters.
-During this iterative process, a cell may even be reassigned to a different lineage if it is significantly closer to the corresponding curve.
-This makes `slingshot` less reliant on the original clustering and generally more stable.
+This makes them more stable and better suited for assigning cells to lineages.
 The final curves are shown in Figure \@ref(fig:curves).
@@
 ```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
-pairs(lineages, type="curves", col = colorCl)
+lineages <- getCurves(lineages)
+pairs(lineages, type="curves", col = pal[cl])
 ```
-
-The `show` method of the `SlingshotDataSet` class displays useful information on the inferred lineages.
 
@@
 
-As an alternative, we could have incorporated the MDS results into the `ClusterExperiment` object and applied `slingshot` directly to it.
-Here, we need to specify that we want to use the MDS results, because `slingshot` would otherwise use the first element of the `reducedDims` list (in this case, the 10-dimensional `W` matrix from `zinbwave`).
-
-```{r sling_sce}
-reducedDim(dataFilt, "MDS") <- mds$points
-dataFilt <- slingshot(dataFilt, reducedDim = "MDS", start.clus = "c1")
-dataFilt
-colData(dataFilt)[,grep("^sling", colnames(colData(dataFilt)))]
-```
-
-The result of `slingshot` applied to a `ClusterExperiment` object is still of class `ClusterExperiment`.
-Note that we did not specify a set of cluster labels, implying that `slingshot` should use the default `primaryCluster` vector.
-
-In the workflow, we recover the right ordering of the clusters using the unsupervised version of slingshot.
-However, in some other cases, the algorithm may need more guidance to find the correct ordering.
-`slingshot` has the option for the user to provide known end cluster(s).
+In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot.
+However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering.
+`getLineages` has the option for the user to provide known end cluster(s).
 Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c7` represent terminal cell fates.
@@
 ```{r slingshotsupervised, eval=FALSE}
-dataFiltSup <- slingshot(dataFilt, reducedDim = "MDS", start.clus = "c1",
-                          end.clus = c("c3", "c6", "c2"))
+lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
+                         end.clus = c("c3", "c7"))
+lineagees <- getCurves(lineages)
+pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
+pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], 
+      show.constraints = TRUE)
+
+lineages
 ```
@@
 We then can test the null hypothesis of no change over time for each gene using the `gam` package.
-
-Generalized Additive Models (GAMs) [@gam] are a class of models that replaces the usual linear function of the coefficients with a sum of smooth functions.
-In the code below, we use a special case of the GAM in which only one covariate (pseudotime) is present.
-This correspond to a simple loess regression, but with the added advantage of providing a framework for the testing of the significance of the association between pseudotime and expression.
-
 We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
@@
 ```{r fitgam}
-t <- colData(dataFilt)$slingPseudotime_1
-y <- transformData(dataFilt)
+t <- pseudotime(lineages)[,1]
+y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c4")]
 gam.pval <- apply(y,1,function(z){
@@
 topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
+heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
+heatclus <- cl[order(t, na.last = NA)]
+ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
 
-dataFilt1 <- dataFilt[,!is.na(t)]
-orderSamples(dataFilt1)<-order(t[!is.na(t)])
+#match to existing colors
+cols <- clusterLegend(ceObj)$combineMany[, "color"]
+names(cols) <- clusterLegend(ceObj)$combineMany[, "name"]
+clusterLegend(ce)$cluster1[, "color"] <- cols[clusterLegend(ce)$cluster1[, "name"]]
 
-plotHeatmap(dataFilt1[topgenes,], clusterSamplesData = "orderSamplesValue", 
-            breaks = .99)
+plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
 ```
-
-## Alternative approaches
-
-Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years.
-The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach.
-For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above.
-This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of each step will be readable by the next step.
-
-An alternative normalization strategy is to compute gene and cell-specific scaling factors (e.g., `SCNorm` [@bacher2017scnorm]).
-Similarly, `monocle` [@trapnell2014dynamics] or `TSCAN` [@ji2016tscan] are alternatives for cell lineage inference.
-Finally, the software suites `Seurat` [@seurat] (available on CRAN) and `scde` [@kharchenko2014bayesian] provide alternative pipelines for cell type discovery, differential expression, and gene set analysis.
-A comprehensive list of tools developed for the analysis of scRNA-seq data is available at https://www.scrna-tools.org [@zappia2018exploring].
 
@@
 
-Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, datasets made of tens of thousands of cells are now routine.
-In fact, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells.
-Hence, there is a growing need for scalable methods.
+In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model.
+In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference.
+In addition, we are extending ZINB-WaVE to identify differentially expressed genes, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns.
 
-When datasets are too large to be fully stored in memory, HDF5 files [@folk2011overview] can be used to keep the data on disk and realize in memory only the subset of data needed at each step.
-We are in the process of updating the `clusterExperiment` and `zinbwave` packages to work with HDF5 files and we are developing more scalable methods to deal with massive datasets.
+We are also developing a downstream method to identify genes that are DE either within or between lineages inferred from Slingshot.
+Whereas the GAM presented here can only identify dynamic genes within a single lineage, many researchers may be interested in finding genes that exhibit different expression patterns between lineages or identifying those genes which differentiate the earliest.
+For questions such as these, we are developing new, spline- and negative binomial-based methods for comparing gene dynamics between branching trajectories.
 
-# Conclusions
+Finally, a new S4 class called `SingleCellExperiment` is currently under development (https://github.com/drisso/SingleCellExperiment).
+This new class is essentially a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension.
+This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step.
+Once the implementation of this class is stable, we would like to incorporate it to the workflow.
+
+# Conclusion
 
@@
 It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages.
-The workflow is general and flexible, allowing the user to substitute the statistical method used in each step by a different method.
+The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method.
 We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.
@@
 The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow.
-The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available at, respectively, https://bioconductor.org/packages/scone, https://bioconductor.org/packages/zinbwave, https://bioconductor.org/packages/clusterExperiment, and https://bioconductor.org/packages/slingshot.
+The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available at, respectively, https://bioconductor.org/packages/scone, https://bioconductor.org/packages/zinbwave, https://bioconductor.org/packages/clusterExperiment, and https://github.com/kstreet13/slingshot.
 
@@
 
-FP, DR, KS, and EP performed the data analysis and wrote the code portions of the workflow.
-DR, FP, and SD wrote the text portion of the workflow, with contributions from the other three authors.
+FP, DR, KS, and EP performed the data analaysis and wrote the code portions of the workflow.
+FP and SD wrote the text portion of the workflow, with contributions from the other three authors.
 DR, EP, and SD supervised the research.

### 2018-08-23 workflow/workflow.Rmd
Added back Kelly's paragraph

--- 2018-08-20 workflow/workflow.Rmd
@@
 
+We are also developing a downstream method to identify genes that are DE either within or between lineages inferred from Slingshot.
+Whereas the GAM presented here can only identify dynamic genes within a single lineage, many researchers may be interested in finding genes that exhibit different expression patterns between lineages or identifying those genes which differentiate the earliest.
+For questions such as these, we are developing new, spline- and negative binomial-based methods for comparing gene dynamics between branching trajectories.
+
 # Conclusions

### 2018-08-23 workflow/workflow.Rmd
Start addressing Sandrine points

@@
 
-There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); [*** SD: Do you mean the "Sox2-eGFP gene"?] the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used for quality control purposes.
+There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); the Sox2-eGFP transgene was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used for quality control purposes.
 
@@
 To compute the gene-wise variances and keep only the most variable genes, we use two functions from the `clusterExperiment` package.
-In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`) and store the variance values in the `rowData` slot of the object [*** SD: Which object?].
+In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`) and store the variance values in the `rowData` slot of the resulting `SingleCellExperiment` object.
 `filterData` will then use the computed values to keep only the desired number of genes.

### 2018-08-27 workflow/workflow.Rmd
Address comments

@@
 ```{r}
-# added chunk of code
-library(scater)
-sce <- calculateQCMetrics(
-    dataObj,
-    feature_controls = list(
-        ERCC = isSpike(dataObj, "ERCC")
-    )
-)
-cData <- colData(sce)
-cData <- as.data.frame(cData)
-ggplot(cData, 
-       aes(x = total_counts, 
-           y = total_counts_ERCC,
-           col = batch)) + 
-  geom_point()
-```
-
-```{r}
 dataObj <- dataObj[!isSpike(dataObj),]
@@
 
-```{r zinb}
+```{r zinb, eval=FALSE}
 print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
@@
 
+```{r}
+load("dataObj_afterZinbwave.rda")
+```
+
 ### Normalization
 
-The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
+The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`normalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates.
 Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots).
@@
 We set the normalized values as our first assay, so that they become the default in the plots below.
+Note that one can specify which assay to plot with the `whichAssay` parameter in the `clusterExperiment` visualization functions.
 
@@
 The most important parameter to choose for the dimensionality reduction step is the number of hidden covariates to be included in the model, i.e., the number of columns of the hidden matrix `W`, which we denote by `K`.
-One way to select the optimal value of `K` is to compare different values in terms of the Akaike information criterion (AIC) or the Bayesian information criterion (BIC).
-The `zinbwave` package provides two functions to compute such quantities, `zinbAIC` and `zinbBIC`.
+As with any dimensionality reduction technique, choosing the number of factors `K` is far from trivial.
+There is no "one-fits-all" solution and the researcher may want to try a few different values of `K` and compare the results.
+
+The `zinbwave` package provides two functions that may be used to guide the choice of `K`.
+The functions `zinbAIC` and `zinbBIC` compute the Akaike information criterion (AIC) and the Bayesian information criterion (BIC), respectively, that can be used to compare models with different values of `K` (smaller values of AIC and BIC indicate a better fit).
 
@@
 
-```{r}
+```{r, eval=FALSE}
 k_values <- c(1:5, 10)
@@
 ```{r bic, fig.cap="ZINB-WaVE: BIC of model for different values of K."}
-k_values <- c(1: 5, 10)
+k_values <- c(1:5, 10)
+fit <- readRDS("Fit_allK.rds")
 plot(k_values, sapply(fit, zinbBIC, t(counts(dataObj))), 
@@
 
-In this case, the BIC suggests a value of `K = 3` (Figure \@ref(fig:bic))), while the AIC suggests a value of `K >= 10` (Figure \@ref(fig:aic))).
+In this case, the BIC suggests a value of `K = 3` (Figure \@ref(fig:bic)), while the AIC suggests a value of `K >= 10` (Figure \@ref(fig:aic)).
 
-Note that since ZINB-WaVE is linear in the number of genes and the number of cells, but quadratic on `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets.
+Depending on the application and on the interpretation of the latent space, we may want to include only a few or several factors.
+Here, we hope that our reduced space will contain all the biological signal of interest, and selecting a large number of factors seems preferable.
+The ZINB-WaVE model can also be used to infer unknown normalization factors (similar to RUV [@risso2014normalization]); in such case, a smaller value of `K` would be more appropriate.
+
+Finally, note that since ZINB-WaVE is linear in the number of genes and the number of cells, but quadratic in `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets.
 In such cases, choosing a single value of `K` may be needed.
-In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets.
+In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets, when the goal is dimensionality reduction.
 
@@
 
-```{r rsec_50}
+```{r rsec_50, eval=FALSE}
 print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
@@
                                 consensusMinSize = 10)))
+```
+
+```{r}
+load("dataObj_afterRSEC.rda")
 ```
@@
 Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years.
-The use of the `SingleCellExperiment` class, which is the shared "language" [*** SD: Better term than "language"?] among Bioconductor's scRNA-seq software packages, makes it extremely easy to replace one of the steps above by a different approach.
+The use of the `SingleCellExperiment` class, which is the shared data container among Bioconductor's scRNA-seq software packages, makes it extremely easy to replace one of the steps above by a different approach.
 For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global-scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above.

### 2018-08-28 workflow/workflow.Rmd
Update workflow.Rmd

@@
 As with any dimensionality reduction technique, choosing the number of factors `K` is far from trivial.
-There is no "one-fits-all" solution and the researcher may want to try a few different values of `K` and compare the results.
+There is no "one-size-fits-all" solution and the researcher may want to try a few different values of `K` and compare the results.
 
 The `zinbwave` package provides two functions that may be used to guide the choice of `K`.
-The functions `zinbAIC` and `zinbBIC` compute the Akaike information criterion (AIC) and the Bayesian information criterion (BIC), respectively, that can be used to compare models with different values of `K` (smaller values of AIC and BIC indicate a better fit).
+The functions `zinbAIC` and `zinbBIC`, which compute the Akaike information criterion (AIC) and the Bayesian information criterion (BIC), respectively, can be used to compare models with different values of `K` (smaller values of AIC and BIC indicate a better fit).
 
@@
 
-Depending on the application and on the interpretation of the latent space, we may want to include only a few or several factors.
-Here, we hope that our reduced space will contain all the biological signal of interest, and selecting a large number of factors seems preferable.
-The ZINB-WaVE model can also be used to infer unknown normalization factors (similar to RUV [@risso2014normalization]); in such case, a smaller value of `K` would be more appropriate.
+Depending on the application and on the interpretation of the latent space, we may want to include different numbers of hidden factors.
+Here, the goal is for the reduced space to capture the biological signal of interest; selecting a large number of factors therefore seems preferable.
+The ZINB-WaVE model can also be used to infer unknown normalization factors (similar to RUV [@risso2014normalization]); in such a case, a smaller value of `K` would be more appropriate.
 
